### 2024-12-20 foundations_of_community_ecology.Rmd
Initialise all of the relevant files

---
title: "Foundations of community ecology"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{fancyhdr}
bibliography: refs.bib
link-citations: yes
linkcolor: blue
biblio-style: apalike
---

We have discovered the fundamental equation that provides a complete description of eco-evolutionary change in any system,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
\tag{1}
$$

In our previous efforts, we derived both the Price equation and the birth-death model from the above.
Here we can derive a fundamental model of community ecology, a discrete time generalised Lotka-Volterra equations leading to modern coexistence theory.
We can do this by partitioning the sum total change $\Omega$ into change caused by density-independent and density-dependent effects,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\left(1 - \sum_{j = 1}^{N}\left(a_{ij}\left(z_{i} + \Delta z_{i} \right) \right) \right).
\tag{2}
$$

In eqn 2, $a_{ij}$ is the effect that $j$ has on the contribution of $i$ to $\Omega$.
Note that we are, again, not relaxing an assumption in eqn 1, nor are we adding something that did not already exist to the model.
Equation 1 is already a complete description of eco-evolutionary change, so it included both density-independent and density-dependent effects. 
The new expression in parentheses is only splitting the total change into density-independent versus density-dependent effects.
As with the previous case of ecology, we can set $z_{i} = 1$ and $\Delta z_{i} = 0$, which allows us to simplify,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
\tag{3}
$$

Now, assume that every individual $j$ in the population has an identical effect on $i$, such that $a_{ij} = \alpha$.
With the conceptual understanding that $\alpha$ is then the per capita effect of $j$ on $i$, we can rewrite eqn 3,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}\alpha \right).
$$

Because we then sum up $\alpha$ a total of $N$ times, it then follows,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \alpha N \right).
$$

Now, we can again assume that individual births ($b = \beta_{i}$) and deaths ($d = \delta_{i}$) are identical for all $i$,

$$\Omega = \sum_{i=1}^{N} \left(b -d + 1 \right)\left(1 - \alpha N \right).
$$

Because we define $\lambda = b - d + 1$,

$$\Omega = \sum_{i=1}^{N} \lambda\left(1 - \alpha N \right).
$$

We now have removed all subscripts $i$ from the right-hand side, and we can rewrite,

$$\Omega =  N \lambda\left(1 - \alpha N \right).
$$

Now is a good time to define $\Omega$ again and add subscripts to $N$,

$$N_{t+1} =  N_{t} \lambda\left(1 - \alpha N_{t} \right).
\tag{4}
$$

Equation 4 gives us a discrete time logistic growth equation. 
This is not typically the way a discrete time equation is expressed, which made me a bit nervous at first, but there is nothing wrong with it either.

We can do some simple simulations to confirm that it behaves as it should.
I am going to make a lot of common simplifying assumptions in what follows (rather, I already have in eqn 4).
I assume that $\alpha$ and $\lambda$ are constant and do not change with density.
Suppose then that we have a population of just two individuals, $N_{t} = 2$. 
With density independent effects, assume that the population is doubling from $t$ to $t + 1$, so $\lambda = 2$.
Now assume that the effect of each individual in the population is to decrease each other individual's growth by $\alpha = 0.01$.
We then have $N_{t+1} =  2 N_{t} \left(1 - 0.01 N_{t} \right)$.
We can simulate this over 10 time steps.

```{r}
NN <- 2;
LL <- 2;
aa <- 0.01;
tt <- 10;
i  <- 1;
while(i < tt){
    Nt        <- NN[i]*LL*(1 - aa * NN[i]);
    NN[i + 1] <- Nt;
    i         <- i + 1;
}
print(NN);
plot(x = 1:10, y = NN, type = "b", pch = 20, lwd = 2,
     xlab = "Time step",
     ylab = "Population density (N)");
points(x = 1:10, y = NN, type = "l", pch = 20, lwd = 2);
```

The population increases exponentially, then levels off at $N = 50$, at which point the population growth $N\lambda = 2(50) = 100$ is balanced by the total effect of competition $\alpha N = 50^{2}(2)(0.01) = 50$, such that $100 - 50$ keeps the population at $50$.

I was a bit bothered by the unusual formulation at first, but now I am feeling better about it.
I am convinced it is correct, albeit not the typically way that researchers tend to model population change.
But usually modellers will switch to continuous time models for logistic growth and generalised Lotka-Volterra. 
And for discrete growth, they will usually do something like $N_{t+1} = N_{t} + N_{t}\lambda(1 - N_{t}/K)$, where $K$ is the carrying capacity.
But this is effort is not derived from first principles, I do not think, in the same way as the above eqn 4.

Recovering proper logistic growth
=======================================

Note that we can also recover a different logistic growth equation if we start at eqn 3, but then assume that $r = b - d$, and density dependent effects only apply to birth and death,

$$\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
\tag{5}
$$

This seems more reasonable.
The effect of an individual $j$ on a focal individual $i$ has to be either on birth or death.
The 1 is a constant, and I do not think it should be affected.
Mathematically, we can expand eqn 5.

$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + \left(1 - \sum_{j = 1}^{N}a_{ij} \right).$$

Biologically, there is a clear way to resolve this, though I think some caution is needed here with the mathematics.
It is biologically reasonable that $a_{ij} = 0$ in the second term for all $j$.
If $a_{ij}$ only applies to $r_{i}$ such that $a_{ij} = 0$ for all $i$ and $j$ in the second term, then,

$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + 1.$$

If we can get to this point, then we can make the same assumption that we did around eqn 3, such that every individual $j$ has an identical effect on $i$, so $a_{ij} = \alpha$.
It then follows,

$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}\alpha \right) + 1.$$

From here, we then sum up the $\alpha$ values,

$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \alpha N \right) + 1.$$

If we assume that $r_{i} = r$ for all $i$, then we are also summing $r(1 - \alpha N) + 1$ a total of $N$ times,

$$\Omega = N \left( r_{i}\left(1 - \alpha N \right) + 1\right).$$

By setting $\Omega = N_{t+1}$ and $N = N_{t}$, we can then recover,

$$N_{t+1} = N_{t} + rN_{t}\left(1 - \alpha N_{t}\right)$$

Again, in this case, $\alpha$ reflects interactions among individuals only affecting birth and death.
This is a standard discrete time generalised Lotka-Volterra equation, as appears in @Chesson2000b.
To recover community ecology, it should be a standard matter of defining two sets $J$ and $K$ representing different species,

$$\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j \in J}a_{ij} - \sum_{k \in K}a_{ik} \right).$$

Summations can be made over different scales too, but I think that this gets us solidly to @Chesson2000b.

References 
==========================

### 2024-12-20 foundations_of_community_ecology.tex
Initialise all of the relevant files

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Foundations of community ecology},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Foundations of community ecology}
\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
\date{{[}1{]} Department of Biological and Environmental Sciences,
University of Stirling, Stirling, Scotland {[}2{]} Department of
Philosophy, University of Valencia, Valencia, Spain {[}*{]}
Corresponding author:
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
{[}a{]} Equal contribution}

\begin{document}
\maketitle

We have discovered the fundamental equation that provides a complete
description of eco-evolutionary change in any system,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
\tag{1}
\]

In our previous efforts, we derived both the Price equation and the
birth-death model from the above. Here we can derive a fundamental model
of community ecology, a discrete time generalised Lotka-Volterra
equations leading to modern coexistence theory. We can do this by
partitioning the sum total change \(\Omega\) into change caused by
density-independent and density-dependent effects,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\left(1 - \sum_{j = 1}^{N}\left(a_{ij}\left(z_{i} + \Delta z_{i} \right) \right) \right).
\tag{2}
\]

In eqn 2, \(a_{ij}\) is the effect that \(j\) has on the contribution of
\(i\) to \(\Omega\). Note that we are, again, not relaxing an assumption
in eqn 1, nor are we adding something that did not already exist to the
model. Equation 1 is already a complete description of eco-evolutionary
change, so it included both density-independent and density-dependent
effects. The new expression in parentheses is only splitting the total
change into density-independent versus density-dependent effects. As
with the previous case of ecology, we can set \(z_{i} = 1\) and
\(\Delta z_{i} = 0\), which allows us to simplify,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
\tag{3}
\]

Now, assume that every individual \(j\) in the population has an
identical effect on \(i\), such that \(a_{ij} = \alpha\). With the
conceptual understanding that \(\alpha\) is then the per capita effect
of \(j\) on \(i\), we can rewrite eqn 3,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}\alpha \right).
\]

Because we then sum up \(\alpha\) a total of \(N\) times, it then
follows,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \alpha N \right).
\]

Now, we can again assume that individual births (\(b = \beta_{i}\)) and
deaths (\(d = \delta_{i}\)) are identical for all \(i\),

\[\Omega = \sum_{i=1}^{N} \left(b -d + 1 \right)\left(1 - \alpha N \right).
\]

Because we define \(\lambda = b - d + 1\),

\[\Omega = \sum_{i=1}^{N} \lambda\left(1 - \alpha N \right).
\]

We now have removed all subscripts \(i\) from the right-hand side, and
we can rewrite,

\[\Omega =  N \lambda\left(1 - \alpha N \right).
\]

Now is a good time to define \(\Omega\) again and add subscripts to
\(N\),

\[N_{t+1} =  N_{t} \lambda\left(1 - \alpha N_{t} \right).
\tag{4}
\]

Equation 4 gives us a discrete time logistic growth equation. This is
not typically the way a discrete time equation is expressed, which made
me a bit nervous at first, but there is nothing wrong with it either.

We can do some simple simulations to confirm that it behaves as it
should. I am going to make a lot of common simplifying assumptions in
what follows (rather, I already have in eqn 4). I assume that \(\alpha\)
and \(\lambda\) are constant and do not change with density. Suppose
then that we have a population of just two individuals, \(N_{t} = 2\).
With density independent effects, assume that the population is doubling
from \(t\) to \(t + 1\), so \(\lambda = 2\). Now assume that the effect
of each individual in the population is to decrease each other
individual's growth by \(\alpha = 0.01\). We then have
\(N_{t+1} = 2 N_{t} \left(1 - 0.01 N_{t} \right)\). We can simulate this
over 10 time steps.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{NN }\OtherTok{\textless{}{-}} \DecValTok{2}\NormalTok{;}
\NormalTok{LL }\OtherTok{\textless{}{-}} \DecValTok{2}\NormalTok{;}
\NormalTok{aa }\OtherTok{\textless{}{-}} \FloatTok{0.01}\NormalTok{;}
\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{10}\NormalTok{;}
\NormalTok{i  }\OtherTok{\textless{}{-}} \DecValTok{1}\NormalTok{;}
\ControlFlowTok{while}\NormalTok{(i }\SpecialCharTok{\textless{}}\NormalTok{ tt)\{}
\NormalTok{    Nt        }\OtherTok{\textless{}{-}}\NormalTok{ NN[i]}\SpecialCharTok{*}\NormalTok{LL}\SpecialCharTok{*}\NormalTok{(}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ aa }\SpecialCharTok{*}\NormalTok{ NN[i]);}
\NormalTok{    NN[i }\SpecialCharTok{+} \DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ Nt;}
\NormalTok{    i         }\OtherTok{\textless{}{-}}\NormalTok{ i }\SpecialCharTok{+} \DecValTok{1}\NormalTok{;}
\NormalTok{\}}
\FunctionTok{print}\NormalTok{(NN);}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1]  2.000000  3.920000  7.532672 13.930521 23.979854 36.459040 46.332848
##  [8] 49.731040 49.998553 50.000000
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10}\NormalTok{, }\AttributeTok{y =}\NormalTok{ NN, }\AttributeTok{type =} \StringTok{"b"}\NormalTok{, }\AttributeTok{pch =} \DecValTok{20}\NormalTok{, }\AttributeTok{lwd =} \DecValTok{2}\NormalTok{,}
     \AttributeTok{xlab =} \StringTok{"Time step"}\NormalTok{,}
     \AttributeTok{ylab =} \StringTok{"Population density (N)"}\NormalTok{);}
\FunctionTok{points}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10}\NormalTok{, }\AttributeTok{y =}\NormalTok{ NN, }\AttributeTok{type =} \StringTok{"l"}\NormalTok{, }\AttributeTok{pch =} \DecValTok{20}\NormalTok{, }\AttributeTok{lwd =} \DecValTok{2}\NormalTok{);}
\end{Highlighting}
\end{Shaded}

\includegraphics{foundations_of_community_ecology_files/figure-latex/unnamed-chunk-1-1.pdf}

The population increases exponentially, then levels off at \(N = 50\),
at which point the population growth \(N\lambda = 2(50) = 100\) is
balanced by the total effect of competition
\(\alpha N = 50^{2}(2)(0.01) = 50\), such that \(100 - 50\) keeps the
population at \(50\).

I was a bit bothered by the unusual formulation at first, but now I am
feeling better about it. I am convinced it is correct, albeit not the
typically way that researchers tend to model population change. But
usually modellers will switch to continuous time models for logistic
growth and generalised Lotka-Volterra. And for discrete growth, they
will usually do something like
\(N_{t+1} = N_{t} + N_{t}\lambda(1 - N_{t}/K)\), where \(K\) is the
carrying capacity. But this is effort is not derived from first
principles, I do not think, in the same way as the above eqn 4.

\hypertarget{recovering-proper-logistic-growth}{%
\section{Recovering proper logistic
growth}\label{recovering-proper-logistic-growth}}

Note that we can also recover a different logistic growth equation if we
start at eqn 3, but then assume that \(r = b - d\), and density
dependent effects only apply to birth and death,

\[\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
\tag{5}
\]

This seems more reasonable. The effect of an individual \(j\) on a focal
individual \(i\) has to be either on birth or death. The 1 is a
constant, and I do not think it should be affected. Mathematically, we
can expand eqn 5.

\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + \left(1 - \sum_{j = 1}^{N}a_{ij} \right).\]

Biologically, there is a clear way to resolve this, though I think some
caution is needed here with the mathematics. It is biologically
reasonable that \(a_{ij} = 0\) in the second term for all \(j\). If
\(a_{ij}\) only applies to \(r_{i}\) such that \(a_{ij} = 0\) for all
\(i\) and \(j\) in the second term, then,

\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + 1.\]

If we can get to this point, then we can make the same assumption that
we did around eqn 3, such that every individual \(j\) has an identical
effect on \(i\), so \(a_{ij} = \alpha\). It then follows,

\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}\alpha \right) + 1.\]

From here, we then sum up the \(\alpha\) values,

\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \alpha N \right) + 1.\]

If we assume that \(r_{i} = r\) for all \(i\), then we are also summing
\(r(1 - \alpha N) + 1\) a total of \(N\) times,

\[\Omega = N \left( r_{i}\left(1 - \alpha N \right) + 1\right).\]

By setting \(\Omega = N_{t+1}\) and \(N = N_{t}\), we can then recover,

\[N_{t+1} = N_{t} + rN_{t}\left(1 - \alpha N_{t}\right)\]

Again, in this case, \(\alpha\) reflects interactions among individuals
only affecting birth and death. This is a standard discrete time
generalised Lotka-Volterra equation, as appears in Chesson
(\protect\hyperlink{ref-Chesson2000b}{2000}). To recover community
ecology, it should be a standard matter of defining two sets \(J\) and
\(K\) representing different species,

\[\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j \in J}a_{ij} - \sum_{k \in K}a_{ik} \right).\]

Summations can be made over different scales too, but I think that this
gets us solidly to Chesson (\protect\hyperlink{ref-Chesson2000b}{2000}).

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Chesson2000b}{}}%
Chesson, Peter L. 2000. {``{Mechanisms of maintenance of species
diversity}.''} \emph{Annual Review of Ecology and Systematics} 31:
343--66.

\end{CSLReferences}

\end{document}

### 2024-12-20 foundations_of_ecoevo.Rmd
Initialise all of the relevant files

---
title: "Foundations of ecological and evolutionary change"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{fancyhdr}
bibliography: refs.bib
link-citations: yes
linkcolor: blue
biblio-style: apalike
---

Abstract
=====================================

**Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density.**
**The deep interdependence between ecology and evolution is well-established, but much theory in each discipline has been developed in isolation.**
**While recent work has accomplished eco-evolutionary synthesis, a gap remains between the logical foundations of ecology and evolution.**
**We bridge this gap with a new equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived.**
**We thereby unify the fundamental equations of population ecology and biological evolution under a general framework.**
**Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness, shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness, respectively, and links this change to ecosystem function.**
**Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.**

Introduction
=====================================

Theoretical unification of phenomena is a powerful tool for scientific advancement. 
Such unification has been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions. 
In theoretical physics, Newton's unification of celestial and terrestrial motion, and Maxwell's unification of electricity and magnetism, are among the most important advances in scientific history. 
Likewise, in the life sciences, the modern synthesis of natural selection and Mendelian inheritance revolutionised how mechanisms underlying evolution and genetics are understood and applied [@Gayon2019].
The analogy to physics is imperfect because evolution encompasses non-biological systems with non-genetic inheritance [e.g., @Price1995; @Shennan2011; @MacCallum2012; @Lewens2015], but to illustrate our point, we restrict our focus to biological evolution.
Following the modern synthesis, patterns in evolutionary biology and genetics could be interpreted through a common framework.

Selection, mutation, recombination, migration, and drift all have predictable consequences for both phenotypes and genotypes in a population.
These mechanisms are therefore foundational to understanding evolution and genetics, and their roles in explaining evolutionary and genetic phenomena are no more conceptually pluralistic than that of gravity in explaining celestial and terrestrial motion.
Instead, biologists understand that each of these mechanisms will jointly affect genetic composition and evolutionary change in a population.
For example, all else being equal, drift will lead to a loss of both heterozygosity and adaptive phenotypes.
This constrains the possible relations that can exist between genetic composition and evolutionary change and therefore clarifies the scope of hypotheses that are biologically coherent [@Caswell1988].
It would obviously be mistaken to model a population in which drift had inconsistent effects on evolution and genetics, but such mistakes would surely be common without the foundational theory required to unify the two subjects.
The absence of this theory is almost unthinkable in modern biology.
Evolution and genetics are arguably as intertwined in biology as electricity and magnetism are in physics.

In contrast, theory underlying ecology and evolution has not been so tightly intertwined.
Until recently, population ecology and evolution in particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and evolutionary phenomena [e.g., @Lion2018; @Patel2018; @Govaert2019], the foundational assumptions and equations of ecology and evolution in these models remain mostly separated conceptually. 
But like the mechanisms driving evolutionary and genetic change, those that drive ecological and evolutionary change are well-established with predictable joint consequences.
In evolution, natural selection, drift, and gene flow are all instantiated through a combination of different ecological factors [@Connor2004].
Most notably, change in population density and phenotype  are both caused by individual birth and death [@Turchin2001].
This should allow biologists to place ecological and evolutionary change on a shared theoretical foundation.

We believe that such a foundation exists for population ecology and evolutionary biology, and that it can bridge the fundamental equations of both subjects and serve as a starting point for understanding change in any population.
@Turchin2001 argues for a simple birth and death model (Box 1) as a fundamental equation of population ecology, and for exponential growth to consequently be as fundamental to ecology as Newton's first law of motion is to physics.
@Turchin2001 argues that general principles are needed to establish a logical foundation for population ecology. 
In evolutionary biology, the Price equation (Box 2) is almost universally regarded as fundamental [@Price1970; @Rice2004; @Gardner2008; @Frank2015a;  @Luque2016; @Luque2021].
It exhaustively and exactly describes evolutionary change for any closed evolving population [@Price1970;  @Luque2016; @Lehtonen2020], and it serves as the starting point for all other fundamental equations in evolution [@Queller2017].
We believe that this logical foundation can be made sufficiently broad to encompass both population and evolutionary change.

We propose an equation from which the fundamental ecological and evolutionary equations can be derived.
Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics (e.g., average excess) or quantitative genetics (e.g., Breeder's equation) can be derived from the Price equation [@Queller2017].
We hope that our equation can provide a starting point for building a shared logical framework between ecology and evolution.

*******************

> **Box 1:** The number of individuals in any closed population ($N$) at any given time ($t + 1$) is determined by the existing number of individuals ($N_{t}$), plus the number of births ($Births$) minus the number of deaths ($Deaths$), $$N_{t+1} = N_{t} + Births - Deaths.$$ This equation is necessarily true for any closed population. Despite its simplicity, it is a general equation for defining population change and a starting point for understanding population ecology. @Turchin2001 notes that a consequence of this fundamental equation is the tendency for populations to grow exponentially (technically geometrically in the above case where time is discrete). This inherent underlying tendency towards exponential growth persists even as the complexities of real populations, such as structure, stochasticity, or density-dependent effects are added to population models [@Turchin2001]. Given the assumption that all individuals in the population are identical, a per capita rate of birth, $Births = bN_{t}$ and death, $Deaths = dN_{t}$ can be defined. Rearranging and defining $\lambda = 1 + b - d$ gives, $N_{t+1} = N_{t}\lambda$. Here $\lambda$ is the finite rate of increase [@Gotelli2001], and note that because $0 \leq d \leq 1$, $\lambda \geq 0$. Verbally, the change in size of any closed population equals its existing size times its finite rate of increase.

*******************

*******************

> **Box 2:** The Price equation is an abstract formula to represent evolutionary change. Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004]. Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's absolute fitness, and $\bar{w}$ average population fitness. On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction. The second term is the expected value of $\Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012]. A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection. The Price equation adds an expectation term and abstracts away from any specific mechanisms of replication or reproduction, or mechanisms of inheritance. Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness. The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fischer's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018]. This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023]. In contrast, some researchers consider the Price equation just a triviality (even tautological), and useless without further modelling assumptions [@VanVeelen2005; @VanVeelen2012]. The debate remains open [@VanVeelen2020; @Baravalle2024].

*******************

A foundation for biological evolution and population ecology
====================================================

To fully unify biological evolution and population ecology, we must reconcile the general equation for population change (Box 1) with the Price equation (Box 2).
Previous theory has produced mathematical synthesis between ecological and evolutionary equations [e.g., @Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
For example, @Page2002 demonstrate a mathematical equivalence between the Price equation and a general equation for community dynamics. 
In both equations, the relative frequencies of different ecological (species) or evolutionary (phenotype) variants map to their respective ecological (population growth) or evolutionary (fitness) change in a way that is mathematically equivalent.
@Collins2009 use the Price equation to build a general quantitative framework that partitions phenotypic change into ecological, evolutionary, and physiological components.
The average change in a community-wide trait (e.g., carbon uptake) is exactly and generally partitioned into change attributable to changes in species composition, expected evolution of lineages within species, and expected change in lineage physiology across all species.
Both @Page2002 and @Collins2009 use relative species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence [@Page2002] or expected change in a community-wide trait [@Collins2009], but a different approach is needed to recover absolute change in population size and thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of biological change [@Price1970; @Frank1997; @Gardner2008; @Luque2016; @Lehtonen2018; @Queller2017].
It has also been highly useful for integrating evolutionary theory across disciplines [@Fox2006; @Brantingham2007; @MacCallum2012; @Frank2015; @Godsoe2021; @Ulrich2024].
These properties would seem to make it an intuitive starting point for a logical foundation of ecology and evolution, perhaps through some kind of mathematical equivalence [e.g., @Page2002] or addition of terms [e.g., @Collins2009], or through the use of its recursive structure [e.g., @Kerr2009; @Frank2012].
But despite its flexibility, the Price equation still relies on relative frequencies, which must by definition sum to one [@Frank2015].
This is because the Price equation describes the average change in a population; the frequency of entities is scaled thereby conserving total probability [@Frank2015; @Frank2016].
But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided in a fundamental equation of ecology and evolution.

We therefore begin with the most fundamental axioms underlying the ecology and evolution of living systems [@Rice2004; @Rice2009].
In such systems, diversity is discontinuous and can be defined in terms of discrete entities [@Dobzhansky1970].
Our framework is general enough that entities can be anything discrete, but we will focus on each entity $i$ as an individual organism.
Individuals give rise to new individuals through birth such that $\beta_{i}$ is the number of births attributable to $i$.
Individuals are removed from the population through death such that $\delta_{i}$ is an indicator variable that takes a value of 1 (death of $i$) or 0 (persistence of $i$).
All individuals are defined by some characteristic $z_{i}$, and $\Delta z_{i}$ defines any change in $z_{i}$ from one time step $t$ to the next $t + 1$.
The total number of individuals in the population is $N$.
From this foundation, we can define $\Omega$ to be a summed characteristic across $N$ entities,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
\tag{1}
$$

From eqn 1, we can derive the most fundamental equations of population ecology (Box 1) and evolutionary biology (Box 2) through an appropriate interpretation of $z_{i}$.

Population ecology
====================================================

To recover the general equation for population ecology (Box 1), we define $z_{i}$ as the identity of $i$ belonging to the population.
In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
In this case,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right).$$

Since we assume that individuals are identical, we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
Summing from 1 to $N$, we can rewrite the above,

$$\Omega = N\left(1 + \beta - \delta \right).$$

We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$ and note that $N\beta$, $N\delta$, and $N$ are the total births, total deaths, and size of the population at $t$, respectively,

$$N_{t+1} = N_{t} + Births - Deaths.
\tag{2}
$$

If we define $\lambda = 1 + \beta - \delta$ (Box 1), then we can rewrite,

$$N_{t+1} = N_{t}\lambda.
\tag{3}
$$

We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [eqn 3, @Turchin2001].

Evolutionary biology
====================================================

Recovering the Price equation requires a few more steps.
We start by defining individual fitness,

$$w_{i} = \beta_{i} - \delta_{i} + 1.
\tag{4}
$$

In this definition, the longevity of the individual matters.
An individual that survives from $t$ to $t + 1$ has a higher fitness than one that dies, even if both have the same reproductive output.
With this definition of fitness (eqn 4), we substitute,

$$\Omega = \sum_{i=1}^{N} \left(w_{i}z_{i} + w_{i}\Delta z_{i} \right).
\tag{5}
$$

We can break eqn 5 down further and multiply each side by $1/N$,

$$\frac{1}{N}\Omega = \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}z_{i} \right) + \frac{1}{N}\sum_{i=1}^{N}\left( w_{i}\Delta z_{i} \right).
\tag{6}
$$

We can rewrite the terms on the right-hand side of eqn 6 as expected values and remove the subscripts,

$$\frac{1}{N}\Omega = \mathrm{E}\left(w z \right) + \mathrm{E}\left( w \Delta z  \right).
\tag{7}
$$

Now we must consider the total conservation of probability [@Frank2015; @Frank2016].
In eqn 7, $\Omega$ is the total sum trait values ($z_{i}$) across the entire population at $t + 1$ divided by the number of individuals ($N$) in the population at $t$.
But the size of the population can change from $t$ to $t + 1$.
To recover mean trait change for the Price equation (and therefore conserve total probability), we need to account for this change in population size.
The left-hand side of eqn 7 describes contributions to the sum trait value from the new population at $t + 1$.
But we cannot treat $\Omega/N$ as the mean of $z$ at $t+1$ ($\bar{z}'$) because we need to weigh $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.

For example, if mean fitness at $t$ is 2 (i.e., $\bar{w} = 2$), then half as many individuals will have contributed to $\Omega$ in $t + 1$ than would have if $\bar{w} = 1$ (i.e., there are $N$ individuals at $t$ and $2N$ individuals at $t + 1$).
We therefore need to multiply the mean trait value $\bar{z}'$ (at $t + 1$) by the mean fitness $\bar{w}$ (at $t$) to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$.
Consequently,

$$\Omega = N\bar{w}\bar{z}'
\tag{8}
$$

Equation 8 conserves the total probability and now clarifies $\Omega$ as a summed trait value, which equals expected population growth at $t$ times mean trait value at $t + 1$.
This is consistent with the population ecology derivation from the previous section where $z_{i} = 1$ by definition, and $\Omega = N_{t+1}$ (note $\lambda = \bar{w}$).
We can therefore rewrite eqn 7,

$$\bar{w}\bar{z}' = \mathrm{E}\left(w z \right) + \mathrm{E}\left( w \Delta z  \right).
\tag{9}
$$

We can rearrange eqn 9 to derive the Price equation by expressing covariance as, $\mathrm{Cov}(X,Y) = \mathrm{E}(XY) - \mathrm{E}(X)\mathrm{E}(Y)$, and therefore $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
Substituting into eqn 9,

$$\bar{w}\bar{z}' = \mathrm{Cov}\left(w ,z \right) + \bar{w}\bar{z} + \mathrm{E}\left( w \Delta z  \right).$$

From here,

$$\bar{w}\left(\bar{z}' - \bar{z}\right) = \mathrm{Cov}\left(w ,z \right) + \mathrm{E}\left( w \Delta z  \right).$$

Since $\Delta \bar{z} = \left(\bar{z}' - \bar{z}\right)$,

$$\bar{w}\Delta \bar{z} = \mathrm{Cov}\left(w ,z \right) + \mathrm{E}\left( w \Delta z  \right).
\tag{10}
$$

From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 1) and the fundamental equation of evolution (eqn 10; Box 2).
We might also consider $\Omega$ to be a quantity of ecosystem function, in that it is the summed effect of the traits ($z_{i}$) of individual entities ($i$) in the population.
For example, $z$ might be interpreted as biomass or photosynthetic rate.

Discussion
====================================================

A classical sign of scientific progress is the ability to connect disparate theories and models to show how empirical and theoretical models are logical (mathematical) consequences of more fundamental ones [@Nagel1961].
Conceptual unification therefore has a critical role in advancing scientific theory [@Morrison2000].
Showing how apparently disparate phenomena follow from the same shared principles reveals what is most fundamental about the natural world and provides a foundation for further scientific investigation and the construction of more coherent and predictive models.
Our eqn 1 provides a foundation for the conceptual unification of the fundamental equations of population ecology and evolution.

Rather than making simplifying assumptions, as is the approach for most ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 1) and principle of exponential growth ($N_{t+1} = \lambda N_{t}$).
By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 2).
We therefore deliver a general framework to unify fundamental equations of ecological and evolutionary change.

The Price equation provides a complete and exact description of evolution in any closed evolving system [@Price1970; @Frank2012].
It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [e.g., individuals, alleles, @Price1970; @Gardner2008; @Luque2016].
This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation [e.g., group, individual, @Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
Because of its abstract nature and lack of any system-specific assumptions, the Price equation makes no predictions about what will happen in any particular system [@Gardner2020].
Its role is not to predict, but to formally and completely define and separate components of evolutionary change. 
The same is true of the general equation for population change (eqns 2 and 3), at least as we have used it here where it serves to define what population growth means in ecology.
This equation formally and completely describes population change in terms of births and deaths.
In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population.
We anticipate that this will be useful for eco-evolutionary theory in the same way that the Price equation is useful for evolutionary theory: facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.

<!---

Our unification and derivation of fundamental equations of ecology and evolution recovers and formalises well-known properties of biological systems.
The expressions $\beta_{i} - \delta_{i} + 1$ and $z_{i} - \Delta z_{i}$ can be interpreted to represent the ecological and evolutionary components of a system, respectively.
When focused only on change in population size, we set $z_{i} = 1$ and $\Delta z_{i} = 0$, then we can derive the fundamentals of population ecology without considering evolution.
Interestingly, the same is not true when focused only on evolutionary change.
If we set $\beta_{i} = 0$ and $\delta_{i} = 0$ (no birth or death), then there is no variance in fitness ($w_{i} = 1$ for all individuals), and evolution does not occur (note that evolution can still occur in a population at equilibrium where the birth rate and death rate are the same).
Fundamentally, evolutionary change is therefore enacted through ecological change, i.e., individual births and deaths, but ecological change does not necessarily imply evolution.
And while the link to population change is intuitive from the structure of eqn 1, the link to the Price equation is much less intuitive, especially after eqn 5 in our derivation.
The Price equation describes change in any evolving system, not just biological systems [@Fox2006; @Brantingham2007; @Bell2009a; @MacCallum2012; @Frank2015].
It is perhaps unsurprising that a derivation of ecological and evolutionary change from a common starting point requires relatively more labour to recover the Price equation.

--->

Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ [Box 1, @Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 2).
The population growth equation $N_{t+1} = N_{t}\lambda$ could therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
We do not claim to be the first to notice this equivalence between population growth and mean population fitness.
Indeed, the relationship between population growth rate and evolutionary fitness has been proposed and applied many times before [e.g., @Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
For example, @Lande1982 explicitly concludes mean absolute fitness per unit time is $\bar{w} = e^{r}$, where $r = ln(\lambda)$.
We show this from first principles and clarify the relationship between fitness and population growth.
Over an arbitrary length of time, fitness is properly defined as $w_{i} = \beta_{i} - \delta_{i} + 1$.
Over an individual's lifetime (which, by definition, includes death), fitness is therefore $\beta_{i}$.
Interestingly, the rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)$ [i.e., Fisher's fundamental theorem, @Frank1997; @Rice2004; @Queller2017].

Our unification may also help explain, at least partially, some of the success of classical population genetic models. 
For decades, population genetics (and to some extent quantitative genetics) has been accused of being a reductionistic view of evolution, reducing everything to changes in allele frequencies and abstracting away from individuals and their environments [e.g., the ecological interactions, @MacColl2011]. 
This has been a line of argumentation by some defenders of the so-called Extended Synthesis [@Pigliucci2009], especially in relation to niche construction [@Odling-smee2003].
Famously, @Mayr1960 characterized population genetics as a simple input and output of genes, analogous to "the adding of certain beans to a beanbag and the withdrawing of others" (also called "beanbag genetics").

Historical critics of population genetics could not articulate a clear explanation for why it works so well despite all of its idealisations and simplifications.
From the Price equation, we are able to recover classical population and quantitative genetic models [@Queller2017] and develop new ones [@Rice2004; @Rice2020; @Luque2016; @Lion2018].
Our eqn 1 contains ecology at its core, and we show how the Price equation logically follows from it after accounting for absolute population growth (eqn 8).
We therefore conclude that population and quantitative genetic equations contain ecology (no matter how hidden), and the ecological nature of evolution is implicit in population and quantitative genetic models.

We have focused on the dynamics of a closed population, and in doing so leave ecological and evolutionary change attributable to migration for future work.
In population ecology, immigration and emigration can be incorporated by adding a term for each to the right-hand side of the equation in Box 1 [@Gotelli2001].
In evolution, because the Price equation relies on mapping ancestor-descendant relationships, accounting for migration is more challenging.
@Kerr2009 demonstrate how the Price equation can be extended to allow for arbitrary links between ancestors and descendants, thereby extending the Price equation to allow for immigration and emigration. 
@Frank2012 presents a simplified version of @Kerr2009 that allows some fraction of descendants to be unconnected to ancestors.
In both ecology and evolution, accounting for migration is done through the use of additional terms on the right-hand side of the equations.

<!---

PROBLEM HERE: TRYING TO RECOVER THE LOGISTIC GROWTH IN DISCRETE TIME

=======================================

Our fundamental equation is complete and exact.
It therefore implicitly includes any effects of density dependence in population growth (see Box 1).
Starting from eqn 1, we can partition ecological change into density-independent and density-dependent effects by multiplying the expression $\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$ within the summation by $\left(1 + \sum_{j=1}^{N} \alpha_{ij}z_{j}\right)$, where $j$ indexes the same individuals as $i$ and $\alpha_{ij}$ defines the effect $j$ has on the contribution of $i$ to $\Omega$. 
This partition can be used to recover

OR............. BELIEVE THAT THE BELOW IS THE CORRECT ONE!!

Our fundamental equation is complete and exact.
It therefore implicitly includes any effects of density dependence in population growth (see Box 1).
Starting from eqn 1, we can partition ecological change into density-independent and density-dependent effects by first defining $a_{ij}$ as the effect that the characteristic of $j$ ($z_{j}$) has on the contribution of $i$ to $\Omega$. 
Within the summation of eqn 1, we can then include a new term summing across all individuals in the population, $\sum_{j=1}^{N} a_{ij}z_{j}\right)$.
This sums the effect of each individual on a focal individual $i$.
If we define $\alpha = E\left( a_{ij}\right)$, then we can

--->

Our fundamental equation is complete and exact.
It therefore implicitly includes any effects of density dependence on population growth (see Box 1), or any social effects on evolutionary change (see Box 2).
We can make both of these effects explicit with the same partition in the summation on the right-hand side of eqn 1.
To do this, we multiply the expression $\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$ by $\left(1 - \sum_{j=1}^{N} a_{ij}\left(z_{j} + \Delta z_{j}\right)\right)$, where $j$ indexes the same individuals as $i$, and $a_{ij}$ defines the effect $j$ has on the contribution of $i$ to $\Omega$. 
In evolutionary terms, $a_{ij}$ defines how $j$ modulates the fitness of $i$ ($w_{i}$).
In ecological terms, $a_{ij}$ defines how $j$ modulates the contribution of $i$ to total population growth.
If we define $\alpha$ as the summed effect of all individuals on a focal $i$ (i.e., $\alpha = \sum_{j=1}^{N} a_{ij}$), then we can partition ecological change into density-independent and density-dependent effects and derive a discrete time logistic growth equation.
We save these derivations for a future investigations focused on evolutionary and community ecology.

We have shown that we can derive the fundamental equations of population ecology and biological evolution from a single unifying equation.
Lastly, we propose our eqn 1 as a potential starting point for defining ecosystem function and further conceptual unification between ecology, evolution, and ecosystem function.
The Price equation has previously been used to investigate ecosystem function [@Loreau2001; @Fox2006], but not with any attempt towards conceptual unification with evolutionary biology.
For example, @Fox2006 applied the abstract properties of the Price equation to partition total change in ecosystem function into separate components attributable to species richness, species composition, and context dependent effects.
This approach provides a framework for comparing the effects of biodiversity on ecosystem function in empirical systems [e.g., @Fox2006; @Winfree2015; @Mateo-Tomas2017].
Instead, our eqn 1 defines $\Omega$ as total ecosystem function contributed by a focal population (e.g., biomass production, decomposition, photosynthetic rate).
It is therefore possible to investigate ecological, evolutionary, and ecosystem function change from the same shared framework.

<!---

NOTE 1
==================

Need to discuss that evolution *is* possible without ecology, if we imagine a set of RNA molecules that are all identical in an environment with no birth or death. We might have some change in RNA sequence (and therefore structure) that is evolution; this evolution will be in the \Delta z as non-zero -- the covariance of the Price equation will still be zero, but the expectation term will be nonzero because expected fitness is 1 (because you get credit for surviving from time step t to t+1). Hence, no adaptive evolution (selection) is possible without ecology, but mutation can happen (or selection at lower levels of biological organisation).

So evolution is possible without ecology and vice versa, however unlikely.

NOTE 2
================

We can partition \Delta z in the original equation however we want, and it just gets back to new terms attached to the Price equation at the end

Need to try to derive what happens when we partition fitness into birth, death, and survival component

--->

Acknowledgements
====================================================

This manuscript was supported by joint funding between the French Foundation for Research on Biodiversity (FRB) Centre for the Synthesis and Analysis of Biodiversity (CESAB) and the German Centre for Integrative Biodiversity Research (sDiv).
It was written as part of the Unification of Modern Coexistence Theory and Price Equation (UNICOP) project.
Victor J. Luque was also supported by the Spanish Ministry of Science and Innovation (Project: PID2021-128835NB-I00), and the Conselleria  d’Innovaci$\'{o}$, Universitats, Ci$\`{e}$ncia i Societat Digital – Generalitat  Valenciana (Project: CIGE/2023/16).
We are grateful for many conversations with S$\'{e}$bastien Lion, Kelsey Lyberger, Swati Patel, and especially Lynn Govaert, whose questions helped us clarify the relationship between population growth and fitness.
Brad Duthie would also like to thank Brent Danielson and Stan Harpole.
Victor J. Luque would also like to thank Lorenzo Baravalle, Pau Carazo, Santiago Ginnobili, Manuel Serra, and Ariel Roff$\'{e}$.

References
====================================================

### 2024-12-20 foundations_of_ecoevo.tex
Initialise all of the relevant files

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Foundations of ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Foundations of ecological and evolutionary change}
\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
\date{{[}1{]} Department of Biological and Environmental Sciences,
University of Stirling, Stirling, Scotland {[}2{]} Department of
Philosophy, University of Valencia, Valencia, Spain {[}*{]}
Corresponding author:
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
{[}a{]} Equal contribution}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

\textbf{Biological evolution is realised through the same mechanisms of
birth and death that underlie change in population density.} \textbf{The
deep interdependence between ecology and evolution is well-established,
but much theory in each discipline has been developed in isolation.}
\textbf{While recent work has accomplished eco-evolutionary synthesis, a
gap remains between the logical foundations of ecology and evolution.}
\textbf{We bridge this gap with a new equation that defines a summed
value for a characteristic across individuals in a population, from
which the fundamental equations of population ecology and evolutionary
biology (the Price equation) are derived.} \textbf{We thereby unify the
fundamental equations of population ecology and biological evolution
under a general framework.} \textbf{Our unification further demonstrates
the equivalence between mean population growth rate and evolutionary
fitness, shows how ecological and evolutionary change are reflected in
the first and second statistical moments of fitness, respectively, and
links this change to ecosystem function.} \textbf{Finally, we outline
how our proposed framework can be used to unify social evolution and
density-dependent population growth.}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\protect\hyperlink{ref-Kitcher1993}{Kitcher
1993}), and its value is perhaps most evident in reconciling unconnected
models and revealing new and unexpected empirical predictions. In
theoretical physics, Newton's unification of celestial and terrestrial
motion, and Maxwell's unification of electricity and magnetism, are
among the most important advances in scientific history. Likewise, in
the life sciences, the modern synthesis of natural selection and
Mendelian inheritance revolutionised how mechanisms underlying evolution
and genetics are understood and applied
(\protect\hyperlink{ref-Gayon2019}{Gayon and Huneman 2019}). The analogy
to physics is imperfect because evolution encompasses non-biological
systems with non-genetic inheritance (e.g.,
\protect\hyperlink{ref-Price1995}{Price 1995};
\protect\hyperlink{ref-Shennan2011}{Shennan 2011};
\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
\protect\hyperlink{ref-Lewens2015}{Lewens 2015}), but to illustrate our
point, we restrict our focus to biological evolution. Following the
modern synthesis, patterns in evolutionary biology and genetics could be
interpreted through a common framework.

Selection, mutation, recombination, migration, and drift all have
predictable consequences for both phenotypes and genotypes in a
population. These mechanisms are therefore foundational to understanding
evolution and genetics, and their roles in explaining evolutionary and
genetic phenomena are no more conceptually pluralistic than that of
gravity in explaining celestial and terrestrial motion. Instead,
biologists understand that each of these mechanisms will jointly affect
genetic composition and evolutionary change in a population. For
example, all else being equal, drift will lead to a loss of both
heterozygosity and adaptive phenotypes. This constrains the possible
relations that can exist between genetic composition and evolutionary
change and therefore clarifies the scope of hypotheses that are
biologically coherent (\protect\hyperlink{ref-Caswell1988}{Caswell
1988}). It would obviously be mistaken to model a population in which
drift had inconsistent effects on evolution and genetics, but such
mistakes would surely be common without the foundational theory required
to unify the two subjects. The absence of this theory is almost
unthinkable in modern biology. Evolution and genetics are arguably as
intertwined in biology as electricity and magnetism are in physics.

In contrast, theory underlying ecology and evolution has not been so
tightly intertwined. Until recently, population ecology and evolution in
particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and
evolutionary phenomena (e.g., \protect\hyperlink{ref-Lion2018}{Lion
2018}; \protect\hyperlink{ref-Patel2018}{Patel, Cortez, and Schreiber
2018}; \protect\hyperlink{ref-Govaert2019}{Govaert et al. 2019}), the
foundational assumptions and equations of ecology and evolution in these
models remain mostly separated conceptually. But like the mechanisms
driving evolutionary and genetic change, those that drive ecological and
evolutionary change are well-established with predictable joint
consequences. In evolution, natural selection, drift, and gene flow are
all instantiated through a combination of different ecological factors
(\protect\hyperlink{ref-Connor2004}{Connor and Hartl 2004}). Most
notably, change in population density and phenotype are both caused by
individual birth and death (\protect\hyperlink{ref-Turchin2001}{Turchin
2001}). This should allow biologists to place ecological and
evolutionary change on a shared theoretical foundation.

We believe that such a foundation exists for population ecology and
evolutionary biology, and that it can bridge the fundamental equations
of both subjects and serve as a starting point for understanding change
in any population. Turchin (\protect\hyperlink{ref-Turchin2001}{2001})
argues for a simple birth and death model (Box 1) as a fundamental
equation of population ecology, and for exponential growth to
consequently be as fundamental to ecology as Newton's first law of
motion is to physics. Turchin
(\protect\hyperlink{ref-Turchin2001}{2001}) argues that general
principles are needed to establish a logical foundation for population
ecology. In evolutionary biology, the Price equation (Box 2) is almost
universally regarded as fundamental
(\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Frank2015a}{Frank 2017};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021}). It
exhaustively and exactly describes evolutionary change for any closed
evolving population (\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Lehtonen2020}{Lehtonen, Okasha, and Helanterä
2020}), and it serves as the starting point for all other fundamental
equations in evolution (\protect\hyperlink{ref-Queller2017}{Queller
2017}). We believe that this logical foundation can be made sufficiently
broad to encompass both population and evolutionary change.

We propose an equation from which the fundamental ecological and
evolutionary equations can be derived. Derivation follows by adding
assumptions that are specific to population ecology or evolution in the
same way that key equations of population genetics (e.g., average
excess) or quantitative genetics (e.g., Breeder's equation) can be
derived from the Price equation
(\protect\hyperlink{ref-Queller2017}{Queller 2017}). We hope that our
equation can provide a starting point for building a shared logical
framework between ecology and evolution.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{quote}
\textbf{Box 1:} The number of individuals in any closed population
(\(N\)) at any given time (\(t + 1\)) is determined by the existing
number of individuals (\(N_{t}\)), plus the number of births
(\(Births\)) minus the number of deaths (\(Deaths\)),
\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
for any closed population. Despite its simplicity, it is a general
equation for defining population change and a starting point for
understanding population ecology. Turchin
(\protect\hyperlink{ref-Turchin2001}{2001}) notes that a consequence of
this fundamental equation is the tendency for populations to grow
exponentially (technically geometrically in the above case where time is
discrete). This inherent underlying tendency towards exponential growth
persists even as the complexities of real populations, such as
structure, stochasticity, or density-dependent effects are added to
population models (\protect\hyperlink{ref-Turchin2001}{Turchin 2001}).
Given the assumption that all individuals in the population are
identical, a per capita rate of birth, \(Births = bN_{t}\) and death,
\(Deaths = dN_{t}\) can be defined. Rearranging and defining
\(\lambda = 1 + b - d\) gives, \(N_{t+1} = N_{t}\lambda\). Here
\(\lambda\) is the finite rate of increase
(\protect\hyperlink{ref-Gotelli2001}{Gotelli 2001}), and note that
because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
size of any closed population equals its existing size times its finite
rate of increase.
\end{quote}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{quote}
\textbf{Box 2:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\protect\hyperlink{ref-Price1970}{Price 1970},
\protect\hyperlink{ref-Price1972}{1972}), it postulates some basic
properties that all evolutionary systems must satisfy: change over time,
ancestor and descendant relations, and a character or phenotype
(\protect\hyperlink{ref-Rice2004}{Rice 2004}). Using simple algebraic
language, the Price equation represents evolutionary change with the
predominant notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Okasha2006}{Okasha 2006};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). A more specific version
of the covariance term was already known within the quantitative and
population genetics tradition
(\protect\hyperlink{ref-Robertson1966}{Robertson 1966}), usually
representing the action of natural selection. The Price equation adds an
expectation term and abstracts away from any specific mechanisms of
replication or reproduction, or mechanisms of inheritance. Its
definitional nature and lack of substantive biological assumptions has
been portrayed both as a strength (\protect\hyperlink{ref-Rice2004}{Rice
2004}; \protect\hyperlink{ref-Frank2012}{Frank 2012};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Baravalle2022}{Baravalle and Luque 2022}), and
its greatest weakness. The abstract nature of the Price equation places
it at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fischer's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\protect\hyperlink{ref-Queller2017}{Queller 2017};
\protect\hyperlink{ref-Lehtonen2016}{Lehtonen 2016},
\protect\hyperlink{ref-Lehtonen2018}{2018}). This abstractness is also
key to developing a more general view of evolution
(\protect\hyperlink{ref-Rice2020}{Rice 2020};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021};
\protect\hyperlink{ref-Edelaar2023}{Edelaar, Otsuka, and Luque 2023}).
In contrast, some researchers consider the Price equation just a
triviality (even tautological), and useless without further modelling
assumptions (\protect\hyperlink{ref-VanVeelen2005}{van Veelen 2005};
\protect\hyperlink{ref-VanVeelen2012}{van Veelen et al. 2012}). The
debate remains open (\protect\hyperlink{ref-VanVeelen2020}{van Veelen
2020}; \protect\hyperlink{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{a-foundation-for-biological-evolution-and-population-ecology}{%
\section{A foundation for biological evolution and population
ecology}\label{a-foundation-for-biological-evolution-and-population-ecology}}

To fully unify biological evolution and population ecology, we must
reconcile the general equation for population change (Box 1) with the
Price equation (Box 2). Previous theory has produced mathematical
synthesis between ecological and evolutionary equations (e.g.,
\protect\hyperlink{ref-Page2002}{Page and Nowak 2002};
\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but not
established a unified conceptual framework for joint ecological and
evolutionary change. For example, Page and Nowak
(\protect\hyperlink{ref-Page2002}{2002}) demonstrate a mathematical
equivalence between the Price equation and a general equation for
community dynamics. In both equations, the relative frequencies of
different ecological (species) or evolutionary (phenotype) variants map
to their respective ecological (population growth) or evolutionary
(fitness) change in a way that is mathematically equivalent. Collins and
Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use the Price
equation to build a general quantitative framework that partitions
phenotypic change into ecological, evolutionary, and physiological
components. The average change in a community-wide trait (e.g., carbon
uptake) is exactly and generally partitioned into change attributable to
changes in species composition, expected evolution of lineages within
species, and expected change in lineage physiology across all species.
Both Page and Nowak (\protect\hyperlink{ref-Page2002}{2002}) and Collins
and Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use relative
species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence
(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002}) or expected
change in a community-wide trait
(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but a
different approach is needed to recover absolute change in population
size and thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of
biological change (\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Frank1997}{Frank 1997};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Lehtonen2018}{Lehtonen 2018};
\protect\hyperlink{ref-Queller2017}{Queller 2017}). It has also been
highly useful for integrating evolutionary theory across disciplines
(\protect\hyperlink{ref-Fox2006}{Fox 2006};
\protect\hyperlink{ref-Brantingham2007}{Brantingham 2007};
\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
\protect\hyperlink{ref-Frank2015}{Frank 2015};
\protect\hyperlink{ref-Godsoe2021}{Godsoe et al. 2021};
\protect\hyperlink{ref-Ulrich2024}{Ulrich et al. 2024}). These
properties would seem to make it an intuitive starting point for a
logical foundation of ecology and evolution, perhaps through some kind
of mathematical equivalence (e.g., \protect\hyperlink{ref-Page2002}{Page
and Nowak 2002}) or addition of terms (e.g.,
\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), or
through the use of its recursive structure (e.g.,
\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). But despite its
flexibility, the Price equation still relies on relative frequencies,
which must by definition sum to one
(\protect\hyperlink{ref-Frank2015}{Frank 2015}). This is because the
Price equation describes the average change in a population; the
frequency of entities is scaled thereby conserving total probability
(\protect\hyperlink{ref-Frank2015}{Frank 2015},
\protect\hyperlink{ref-Frank2016}{2016}). But to recover the fundamental
principle of exponential population growth
(\protect\hyperlink{ref-Turchin2001}{Turchin 2001}), this scaling must
be avoided in a fundamental equation of ecology and evolution.

We therefore begin with the most fundamental axioms underlying the
ecology and evolution of living systems
(\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Rice2009}{Rice and Papadopoulos 2009}). In such
systems, diversity is discontinuous and can be defined in terms of
discrete entities (\protect\hyperlink{ref-Dobzhansky1970}{Dobzhansky
1970}). Our framework is general enough that entities can be anything
discrete, but we will focus on each entity \(i\) as an individual
organism. Individuals give rise to new individuals through birth such
that \(\beta_{i}\) is the number of births attributable to \(i\).
Individuals are removed from the population through death such that
\(\delta_{i}\) is an indicator variable that takes a value of 1 (death
of \(i\)) or 0 (persistence of \(i\)). All individuals are defined by
some characteristic \(z_{i}\), and \(\Delta z_{i}\) defines any change
in \(z_{i}\) from one time step \(t\) to the next \(t + 1\). The total
number of individuals in the population is \(N\). From this foundation,
we can define \(\Omega\) to be a summed characteristic across \(N\)
entities,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
\tag{1}
\]

From eqn 1, we can derive the most fundamental equations of population
ecology (Box 1) and evolutionary biology (Box 2) through an appropriate
interpretation of \(z_{i}\).

\hypertarget{population-ecology}{%
\section{Population ecology}\label{population-ecology}}

To recover the general equation for population ecology (Box 1), we
define \(z_{i}\) as the identity of \(i\) belonging to the population.
In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
member of the population. Further, we assume that individuals do not
change species by setting \(\Delta z_{i} = 0\). In this case,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right).\]

Since we assume that individuals are identical, we can drop the
subscript \(i\) such that \(\beta_{i} = \beta\) and
\(\delta_{i} = \delta\). Summing from 1 to \(N\), we can rewrite the
above,

\[\Omega = N\left(1 + \beta - \delta \right).\]

We can now interpret \(\Omega\) as the population size at \(t+1\),
\(N_{t+1}\) and note that \(N\beta\), \(N\delta\), and \(N\) are the
total births, total deaths, and size of the population at \(t\),
respectively,

\[N_{t+1} = N_{t} + Births - Deaths.
\tag{2}
\]

If we define \(\lambda = 1 + \beta - \delta\) (Box 1), then we can
rewrite,

\[N_{t+1} = N_{t}\lambda.
\tag{3}
\]

We therefore recover the general equation for population ecology (eqn 2)
and the fundamental property of exponential growth in populations (eqn
3, \protect\hyperlink{ref-Turchin2001}{Turchin 2001}).

\hypertarget{evolutionary-biology}{%
\section{Evolutionary biology}\label{evolutionary-biology}}

Recovering the Price equation requires a few more steps. We start by
defining individual fitness,

\[w_{i} = \beta_{i} - \delta_{i} + 1.
\tag{4}
\]

In this definition, the longevity of the individual matters. An
individual that survives from \(t\) to \(t + 1\) has a higher fitness
than one that dies, even if both have the same reproductive output. With
this definition of fitness (eqn 4), we substitute,

\[\Omega = \sum_{i=1}^{N} \left(w_{i}z_{i} + w_{i}\Delta z_{i} \right).
\tag{5}
\]

We can break eqn 5 down further and multiply each side by \(1/N\),

\[\frac{1}{N}\Omega = \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}z_{i} \right) + \frac{1}{N}\sum_{i=1}^{N}\left( w_{i}\Delta z_{i} \right).
\tag{6}
\]

We can rewrite the terms on the right-hand side of eqn 6 as expected
values and remove the subscripts,

\[\frac{1}{N}\Omega = \mathrm{E}\left(w z \right) + \mathrm{E}\left( w \Delta z  \right).
\tag{7}
\]

Now we must consider the total conservation of probability
(\protect\hyperlink{ref-Frank2015}{Frank 2015},
\protect\hyperlink{ref-Frank2016}{2016}). In eqn 7, \(\Omega\) is the
total sum trait values (\(z_{i}\)) across the entire population at
\(t + 1\) divided by the number of individuals (\(N\)) in the population
at \(t\). But the size of the population can change from \(t\) to
\(t + 1\). To recover mean trait change for the Price equation (and
therefore conserve total probability), we need to account for this
change in population size. The left-hand side of eqn 7 describes
contributions to the sum trait value from the new population at
\(t + 1\). But we cannot treat \(\Omega/N\) as the mean of \(z\) at
\(t+1\) (\(\bar{z}'\)) because we need to weigh \(N\) by the mean
fitness of the population at \(t\) to account for any change in
population size from \(t\) to \(t+1\).

For example, if mean fitness at \(t\) is 2 (i.e., \(\bar{w} = 2\)), then
half as many individuals will have contributed to \(\Omega\) in
\(t + 1\) than would have if \(\bar{w} = 1\) (i.e., there are \(N\)
individuals at \(t\) and \(2N\) individuals at \(t + 1\)). We therefore
need to multiply the mean trait value \(\bar{z}'\) (at \(t + 1\)) by the
mean fitness \(\bar{w}\) (at \(t\)) to recover the mean contribution of
the \(N\) individuals at \(t\) to the total \(\Omega\). Consequently,

\[\Omega = N\bar{w}\bar{z}'
\tag{8}
\]

Equation 8 conserves the total probability and now clarifies \(\Omega\)
as a summed trait value, which equals expected population growth at
\(t\) times mean trait value at \(t + 1\). This is consistent with the
population ecology derivation from the previous section where
\(z_{i} = 1\) by definition, and \(\Omega = N_{t+1}\) (note
\(\lambda = \bar{w}\)). We can therefore rewrite eqn 7,

\[\bar{w}\bar{z}' = \mathrm{E}\left(w z \right) + \mathrm{E}\left( w \Delta z  \right).
\tag{9}
\]

We can rearrange eqn 9 to derive the Price equation by expressing
covariance as,
\(\mathrm{Cov}(X,Y) = \mathrm{E}(XY) - \mathrm{E}(X)\mathrm{E}(Y)\), and
therefore
\(\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)\).
Substituting into eqn 9,

\[\bar{w}\bar{z}' = \mathrm{Cov}\left(w ,z \right) + \bar{w}\bar{z} + \mathrm{E}\left( w \Delta z  \right).\]

From here,

\[\bar{w}\left(\bar{z}' - \bar{z}\right) = \mathrm{Cov}\left(w ,z \right) + \mathrm{E}\left( w \Delta z  \right).\]

Since \(\Delta \bar{z} = \left(\bar{z}' - \bar{z}\right)\),

\[\bar{w}\Delta \bar{z} = \mathrm{Cov}\left(w ,z \right) + \mathrm{E}\left( w \Delta z  \right).
\tag{10}
\]

From eqn 1, which describes fundamental birth and death processes in a
population, we can derive both the most fundamental model of population
ecology (eqn 2; Box 1) and the fundamental equation of evolution (eqn
10; Box 2). We might also consider \(\Omega\) to be a quantity of
ecosystem function, in that it is the summed effect of the traits
(\(z_{i}\)) of individual entities (\(i\)) in the population. For
example, \(z\) might be interpreted as biomass or photosynthetic rate.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

A classical sign of scientific progress is the ability to connect
disparate theories and models to show how empirical and theoretical
models are logical (mathematical) consequences of more fundamental ones
(\protect\hyperlink{ref-Nagel1961}{Nagel 1961}). Conceptual unification
therefore has a critical role in advancing scientific theory
(\protect\hyperlink{ref-Morrison2000}{Morrison 2000}). Showing how
apparently disparate phenomena follow from the same shared principles
reveals what is most fundamental about the natural world and provides a
foundation for further scientific investigation and the construction of
more coherent and predictive models. Our eqn 1 provides a foundation for
the conceptual unification of the fundamental equations of population
ecology and evolution.

Rather than making simplifying assumptions, as is the approach for most
ecological and evolutionary models, we focus on fundamental axioms that
are universal to closed biological systems: discrete individuals, birth,
death, and change over time. We define an abstract sum (\(\Omega\)), to
which all individuals within the population contribute. From the simple
assumptions of population identity (\(z_{i} = 1\)) and invariability
(\(\Delta z_{i} = 0\)), we recover the most general equation of
population ecology (Box 1) and principle of exponential growth
(\(N_{t+1} = \lambda N_{t}\)). By defining individual fitness
(\(w_{i}\)) and applying the total conservation of probability to
individual frequencies (\protect\hyperlink{ref-Frank2015}{Frank 2015},
\protect\hyperlink{ref-Frank2016}{2016}), we recover the most
fundamental equation of evolution (Box 2). We therefore deliver a
general framework to unify fundamental equations of ecological and
evolutionary change.

The Price equation provides a complete and exact description of
evolution in any closed evolving system
(\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). It is derived by
rearranging the mathematical notation defining changes in the
frequencies and characteristics of any type of entity (e.g.,
individuals, alleles, \protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Luque2016}{Luque 2017}). This derivation
partitions total characteristic change into different components, making
it possible to isolate evolutionary mechanisms (e.g., selection) and
levels of biological organisation (e.g., group, individual,
\protect\hyperlink{ref-Frank1995}{Frank 1995},
\protect\hyperlink{ref-Frank2012}{2012};
\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Okasha2020}{Okasha and Otsuka 2020}). Because of
its abstract nature and lack of any system-specific assumptions, the
Price equation makes no predictions about what will happen in any
particular system (\protect\hyperlink{ref-Gardner2020}{Gardner 2020}).
Its role is not to predict, but to formally and completely define and
separate components of evolutionary change. The same is true of the
general equation for population change (eqns 2 and 3), at least as we
have used it here where it serves to define what population growth means
in ecology. This equation formally and completely describes population
change in terms of births and deaths. In eqn 1, we therefore have a
fundamental equation from which we can derive complete ecological and
evolutionary change in any closed biological population. We anticipate
that this will be useful for eco-evolutionary theory in the same way
that the Price equation is useful for evolutionary theory: facilitating
specific model development and identifying new conceptual insights,
unresolved errors, and sources of model disagreements.

Our unification demonstrates the equivalence between the finite rate of
increase \(\lambda\) (Box 1, \protect\hyperlink{ref-Gotelli2001}{Gotelli
2001}) and population mean evolutionary fitness \(\bar{w}\) (Box 2). The
population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
rewritten as \(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first
to notice this equivalence between population growth and mean population
fitness. Indeed, the relationship between population growth rate and
evolutionary fitness has been proposed and applied many times before
(e.g., \protect\hyperlink{ref-Fisher1930}{Fisher 1930};
\protect\hyperlink{ref-Charlesworth1980}{Charlesworth 1980};
\protect\hyperlink{ref-Lande1982}{Lande 1982};
\protect\hyperlink{ref-Roff2008}{Roff 2008};
\protect\hyperlink{ref-Lion2018}{Lion 2018}). For example, Lande
(\protect\hyperlink{ref-Lande1982}{1982}) explicitly concludes mean
absolute fitness per unit time is \(\bar{w} = e^{r}\), where
\(r = ln(\lambda)\). We show this from first principles and clarify the
relationship between fitness and population growth. Over an arbitrary
length of time, fitness is properly defined as
\(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an individual's lifetime
(which, by definition, includes death), fitness is therefore
\(\beta_{i}\). Interestingly, the rate of change in ecology and
evolution are reflected in the first and second statistical moments of
fitness, respectively. Population growth rate reflects mean fitness
\(\bar{w}\), while the rate of evolutionary change reflects the variance
in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem,
\protect\hyperlink{ref-Frank1997}{Frank 1997};
\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Queller2017}{Queller 2017}).

Our unification may also help explain, at least partially, some of the
success of classical population genetic models. For decades, population
genetics (and to some extent quantitative genetics) has been accused of
being a reductionistic view of evolution, reducing everything to changes
in allele frequencies and abstracting away from individuals and their
environments (e.g., the ecological interactions,
\protect\hyperlink{ref-MacColl2011}{MacColl 2011}). This has been a line
of argumentation by some defenders of the so-called Extended Synthesis
(\protect\hyperlink{ref-Pigliucci2009}{Pigliucci 2009}), especially in
relation to niche construction
(\protect\hyperlink{ref-Odling-smee2003}{Odling-Smee, Laland, and
Feldman 2003}). Famously, Mayr (\protect\hyperlink{ref-Mayr1960}{1959})
characterized population genetics as a simple input and output of genes,
analogous to ``the adding of certain beans to a beanbag and the
withdrawing of others'' (also called ``beanbag genetics'').

Historical critics of population genetics could not articulate a clear
explanation for why it works so well despite all of its idealisations
and simplifications. From the Price equation, we are able to recover
classical population and quantitative genetic models
(\protect\hyperlink{ref-Queller2017}{Queller 2017}) and develop new ones
(\protect\hyperlink{ref-Rice2004}{Rice 2004},
\protect\hyperlink{ref-Rice2020}{2020};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Lion2018}{Lion 2018}). Our eqn 1 contains ecology
at its core, and we show how the Price equation logically follows from
it after accounting for absolute population growth (eqn 8). We therefore
conclude that population and quantitative genetic equations contain
ecology (no matter how hidden), and the ecological nature of evolution
is implicit in population and quantitative genetic models.

We have focused on the dynamics of a closed population, and in doing so
leave ecological and evolutionary change attributable to migration for
future work. In population ecology, immigration and emigration can be
incorporated by adding a term for each to the right-hand side of the
equation in Box 1 (\protect\hyperlink{ref-Gotelli2001}{Gotelli 2001}).
In evolution, because the Price equation relies on mapping
ancestor-descendant relationships, accounting for migration is more
challenging. Kerr and Godfrey-Smith
(\protect\hyperlink{ref-Kerr2009}{2009}) demonstrate how the Price
equation can be extended to allow for arbitrary links between ancestors
and descendants, thereby extending the Price equation to allow for
immigration and emigration. Frank
(\protect\hyperlink{ref-Frank2012}{2012}) presents a simplified version
of Kerr and Godfrey-Smith (\protect\hyperlink{ref-Kerr2009}{2009}) that
allows some fraction of descendants to be unconnected to ancestors. In
both ecology and evolution, accounting for migration is done through the
use of additional terms on the right-hand side of the equations.

Our fundamental equation is complete and exact. It therefore implicitly
includes any effects of density dependence on population growth (see Box
1), or any social effects on evolutionary change (see Box 2). We can
make both of these effects explicit with the same partition in the
summation on the right-hand side of eqn 1. To do this, we multiply the
expression
\(\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\)
by
\(\left(1 - \sum_{j=1}^{N} a_{ij}\left(z_{j} + \Delta z_{j}\right)\right)\),
where \(j\) indexes the same individuals as \(i\), and \(a_{ij}\)
defines the effect \(j\) has on the contribution of \(i\) to \(\Omega\).
In evolutionary terms, \(a_{ij}\) defines how \(j\) modulates the
fitness of \(i\) (\(w_{i}\)). In ecological terms, \(a_{ij}\) defines
how \(j\) modulates the contribution of \(i\) to total population
growth. If we define \(\alpha\) as the summed effect of all individuals
on a focal \(i\) (i.e., \(\alpha = \sum_{j=1}^{N} a_{ij}\)), then we can
partition ecological change into density-independent and
density-dependent effects and derive a discrete time logistic growth
equation. We save these derivations for a future investigations focused
on evolutionary and community ecology.

We have shown that we can derive the fundamental equations of population
ecology and biological evolution from a single unifying equation.
Lastly, we propose our eqn 1 as a potential starting point for defining
ecosystem function and further conceptual unification between ecology,
evolution, and ecosystem function. The Price equation has previously
been used to investigate ecosystem function
(\protect\hyperlink{ref-Loreau2001}{Loreau and Hector 2001};
\protect\hyperlink{ref-Fox2006}{Fox 2006}), but not with any attempt
towards conceptual unification with evolutionary biology. For example,
Fox (\protect\hyperlink{ref-Fox2006}{2006}) applied the abstract
properties of the Price equation to partition total change in ecosystem
function into separate components attributable to species richness,
species composition, and context dependent effects. This approach
provides a framework for comparing the effects of biodiversity on
ecosystem function in empirical systems (e.g.,
\protect\hyperlink{ref-Fox2006}{Fox 2006};
\protect\hyperlink{ref-Winfree2015}{Winfree et al. 2015};
\protect\hyperlink{ref-Mateo-Tomas2017}{Mateo-Tomás et al. 2017}).
Instead, our eqn 1 defines \(\Omega\) as total ecosystem function
contributed by a focal population (e.g., biomass production,
decomposition, photosynthetic rate). It is therefore possible to
investigate ecological, evolutionary, and ecosystem function change from
the same shared framework.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

This manuscript was supported by joint funding between the French
Foundation for Research on Biodiversity (FRB) Centre for the Synthesis
and Analysis of Biodiversity (CESAB) and the German Centre for
Integrative Biodiversity Research (sDiv). It was written as part of the
Unification of Modern Coexistence Theory and Price Equation (UNICOP)
project. Victor J. Luque was also supported by the Spanish Ministry of
Science and Innovation (Project: PID2021-128835NB-I00), and the
Conselleria d'Innovaci\(\'{o}\), Universitats, Ci\(\`{e}\)ncia i
Societat Digital -- Generalitat Valenciana (Project: CIGE/2023/16). We
are grateful for many conversations with S\(\'{e}\)bastien Lion, Kelsey
Lyberger, Swati Patel, and especially Lynn Govaert, whose questions
helped us clarify the relationship between population growth and
fitness. Brad Duthie would also like to thank Brent Danielson and Stan
Harpole. Victor J. Luque would also like to thank Lorenzo Baravalle, Pau
Carazo, Santiago Ginnobili, Manuel Serra, and Ariel Roff\(\'{e}\).

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Baravalle2022}{}}%
Baravalle, Lorenzo, and Victor J. Luque. 2022. {``Towards a Pricean
Foundation for Cultural Evolutionary Theory.''} \emph{Theoria} 37 (2):
209--31. \url{https://doi.org/10.1387/theoria.21940}.

\leavevmode\vadjust pre{\hypertarget{ref-Baravalle2024}{}}%
Baravalle, Lorenzo, Ariel Roffé, Victor J. Luque, and Santiago
Ginnobili. 2024. {``The Value of Price.''} In \emph{Biological Theory}.
\url{https://doi.org/10.1007/s13752-024-00482-4}.

\leavevmode\vadjust pre{\hypertarget{ref-Brantingham2007}{}}%
Brantingham, P. Jeffrey. 2007. {``{A unified evolutionary model of
archaeological style and function based on the Price equation}.''}
\emph{American Antiquity} 72 (3): 395--416.
\url{https://doi.org/10.2307/40035853}.

\leavevmode\vadjust pre{\hypertarget{ref-Caswell1988}{}}%
Caswell, Hal. 1988. {``Theory and Models in Ecology: A Different
Perspective.''} \emph{Ecological Modelling} 43 (1-2): 33--44.
\url{https://doi.org/10.1016/0304-3800(88)90071-3}.

\leavevmode\vadjust pre{\hypertarget{ref-Charlesworth1980}{}}%
Charlesworth, Brian. 1980. \emph{Evolution in Age-Structured
Populations}. Cambridge Studies in Mathematical Biology. Cambridge:
Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Collins2009}{}}%
Collins, Sinéad, and Andy Gardner. 2009. {``{Integrating physiological,
ecological and evolutionary change: A Price equation approach}.''}
\emph{Ecology Letters} 12 (8): 744--57.
\url{https://doi.org/10.1111/j.1461-0248.2009.01340.x}.

\leavevmode\vadjust pre{\hypertarget{ref-Connor2004}{}}%
Connor, J., and Daniel L. Hartl. 2004. \emph{{A premier of ecological
genetics}}. Sinauer Associates Incorporated.

\leavevmode\vadjust pre{\hypertarget{ref-Dobzhansky1970}{}}%
Dobzhansky, Theodosius. 1970. \emph{Genetics of the Evolutionary
Process}. Vol. 139. Columbia University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Edelaar2023}{}}%
Edelaar, Pim, Jun Otsuka, and Victor J. Luque. 2023. {``{A generalised
approach to the study and understanding of adaptive evolution}.''}
\emph{Biological Reviews} 98 (1): 352--75.
\url{https://doi.org/10.1111/brv.12910}.

\leavevmode\vadjust pre{\hypertarget{ref-Fisher1930}{}}%
Fisher, Ronald A. 1930. \emph{The Genetical Theory of Natural
Selection}. Oxford University Press, Oxford, UK.

\leavevmode\vadjust pre{\hypertarget{ref-Fox2006}{}}%
Fox, Jeremy W. 2006. {``{Using the price equation to partition the
effects of biodiversity loss on ecosystem function}.''} \emph{Ecology}
87 (11): 2687--96.
\url{https://doi.org/10.1890/0012-9658(2006)87\%5B2687:utpetp\%5D2.0.co;2}.

\leavevmode\vadjust pre{\hypertarget{ref-Frank1995}{}}%
Frank, Steven A. 1995. {``{George Price's contributions to evolutionary
genetics}.''} \emph{Journal of Theoretical Biology} 175: 373--88.
\url{https://doi.org/10.1006/jtbi.1995.0148}.

\leavevmode\vadjust pre{\hypertarget{ref-Frank1997}{}}%
---------. 1997. {``{The Price equation, Fisher's fundamental theorem,
kin selection, and causal analysis}.''} \emph{Evolution} 51 (6):
1712--29. \url{https://doi.org/10.1111/j.1558-5646.1997.tb05096.x}.

\leavevmode\vadjust pre{\hypertarget{ref-Frank2012}{}}%
---------. 2012. {``{Natural selection. IV. The Price equation}.''}
\emph{Journal of Evolutionary Biology} 25: 1002--19.
\url{https://doi.org/10.1111/j.1420-9101.2012.02498.x}.

\leavevmode\vadjust pre{\hypertarget{ref-Frank2015}{}}%
---------. 2015. {``{D'Alembert's direct and inertial forces acting on
populations: The price equation and the fundamental theorem of natural
selection}.''} \emph{Entropy} 17 (10): 7087--7100.
\url{https://doi.org/10.3390/e17107087}.

\leavevmode\vadjust pre{\hypertarget{ref-Frank2016}{}}%
---------. 2016. {``{Common probability patterns arise from simple
invariances}.''} \emph{Entropy} 18 (5): 1--22.
\url{https://doi.org/10.3390/e18050192}.

\leavevmode\vadjust pre{\hypertarget{ref-Frank2015a}{}}%
---------. 2017. {``{Universal expressions of population change by the
Price equation: natural selection, information, and maximum entropy
production}.''} \emph{Ecology and Evolution}, no. February: 1--16.
\url{https://doi.org/10.1002/ece3.2922}.

\leavevmode\vadjust pre{\hypertarget{ref-Gardner2008}{}}%
Gardner, Andy. 2008. {``{The Price equation}.''} \emph{Current Biology}
18 (5): 198--202. \url{https://doi.org/10.1016/j.cub.2008.01.005}.

\leavevmode\vadjust pre{\hypertarget{ref-Gardner2020}{}}%
---------. 2020. {``{Price's equation made clear}.''}
\emph{Philosophical Transactions of the Royal Society B: Biological
Sciences} 375 (1797): 20190361.
\url{https://doi.org/10.1098/rstb.2019.0361}.

\leavevmode\vadjust pre{\hypertarget{ref-Gayon2019}{}}%
Gayon, Jean, and Philippe Huneman. 2019. {``The Modern Synthesis:
Theoretical or Institutional Event?''} \emph{Journal of the History of
Biology} 52 (4): 519--35.
\url{https://doi.org/10.1007/s10739-019-09569-2}.

\leavevmode\vadjust pre{\hypertarget{ref-Godsoe2021}{}}%
Godsoe, William, Katherine E. Eisen, Daniel Stanton, and Katherine M.
Sirianni. 2021. {``Selection and Biodiversity Change.''}
\emph{Theoretical Ecology} 14 (3): 367--79.
\url{https://doi.org/10.1007/s12080-020-00478-3}.

\leavevmode\vadjust pre{\hypertarget{ref-Gotelli2001}{}}%
Gotelli, Nicholas J. 2001. {``A Primer of Ecology. Sinauer Associate.''}
\emph{Inc. Sunderland, MA}.

\leavevmode\vadjust pre{\hypertarget{ref-Govaert2019}{}}%
Govaert, Lynn, Emanuel A. Fronhofer, Sébastien Lion, Christophe
Eizaguirre, Dries Bonte, Martijn Egas, Andrew P. Hendry, et al. 2019.
{``{Eco-evolutionary feedbacks---Theoretical models and
perspectives}.''} \emph{Functional Ecology} 33 (1): 13--30.
\url{https://doi.org/10.1111/1365-2435.13241}.

\leavevmode\vadjust pre{\hypertarget{ref-Kerr2009}{}}%
Kerr, Benjamin, and Peter Godfrey-Smith. 2009. {``{Generalization of the
price equation for evolutionary change}.''} \emph{Evolution} 63 (2):
531--36. \url{https://doi.org/10.1111/j.1558-5646.2008.00570.x}.

\leavevmode\vadjust pre{\hypertarget{ref-Kitcher1993}{}}%
Kitcher, Philip. 1993. \emph{{The advancement of science}}. New York:
Oxford University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Lande1982}{}}%
Lande, Russell. 1982. {``{A quantitative genetic theory of life history
evolution}.''} \emph{Ecology} 63 (3): 607--15.
\url{https://doi.org/10.2307/1936778}.

\leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2016}{}}%
Lehtonen, Jussi. 2016. {``{Multilevel selection in kin selection
language}.''} \emph{Trends in Ecology and Evolution} xx: 1--11.
\url{https://doi.org/10.1016/j.tree.2016.07.006}.

\leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2018}{}}%
---------. 2018. {``{The Price equation, gradient dynamics, and
continuous trait game theory}.''} \emph{American Naturalist} 191 (1):
146--53. \url{https://doi.org/10.1086/694891}.

\leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2020}{}}%
Lehtonen, Jussi, Samir Okasha, and Heikki Helanterä. 2020. {``{Fifty
years of the Price equation}.''} \emph{Philosophical Transactions of the
Royal Society B: Biological Sciences} 375 (1797): 20190350.
\url{https://doi.org/10.1098/rstb.2019.0350}.

\leavevmode\vadjust pre{\hypertarget{ref-Lewens2015}{}}%
Lewens, Tim. 2015. \emph{Cultural Evolution: Conceptual Challenges}.
Oxford University Press.
\url{https://doi.org/10.1093/acprof:oso/9780199674183.001.0001}.

\leavevmode\vadjust pre{\hypertarget{ref-Lion2018}{}}%
Lion, Sébastien. 2018. {``{Theoretical approaches in evolutionary
ecology: environmental feedback as a unifying perspective}.''}
\emph{American Naturalist} 191 (1).
\url{https://doi.org/10.1086/694865}.

\leavevmode\vadjust pre{\hypertarget{ref-Loreau2001}{}}%
Loreau, Michel, and Andy Hector. 2001. {``{Partitioning selection and
complementarity in biodiversity experiments}.''} \emph{Nature} 412:
72--76. \url{https://doi.org/10.1038/35083573}.

\leavevmode\vadjust pre{\hypertarget{ref-Luque2016}{}}%
Luque, Victor J. 2017. {``{One equation to rule them all: a
philosophical analysis of the Price equation}.''} \emph{Biology and
Philosophy} 32 (1): 1--29.
\url{https://doi.org/10.1007/s10539-016-9538-y}.

\leavevmode\vadjust pre{\hypertarget{ref-Luque2021}{}}%
Luque, Victor J., and Lorenzo Baravalle. 2021. {``{The mirror of
physics: on how the price equation can unify evolutionary biology}.''}
\emph{Synthese} 199: 12439--62.
\url{https://doi.org/10.1007/s11229-021-03339-6}.

\leavevmode\vadjust pre{\hypertarget{ref-MacCallum2012}{}}%
MacCallum, Robert M., Matthias Mauch, Austin Burt, and Armand M. Leroi.
2012. {``{Evolution of music by public choice}.''} \emph{Proceedings of
the National Academy of Sciences} 109 (30): 12081--86.
\url{https://doi.org/10.5061/dryad.h0228}.

\leavevmode\vadjust pre{\hypertarget{ref-MacColl2011}{}}%
MacColl, Andrew D. C. 2011. {``{The ecological causes of evolution}.''}
\emph{Trends in Ecology and Evolution} 26 (10): 514--22.
\url{https://doi.org/10.1016/j.tree.2011.06.009}.

\leavevmode\vadjust pre{\hypertarget{ref-Mateo-Tomas2017}{}}%
Mateo-Tomás, Patricia, Pedro P. Olea, Marcos Moleón, Nuria Selva, and
José A. Sánchez-Zapata. 2017. {``{Both rare and common species support
ecosystem services in scavenger communities}.''} \emph{Global Ecology
and Biogeography} 26 (12): 1459--70.
\url{https://doi.org/10.1111/geb.12673}.

\leavevmode\vadjust pre{\hypertarget{ref-Mayr1960}{}}%
Mayr, Ernst. 1959. {``Where Are We? Genetics and Twentieth Century
Darwinism.''} In \emph{Cold Spring Harbor Symposia on Quantitative
Biology}, 24:1--14. \url{https://doi.org/10.1101/SQB.1959.024.01.003}.

\leavevmode\vadjust pre{\hypertarget{ref-Morrison2000}{}}%
Morrison, Margaret. 2000. \emph{Unifying Scientific Theories: Physical
Concepts and Mathematical Structures}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Nagel1961}{}}%
Nagel, Ernest. 1961. \emph{The Structure of Science: Problems in the
Logic of Scientific Explanation}. New York, NY, USA: Harcourt, Brace \&
World.

\leavevmode\vadjust pre{\hypertarget{ref-Odling-smee2003}{}}%
Odling-Smee, F. John, Kevin N. Laland, and Marcus W. Feldman. 2003.
\emph{Niche Construction: The Neglected Process in Evolution}. Princeton
University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Okasha2006}{}}%
Okasha, Samir. 2006. \emph{Evolution and the Levels of Selection}.
Oxford University Press.
\url{https://doi.org/10.1093/acprof:oso/9780199267972.001.0001}.

\leavevmode\vadjust pre{\hypertarget{ref-Okasha2020}{}}%
Okasha, Samir, and Jun Otsuka. 2020. {``{The Price equation and the
causal analysis of evolutionary change}.''} \emph{Philosophical
Transactions of the Royal Society B: Biological Sciences} 375 (1797):
20190365. \url{https://doi.org/10.1098/rstb.2019.0365}.

\leavevmode\vadjust pre{\hypertarget{ref-Page2002}{}}%
Page, Karen M., and Martin A. Nowak. 2002. {``Unifying Evolutionary
Dynamics.''} \emph{Journal of Theoretical Biology} 219 (1): 93--98.
\url{https://doi.org/10.1006/jtbi.2002.3112}.

\leavevmode\vadjust pre{\hypertarget{ref-Patel2018}{}}%
Patel, Swati, Michael H. Cortez, and Sebastian J. Schreiber. 2018.
{``{Partitioning the effects of eco-evolutionary feedbacks on community
stability}.''} \emph{American Naturalist} 191 (3): 1--29.
\url{https://doi.org/10.1101/104505}.

\leavevmode\vadjust pre{\hypertarget{ref-Pigliucci2009}{}}%
Pigliucci, Massimo. 2009. {``{An extended synthesis for evolutionary
biology}.''} \emph{Annals of the New York Academy of Sciences} 1168:
218--28. \url{https://doi.org/10.1111/j.1749-6632.2009.04578.x}.

\leavevmode\vadjust pre{\hypertarget{ref-Price1970}{}}%
Price, George R. 1970. {``{Selection and covariance}.''}
\url{https://doi.org/10.1038/227520a0}.

\leavevmode\vadjust pre{\hypertarget{ref-Price1972}{}}%
---------. 1972. {``{Extension of covariance selection mathematics}.''}
\emph{Annals of Human Genetics} 35 (4): 485--90.
\url{https://doi.org/10.1111/j.1469-1809.1957.tb01874.x}.

\leavevmode\vadjust pre{\hypertarget{ref-Price1995}{}}%
---------. 1995. {``{The nature of selection}.''} \emph{Journal of
Theoretical Biology} 175 (3): 389--96.
\url{https://doi.org/10.1006/jtbi.1995.0149}.

\leavevmode\vadjust pre{\hypertarget{ref-Queller2017}{}}%
Queller, David C. 2017. {``{Fundamental theorems of evolution}.''}
\emph{American Naturalist} 189 (4): 000--000.
\url{https://doi.org/10.1086/690937}.

\leavevmode\vadjust pre{\hypertarget{ref-Rice2004}{}}%
Rice, Sean H. 2004. \emph{{Evolutionary theory: mathematical and
conceptual foundations}}. Sinauer Associates Incorporated.

\leavevmode\vadjust pre{\hypertarget{ref-Rice2020}{}}%
---------. 2020. {``{Universal rules for the interaction of selection
and transmission in evolution}.''} \emph{Philosophical Transactions of
the Royal Society B: Biological Sciences} 375 (1797).
\url{https://doi.org/10.1098/rstb.2019.0353}.

\leavevmode\vadjust pre{\hypertarget{ref-Rice2009}{}}%
Rice, Sean H., and Anthony Papadopoulos. 2009. {``{Evolution with
stochastic fitness and stochastic migration}.''} \emph{PLoS One} 4 (10).
\url{https://doi.org/10.1371/journal.pone.0007130}.

\leavevmode\vadjust pre{\hypertarget{ref-Robertson1966}{}}%
Robertson, Alan. 1966. {``{A mathematical model of the culling process
in dairy cattle}.''} \emph{Animal Science} 8 (1): 95--108.
\url{https://doi.org/10.1017/S0003356100037752}.

\leavevmode\vadjust pre{\hypertarget{ref-Roff2008}{}}%
Roff, Derek A. 2008. {``{Defining fitness in evolutionary models}.''}
\emph{Journal of Genetics} 87 (4): 339--48.
\url{https://doi.org/10.1007/s12041-008-0056-9}.

\leavevmode\vadjust pre{\hypertarget{ref-Shennan2011}{}}%
Shennan, Stephen. 2011. {``{Descent with modification and the
archaeological record}.''} \emph{Philosophical Transactions of the Royal
Society B: Biological Sciences} 366 (1567): 1070--79.
\url{https://doi.org/10.1098/rstb.2010.0380}.

\leavevmode\vadjust pre{\hypertarget{ref-Turchin2001}{}}%
Turchin, Peter. 2001. {``Does Population Ecology Have General Laws?''}
\emph{Oikos} 94 (1): 17--26.
\url{https://doi.org/10.1034/j.1600-0706.2001.11310.x}.

\leavevmode\vadjust pre{\hypertarget{ref-Ulrich2024}{}}%
Ulrich, Werner, Nicholas J. Gotelli, Giovanni Strona, and William
Godsoe. 2024. {``Reconsidering the Price Equation: Benchmarking the
Analytical Power of Additive Partitioning in Ecology.''}
\emph{Ecological Modelling} 491: 110695.
\url{https://doi.org/10.1016/j.ecolmodel.2024.110695}.

\leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2005}{}}%
van Veelen, Matthijs. 2005. {``{On the use of the Price equation}.''}
\emph{Journal of Theoretical Biology} 237 (4): 412--26.
\url{https://doi.org/10.1016/j.jtbi.2005.04.026}.

\leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2020}{}}%
---------. 2020. {``{The problem with the Price equation}.''}
\emph{Philosophical Transactions of the Royal Society B: Biological
Sciences} 375 (1797): 20190355.
\url{https://doi.org/10.1098/rstb.2019.0355}.

\leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2012}{}}%
van Veelen, Matthijs, Julián García, Maurice W. Sabelis, and Martijn
Egas. 2012. {``{Group selection and inclusive fitness are not
equivalent; the Price equation vs. models and statistics}.''}
\emph{Journal of Theoretical Biology} 299 (April): 64--80.
\url{https://doi.org/10.1016/j.jtbi.2011.07.025}.

\leavevmode\vadjust pre{\hypertarget{ref-Winfree2015}{}}%
Winfree, Rachael, Jeremy W. Fox, Neal M. Williams, James R. Reilly, and
Daniel P. Cariveau. 2015. {``{Abundance of common species, not species
richness, drives delivery of a real-world ecosystem service}.''}
\emph{Ecology Letters} 18: 626--35.
\url{https://doi.org/10.1111/ele.12424}.

\end{CSLReferences}

\end{document}

### 2024-12-20 foundations_talk.Rmd
Initialise all of the relevant files

---
title: "Foundations of ecological and evolutionary change"
author: "Brad Duthie and Victor Luque"
output:
  beamer_presentation:
    theme: "default"
    colortheme: "default"
    fonttheme: "default"
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \usepackage{hyperref}
- \usepackage{caption}
- \usepackage[normalem]{ulem}
- \definecolor{links}{HTML}{2A1B81}
- \hypersetup{colorlinks,linkcolor=,urlcolor=links}
colorlinks: true
linkcolor: blue
urlcolor: blue
---

```{r, echo = FALSE}
library(knitr);
opts_chunk$set(echo = FALSE);
```

## Why people do science, particularly biology

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/magnifying.jpg");
```

\footnotetext[1]{Image: Gillette, B. 1972. \href{https://commons.wikimedia.org/wiki/File:GIRL_USES_A_MAGNIFYING_GLASS_TO_STUDY_PLANT_LIFE_IN_THE_TUNDRA_OF_THE_ROCKY_MOUNTAINS._THE_DENVER_PTA_SPONSORED_A..._-_NARA_-_543740.jpg}{Public Domain}}

## Unification of celestial and terrestrial motion

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/solar.jpg");
```

\footnotetext[1]{Image: Smith, H and L Generosa (2006). \href{https://commons.wikimedia.org/wiki/File:Solar_sys8.jpg}{Public Domain}}

## Complexity of the biological sciences

```{r, out.width = "90%", echo = FALSE, fig.cap = ""}
include_graphics("img/bee.jpg");
```

\footnotetext[1]{Image: John Severns (2006). \href{https://commons.wikimedia.org/wiki/File:European_honey_bee_extracts_nectar.jpg}{Public Domain}}

## Complexity of the biological sciences

```{r, out.width = "90%", echo = FALSE, fig.cap = ""}
include_graphics("img/food_web.png");
```

\footnotetext[1]{Darnis, G, et al. (2012). Current state and trends in Canadian Arctic marine ecosystems: II. Heterotrophic food web, pelagic-benthic coupling, and biodiversity. Climatic Change, 115, 179-205.}
\footnotetext[1]{Image: Darnis, G, et al. (2012). \href{https://commons.wikimedia.org/wiki/File:Arctic_Marine_Food_Web.png}{Public Domain}}

## Unification of evolution and genetics

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{center}
\textbf{Darwinian evolution}
\end{center}

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/ithink.png");
```

\end{column}
\begin{column}{0.5\textwidth}  
\begin{center}
\textbf{Mendelian genetics}
\end{center}

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/punnet_square_2.png");
```

\end{column}
\end{columns}

\footnotetext[1]{Image (left): \href{https://commons.wikimedia.org/wiki/File:Darwin_tree.png}{Public Domain}}
\footnotetext[2]{Image (right): Madeleine Price Ball (2007).  \href{https://commons.wikimedia.org/wiki/File:Punnett_square_mendel_flowers.svg}{Public Domain}}

## Unification of evolution and genetics

\Huge

$$p^{2} + 2pq + q^{2} = 1$$
\vspace{0.5mm}
\hrule
\vspace{1mm}

\Large

\begin{itemize}
\item No natural selection
\item No mutation
\item No migration (no gene flow)
\item Infinite population size
\item Mating is random
\item Non-overlapping generations
\end{itemize}

## General theory of community ecology

\LARGE

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{center}
\textbf{Population genetics}
\vspace{2mm}
\hrule
\vspace{2mm}
\end{center}

\begin{itemize}
\item Selection
\item Mutation
\item Gene flow
\item Genetic drift
\end{itemize}

\end{column}
\begin{column}{0.5\textwidth}  
\begin{center}
\textbf{Community ecology}
\vspace{2mm}
\hrule
\vspace{2mm}
\end{center}

\begin{itemize}
\item Species selection
\item Disperal
\item Drift
\item Speciation
\end{itemize}

\end{column}
\end{columns}
\vspace{3mm}
\hrule

General community ecology theory focused on high-level processes$^{1}$

\footnotetext[1]{Vellend, M (2016). The theory of ecological communities. Monographs in population biology. Princeton University Press.}

## Eco-evolutionary models

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/fig1_b.png");
```

## Eco-evolutionary models

\Large

\begin{columns}
\begin{column}{0.5\textwidth}

Lion$^{1}$ (2018)
\hrule

\vspace{2mm}

\textbf{Ecology:}

$$\frac{dn_{i}}{dt} = r_{i}\left(\mathbf{E}\right)n_{i}$$

\vspace{4mm}
\textbf{Evolution:}

$$\frac{df_{i}}{dt} = f_{i}\left(r_{i} \left(\mathbf{E} \right) - \bar{r} \left(\mathbf{E} \right) \right)$$

\end{column}
\begin{column}{0.5\textwidth}

\pause

Patel et al.$^{2}$ (2018)
\hrule

\vspace{2mm}

\textbf{Ecology:}

$$\frac{dN_{i}}{dt} = N_{i}f_{i}\left(N, x\right)$$

\vspace{2mm}
\textbf{Evolution:}

$$\frac{dx_{j}}{dt} = \epsilon g_{j}\left(N, x\right)$$

\end{column}
\end{columns}

\footnotetext[1]{Lion, S (2018). Theoretical approaches in evolutionary ecology: environmental feedback as a unifying perspective. Am. Nat., 191:21-44}
\footnotetext[2]{Patel, S, Cortez, MH, \& Schreiber, SJ (2018). Partitioning the effects of eco-evolutionary feedbacks on community stability. The Am. Nat., 191:381-394.}

## Conceptual unification of ecology and evolution

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/fig1_b.png");
```

## Conceptual unification of ecology and evolution

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/fig1.png");
```

## Unification of modern coexistence theory \& Price equation

Joint funding call between French Foundation for Research on Biodiversity and German Centre for Integrative Biodiversity Research (iDiv): CESAB-sDIV SYNERGY

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/unicop_team.jpg");
```

The UNICOP team in Montpellier (summer 2023)

## A different role of theory (not doing modelling)

\LARGE

- Not making assumptions or predictions \pause
- Defining population change in an eco-evo context \pause
- Framework for models (meta-modelling) \pause
- A rough analogy,

$$y = \beta_{0} + \beta_{1}x$$

\pause\vspace{5mm}

- Or Newton's second law, $F = ma$

## Newtonian mechanics as a theory net

```{r, out.width = "110%", echo = FALSE, fig.cap = ""}
include_graphics("img/physics.png");
```

\footnotetext[1]{Luque, VJ, \& L Baravalle (2021). The mirror of physics: on how the Price equation can unify evolutionary biology. Synthese, 0123456789.}

## The fundamental equation of evolution

\Huge

$$\bar{w}\Delta\bar{z} = Cov\left(w, z\right) + E\left(w\Delta z\right)$$

\footnotetext[1]{Price, GR (1970). Selection and covariance. Nature 227:520–521.}
\footnotetext[2]{Luque, VJ (2017). One equation to rule them all: a philosophical analysis of the Price equation. Biology and Philosophy, 32:1–29.}
\footnotetext[3]{Lehtonen, J, Okasha, S, \& Helanterä, H (2020). Fifty years of the Price equation. Philosophical Transactions of the Royal Society B: Biological Sciences, 375:20190350.}
\footnotetext[4]{Frank, SA (2012). Natural selection. IV. The Price equation. Journal of Evolutionary Biology, 25, 1002–1019.}
\footnotetext[5]{Lion, S (2018). From the Price equation to the selection gradient in class-structured populations: a quasi-equilibrium route. Journal of Theoretical Biology, 447, 178–189.}

## The fundamental equation of evolution

```{r, out.width = "110%", echo = FALSE, fig.cap = ""}
include_graphics("img/derivation.png");
```

\footnotetext[1]{Luque, VJ (2017). One equation to rule them all: a philosophical analysis of the Price equation. Biology and Philosophy, 32:1–29.}

## The fundamental equation of evolution

```{r, out.width = "110%", echo = FALSE, fig.cap = ""}
include_graphics("img/synthesis.png");
```

\footnotetext[1]{Luque, VJ, \& L Baravalle (2021). The mirror of physics: on how the Price equation can unify evolutionary biology. Synthese, 0123456789.}

## Recursive properties of the Price equation

\Huge

$$\bar{w}\Delta\bar{z} = Cov\left(w, z\right) + E\left(w\Delta z\right)$$

\footnotetext[1]{Price, GR (1970). Selection and covariance. Nature 227:520–521.}
\footnotetext[2]{Kerr, B, \& P Godfrey-Smith (2009). Generalization of the price equation for evolutionary change. Evolution, 63:531–536.}
\footnotetext[3]{Frank, SA (2012). Natural selection. IV. The Price equation. Journal of Evolutionary Biology, 25, 1002–1019.}

## Derive other fundamental equations

```{r, out.width = "90%", echo = FALSE, fig.cap = ""}
include_graphics("img/queller2017.png");
```

\footnotetext[1]{Queller, DC (2017). Fundamental theorems of evolution. The American Naturalist, 189:345-353.}

## Generality of the Price equation

\Large

**Evolution of non-biological systems**

- Archaeological artefacts$^{1}$
- Musical properties$^{2}$

\pause

**Synthesis across disciplines**

- D'Alembert's principle of mechanics$^{3}$
- Information theory$^{4}$

\footnotetext[1]{Brantingham, PJ (2007). A unified evolutionary model of archaeological style and function based on the Price equation. American Antiquity 72:395–416.}
\footnotetext[2]{MacCallum, RM, et al. (2012). Evolution of music by public choice. PNAS 109:12081–12086.}
\footnotetext[3]{Frank, SA (2015). D'Alembert’s direct and inertial forces acting on populations: The price equation and the fundamental theorem of natural selection. Entropy 17:7087–7100.}
\footnotetext[3]{Frank, SA (2012). Natural selection. V. How to read the fundamental equations of evolutionary change in terms of information theory. Journal of Evolutionary Biology, 25:2377-2396.}

## Conceptual unification of ecology and Price equation

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/fig1.png");
```

## Links between ecology and the Price equation

**Ecosystem function loss due to species loss$^{1}$**

\Large

$$\Delta T = \bar{z}\Delta s + Sp(w, z) + \sum_{i}w_{i}\Delta z_{i}$$

\normalsize\pause

**Community trait change: Physiology, evolution, ecology$^{2}$**

\Large

$$\Delta \bar{z} = E_{I}\left(E_{Ji} \left(\Delta z_{ij} \right) \right) + E_{I}\left(cov_{Ji} \left(w_{ij}, z'_{ij} \right) \right) + cov_{I}\left(w_{i}, z'_{i}\right)$$

\footnotetext[1]{Fox, JW (2006). Using the price equation to partition the effects of biodiversity loss on ecosystem function. Ecology, 87(11), 2687-2696.}
\footnotetext[2]{Collins, S \& Gardner, A (2009). Integrating physiological, ecological and evolutionary change: A Price equation approach. Ecology Letters, 12(8), 744-757.}

## Links between ecology and the Price equation

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/Page_Nowak.png");
```

\LARGE

- Figure 1 from Page \& Nowak$^{1}$
- Relies on \textit{relative} species frequenies

\footnotetext[1]{Page, KM \& Nowak, MA (2002). Unifying evolutionary dynamics. Journal of Theoretical Biology, 219(1), 93–98.}

## Reconciling evolutionary change and population growth {.t}

\Large

- Start with Price equation and add ecology?
- Start elsewhere and derive the Price equation?

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/lost_in_woods.jpg");
```

\footnotetext[1]{Image: Eneas De Troya (2006).  \href{https://commons.wikimedia.org/wiki/File:Lost_in_a_Virtual_Wood_(177504871).jpg}{CC BY 2.0}}

## The fundamental equation of ecology

\Huge

- Evolution: Price Equation
- Ecology: Births and deaths

$$N_{t+1} = N_{t} + Births - Deaths$$

\footnotetext[1]{Gotelli, NJ (2001). A Primer of Ecology. Sinauer Associate. Inc. Sunderland, MA.}
\footnotetext[2]{Rockwood, LL (2006). Introduction to population ecology. Blacwell. Malden, MA}
\footnotetext[3]{Turchin, P (2001). Does population ecology have general laws? Oikos, 94(1), 17-26 \href{https://doi.org/10.1034/j.1600-0706.2001.11310.x}{[PDF]}}

## Finite rate of increase for a population {.t}

\Huge

$$N_{t+1} = N_{t}(1 + b - d)$$

$$\lambda = (1 + b - d)$$

$$N_{t+1} = N_{t}\lambda$$

## Finite rate of increase for a population

```{r, echo = FALSE, fig.height = 7, fig.width = 9}
BB    <- 1.05;
No    <- 100;
tt    <- 1:100;
NN    <- rep(x = NA, times = 100);
NN[1] <- No;
for(i in 2:100){
  NN[i] <- NN[i - 1] * BB;
}
par(mar = c(6, 6, 1, 1));
plot(x = 1:100, y = NN, type = "l", lwd = 6, xlab = "Time (t)", 
     ylab = "Population Size (N)",
     cex.axis = 1.5, cex.lab = 2.5, xaxt = "n", yaxt = "n")
```

## Reconciling foundations of evolution and ecology

\Huge

**Evolution:** Price Equation

\vspace{-10mm}

$$\bar{w}\Delta\bar{z} = Cov\left(w, z\right) + E\left(w\Delta z\right)$$

\hrule

\vspace{5mm}

**Ecology:** Exponential growth

\vspace{-10mm}

$$N_{t+1} = N_{t} + Births - Deaths$$

## The unifying equation

\LARGE

- No systematic approach
- Individual birth, death, characteristics
- Deductive exploration

\vspace{5mm}
\hrule
\pause
\vspace{5mm}

\Huge

As expected, the answer is obvious in hindsight.

## A fundamental equation for ecology and evolution {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule

\pause
\LARGE

- $\Omega$: Some summed quantity
- $N$: Population size
- $i$: Individuals in population
- $\beta$: Births attributable to $i$
- $\delta$: Death indicator variable
- $z_{i}$: Character of $i$

## Ecology: Character species identity (z = 1), ($\Delta$z = 0) {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule \pause

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)$$

$$N_{t+1} = N_{t} + Births - Deaths$$

$$N_{t+1} = N_{t} \left(1 + \beta - \delta \right) = N_{t}\lambda$$

## Evolution: Character individual trait ($z_{i}$) {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule \pause

Define fitness, $w_{i} = \beta_{i} - \delta_{i} + 1$. Substitute,

$$\Omega = \sum_{i=1}^{N} \left(w_{i}z_{i} + w_{i}\Delta z_{i} \right)$$

$$\frac{1}{N}\Omega = \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}z_{i}\right) + \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}\Delta z_{i} \right)$$

## Evolution: Character individual trait ($z_{i}$) {.t}

\LARGE

$$\frac{1}{N}\Omega = \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}z_{i}\right) + \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}\Delta z_{i} \right)$$

\vspace{5mm}
\hrule \pause

Rewrite as expected values,

$$\frac{1}{N}\Omega = E\left(wz\right) + E\left(w\Delta z \right).$$

## Evolution: Character individual trait ($z_{i}$) {.t}

\LARGE

$$\frac{1}{N}\Omega = E\left(wz\right) + E\left(w\Delta z \right).$$

\vspace{5mm}
\hrule \pause

Tricky step is to note $\Omega = N\bar{w}\bar{z}'$ because we need to conserve total probability$^{1,2}$, so,

$$\bar{w}\bar{z}' = E\left(wz\right) + E\left(w\Delta z \right).$$

\footnotetext[1]{Frank, Steven A (2015) "D'Alembert's direct and inertial forces acting on populations: The Price equation and the fundamental theorem of natural selection." Entropy 17: 7087–7100.}
\footnotetext[2]{Frank, Steven A (2016) "Common probability patterns arise from simple invariances." Entropy 18 (5): 1–22.}

## Evolution: Character individual trait ($z_{i}$) {.t}

\LARGE

$$\bar{w}\bar{z}' = E\left(wz\right) + E\left(w\Delta z \right).$$

\vspace{5mm}
\hrule \pause

Since $COV(X, Y) = E(XY) - E(X)E(Y)$,

$$\bar{w}\bar{z}' = Cov(w, z) + \bar{w}\bar{z} + E\left(w\Delta z \right).$$

$$\bar{w}\left(\bar{z}' - \bar{z}\right) = Cov(w, z) + E\left(w\Delta z \right).$$

## Evolution: Character individual trait ($z_{i}$) {.t}

\LARGE

$$\bar{w}\left(\bar{z}' - \bar{z}\right) = Cov(w, z) + E\left(w\Delta z \right).$$

\vspace{5mm}
\hrule \pause

Since $\Delta z = \bar{z}' - \bar{z}$,

$$\bar{w}\Delta\bar{z} = Cov(w, z) + E\left(w\Delta z \right).$$

We have therefore derived Price.

## A formal definition of eco-evolutionary change {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule

Formalises ecology \& evolution \pause

- No simplifying assumptions \pause
- Complete and exact (closed populations) \pause
- Ecology: $\beta_{i} - \delta_{i} + 1$ \pause
- Evolution: $z_{i} + \Delta z_{i}$ \pause
- Ecology required for selection or drift

## Confirmations from a formal definition {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule

Formalises fitness and population growth: \pause

- **Fitness:** $\bar{w} = \lambda = \beta - \delta + 1$ \pause
- Derives assumed $N_{t+1} = N_{t}\lambda = N_{t}\bar{w}$ \pause
- Lifetime fitness is still $\beta_{i}$ \pause
- Starting point for further unification \pause
- Build upon for community ecology

## An elegance in two fits {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{2mm}
\hrule

Fisher's fundamental theorem$^{1,2}$ $w = z$, \pause

$$\Delta \bar{w} = \frac{1}{\bar{w}}Var(w)$$

This is because $Cov(w, w) = Var(w)$.

\footnotetext[1]{Frank, SA (1997). "The Price equation, Fisher’s fundamental theorem, kin selection, and causal analysis." Evolution 51:1712–29.}
\footnotetext[1]{Queller, DC (2017). Fundamental theorems of evolution. The American Naturalist, 189:345-353.}

## An elegance in two fits {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule

**What this means is the following:** \pause

- Rate of population growth is the first statistical moment of fitness  \pause
- Rate of adaptive evolution is the second statistical moment of fitness

## Bonus for ecosystem function {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule

We can define z as a trait (e.g., photosynthetic rate, decomposition rate, biomass) and recover $\Omega$ as a population's contribution to ecosystem function.

## Foundations of ecological and evolutionary change {.t}

\LARGE

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$

\vspace{5mm}
\hrule

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/foundations.png");
```

\footnotetext[1]{Duthie, AB, \& Luque, VJ (2024). Foundations of ecological and evolutionary change. arXiv preprint arXiv:2409.10766.}

### 2024-12-20 scratch.Rmd
Initialise all of the relevant files

---
title: "R Notebook"
output: html_notebook
---

@Turchin2001 argues convincingly for exponential (or geometric, if discrete) growth being the first law of population dyanmics.
It certainly is universal, at least as it is expressed in a similar fashion to Newton's laws of motion.
This paper also makes the case for why the logistic growth model is not so universal, and really just a special case.
But the basic $N = B - D$ is clear, and there is a nice counterfactual for spontaneous generation and how its existence would change this fundamental equation.
Basically, in our own equation below, we are making the assumption (without stating it, though we should) that individuals do not spontaneously generate; they must be introduced from existing individuals $i$ through birth.
@Turchin2001 does a good job of explaining why $N = B - D$ is really the best we do.
He distinguishes between postulates (axioms) and predictions (theorems), and he insists that "we need a set of general principles to place populations on a firm logical foundation".
I believe that we need a set of general principles to place population ecology and evolution on *the same* shared foundation.

### 2024-12-20 tuesday_seminar.Rmd
Initialise all of the relevant files

--- 2024-12-20 foundations_talk.Rmd
@@
 ---
-title: "Foundations of ecological and evolutionary change"
-author: "Brad Duthie and Victor Luque"
+title: "The Unicorn Equation: Conceptual unification of ecology, evolution, and ecosystem function: TESTING"
 output:
@@
 
+## BES Tuesday seminars this autumn
+
+\Large
+
+**TEG: LT W1 from 13:00-14:00** (until 26 NOV)
+
+\large
+
+\vspace{5mm}
+
+| Date    |  Speaker           | | Date    | Speaker            |
+|---------|--------------------|-|---------|--------------------|
+| 24 SEP  | Heather Anderson   | | 05 NOV  | Reuben Nowell      |
+| 01 OCT  | **Open**           | | 12 NOV  | Ruth Wright        |
+| 08 OCT  | Matt Tinsley       | | 19 NOV  | Zarah Pattison     |
+| 15 OCT  | Carmen Carmona     | | 26 NOV  | Kirsty Park        |
+| 22 OCT  | Mahtab Yaghouti    | | 03 DEC  | Emily Waddell      |
+| 29 OCT  | Ponsarut Boonchuay | | 10 DEC  | Colin Bull         |
+
 ## Why people do science, particularly biology
@@
 
-## Complexity of the biological sciences
+## Big questions in biology
 
-```{r, out.width = "90%", echo = FALSE, fig.cap = ""}
-include_graphics("img/bee.jpg");
+```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
+include_graphics("img/cvs.png");
 ```
 
-\footnotetext[1]{Image: John Severns (2006). \href{https://commons.wikimedia.org/wiki/File:European_honey_bee_extracts_nectar.jpg}{Public Domain}}
+## Big questions in biology
 
-## Complexity of the biological sciences
-
-```{r, out.width = "90%", echo = FALSE, fig.cap = ""}
-include_graphics("img/food_web.png");
+```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
+include_graphics("img/iowa_state.png");
 ```
 
-\footnotetext[1]{Darnis, G, et al.
-(2012).
-Current state and trends in Canadian Arctic marine ecosystems: II.
-Heterotrophic food web, pelagic-benthic coupling, and biodiversity.
-Climatic Change, 115, 179-205.}
-\footnotetext[1]{Image: Darnis, G, et al. (2012). \href{https://commons.wikimedia.org/wiki/File:Arctic_Marine_Food_Web.png}{Public Domain}}
-
-## Unification of evolution and genetics
-
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{center}
-\textbf{Darwinian evolution}
-\end{center}
+## Call for proposals: joint sDiv-CESAB
 
 ```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/ithink.png");
+include_graphics("img/synergy_call.png");
 ```
 
-\end{column}
-\begin{column}{0.5\textwidth}  
-\begin{center}
-\textbf{Mendelian genetics}
-\end{center}
-
-```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/punnet_square_2.png");
-```
-
-\end{column}
-\end{columns}
-
-\footnotetext[1]{Image (left): \href{https://commons.wikimedia.org/wiki/File:Darwin_tree.png}{Public Domain}}
-\footnotetext[2]{Image (right): Madeleine Price Ball (2007).  \href{https://commons.wikimedia.org/wiki/File:Punnett_square_mendel_flowers.svg}{Public Domain}}
-
-## Unification of evolution and genetics
-
-\Huge
-
-$$p^{2} + 2pq + q^{2} = 1$$
-\vspace{0.5mm}
-\hrule
-\vspace{1mm}
-
-\Large
-
-\begin{itemize}
-\item No natural selection
-\item No mutation
-\item No migration (no gene flow)
-\item Infinite population size
-\item Mating is random
-\item Non-overlapping generations
-\end{itemize}
-
-## General theory of community ecology
-
-\LARGE
-
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{center}
-\textbf{Population genetics}
-\vspace{2mm}
-\hrule
-\vspace{2mm}
-\end{center}
-
-\begin{itemize}
-\item Selection
-\item Mutation
-\item Gene flow
-\item Genetic drift
-\end{itemize}
-
-\end{column}
-\begin{column}{0.5\textwidth}  
-\begin{center}
-\textbf{Community ecology}
-\vspace{2mm}
-\hrule
-\vspace{2mm}
-\end{center}
-
-\begin{itemize}
-\item Species selection
-\item Disperal
-\item Drift
-\item Speciation
-\end{itemize}
-
-\end{column}
-\end{columns}
-\vspace{3mm}
-\hrule
-
-General community ecology theory focused on high-level processes$^{1}$
-
-\footnotetext[1]{Vellend, M (2016). The theory of ecological communities. Monographs in population biology. Princeton University Press.}
-
-## Eco-evolutionary models
-
-```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/fig1_b.png");
-```
-
-## Eco-evolutionary models
-
-\Large
-
-\begin{columns}
-\begin{column}{0.5\textwidth}
-
-Lion$^{1}$ (2018)
-\hrule
-
-\vspace{2mm}
-
-\textbf{Ecology:}
-
-$$\frac{dn_{i}}{dt} = r_{i}\left(\mathbf{E}\right)n_{i}$$
-
-\vspace{4mm}
-\textbf{Evolution:}
-
-$$\frac{df_{i}}{dt} = f_{i}\left(r_{i} \left(\mathbf{E} \right) - \bar{r} \left(\mathbf{E} \right) \right)$$
-
-\end{column}
-\begin{column}{0.5\textwidth}
-
-\pause
-
-Patel et al.$^{2}$ (2018)
-\hrule
-
-\vspace{2mm}
-
-\textbf{Ecology:}
-
-$$\frac{dN_{i}}{dt} = N_{i}f_{i}\left(N, x\right)$$
-
-\vspace{2mm}
-\textbf{Evolution:}
-
-$$\frac{dx_{j}}{dt} = \epsilon g_{j}\left(N, x\right)$$
-
-\end{column}
-\end{columns}
-
-\footnotetext[1]{Lion, S (2018). Theoretical approaches in evolutionary ecology: environmental feedback as a unifying perspective. Am. Nat., 191:21-44}
-\footnotetext[2]{Patel, S, Cortez, MH, \& Schreiber, SJ (2018). Partitioning the effects of eco-evolutionary feedbacks on community stability. The Am. Nat., 191:381-394.}
-
-## Conceptual unification of ecology and evolution
-
-```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/fig1_b.png");
-```
-
-## Conceptual unification of ecology and evolution
+## Call for proposals: joint sDiv-CESAB
 
@@
 
-## Unification of modern coexistence theory \& Price equation
+## The UNICOP project
 
-Joint funding call between French Foundation for Research on Biodiversity and German Centre for Integrative Biodiversity Research (iDiv): CESAB-sDIV SYNERGY
+\Huge
 
-```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/unicop_team.jpg");
-```
-
-The UNICOP team in Montpellier (summer 2023)
+Conceptual unification of modern coexistence theory and the Price equation
 
@@
 
+- Framework for making models (meta-modelling) \pause
 - Not making assumptions or predictions \pause
-- Defining population change in an eco-evo context \pause
-- Framework for models (meta-modelling) \pause
-- A rough analogy,
+- Not the kind of thing we test, e.g.,
 
-$$y = \beta_{0} + \beta_{1}x$$
+$$y_{i} = \beta_{0} + \beta_{1}x_{i} + \epsilon_{i}$$
 
@@
 
-\footnotetext[1]{Queller, DC (2017). Fundamental theorems of evolution. The American Naturalist, 189:345-353.}
+\footnotetext[1]{Queller, D. C. (2017). Fundamental theorems of evolution. The American Naturalist, 189:345-353.}
 
-## Generality of the Price equation
+## My goal: starting point is unclear
+
+\LARGE
+
+- Figure out how to unify the Price equation with ecological change \pause
+- Start with the Price equation and add ecology? \pause
+- Start somewhere else and derive Price? \pause
+- Is this even possible? \pause
+- If it's possible, it will look obvious in hindsight
+
+## The UNICOP team
+
+```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
+include_graphics("img/team.jpg");
+```
+
+## Leipzig in September 2022
+
+```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
+include_graphics("img/leipzig_2022.jpg");
+```
+
+## Leipzig in September 2022
+
+\Huge
+
+- Confusion about the question
+- Separated project aims
+- An attempt at an answer: \pause
 
@@
 
-**Evolution of non-biological systems**
-
-- Archaeological artefacts$^{1}$
-- Musical properties$^{2}$
+$$iN + \Delta\bar{z} = i\left(1 + \bar{w} - \bar{\delta} \right)N + \frac{1}{\bar{w}}Cov\left(z, w\right) + \frac{1}{\bar{w}} E\left(w\Delta z \right).$$
 
@@
 
-**Synthesis across disciplines**
+- "It's not *wrong*" - S$\'{e}$b \pause
+- "Brad, that's cheating" -Th$\'{e}$o
 
-- D'Alembert's principle of mechanics$^{3}$
-- Information theory$^{4}$
+## Montpellier in June 2023: Not much progress
 
-\footnotetext[1]{Brantingham, PJ (2007). A unified evolutionary model of archaeological style and function based on the Price equation. American Antiquity 72:395–416.}
-\footnotetext[2]{MacCallum, RM, et al. (2012). Evolution of music by public choice. PNAS 109:12081–12086.}
-\footnotetext[3]{Frank, SA (2015). D'Alembert’s direct and inertial forces acting on populations: The price equation and the fundamental theorem of natural selection. Entropy 17:7087–7100.}
-\footnotetext[3]{Frank, SA (2012). Natural selection. V. How to read the fundamental equations of evolutionary change in terms of information theory. Journal of Evolutionary Biology, 25:2377-2396.}
+```{r, out.width = "85%", echo = FALSE, fig.cap = ""}
+include_graphics("img/montpellier.jpg");
+```
 
-## Conceptual unification of ecology and Price equation
+## On the backburner until January 2024
 
 ```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/fig1.png");
+include_graphics("img/leipzig_more_snow.jpg");
 ```
 
-## Links between ecology and the Price equation
+## On the backburner until January 2024
 
-**Ecosystem function loss due to species loss$^{1}$**
+```{r, out.width = "160%", echo = FALSE, fig.cap = ""}
+include_graphics("img/leipzig_snow.jpg");
+```
 
-\Large
-
-$$\Delta T = \bar{z}\Delta s + Sp(w, z) + \sum_{i}w_{i}\Delta z_{i}$$
-
-\normalsize\pause
-
-**Community trait change: Physiology, evolution, ecology$^{2}$**
-
-\Large
-
-$$\Delta \bar{z} = E_{I}\left(E_{Ji} \left(\Delta z_{ij} \right) \right) + E_{I}\left(cov_{Ji} \left(w_{ij}, z'_{ij} \right) \right) + cov_{I}\left(w_{i}, z'_{i}\right)$$
-
-\footnotetext[1]{Fox, JW (2006). Using the price equation to partition the effects of biodiversity loss on ecosystem function. Ecology, 87(11), 2687-2696.}
-\footnotetext[2]{Collins, S \& Gardner, A (2009). Integrating physiological, ecological and evolutionary change: A Price equation approach. Ecology Letters, 12(8), 744-757.}
-
-## Links between ecology and the Price equation
+## Working day and night back and forth
 
 ```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/Page_Nowak.png");
+include_graphics("img/leipzig_2024.jpg");
 ```
 
-\LARGE
-
-- Figure 1 from Page \& Nowak$^{1}$
-- Relies on \textit{relative} species frequenies
-
-\footnotetext[1]{Page, KM \& Nowak, MA (2002). Unifying evolutionary dynamics. Journal of Theoretical Biology, 219(1), 93–98.}
-
-## Reconciling evolutionary change and population growth {.t}
-
-\Large
-
-- Start with Price equation and add ecology?
-- Start elsewhere and derive the Price equation?
-
-```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/lost_in_woods.jpg");
-```
-
-\footnotetext[1]{Image: Eneas De Troya (2006).  \href{https://commons.wikimedia.org/wiki/File:Lost_in_a_Virtual_Wood_(177504871).jpg}{CC BY 2.0}}
-
-## The fundamental equation of ecology
+## Finally a way forward, and fundamental ecology
 
@@
 - Evolution: Price Equation
-- Ecology: Births and deaths
+- Ecology: Exponential growth
 
@@
 
-## Reconciling foundations of evolution and ecology
+## Finally a way forward, and fundamental ecology
 
@@
 
-## The unifying equation
-
-\LARGE
-
-- No systematic approach
-- Individual birth, death, characteristics
-- Deductive exploration
-
-\vspace{5mm}
-\hrule
-\pause
-\vspace{5mm}
+## Nearly there: the unicorn equation
 
@@
 
-## A formal definition of eco-evolutionary change {.t}
+## Who cares? {.t}
 
@@
 - No simplifying assumptions \pause
-- Complete and exact (closed populations) \pause
+- Complete and exact for all closed populations \pause
 - Ecology: $\beta_{i} - \delta_{i} + 1$ \pause
-- Evolution: $z_{i} + \Delta z_{i}$ \pause
-- Ecology required for selection or drift
+- Evolution: $z_{i} + \Delta z_{i}$
 
-## Confirmations from a formal definition {.t}
+## Who cares? {.t}
 
@@
 
-- **Fitness:** $\bar{w} = \lambda = \beta - \delta + 1$ \pause
+- **Fitness:** $\bar{w} = \lambda = 1 + \beta + \delta$ \pause
 - Derives assumed $N_{t+1} = N_{t}\lambda = N_{t}\bar{w}$ \pause
@@
 - Starting point for further unification \pause
-- Build upon for community ecology
+- Ecology hiding in population genetics
 
-## An elegance in two fits {.t}
+## The most elegant part {.t}
 
@@
 
-\vspace{2mm}
+\vspace{5mm}
 \hrule
@@
 
-## An elegance in two fits {.t}
+## The most elegant part {.t}
 
@@
 We can define z as a trait (e.g., photosynthetic rate, decomposition rate, biomass) and recover $\Omega$ as a population's contribution to ecosystem function.
-
-## Foundations of ecological and evolutionary change {.t}
-
-\LARGE
-
-$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$$
-
-\vspace{5mm}
-\hrule
-
-```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
-include_graphics("img/foundations.png");
-```
-
-\footnotetext[1]{Duthie, AB, \& Luque, VJ (2024). Foundations of ecological and evolutionary change. arXiv preprint arXiv:2409.10766.}

### 2024-12-20 unifying_equation.Rmd
Initialise all of the relevant files

---
title: "A unifying equation of ecology and evolution"
author: "Brad and Victor"
date: "2024-03-05"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
bibliography: refs.bib
---

Background
=====================================

Theoretical unification of phenomena once assumed to be disparate is a powerful tool for scientific advancement. 
Such unification has therefore been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions. 
In theoretical physics, Newton's unification of celestial and terrestrial motion, and Maxwell's unification of electricity and magnetism, are among the most important advances in scientific history. 
Likewise, in the life sciences, the modern synthesis of natural selection and Mendelian inheritance revolutionised how mechanisms underlying evolution are understood and applied.

Three major evolutionary mechanisms (natural selection, genetic drift, and gene flow) are instantiated through a combination of different ecological factors [@Connor2004].
Most notably, change in both the traits and size of a closed population are caused by individual birth and death.

Nevertheless, until relatively recently, evolutionary biology and population ecology have taken almost parallel paths for developing theory.
Although many models now exist to connect evolutionary and ecological phenomena [e.g., @Lion2018; @Patel2018; @Govaert2019], the foundational assumptions and equations of evolution and ecology in these models remain mostly separated conceptually. 
Here we propose the foundation for a different approach. 
Instead of joining existing fundamental models of ecology and evolution, we propose a new equation from which the most general equations of population ecology, evolution, and ecosystem function can be derived.

The fundamental equation of ecology, evolution, and ecosystem function
====================================================

Our objective is to start with the most fundamental axioms underlying both ecology and evolution in all systems [@Rice2004; @Rice2009].
All such systems include entities (e.g., individuals, but our framework is sufficiently general such that entities can be anything discrete).
Entities are introduced into the population through birth and removed from it through death.
Entities are defined by some property, which might be a species identity (ecology) or a trait (evolution).

Let $\beta_{i}$ be the number of births attributable to entity $i$.
Let $\delta_{i}$ indicate the death (1) or survival (0) of $i$.
Let $z_{i}$ be a property attributable to $i$, and $\Delta z_{i}$ be any change in $z_{i}$ from one time step $t$ to the next $t + 1$.
Let the number of entities in the population be $N$.
Finally, let $\Omega$ be a summed property across $N$ entities,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i}\left(z_{i} + \Delta z_{i} \right) - \delta_{i}\left(z_{i} + \Delta z_{i} \right) + \left(z_{i} + \Delta z_{i} \right)\right).$$

From the above equation, we can derive the most fundamental equations of population ecology and evolutionary biology through an appropriate interpretation of $z_{i}$.

Population ecology
====================================================

Assume that $z_{i}$ defines the identity of $i$ belonging to the population.
In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
Further, we make the standard population ecology assumption that entities (now specifically individual organisms) are identical and do not change species.
To define this formally, we set $\Delta z_{i} = 0$ and note that we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
In this case, we have,

$$\Omega = \sum_{i=1}^{N} \left(\beta - \delta + 1 \right).$$

Summing up from 1 to $N$, we can rewrite the above,

$$\Omega = N\left(\beta - \delta + 1 \right).$$

We can now interpret $\Omega$ as the population size in $t+1$, $N_{t+1}$, and note that $N\beta$, $N\delta$, and $N$ are the total births, total deaths, and size of the population at $t$, respectively,

$$N_{t+1} = births - deaths + N_{t}.$$

Alternatively, if we define $\lambda = 1 + \beta - \delta$, then we can rewrite,

$$N_{t+1} = N_{t}\lambda.$$

This equation is fundamental in the sense that the change in size of every closed population is attributable to births minus deaths. 
We can build upon this by structuring the population (e.g., into age or stage classes), or by including space or density-dependent effects, but all of these different mechanisms of population change are ultimately realised through the birth and death of individuals.
Consequently, the above equation describes the change in every closed population from one time step to the next.

It is worth pointing out now that $\lambda$ is going to come out as equivalent to mean fitness $\bar{w}$ in working out the link to the Price equation below.
We could therefore write the above as, $N_{t+1} = N_{t}\bar{w}$ with $w = \bar{w}$ then being the fitness of all individuals in the population.
Next, we will derive the Price equation from the fundamental equation.

Evolutionary change
====================================================

Deriving Price takes a few more steps.
We can start by rearranging the original equation,

$$\Omega = \sum_{i=1}^{N} \left(z_{i}\left(\beta_{i} - \delta_{i} + 1\right) + \Delta z_{i}\left(\beta_{i} - \delta_{i} + 1\right) \right).$$

As noted in the previous section, we can define fitness,

$$w_{i} = \beta_{i} - \delta_{i} + 1$$

The fitness of an individual from one time step to the next equals the number of offspring of the individual ($\beta_{i}$), minus 1 if the individual has died ($\delta_{i} = 1$) or 0 if the individual has not died, plus 1.
Note that in addition to offspring number, the longevity of the individual matters in terms of fitness.
If the individual dies, then fitness is simply $\beta_{i}$.
If the individual lives, then fitness is $\beta_{i} + 1$.
This is conceptually satisfying because an individual that lives from $t$ to $t + 1$ should have higher fitness than an individual that dies, even if both have the same reproductive output.
With this definition of fitness, we substitute,

$$\Omega = \sum_{i=1}^{N} \left(w_{i}z_{i} + w_{i}\Delta z_{i} \right).$$

We can break this down further,

$$\Omega = \sum_{i=1}^{N} \left(w_{i}z_{i} \right) + \sum_{i=1}^{N}\left( w_{i}\Delta z_{i} \right).$$

We can multiply each side by $1/N$,

$$\frac{1}{N}\Omega = \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}z_{i} \right) + \frac{1}{N}\sum_{i=1}^{N}\left( w_{i}\Delta z_{i} \right).$$

On the right-hand side of the equation above, this gives us the expected values within the summation,

$$\frac{1}{N}\Omega = E\left(w_{i}z_{i}\right) + E\left( w_{i}\Delta z_{i} \right).$$

The left-hand side requires a bit more explanation.
What we have above is the total sum trait values $z$ across the entire population at $t + 1$ divided by the number of individuals ($N$) in the population at $t$.
What we need is the mean $z$ across $N$ in our *original* population. 
In other words, dividing $\Omega$ by $N$ will include contributions to the sum trait value from the *new* population at $t + 1$ on the left-hand side of the equation.
We therefore cannot treat $\Omega/N$ as the mean of $z'$ because we need to weigh $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.
For example, if mean fitness is 2 (i.e., $\bar{w} = 2$), then half as many individuals from $t$ will have contributed to $\Omega$ in $t + 1$ (i.e., there are $N$ individuals at $t$ and $2N$ individuals at $t + 1$).
We therefore need to multiply the mean trait value of $\bar{z}'$ (at $t + 1$) by the mean fitness of $\bar{w}$ at $t$ to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$.
Consequently, $\Omega = N\bar{w}\bar{z}'$.
That is, the number of individuals at $t$, times the mean fitness at $t$, times the mean of $z$ at $t + 1$ ($\bar{z}'$).
We can therefore rewrite,

$$\bar{w}\bar{z}' = E\left(w_{i}z_{i}\right) + E\left( w_{i}\Delta z_{i} \right).$$

Rearranging this now to derive the Price equation is not difficult. 
First, note that we can express covariance as, $Cov(X,Y) = E(XY) - E(X)E(Y)$, and therefore $E(XY) = Cov(X,Y) + E(X)E(Y)$.
Substituting into the above,

$$\bar{w}\bar{z}' = Cov\left(w_{i},z_{i}\right) + \bar{w}\bar{z} + E\left( w_{i}\Delta z_{i} \right).$$

From here,

$$\bar{w}\bar{z}' - \bar{w}\bar{z} = Cov\left(w_{i},z_{i}\right) + E\left( w_{i}\Delta z_{i} \right).$$

Pulling fitness out of the left-hand side,

$$\bar{w}\left(\bar{z}' - \bar{z}\right) = Cov\left(w_{i},z_{i}\right) + E\left( w_{i}\Delta z_{i} \right).$$

Since we define $\Delta z_{i} = \left(z_{i}' - z_{i}\right)$,

$$\bar{w}\Delta z_{i} = Cov\left(w_{i},z_{i}\right) + E\left( w_{i}\Delta z_{i} \right).$$

From the starting point of the first equation for $\Omega$, which describes fundamental birth and death processes in a population, we can derive both the most fundamental geomemetric growth model of population ecology and the fundamental equation of evolution.
By itself, we could also consider $\Omega$ to be a quantity of ecosystem function, in that it is the summed effect of the traits ($z_{i}$) of individual entities ($i$) in the population.
For example, $z$ might be interpreted as biomass or photosynthetic rate.

Discussion
====================================================

To my knowledge, a fundamental equation from which geometric growth, the Price equation, and ecosystem function models can be derived has not been attempted or demonstrated.
@Page2002 showed some nice equivalences between replicator equations, Lotka-Volterra, and the Price equation, but not in quite the unifying way that I think we show here.

I believe that there is conceptual and philosophical merit in this work.
The equation certainly was not obvious at the outset of this effort, although if correct and useful, it will almost certainly appear obvious in hindsight.
In working through this result, there are some issues that I think become clearer with this unifying approach.
For example, the equivalence between mean fitness ($\bar{w}$) and population growth rate ($\lambda$) is confirmed from first principles [@Lion2018].

Assumptions of ecological and evolutionary models are also made clearer, I think.
And with this formalism, we can more clearly identify the kinds of processes that are enacted through ecology, evolution, or both.
For example, demographic stochasticity, which is defined by fluctuations caused by the probabilistic nature of birth and death, will clearly affect $\Omega$ and therefore have joint effects on population ecology and evolution.
Any model that includes demographic stochasticity should therefore do so in a way that has consistent and simultaneous effects on both population and evolutionary dynamics.
In contrast, some processes are relevant to only ecology or evolution.

In population ecology, the assumption that individuals within a species are identical and do not change is formalised by setting $z_{i} = 1$, $\Delta z_{i} = 0$, $\beta_{i} = \beta$, and $\delta_{i} = \delta$. 
Expansions from the equation for geometric growth could be made to model population structure.
The Price equation is established as the fundamental equation of evolution, and our work here shows how evolutionary change is grounded in the birth and death of discrete entities.

There are other avenues of potential interest.
For example, it is potentially interesting that the derivation of geometric growth takes considerably fewer steps than the Price equation.
It also seems relevant that neither the ecology nor evolution seem to take precedence, in the sense that one is more fundamental and derived from the other.
It also might be interesting to think more about the conservation of total probability [@Frank2020] and where it first becomes relevant above.

References
====================================================

### 2024-12-24 foundations_of_ecoevo.Rmd
Edits right before changing referencing style for nature

@@
 ---
-title: "Foundations of ecological and evolutionary change"
+title: "Darwin's Dream: Unifying ecological and evolutionary change"
 author: |
@@
 - \usepackage{fancyhdr}
+- \usepackage{lineno}
+- \linenumbers
+- \modulolinenumbers[2]
+linestretch: 2
 bibliography: refs.bib
@@
 **Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.**
+
+**Key words:** Ecology, Evolution, Eco-Evolutionary Theory, Fundamental Theorem, Price Equation, Population Growth
+
+Non-technical Summary
+=====================================
+
+Biologists have long known that the ecology of population size is inextricably linked with the evolution of population traits.
+Nevertheless, the foundational theory underlying ecological and evolutionary change have been historically separated, with recent attempts to bridge this theoretical divide using eco-evolutionary modelling.
+We discover a new equation from first principles that provides a universal and complete definition for eco-evolutionary change.
+This equation thereby offers a unified framework for ecology and evolution, which also recovers the foundational theory of ecology and evolution as developed in isolation.
 
@@
 We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
-From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 1) and principle of exponential growth ($N_{t+1} = \lambda N_{t}$).
+From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 1) and principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
 By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 2).
@@
 We anticipate that this will be useful for eco-evolutionary theory in the same way that the Price equation is useful for evolutionary theory: facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.
-
-<!---
-
-Our unification and derivation of fundamental equations of ecology and evolution recovers and formalises well-known properties of biological systems.
-The expressions $\beta_{i} - \delta_{i} + 1$ and $z_{i} - \Delta z_{i}$ can be interpreted to represent the ecological and evolutionary components of a system, respectively.
-When focused only on change in population size, we set $z_{i} = 1$ and $\Delta z_{i} = 0$, then we can derive the fundamentals of population ecology without considering evolution.
-Interestingly, the same is not true when focused only on evolutionary change.
-If we set $\beta_{i} = 0$ and $\delta_{i} = 0$ (no birth or death), then there is no variance in fitness ($w_{i} = 1$ for all individuals), and evolution does not occur (note that evolution can still occur in a population at equilibrium where the birth rate and death rate are the same).
-Fundamentally, evolutionary change is therefore enacted through ecological change, i.e., individual births and deaths, but ecological change does not necessarily imply evolution.
-And while the link to population change is intuitive from the structure of eqn 1, the link to the Price equation is much less intuitive, especially after eqn 5 in our derivation.
-The Price equation describes change in any evolving system, not just biological systems [@Fox2006; @Brantingham2007; @Bell2009a; @MacCallum2012; @Frank2015].
-It is perhaps unsurprising that a derivation of ecological and evolutionary change from a common starting point requires relatively more labour to recover the Price equation.
-
---->
 
@@
 
-<!---
-
-PROBLEM HERE: TRYING TO RECOVER THE LOGISTIC GROWTH IN DISCRETE TIME
-
-=======================================
-
-Our fundamental equation is complete and exact.
-It therefore implicitly includes any effects of density dependence in population growth (see Box 1).
-Starting from eqn 1, we can partition ecological change into density-independent and density-dependent effects by multiplying the expression $\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$ within the summation by $\left(1 + \sum_{j=1}^{N} \alpha_{ij}z_{j}\right)$, where $j$ indexes the same individuals as $i$ and $\alpha_{ij}$ defines the effect $j$ has on the contribution of $i$ to $\Omega$.
-This partition can be used to recover
-
-OR............. BELIEVE THAT THE BELOW IS THE CORRECT ONE!!
-
-Our fundamental equation is complete and exact.
-It therefore implicitly includes any effects of density dependence in population growth (see Box 1).
-Starting from eqn 1, we can partition ecological change into density-independent and density-dependent effects by first defining $a_{ij}$ as the effect that the characteristic of $j$ ($z_{j}$) has on the contribution of $i$ to $\Omega$.
-Within the summation of eqn 1, we can then include a new term summing across all individuals in the population, $\sum_{j=1}^{N} a_{ij}z_{j}\right)$.
-This sums the effect of each individual on a focal individual $i$.
-If we define $\alpha = E\left( a_{ij}\right)$, then we can
-
---->
-
 Our fundamental equation is complete and exact.
@@
 
-<!---
-
-NOTE 1
-==================
-
-Need to discuss that evolution *is* possible without ecology, if we imagine a set of RNA molecules that are all identical in an environment with no birth or death.
-We might have some change in RNA sequence (and therefore structure) that is evolution; this evolution will be in the \Delta z as non-zero -- the covariance of the Price equation will still be zero, but the expectation term will be nonzero because expected fitness is 1 (because you get credit for surviving from time step t to t+1).
-Hence, no adaptive evolution (selection) is possible without ecology, but mutation can happen (or selection at lower levels of biological organisation).
-
-So evolution is possible without ecology and vice versa, however unlikely.
-
-NOTE 2
-================
-
-We can partition \Delta z in the original equation however we want, and it just gets back to new terms attached to the Price equation at the end
-
-Need to try to derive what happens when we partition fitness into birth, death, and survival component
-
---->
-
 Acknowledgements
@@
 
+Author Contributions
+====================================================
+
+Both authors came up with the question idea. 
+Following many discussions between the authors, ABD proposed the initial equation with subsequent exploration and development from both authors. 
+Both authors contributed to the writing.
+
+Competing Interests
+====================================================
+
+The authors declare no competing interests.
+
 References

### 2024-12-24 foundations_of_ecoevo.tex
Edits right before changing referencing style for nature

@@
 \usepackage{lmodern}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \hypersetup{
-  pdftitle={Foundations of ecological and evolutionary change},
+  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
   colorlinks=true,
@@
 \usepackage{fancyhdr}
+\usepackage{lineno}
+\linenumbers
+\modulolinenumbers[2]
 \ifLuaTeX
@@
 
-\title{Foundations of ecological and evolutionary change}
+\title{Darwin's Dream: Unifying ecological and evolutionary change}
 \author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
@@
 
+\setstretch{2}
 \hypertarget{abstract}{%
@@
 density-dependent population growth.}
+
+\textbf{Key words:} Ecology, Evolution, Eco-Evolutionary Theory,
+Fundamental Theorem, Price Equation, Population Growth
+
+\hypertarget{non-technical-summary}{%
+\section{Non-technical Summary}\label{non-technical-summary}}
+
+Biologists have long known that the ecology of population size is
+inextricably linked with the evolution of population traits.
+Nevertheless, the foundational theory underlying ecological and
+evolutionary change have been historically separated, with recent
+attempts to bridge this theoretical divide using eco-evolutionary
+modelling. We discover a new equation from first principles that
+provides a universal and complete definition for eco-evolutionary
+change. This equation thereby offers a unified framework for ecology and
+evolution, which also recovers the foundational theory of ecology and
+evolution as developed in isolation.
 
@@
 population ecology (Box 1) and principle of exponential growth
-(\(N_{t+1} = \lambda N_{t}\)). By defining individual fitness
-(\(w_{i}\)) and applying the total conservation of probability to
-individual frequencies (\protect\hyperlink{ref-Frank2015}{Frank 2015},
+(\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
+and applying the total conservation of probability to individual
+frequencies (\protect\hyperlink{ref-Frank2015}{Frank 2015},
 \protect\hyperlink{ref-Frank2016}{2016}), we recover the most
@@
 
+\hypertarget{author-contributions}{%
+\section{Author Contributions}\label{author-contributions}}
+
+Both authors came up with the question idea. Following many discussions
+between the authors, ABD proposed the initial equation with subsequent
+exploration and development from both authors. Both authors contributed
+to the writing.
+
+\hypertarget{competing-interests}{%
+\section{Competing Interests}\label{competing-interests}}
+
+The authors declare no competing interests.
+
 \hypertarget{references}{%

### 2025-05-29 response_to_reviewers.md
Initialise response to reviewer comments, with current manuscript

We thank both reviewers for their valuable comments and suggestions. They helped us to improve the presentation of a number of points that were not entirely clear in the previous version of the manuscript and focus our revision on better communicating the aim of our work. To summarise, it does not appear that the accuracy or novelty of our proposed equation is in question. But the motivation for seeking and reporting it clearly needs to be better communicated. Couching this motivation in terms of major historical breakthroughs in science has set an incorrect and unintended tone; the point was to explain why pursuing abstract unifying equations is a valid and important part of scientific theory (using the strongest cases available), not to suggest that what we've done is field-changing in its importance!

Consequently, the references to major advances in science, which we appreciate came across as self-aggrandising, were motivated primarily by what we feel is a need to justify the importance and validity of a particular kind of theory. We have scaled back on this emphasis, but we feel as though some such justification is still necessary given the nature of comments from both reviewers, with Reviewer 1 closing their review with "[e]xplaining why it might be useful would help too", and Reviewer 2 stating that "if this equation does not produce evolutionary or ecologically meaningful results on its own, it's unclear that it has any utility". While we believe that there is scope to derive pragmatic results from our equation (i.e., new predictions), the primary goal of this manuscript is, unapologetically, about conceptual and theoretical clarity rather than prediction. This is not different from George Price's works from the 1970s. For example, in his 1972 paper he states that his now famous formula "is intended mainly for use in deriving general relations and constructing theories, and to clarify understanding of selection phenomena, rather than for numerical calculation" (Price 1972, 485). In this sense, our manuscript follows a similar path. We ask, what is most fundamental to all eco-evolutionary change? And we deliver a formal answer, the accuracy of which does not appear to be in question. Ideally, all formalisms try to achieve three characteristics at the same: generality, realism, and precision (Levins 1966). However, as Levins (1966) points out, it is not possible to achieve all of them at the same time and one of them needs to be sacrificed. When you theorise the way that we do here, specific (precise) predictions are sacrificed in favour of generality and realism (more on this later in relation to "dynamic sufficiency").

We contend that this kind of theory is important in ecology and evolution, even if it is not of widespread interest or appreciation (Caswell 1988). For example, even if the Price equation served no other purpose than to provide a formal, abstract, link between the fundamental equations of population and quantitative genetics (Queller 2017), it would still be a valid contribution to evolutionary biology, useful for providing conceptual clarity to theoreticians and demonstrating the logical consistency of the field. Hence, while we are optimistic about where our starting point might lead for developing new theory, we do not see this as being the most important or interesting outcome of this work.

Our equation is general and realistic. It is also simple; it couldn't be anything else given what we set out to achieve (and while it is perhaps trivial to verify in hindsight, it was not trivial to find). A fundamental equation of eco-evolutionary change must, by definition, be able to recover the fundamental equations of evolutionary biology (the Price equation) and population ecology (N_{t+1} = N_{t} + Births - Deaths). Our equation thereby defines what eco-evolutionary change is in its broadest sense; verbally, it is an interaction between individual fitness components (birth, death, and identity) and an individual characteristic.

Reviewer 1 comment 1
==================================

The literature that is cited is VERY selective. There is a huge body of literature on linking ecological and evolutionary theory. High-dimensional matrix models have now been routinely used to link ecological and evolutionary dynamics by Barfield, Boots, Childs, Coulson, Ellner, Lion, Rees, Tuljapurkar and others. They have shown how by writing equations to dynamically track distributions of individual-level phenotypic traits and genotypes you can retrieve: the Price equation (including for structured populations) selection differentials and gradients, heritabilities, genetic covariances, the Euler-Lotka equation, equations for population growth, rates of evolutionary change, evolutionary end points, species interactions, and coexistence mechanisms. This literature needs citing and appropriately summarising. (It turns out it is straightforward to retrieve their equation from this modelling framework. if the authors did this, it will help reveal its limitations. One example is their delta zi terms get split into ontogenetic development among surviving individuals and genetic and non-genetic inheritance between parents and their offspring in age-structured populations).

Author response to Reviewer 1 comment 1
=====================================

We have revised our introduction extensively to address this comment and better communicate the goals of our work.

We agree with Reviewer 1 concerning the existence and importance of this literature integrating ecological and evolutionary theory. We are familiar with this literature, but all of the models developed by the authors mentioned by Reviewer 1, to our knowledge, are addressing questions that are fundamentally different from our own. These are highly general tools for integrating ecological and evolutionary dynamics, and we cited some key ones in our original introduction (Lion 2018; Patel, Cortez, and Schrieber 2018; Govaert et al. 2019). Lion (2018) and Patel et al. (2018) are particularly instructive examples because of how general they are: In Lion (2018), equations for ecological [ dn_{i}/dt = r_{i}(E)n_{i} ] and evolutionary [ f_{i}(r_{i}(E) - \bar{r_{i}(E)) ] change are defined as functions of an environmental vector E that can include other population densities or any relevant external variables. In Patel et al. (2018), ecological [ dN_{i}/dt = N_{i} f_{i}(N, x) ] and evolutionary [ dx_{j}/dt = e g_{j}(N, x) ] equations are similarly defined by general functions for population change (f) and trait change (g). Govaert et al. (2019) reviews eco-evolutionary feedback models.

There are, of course, other excellent examples by the authors Reviewer 1 cites.

For example, Barfield, Hold, and Gomulkiewicz (2011) develop a general model of eco-evolutionary dynamics in a stage-structured population. Their model is highly general, defining a multivariate quantitative trait with functions that define transition probabilities across different life stages; it makes some reasonable simplifying assumptions and nicely demonstrates that Lande's result of phenotypes evolving at the same rates across ages also applies to stages. This is an important and exciting theoretical result, but the authors also recognise the importance of putting it in the context of the fundamental equation of evolution: "In this section, we show how evolution in stage-structured populations can be related to a simple, universal law of evolution, Price's theorem (Price 1970), which provides a general statement about any evolutionary system -- including a stage-structured one". It is perhaps relevant to note that Price's (1970) motivation was solely to present a transparent "tool in qualitative evolutionary reasoning".

Lion, Sasaki, and Boots (2023) outline a theoretical framework linking quantitative genetics and adaptive dynamics, which reviews classical approaches to investigating eco-evolutionary dynamics and outlines key challenges for eco-evolutionary modelling. Rees and Ellner (2016) use an integral project model and compare it to an individual-based model (IBM), demonstrating how the former can be used to show why population structure changes are important for selection and generally recover results in the IBM. Coulson and Tuljapurkar (2008) investigate how selection on trait change can be partitioned into different components using an extension of the Price (1970) equation that accounts for age structure.

We don't just point out these particular examples because they are written by authors that Reviewer 1 cites. These examples are also highly general models (with the ones cited in our original manuscript being especially general examples), and typically dynamically sufficient in a way that leads to interesting new eco-evolutionary predictions. They build off of very general functions as a starting point, as in Lion (2018) and Patel et al. (2018), or expand upon more fundamental equations such as that of Price (1970). But these models all have a different goal from ours; to be useful, they make simplifying assumptions to link ecological and evolutionary dynamics.

Reviewer 1 is not correct in stating that these models retrieve the Price equation; the Price equation is not a predictive model but a mathematical identity that formally defines evolutionary change. Hence, it must always be possible, at least in theory, to retrieve the Price equation from any evolutionary model. These models don't retrieve the Price equation so much as the Price equation is necessarily embedded within them. The Price equation is not a simplified version of, e.g., Coulson and Tuljapurkar (2008), which includes a version of the Price equation that takes into account stage structure -- the latter is a specific case that is derived from the more abstract and exact former by making useful simplifying assumptions to generate novel predictions. The Price equation is also not just a generic mathematical function incorporating relevant evolutionary variables (as in, e.g., Lion and Patel examples mentioned above). Price (1970) is derived from first principles in evolutionary theory to define evolution in the same way that N_{t+1} = N_{t} + Births - Deaths defines ecological change.

Similarly, we attempt to define what is most fundamental to eco-evolutionary change. Our equation 1 does this formally, defining eco-evolutionary change in its most abstract and exact form as an interaction between fitness across different demographic components and individual characteristics. From eqn 1, the exact equations defining population and evolutionary change can be recovered. The simplicity of our equation is deceptive, and this is underscored by the fact that despite having never been discovered (to our knowledge), the equation can be used to derive the two most fundamental equations in ecology and evolution. As noted by Frank (1995, p. 381) in reference to Price's work, “[i]t is always difficult, in retrospect, to see the originality and insight of a simple idea”. While we certainly do not claim our work here to be field-changing, or as versatile as modelling tools from the abovementioned papers, we do believe that it is original and insightful in defining eco-evolutionary change.

In summary, are familiar with the literature pointed out by Reviewer 1, and we would even (quite happily) argue that this literature includes models that are highly general tools for modelling complex eco-evolutionary dynamics. The whole point of introducing the paper the way that we did was to illustrate that we are trying to ask a different question than the ones in those papers. We obviously need to try a different approach, and we have done this using extensive revisions in our introduction that address the points made above.

Reviewer 1 comment 2
==================================

The equation that is developed is not dynamically sufficient. This means that without modification it is fairly limited in describing what happens over an individual time step. This is no surprise, because the Price is simply a tautology that is also not dynamically sufficient. The authors may have derived an equation that is worthy of publication. They need to be honest on its limitations (also see point 1).

Author response to Reviewer 1 comment 2
==================================

We now state this more clearly in the text (Line XXX).

Reviewer 1 is of course correct that our eqn 1 is not dynamically sufficient. It couldn't possibly be, for exactly the reason that Reviewer 1 points out. The Price equation is a tool for unifying evolutionary models and providing conceptual clarity, which is necessarily achieved at the cost of any and all predictive power for a given system. Models need to make (simplifying, specific system) assumptions to be predictive (Levins 1966, Frank 2012), and a fundamental equation, by its nature, needs to be free of such assumptions and focus only on what is essential. The worthiness of our equation rests entirely on its power to unify fundamental equations of ecology and evolution and provide conceptual clarity, and its limitations are the same as any other fundamental equation.

Reviewer 1 comment 3
==================================

The authors massively overstate the importance of their equation, starting with the title of "Darwin's dream". There is no evidence I am aware of that Darwin ever thought much about equations. The authors then talk about scientific breakthroughs, attempting to place their equation with other great equations of science. Astonishing self-belief, and not warranted given how much has been done in this space already and the limitations of their formulae. The paper needs toning down significantly. The equation is mildly interesting, but it is not field-changing. This self belief may come from the authors having been working in an echo chamber. This is even hinted at in the acknowledgements and author contributions.

If the authors were to say they have found a way to write an equation that links the Price equation to the equation that describes change in population size from one time step to the next, that would be an honest and critical placement of their work in the broader literature. Explaining why it might be useful would help too.

Author response to Reviewer 1 comment 3
==================================

We have changed the title and reframed the introduction. Our goal was to explain the importance of conceptual unification in the sciences (which, in our experience, is often underappreciated in the life sciences) without overselling our present results. In our revised introduction, we have attempted to tone down our own contribution while also not losing our central point that this kind of theory is important in science (i.e., *useful*, but because it demonstrates logical consistency and provides conceptual clarity -- not just because you can use it in the same way as a predictive model).

In our manuscript, we argue that our equation is fundamental to ecology and evolution. What we mean by this is that the fundamental equations of ecology and evolution can be derived from our eqn 1. What this illustrates is a logical consistency for developing theory, which has previously been assumed but never actually derived. It does not imply that our equation is more *important* than any other equation, and this might be at least partly responsible for what Reviewer 1 interpreted as an inflated self-belief and need to tone the writing down. We do not believe our work to be field-changing, but we do believe that it is more than mildly interesting, and that it is necessary for reasons that are not well-appreciated.

What we have shown is a logical consistency between the fundamental equations of ecology and evolution (and, as far as we can tell, this is not disputed by either reviewer). This logical consistency has been assumed by theoreticians, and perhaps quite reasonably. But as a discipline, it is still important to do our due diligence and show that the theoretical foundations of ecology and evolution are solid. For example, within evolutionary biology, we can model evolutionary change using a population genetics framework or a quantitative genetics framework. In a population genetics framework, the average excess equation describes selection in terms of changes in allele frequencies, whereas in the quantitative genetics framework, the Breeder's equation describes selection in terms of changing quantitative traits (Queller 2017). These two equations are very important for modelling evolutionary change and predicting the consequences of selection. But an important role of theory is to demonstrate how (and why) these two equations are logically consistent, i.e., that they follow without contradiction from the same starting point that formally defines evolutionary change in any system. The Price equation and the framework that it establishes (Gardner 2020, Baravalle et al. 2025) is that starting point; it defines what evolution is in the broadest sense, meaning that any specific model of evolutionary change must be possible to derive from it. If a model of evolution cannot be put into the framework of the Price equation -- at least in principle -- then it is fundamentally flawed.

Hence, while the Price equation has very little direct relevance to the work of most evolutionary biologists, it is fundamental in the sense that it binds together all of evolutionary theory by defining what evolutionary change is (and is not). Our equation 1 fulfils the same criteria: from a deceptively simple starting point, eqn 1 can be used to derive the Price equation itself, but also the fundamental equation of population ecology. It therefore defines what eco-evolutionary change is (and is not), mathematically, at its most fundamental. Verbally, it is the interaction between absolute fitness and some characteristic of an individual.

The fact that the theoretical foundations of ecology and evolution are logically consistent is, of course, hardly a surprise. Had this not been the case, theoretical biologists would surely have noticed by now. But demonstrating this fact from first principles is, we believe, a worthy goal in itself, and something that we have considered exciting work.

Reviewer 2 comment 1  
======================

At the broadest level, I've always understood the Price Equation as explicitly including the birth-death process that would lead to the population growth rate equation from ecology. In the abstract sense, the Price Equation does not require ancestors and descendants to be different, and thus the relation between them necessarily captures births (when they are different) and deaths (when the ancestor is not among the descendants). This relationship allows mean fitness to capture the population growth rate, as first shown by Fisher (1930) and later by folks like Lande (1982) and Rice (2004). Day & Bonduriansky (2011) note that the total number of individuals in each time interval is the sum of all surviving parents and offspring; hence, they are Nt(births – deaths), which is Duthie & Luque’s equation 3. Now, a case could be made that stating this relationship in an obvious and explicit way is useful, but it seems to me that the original formulation of the Price Equation already subsumes population growth dynamics, so erecting an even broader equation isn’t necessary.

Author response to Reviewer 2 comment 1 
=====================

In some sense, it is true that the Price equation subsumes population change; if it did not, then we would not be able to produce a unifying equation from which both the Price equation and the general Nt(birhts - deaths) equation could be derived. However, in the traditional derivation of the Price equation, births, deaths, and population growth are not at all explicit. Instead, the number of individuals (or groups or genes -- whatever the entity is) is simply defined as the sum across n_{i} at t and n'_{i} at t + 1, and frequencies are then immediately defined by n_{i}/n and n'_{i}/n' (Price 1970; Frank 1997; 2012; Gardner 2008). Numbers of entities (n and n') are only mentioned so that the frequencies of entities (q_{i} and q'_{i}) can then be defined, and these frequencies are then used to work out entity fitness. To our knowledge, the Price equation has never been derived from births and deaths or with reference to population growth. Lande (1982) did introduce a demographic model for population demography and dN/dt = rN does appear (eqn 10), but this is in the context of a modelling assumption; the Price equation is then used to calculate a selection gradient. Similar to previously referenced eco-evolutionary models (see our response to Reviewer 1 comments), there is integration of ecology and evolution here, but not a demonstration of the key link from first principles. Similarly, Day and Bonduriansky (2011) define a value W that is the sum of offspring and survival, and this is used within a version of the Price equation to get mean fitness, but a formal link to population growth isn't made, with the goal instead being to predict evolving traits.

Overall, population change appears in models where the Price equation also appears, but what has always been missing is a derivation of both from the same starting point, from first principles of what actually happens in a population. Logically, this has to be possible because the same mechanisms that underlie population change (birth and death) also underlie evolutionary change. But up until now, to the best of our knowledge, the link between population and evolutionary change has always been carefully considered but ultimately ad hoc (as in the abovementioned references and, to our knowledge, all eco-evolutionary models). Given that we are looking at the same mechanisms of population change, we really should be able to demonstrate that both population change and evolutionary change can 'pop out' as intrinsically and necessarily linked from the same starting point.

Figuring this starting point out was not easy; it needed to be done deductively, which inevitably led to many dead ends. Initially, we reasoned that because the Price equation is sufficiently abstract to describe trait change in any population (including groups and higher moments of traits), its recursive properties, and how useful it is for making formal links across disciplines such as physics and information theory (Frank 2015; 2016), we should be able to use it directly to recover population density change. This ended up being a dead end for reasons we will not dwell on (briefly, conservation of total probability -- that entity frequencies need to sum to 1 -- precludes the Price equation from being used to directly to recover absolute population change). We had different issues when starting from population change and trying to recover the Price equation. There are a lot of ways to make coherent models using these avenues of inquiry, but none that recover universal and exact descriptions of population and evolutionary change. The broader equation, our equation 1, is indeed necessary to derive exact and universal eco-evolutionary change from first principles; it thereby serves as a formal definition for eco-evolutionary change in any system. It perhaps appears simple and obvious in hindsight, as if it could be inferred quickly from existing models by anyone giving it a modicum of thought. But we believe this appearance is highly deceptive; if this formal link were trivial to discover, then someone would have surely pointed it out already. After much investigation, we are confident that no one has, and while there are fundamental equations defining ecological and evolutionary change, separately, our work here is unique in establishing a fundamental equation that defines eco-evolutionary change.

In our revision, we try to clarify the need for our equation more clearly.

Reviewer 2 comment 2
======================

This leads me to my second point. The broader equation (1) is defined such that it introduces an abstract quantity (Omega), whose meaning depends on further definitions of variables. In its original formulation (1), it is meaningless. For example, if N = 3, beta = 1, delta = 1, z1 = 1, z2 = 2, and z3 = 3, and Deltaz = 0.5, then the first term in equation 1 reduces to 1, and the summation becomes a sum of z: 1.5 + 2.5 + 3.5 = 7.5, thus, Omega is the sum of z. But if the first term doesn’t reduce to 1, Omega becomes incomprehensible. Let’s say beta_1 = 2, with the other betas remaining as 1. Now,

Omega = (2-1 + 1)(1 + 0.5) + (1-1 + 1)(2 + 0.5) + (1-1 + 1)(3 + 0.5) = 10.5

What is "10.5" exactly? It has no meaning, as it is a unitless composition of both the number of individuals and the sum of the traits they possess. On Line 192, the authors suggest it might be £a quantity of ecosystem function", but I can’t imagine how that could be the case given that what it encompasses (traits and individuals) changes depending on the values of the summation, and different values of traits, birth, and death, would produce identical Omega values.

Author response to Reviewer 2 comment 2
=====================

As Reviewer 2 correctly states, the quantity Omega is abstract and lacks a concrete interpretation or predefined units in eqn 1. This isn't a problem and is actually necessary given what we are trying to achieve, which is to identify the equation from which fundamental equations of ecology and evolution can be logically derived. This requires a very high level of abstraction to encompass any possible eco-evolutionary change.

The variable Omega is abstract and unitless for the exact same reason that 'z' is abstract and unitless in the Price equation. The Price equation requires a very high level of abstraction to be able to describe the change in some characteristic z for any possible evolving system. Units of z could therefore reflect any evolving characteristic (e.g., body size volume, discrete behaviour, phenology, number of hairs on a wing, carbon capture -- any measurable characteristic can be assigned to 'z' in the Price equation). Units of Omega in our equation will be identical to units of z, as used in the Price equation. Omega is just the sum across all individuals.

We appreciate that a concrete example can be helpful, and we thank Reviewer 2 for the suggestion. Here is how we would meaningfully use it: Let this small population of N = 3 be three different plants, which Reviewer 2 has assigned individual one a fecundity of beta = 2 and individuals two and three a fecundity of beta = 1. This population might be interpreted to have non-overlapping generations because death occurs for all individuals (delta = 1). Assume that we are interested in the rate of carbon capture of these plants, so this is the characteristic 'z' that we are measuring (e.g., in kg of carbon dioxide absorbed over some period of time). The DeltaZ = 0.5 would be a bit unusual, but could be caused by a change in carbon capture from t to t + 1 due to changing environmental conditions affecting the plants.

We could then calculate Omega as the summed rate of carbon capture attributable to the population from t to t + 1.

Omega = (2 - 1 + 1)(1 + 0.5) + (1 - 1 + 1)(2 + 0.5) + (1 - 1 + 1)(3 + 0.5) = 9

Note that Omega = 9 in the example Reviewer 2 gave (the 10.5 was just a calculation error). That is, the total carbon capture summed across all individuals is 9 kg. There is no problem with units here; the correct unit is kg, not a combination of individuals and kg.

Counts are unitless, inherently, and 'individual' is a label rather than a unit in this case, regardless of whether it is in beta (births attributable to individual i), delta (death indicator of i), or 1 (the count of i itself).

From the Bureau International des Poids et Mesures (2019) The International System of Units (SI) (9th ed.) (https://www.bipm.org/en/publications/si-brochure/): "There are also some quantities that cannot be described in terms of the seven base quantities of the SI, but are quantities that are a number of entities. Examples are a number of cellular or biomolecular entities, or degeneracy in quantum mechanics. These quantities are also quantities with the unit one", then "The unit one is the neutral element of any system of units – necessary and present automatically. There is no requirement to introduce it formally by decision."

A major part of our work here is to set the fundamental equations of ecology and evolution on as solid of a foundation as possible. And while the point about units that Reviewer 2 raises is thought-provoking (we had to go back to the basics ourselves to work out whether or not this was an issue for our equation), the logic of our equation remains airtight.

Intuitively, note that if we were just interested in measuring total carbon capture in a population, then it wouldn't be a problem to, hypothetically, just sum up how much each individual captures carbon. Similarly, if we wanted to measure total biomass, or decomposition, or pollination in a system (anything, really), it would be entirely sensible (if it were possible in practice) to add up the contributions of each individual to the whole. And this is what we're after with Omega -- the summed value of traits across all individuals.

Now, in the special case in which z = 1 (which is, again, a unitless count), we recover the total number of individuals, as derived in our manuscript. In this case, Omega is interpreted as individual number (N), and,

Omega = (2 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) = 4

Our population size has increased from 3 to 4 because two individuals produced one offspring each (beta = 1), and one produced two (beta = 2). All of those reproducing individuals died (delta = 1).

We thank Reviewer 2 for raising this query; it gave us pause to reflect and be even more deliberate about what our equation is describing. We now mention in our manuscript why this is not an issue by pointing out that individuals are unitless (LINE).

Reviewer 2 comment 3 
======================

Thus, if I'm understanding correctly, equation 1 only becomes meaningful once the steps are taken to reduce it to either the population growth rate equation or the Price Equation. This feels like deriving a fundamental equation in reverse – starting from two known fundamental equations, and then working backwards to piece them together. Which perhaps is a fine approach on its face, but if this equation does not produce evolutionarily or ecologically meaningful results on its own, it's unclear that it has any utility.

Author response to Reviewer 2 comment 3
======================

This is incorrect; eqn 1 is meaningful as written as a definition of eco-evolutionary change. It is, however, correct that the units associated with 'z' (and therefore 'Omega') are unresolved until 'z' is defined (see our response to Reviewer 2 comment 2 above). But this is no different from the Price equation itself. Our equation doesn't need to recover either the population growth rate equation or the Price Equation to be meaningful. On the contrary, the logic of our manuscript is to develop a unifying framework for eco-evolutionary change. Of course it is good and necessary to prove that we can recover the two most fundamental equations of ecology and evolution. But that is not the end of the story. If our equation only worked through specifying an ecological scenario or an evolutionary scenario, but not both, then it would not fulfil its intended purpose.

We can appreciate why it might feel like our equation 1 works backwards from the fundamental equations of ecology and evolution separately. But this isn't the case, logically or methodologically (see our response to Reviewer 2 comment 1). What we show is that the appropriate starting point for defining eco-evolutionary change is the interaction between births/deaths and individual characteristics. Our starting point isn't arbitrary or contingent; it is exact and universal, and no other equation would work as fundamental (anymore than the Price equation or general population change could be substituted by something else as fundamental equations).

We now emphasise more clearly, in words, the interpretation of equation 1 and how it is meaningful even prior to the derivation of population change or the Price equation (LINE XXX).

Reviewer 2 comment 4 
=========================

Lastly, there are a series of statements made in the Abstract that feel like they either are not followed through on or only weakly mentioned in the Discussion. For example, we are told that this unification “shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness”, but this is fleshed out in a single sentence in the Discussion (Line 236–238): “Interestingly, the rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively. Population growth rate reflects mean fitness w ̄ , while the rate of evolutionary change reflects the variance in fitness Var(w).” There appears to be no other mention of this outside the Abstract. Importantly, as the authors note, each of these things are already known – both the equivalence of population growth rate with mean fitness and the rate of evolutionary change being equal to the variance in fitness are found in Fisher (1930).

Author response to Reviewer 2 comment 4 
=========================

We now remove this from the abstract. In the manuscript, we now clarify that our derivation recovers this relationship, which has been pointed out previously.

Reviewer 2 comment 5
=========================

Furthermore, the Abstract claims the unification links ecosystem function to population growth rates and evolutionary change, but this is only speculated on in the last paragraph of the Discussion. The authors cite others who have explicitly done this, suggest that their equation might do the same, but there’s no demonstration of it or suggestion of what it might be. Again, this could be because it’s not obvious to me the connection, so the authors need to be a little more explicit in showing these connections.

Author response to Reviewer 2 comment 5
=========================

We are grateful to Reviewer 2 for pointing this out, and we have revised to make our point about ecosystem function clearer with a more concise explanation of our point in our response to Reviewer 2 comment 2. We now elaborate on how Omega can be interpreted as a measure of ecosystem function for a population (LINE XXX).

Reviewer 2 minor comment 1
=========================

Title – I like the title, but Darwin is never mentioned in the text! I kept expecting to see something about how “Darwin dreamed to unite these fields” or something.

Author response to Reviewer 2 minor comment 1
=========================

We also liked the title, but it is clearly causing confusion. We have switched to 'Foundations of ecological and evolutionary change'.

Reviewer 2 minor comment 2
=========================

Lines 26-28 – I would also cite Betty Smocovitis’ excellent paper here on the unification of biology via evolution: Smocovitis, V. B. (1992). Unifying biology: the evolutionary synthesis and evolutionary biology. Journal of the History of Biology, 25(1), 1-65.

Author response to Reviewer 2 minor comment 2
=========================

This is a good suggestion; we have cited this paper now (Line XXX).

Reviewer 2 minor comment 3
=========================

Lines 45-54 – This historical presentation seems odd to me. My reading of history is that these two fields have always been pretty tightly intertwined; Darwin’s argument relied on Malthusian concepts of population growth, after all, and one of the first concepts of density dependent selection comes from Haldane (1956). But I could be biased. Perhaps a citation after the sentence: “Until recently, population ecology and evolution in particular have taken almost parallel paths for developing theory” to others who have argued this would be helpful.

Author response to Reviewer 2 minor comment 3 
=========================

We are not historians and we don’t claim to have the ultimate answer on this issue. Our reading of history is the following. The theory of evolution by natural selection developed by Darwin is inherently an ecological theory, and the last third of the 19th century saw the development of ecological work rooted on ivolution. In a work devoted to the historical relationship between ecology and evolution, James P. Collins (1986, p. 260) states: "In 1866, for example, Haeckel claimed that the theory of evolution explained and formed the groundwork of ecology. Stephen Forbes argued in 1895 that the whole Darwinian doctrine belonged to ecology. Just after 1900 H. C. Cowles wrote, "If ecology has a place at all in modern biology, certainly one of its great tasks is to unravel the mysteries of adaptation." At that same time, J. G. Needham described ecology as "the study of the phenomena of fitness."

However, that doesn’t imply that ecological theory/practice and evolutionary theory/practice have always been as intertwined as one would think. The progressive specialisation of both areas produced subdisciplines with no evolutionary input/perspective, or evolution (specially adaptation) was just taken for granted, orienting their research to mechanistic/proximate questions (in terms of Mayr and Tinbergen conceptualization of proximate vs. ultimate questions). Moreover, fields such as systems ecology (from Odum to Margalef) could be thought of as a renewed version of the Romantische Naturphilosophie, with little (if any) connection with evolution. Collins (1986, p. 258) cites works from the 1950s claiming for the necessity of an autonomous ecological theory, free of evolutionary thinking. All of this may explain remarks like the following during the 1960s: "The theory of evolution by natural selection is an ecological theory -founded on ecological observation by perhaps the greatest of all ecologists. It has been adopted by and brought up by the science of genetics, and ecologists, being modest people, are apt to forget their distinguished parenthood. Indeed, Darwinian plant ecology has been largely neglected" (Harper 1967 p. 267).

After 1960, evolutionary ecology emerged and many ecology textbooks (Pianka 1974, Roughgarden 1979, Begon et al. 1986/2006) included an evolutionary perspective and entire chapters devoted to evolution. However, we think that the pursuit of a true synthesis of evolutionary and ecological change (eco-evo dynamics) has only taken place (roughly) in the last two decades, with several works trying to develop a more unified framework (as far as we know, only one textbook (Hendry 2017) has summarized these eco-evo efforts). Part of the motivation for this synthesis has been the relatively recent appreciation of concurrent ecological and evolutionary dynamics. From Yamamichi, Ellner, and Hairston Jr. (2023), "Although ecology and evolutionary biology have been neighbouring research areas ever since the age of Darwin, the prevailing assumption had long been that ecological processes occur much faster than evolutionary processes".

One of our critiques of eco-evolutionary theory is that there is no formal integration of the foundational axioms shared by both ecology and evolution. Our take is that we should have an eco-evo formalism tout court, which takes a separate but complementary role to existing eco-evolutionary models, which require simplifying assumptions and typically start with ecology or with evolution and then try to include the other one in the model.

Finally, we also agree with Reviewer 2 that classic evolutionary works (like those developed by Haldane) had ecological ingredients. Actually, as we said in our manuscript, our claim is that, contrary to a spread view among evolutionary biologists (see page XXX of our manuscript for references), population genetics has ecology at its core, though not in an obvious way.

We now clarify this historical view (Line XXX)

Reviewer 2 minor comment 4
=========================

Line 106 – “Fischer” should be spelled “Fisher”.

Author response to Reviewer 2 minor comment 4
=========================

Fixed (Line XXX).

Reviewer 2 minor comment 5 
=========================

Lines 273-274 – If these derivations were included here, it would dramatically clarify my confusion about the usefulness and meaning of Omega in the general formula.

Author response to Reviewer 2 minor comment 5  
=========================

We are reluctant to include too much work still in preparation, or distract from what we believe is the central message of the current manuscript. But we have now included some additional derivations in supporting information.

Author references
==============================

Baravalle, L., Jonathan Roffé, A., Luque, V. J., & Ginnobili, S. (2024). The Value of Price. Biological Theory. https://doi.org/10.1007/s13752-024-00482-4

Barfield, M., Holt, R. D., & Gomulkiewicz, R. (2011). Evolution in stage-structured populations. American Naturalist, 177(4), 397–409. https://doi.org/10.1086/658903

Bureau international Des poids et mesures (2019) The International System of Units (SI).
https://www.bipm.org/documents/20126/41483022/SI-Brochure-9.pdf

Caswell, H. (1988). Theory and models in ecology: a different perspective. Ecological Modelling, 43, 33–44.

Coulson, T., & Tuljapurkar, S. (2008). The dynamics of a quantitative trait in an age-structured population living in a variable environment. American Naturalist, 172(5), 599–612. https://doi.org/10.1086/591693

Levins, R. (1966). The strategy of model building in population biology. In American Naturalist (Vol. 54, Issue 4, pp. 421–431). https://doi.org/10.2307/27836590

Frank, S. A. (1995). George Price’s contributions to evolutionary genetics. Journal of Theoretical Biology, 175, 373–388.

Frank, S. A. (1997). The Price equation, Fisher’s fundamental theorem, kin selection, and causal analysis. Evolution, 51(6), 1712–1729.

Frank, S. A. (2012). Natural selection. IV. The Price equation. Journal of Evolutionary Biology, 25, 1002–1019. https://doi.org/10.1111/j.1420-9101.2012.02498.x

Frank, S. A. (2015). D’Alembert’s direct and inertial forces acting on populations: The price equation and the fundamental theorem of natural selection. Entropy, 17(10), 7087–7100. https://doi.org/10.3390/e17107087

Frank, S. A. (2016). Common probability patterns arise from simple invariances. Entropy, 18(5), 1–22. https://doi.org/10.3390/e18050192

Price, G. R. (1970). Selection and covariance. In Nature (Vol. 227, Issue 5257, pp. 520–521). https://doi.org/10.1038/227520a0

Price, G. R. (1972). Extension of covariance selection mathematics. Annals of Human Genetics, 35(4), 485–490. https://doi.org/10.1111/j.1469-1809.1957.tb01874.x

Queller, D. C. (2017). Fundamental theorems of evolution. American Naturalist, 189(4), 000–000. https://doi.org/10.1086/690937

Lion, S. (2018). Theoretical approaches in evolutionary ecology: environmental feedback as a unifying perspective. American Naturalist, 191(1). https://doi.org/10.1086/694865

Lion, S., Sasaki, A., & Boots, M. (2023). Extending eco-evolutionary theory with oligomorphic dynamics. Ecology Letters, 26(S1), S22–S46. https://doi.org/10.1111/ele.14183

Patel, S., Cortez, M. H., & Schreiber, S. J. (2018). Partitioning the effects of eco-evolutionary feedbacks on community stability. American Naturalist, 191(3), 1–29. https://doi.org/10.1101/104505

Govaert, L., Fronhofer, E. A., Lion, S., Eizaguirre, C., Bonte, D., Egas, M., Hendry, A. P., de Brito Martins, A., Melián, C. J., Raeymaekers, J. A., Ratikainen, I. I., Saether, B. E., Schweitzer, J. A., & Matthews, B. (2019). Eco-evolutionary feedbacks—Theoretical models and perspectives. Functional Ecology, 33(1), 13–30. https://doi.org/10.1111/1365-2435.13241

Rees, M., & Ellner, S. P. (2016). Evolving integral projection models: Evolutionary demography meets eco-evolutionary dynamics. Methods in Ecology and Evolution, 7(2), 157–170. https://doi.org/10.1111/2041-210X.12487

Lande, R. (1982). A quantitative genetic theory of life history evolution. Ecology, 63(3), 607–615. http://www.primes.colostate.edu/Lande 1982.pdf

Rice, S. H. (2004). Evolutionary theory: mathematical and conceptual foundations. Sinauer Associates.

Gardner, A. (2020). Price’s equation made clear. Philosophical Transactions of the Royal Society B: Biological Sciences, 375(1797), 20190361. https://doi.org/10.1098/rstb.2019.0361

Gardner, A. (2008). The Price equation. Current Biology, 18(5), 198–202. https://doi.org/10.1016/j.cub.2008.01.005

Harper, J. L. (1967). A Darwinian Approach to Plant Ecology. In Source: Journal of Ecology (Vol. 55, Issue 2). https://about.jstor.org/terms

Pianka, E. R. (1974). Evolutionary ecology.

Roughgarden, J. (1979). Theory of Population Genetics and Evolutionary Ecology: An Introduction.

Begon, M. et al. (2020). Ecology: From Individuals to Ecosystems.

Hendry, A. P. (2017). Eco-evolutionary dynamics. Princeton university press.

Yamamichi, M., Ellner, S. P., & Hairston, N. G. (2023). Beyond simple adaptation: Incorporating other evolutionary processes and concepts into eco-evolutionary dynamics. Ecology Letters, 26(S1), S16–S21. https://doi.org/10.1111/ele.14197

### 2025-05-29 foundations_of_ecoevo.Rmd
Initialise response to reviewer comments, with current manuscript

@@
 output:
+  word_document: default
   pdf_document:
@@
     df_print: paged
-  word_document: default
 header-includes:
@@
 linkcolor: blue
+csl: the-american-naturalist.csl
 biblio-style: apalike
@@
 
-**Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density.**
-**The deep interdependence between ecology and evolution is well-established, but much theory in each discipline has been developed in isolation.**
-**While recent work has accomplished eco-evolutionary synthesis, a gap remains between the logical foundations of ecology and evolution.**
-**We bridge this gap with a new equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived.**
-**We thereby unify the fundamental equations of population ecology and biological evolution under a general framework.**
-**Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness, shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness, respectively, and links this change to ecosystem function.**
-**Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.**
+Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density. 
+The deep interdependence between ecology and evolution is well-established, but much theory in each discipline has been developed in isolation. 
+While recent work has accomplished eco-evolutionary synthesis, a gap remains between the logical foundations of ecology and evolution. 
+We bridge this gap with a new equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived.
+We thereby unify the fundamental equations of population ecology and biological evolution under a general framework. 
+Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness, shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness, respectively, and links this change to ecosystem function.
+Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.
 
 **Key words:** Ecology, Evolution, Eco-Evolutionary Theory, Fundamental Theorem, Price Equation, Population Growth
+
+<!---
 
@@
 Biologists have long known that the ecology of population size is inextricably linked with the evolution of population traits.
-Nevertheless, the foundational theory underlying ecological and evolutionary change have been historically separated, with recent attempts to bridge this theoretical divide using eco-evolutionary modelling.
+Nevertheless, foundational theories underlying ecological and evolutionary change have been historically separated, with recent attempts to bridge the theoretical divide using eco-evolutionary modelling.
 We discover a new equation from first principles that provides a universal and complete definition for eco-evolutionary change.
 This equation thereby offers a unified framework for ecology and evolution, which also recovers the foundational theory of ecology and evolution as developed in isolation.
+
+--->
 
@@
 Likewise, in the life sciences, the modern synthesis of natural selection and Mendelian inheritance revolutionised how mechanisms underlying evolution and genetics are understood and applied [@Gayon2019].
-The analogy to physics is imperfect because evolution encompasses non-biological systems with non-genetic inheritance [e.g., @Price1995; @Shennan2011; @MacCallum2012; @Lewens2015], but to illustrate our point, we restrict our focus to biological evolution.
+The analogy to physics is imperfect because evolution encompasses non-biological systems with non-genetic inheritance [@Price1995; @Shennan2011; @MacCallum2012; @Lewens2015], but to illustrate our point, we restrict our focus to biological evolution.
 Following the modern synthesis, patterns in evolutionary biology and genetics could be interpreted through a common framework.
@@
 Until recently, population ecology and evolution in particular have taken almost parallel paths for developing theory.
-Although many revelatory models exist to connect ecological and evolutionary phenomena [e.g., @Lion2018; @Patel2018; @Govaert2019], the foundational assumptions and equations of ecology and evolution in these models remain mostly separated conceptually.
+Although many revelatory models exist to connect ecological and evolutionary phenomena [@Lion2018; @Patel2018; @Govaert2019], the foundational assumptions and equations of ecology and evolution in these models remain mostly separated conceptually.
 But like the mechanisms driving evolutionary and genetic change, those that drive ecological and evolutionary change are well-established with predictable joint consequences.
@@
 To fully unify biological evolution and population ecology, we must reconcile the general equation for population change (Box 1) with the Price equation (Box 2).
-Previous theory has produced mathematical synthesis between ecological and evolutionary equations [e.g., @Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
+Previous theory has produced mathematical synthesis between ecological and evolutionary equations [@Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
 For example, @Page2002 demonstrate a mathematical equivalence between the Price equation and a general equation for community dynamics. 
@@
 It has also been highly useful for integrating evolutionary theory across disciplines [@Fox2006; @Brantingham2007; @MacCallum2012; @Frank2015; @Godsoe2021; @Ulrich2024].
-These properties would seem to make it an intuitive starting point for a logical foundation of ecology and evolution, perhaps through some kind of mathematical equivalence [e.g., @Page2002] or addition of terms [e.g., @Collins2009], or through the use of its recursive structure [e.g., @Kerr2009; @Frank2012].
+These properties would seem to make it an intuitive starting point for a logical foundation of ecology and evolution, perhaps through some kind of mathematical equivalence [@Page2002] or addition of terms [@Collins2009], or through the use of its recursive structure [@Kerr2009; @Frank2012].
 But despite its flexibility, the Price equation still relies on relative frequencies, which must by definition sum to one [@Frank2015].
@@
 From eqn 1, we can derive the most fundamental equations of population ecology (Box 1) and evolutionary biology (Box 2) through an appropriate interpretation of $z_{i}$.
+
+<!---
+
+19 MAY: Point out here that by itself, Omega has the same terms as z.
+
+- Verbally, our equation shows interaction between fitness and an individual's characteristic, summed
+- Because it is summing across individuals, some characteristics can be interpreted as measure of ecosystem function
+- Ecosystem function will make the assumption of additivity of z_i across all N
+- We will see that mean fitness is population ecology
+- Variance is evolutionary change
+- Summation is ecosystem function
+
+Eqn 1 captures an invariant relationship between individual fitness and an individual's characteristic
+
+Also note that directly tying fitness to demography
+
+--->
 
@@
 
-We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [eqn 3, @Turchin2001].
+We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [@Turchin2001] (eqn 3).
 
@@
 The Price equation provides a complete and exact description of evolution in any closed evolving system [@Price1970; @Frank2012].
-It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [e.g., individuals, alleles, @Price1970; @Gardner2008; @Luque2016].
-This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation [e.g., group, individual, @Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
+It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [@Price1970; @Gardner2008; @Luque2016] (e.g., individuals, alleles).
+This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation (e.g., group, individual) [@Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
 Because of its abstract nature and lack of any system-specific assumptions, the Price equation makes no predictions about what will happen in any particular system [@Gardner2020].
@@
 
-Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ [Box 1, @Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 2).
+Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ (Box 1) [@Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 2).
 The population growth equation $N_{t+1} = N_{t}\lambda$ could therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
 We do not claim to be the first to notice this equivalence between population growth and mean population fitness.
-Indeed, the relationship between population growth rate and evolutionary fitness has been proposed and applied many times before [e.g., @Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
+Indeed, the relationship between population growth rate and evolutionary fitness has been proposed and applied many times before [@Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
 For example, @Lande1982 explicitly concludes mean absolute fitness per unit time is $\bar{w} = e^{r}$, where $r = ln(\lambda)$.
@@
 Interestingly, the rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
-Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)$ [i.e., Fisher's fundamental theorem, @Frank1997; @Rice2004; @Queller2017].
+Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)$ (i.e., Fisher's fundamental theorem) [@Frank1997; @Rice2004; @Queller2017].
 
 Our unification may also help explain, at least partially, some of the success of classical population genetic models. 
-For decades, population genetics (and to some extent quantitative genetics) has been accused of being a reductionistic view of evolution, reducing everything to changes in allele frequencies and abstracting away from individuals and their environments [e.g., the ecological interactions, @MacColl2011].
+For decades, population genetics (and to some extent quantitative genetics) has been accused of being a reductionistic view of evolution, reducing everything to changes in allele frequencies and abstracting away from individuals and their environments (the ecological interactions) [@MacColl2011].
 This has been a line of argumentation by some defenders of the so-called Extended Synthesis [@Pigliucci2009], especially in relation to niche construction [@Odling-smee2003].
@@
 For example, @Fox2006 applied the abstract properties of the Price equation to partition total change in ecosystem function into separate components attributable to species richness, species composition, and context dependent effects.
-This approach provides a framework for comparing the effects of biodiversity on ecosystem function in empirical systems [e.g., @Fox2006; @Winfree2015; @Mateo-Tomas2017].
+This approach provides a framework for comparing the effects of biodiversity on ecosystem function in empirical systems [@Fox2006; @Winfree2015; @Mateo-Tomas2017].
 Instead, our eqn 1 defines $\Omega$ as total ecosystem function contributed by a focal population (e.g., biomass production, decomposition, photosynthetic rate).
@@
 
+Data Availability
+====================================================
+
+This work does not include any data.
+
 References

### 2025-05-29 foundations_of_ecoevo.tex
Initialise response to reviewer comments, with current manuscript

@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
@@
 \fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
 
@@
 
-\textbf{Biological evolution is realised through the same mechanisms of
-birth and death that underlie change in population density.} \textbf{The
-deep interdependence between ecology and evolution is well-established,
-but much theory in each discipline has been developed in isolation.}
-\textbf{While recent work has accomplished eco-evolutionary synthesis, a
-gap remains between the logical foundations of ecology and evolution.}
-\textbf{We bridge this gap with a new equation that defines a summed
-value for a characteristic across individuals in a population, from
-which the fundamental equations of population ecology and evolutionary
-biology (the Price equation) are derived.} \textbf{We thereby unify the
-fundamental equations of population ecology and biological evolution
-under a general framework.} \textbf{Our unification further demonstrates
-the equivalence between mean population growth rate and evolutionary
-fitness, shows how ecological and evolutionary change are reflected in
-the first and second statistical moments of fitness, respectively, and
-links this change to ecosystem function.} \textbf{Finally, we outline
-how our proposed framework can be used to unify social evolution and
-density-dependent population growth.}
+Biological evolution is realised through the same mechanisms of birth
+and death that underlie change in population density. The deep
+interdependence between ecology and evolution is well-established, but
+much theory in each discipline has been developed in isolation. While
+recent work has accomplished eco-evolutionary synthesis, a gap remains
+between the logical foundations of ecology and evolution. We bridge this
+gap with a new equation that defines a summed value for a characteristic
+across individuals in a population, from which the fundamental equations
+of population ecology and evolutionary biology (the Price equation) are
+derived. We thereby unify the fundamental equations of population
+ecology and biological evolution under a general framework. Our
+unification further demonstrates the equivalence between mean population
+growth rate and evolutionary fitness, shows how ecological and
+evolutionary change are reflected in the first and second statistical
+moments of fitness, respectively, and links this change to ecosystem
+function. Finally, we outline how our proposed framework can be used to
+unify social evolution and density-dependent population growth.
 
@@
 Fundamental Theorem, Price Equation, Population Growth
-
-\hypertarget{non-technical-summary}{%
-\section{Non-technical Summary}\label{non-technical-summary}}
-
-Biologists have long known that the ecology of population size is
-inextricably linked with the evolution of population traits.
-Nevertheless, the foundational theory underlying ecological and
-evolutionary change have been historically separated, with recent
-attempts to bridge this theoretical divide using eco-evolutionary
-modelling. We discover a new equation from first principles that
-provides a universal and complete definition for eco-evolutionary
-change. This equation thereby offers a unified framework for ecology and
-evolution, which also recovers the foundational theory of ecology and
-evolution as developed in isolation.
 
@@
 to physics is imperfect because evolution encompasses non-biological
-systems with non-genetic inheritance (e.g.,
-\protect\hyperlink{ref-Price1995}{Price 1995};
+systems with non-genetic inheritance
+(\protect\hyperlink{ref-Price1995}{Price 1995};
 \protect\hyperlink{ref-Shennan2011}{Shennan 2011};
@@
 Although many revelatory models exist to connect ecological and
-evolutionary phenomena (e.g., \protect\hyperlink{ref-Lion2018}{Lion
-2018}; \protect\hyperlink{ref-Patel2018}{Patel, Cortez, and Schreiber
-2018}; \protect\hyperlink{ref-Govaert2019}{Govaert et al. 2019}), the
+evolutionary phenomena (\protect\hyperlink{ref-Lion2018}{Lion 2018};
+\protect\hyperlink{ref-Patel2018}{Patel et al. 2018};
+\protect\hyperlink{ref-Govaert2019}{Govaert et al. 2019}), the
 foundational assumptions and equations of ecology and evolution in these
@@
 \protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Lehtonen2020}{Lehtonen, Okasha, and Helanterä
-2020}), and it serves as the starting point for all other fundamental
-equations in evolution (\protect\hyperlink{ref-Queller2017}{Queller
-2017}). We believe that this logical foundation can be made sufficiently
-broad to encompass both population and evolutionary change.
+\protect\hyperlink{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it
+serves as the starting point for all other fundamental equations in
+evolution (\protect\hyperlink{ref-Queller2017}{Queller 2017}). We
+believe that this logical foundation can be made sufficiently broad to
+encompass both population and evolutionary change.
 
@@
 breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
-easily derived (\protect\hyperlink{ref-Queller2017}{Queller 2017};
-\protect\hyperlink{ref-Lehtonen2016}{Lehtonen 2016},
-\protect\hyperlink{ref-Lehtonen2018}{2018}). This abstractness is also
-key to developing a more general view of evolution
+easily derived (\protect\hyperlink{ref-Lehtonen2016}{Lehtonen 2016},
+\protect\hyperlink{ref-Lehtonen2018}{2018};
+\protect\hyperlink{ref-Queller2017}{Queller 2017}). This abstractness is
+also key to developing a more general view of evolution
 (\protect\hyperlink{ref-Rice2020}{Rice 2020};
 \protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021};
-\protect\hyperlink{ref-Edelaar2023}{Edelaar, Otsuka, and Luque 2023}).
-In contrast, some researchers consider the Price equation just a
-triviality (even tautological), and useless without further modelling
-assumptions (\protect\hyperlink{ref-VanVeelen2005}{van Veelen 2005};
+\protect\hyperlink{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast,
+some researchers consider the Price equation just a triviality (even
+tautological), and useless without further modelling assumptions
+(\protect\hyperlink{ref-VanVeelen2005}{van Veelen 2005};
 \protect\hyperlink{ref-VanVeelen2012}{van Veelen et al. 2012}). The
@@
 Price equation (Box 2). Previous theory has produced mathematical
-synthesis between ecological and evolutionary equations (e.g.,
-\protect\hyperlink{ref-Page2002}{Page and Nowak 2002};
+synthesis between ecological and evolutionary equations
+(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002};
 \protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but not
@@
 \protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Lehtonen2018}{Lehtonen 2018};
-\protect\hyperlink{ref-Queller2017}{Queller 2017}). It has also been
+\protect\hyperlink{ref-Queller2017}{Queller 2017};
+\protect\hyperlink{ref-Lehtonen2018}{Lehtonen 2018}). It has also been
 highly useful for integrating evolutionary theory across disciplines
@@
 logical foundation of ecology and evolution, perhaps through some kind
-of mathematical equivalence (e.g., \protect\hyperlink{ref-Page2002}{Page
-and Nowak 2002}) or addition of terms (e.g.,
-\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), or
-through the use of its recursive structure (e.g.,
-\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
+of mathematical equivalence (\protect\hyperlink{ref-Page2002}{Page and
+Nowak 2002}) or addition of terms
+(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), or
+through the use of its recursive structure
+(\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
 \protect\hyperlink{ref-Frank2012}{Frank 2012}). But despite its
@@
 We therefore recover the general equation for population ecology (eqn 2)
-and the fundamental property of exponential growth in populations (eqn
-3, \protect\hyperlink{ref-Turchin2001}{Turchin 2001}).
+and the fundamental property of exponential growth in populations
+(\protect\hyperlink{ref-Turchin2001}{Turchin 2001}) (eqn 3).
 
@@
 rearranging the mathematical notation defining changes in the
-frequencies and characteristics of any type of entity (e.g.,
-individuals, alleles, \protect\hyperlink{ref-Price1970}{Price 1970};
+frequencies and characteristics of any type of entity
+(\protect\hyperlink{ref-Price1970}{Price 1970};
 \protect\hyperlink{ref-Gardner2008}{Gardner 2008};
-\protect\hyperlink{ref-Luque2016}{Luque 2017}). This derivation
-partitions total characteristic change into different components, making
-it possible to isolate evolutionary mechanisms (e.g., selection) and
-levels of biological organisation (e.g., group, individual,
-\protect\hyperlink{ref-Frank1995}{Frank 1995},
-\protect\hyperlink{ref-Frank2012}{2012};
+\protect\hyperlink{ref-Luque2016}{Luque 2017}) (e.g., individuals,
+alleles). This derivation partitions total characteristic change into
+different components, making it possible to isolate evolutionary
+mechanisms (e.g., selection) and levels of biological organisation
+(e.g., group, individual) (\protect\hyperlink{ref-Frank1995}{Frank
+1995}, \protect\hyperlink{ref-Frank2012}{2012};
 \protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
@@
 Our unification demonstrates the equivalence between the finite rate of
-increase \(\lambda\) (Box 1, \protect\hyperlink{ref-Gotelli2001}{Gotelli
-2001}) and population mean evolutionary fitness \(\bar{w}\) (Box 2). The
-population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
-rewritten as \(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first
-to notice this equivalence between population growth and mean population
-fitness. Indeed, the relationship between population growth rate and
-evolutionary fitness has been proposed and applied many times before
-(e.g., \protect\hyperlink{ref-Fisher1930}{Fisher 1930};
+increase \(\lambda\) (Box 1)
+(\protect\hyperlink{ref-Gotelli2001}{Gotelli 2001}) and population mean
+evolutionary fitness \(\bar{w}\) (Box 2). The population growth equation
+\(N_{t+1} = N_{t}\lambda\) could therefore be rewritten as
+\(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first to notice
+this equivalence between population growth and mean population fitness.
+Indeed, the relationship between population growth rate and evolutionary
+fitness has been proposed and applied many times before
+(\protect\hyperlink{ref-Fisher1930}{Fisher 1930};
 \protect\hyperlink{ref-Charlesworth1980}{Charlesworth 1980};
@@
 \(\bar{w}\), while the rate of evolutionary change reflects the variance
-in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem,
-\protect\hyperlink{ref-Frank1997}{Frank 1997};
+in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem)
+(\protect\hyperlink{ref-Frank1997}{Frank 1997};
 \protect\hyperlink{ref-Rice2004}{Rice 2004};
@@
 in allele frequencies and abstracting away from individuals and their
-environments (e.g., the ecological interactions,
-\protect\hyperlink{ref-MacColl2011}{MacColl 2011}). This has been a line
-of argumentation by some defenders of the so-called Extended Synthesis
-(\protect\hyperlink{ref-Pigliucci2009}{Pigliucci 2009}), especially in
-relation to niche construction
-(\protect\hyperlink{ref-Odling-smee2003}{Odling-Smee, Laland, and
-Feldman 2003}). Famously, Mayr (\protect\hyperlink{ref-Mayr1960}{1959})
-characterized population genetics as a simple input and output of genes,
-analogous to ``the adding of certain beans to a beanbag and the
-withdrawing of others'' (also called ``beanbag genetics'').
+environments (the ecological interactions)
+(\protect\hyperlink{ref-MacColl2011}{MacColl 2011}). This has been a
+line of argumentation by some defenders of the so-called Extended
+Synthesis (\protect\hyperlink{ref-Pigliucci2009}{Pigliucci 2009}),
+especially in relation to niche construction
+(\protect\hyperlink{ref-Odling-smee2003}{Odling-Smee et al. 2003}).
+Famously, Mayr (\protect\hyperlink{ref-Mayr1960}{1959}) characterized
+population genetics as a simple input and output of genes, analogous to
+``the adding of certain beans to a beanbag and the withdrawing of
+others'' (also called ``beanbag genetics'').
 
@@
 provides a framework for comparing the effects of biodiversity on
-ecosystem function in empirical systems (e.g.,
-\protect\hyperlink{ref-Fox2006}{Fox 2006};
+ecosystem function in empirical systems
+(\protect\hyperlink{ref-Fox2006}{Fox 2006};
 \protect\hyperlink{ref-Winfree2015}{Winfree et al. 2015};
@@
 
+\hypertarget{data-availability}{%
+\section{Data Availability}\label{data-availability}}
+
+This work does not include any data.
+
 \hypertarget{references}{%
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{0}{0}
 \leavevmode\vadjust pre{\hypertarget{ref-Baravalle2022}{}}%
-Baravalle, Lorenzo, and Victor J. Luque. 2022. {``Towards a Pricean
-Foundation for Cultural Evolutionary Theory.''} \emph{Theoria} 37 (2):
-209--31. \url{https://doi.org/10.1387/theoria.21940}.
+Baravalle, L., and V. J. Luque. 2022.
+\href{https://doi.org/10.1387/theoria.21940}{Towards a pricean
+foundation for cultural evolutionary theory}. Theoria 37:209--231.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Baravalle2024}{}}%
-Baravalle, Lorenzo, Ariel Roffé, Victor J. Luque, and Santiago
-Ginnobili. 2024. {``The Value of Price.''} In \emph{Biological Theory}.
-\url{https://doi.org/10.1007/s13752-024-00482-4}.
+Baravalle, L., A. Roffé, V. J. Luque, and S. Ginnobili. 2024.
+\href{https://doi.org/10.1007/s13752-024-00482-4}{The value of price}.
+Biological Theory.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Brantingham2007}{}}%
-Brantingham, P. Jeffrey. 2007. {``{A unified evolutionary model of
-archaeological style and function based on the Price equation}.''}
-\emph{American Antiquity} 72 (3): 395--416.
-\url{https://doi.org/10.2307/40035853}.
+Brantingham, P. J. 2007. \href{https://doi.org/10.2307/40035853}{{A
+unified evolutionary model of archaeological style and function based on
+the Price equation}}. American Antiquity 72:395--416.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Caswell1988}{}}%
-Caswell, Hal. 1988. {``Theory and Models in Ecology: A Different
-Perspective.''} \emph{Ecological Modelling} 43 (1-2): 33--44.
-\url{https://doi.org/10.1016/0304-3800(88)90071-3}.
+Caswell, H. 1988.
+\href{https://doi.org/10.1016/0304-3800(88)90071-3}{Theory and models in
+ecology: A different perspective}. Ecological modelling 43:33--44.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Charlesworth1980}{}}%
-Charlesworth, Brian. 1980. \emph{Evolution in Age-Structured
-Populations}. Cambridge Studies in Mathematical Biology. Cambridge:
-Cambridge University Press.
+Charlesworth, B. 1980. Evolution in age-structured populations.
+Cambridge studies in mathematical biology. Cambridge University Press,
+Cambridge.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Collins2009}{}}%
-Collins, Sinéad, and Andy Gardner. 2009. {``{Integrating physiological,
-ecological and evolutionary change: A Price equation approach}.''}
-\emph{Ecology Letters} 12 (8): 744--57.
-\url{https://doi.org/10.1111/j.1461-0248.2009.01340.x}.
+Collins, S., and A. Gardner. 2009.
+\href{https://doi.org/10.1111/j.1461-0248.2009.01340.x}{{Integrating
+physiological, ecological and evolutionary change: A Price equation
+approach}}. Ecology Letters 12:744--757.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Connor2004}{}}%
-Connor, J., and Daniel L. Hartl. 2004. \emph{{A premier of ecological
-genetics}}. Sinauer Associates Incorporated.
+Connor, J., and D. L. Hartl. 2004. {A premier of ecological genetics}.
+Sinauer Associates Incorporated.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Dobzhansky1970}{}}%
-Dobzhansky, Theodosius. 1970. \emph{Genetics of the Evolutionary
-Process}. Vol. 139. Columbia University Press.
+Dobzhansky, T. 1970. Genetics of the evolutionary process (Vol. 139).
+Columbia University Press.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Edelaar2023}{}}%
-Edelaar, Pim, Jun Otsuka, and Victor J. Luque. 2023. {``{A generalised
-approach to the study and understanding of adaptive evolution}.''}
-\emph{Biological Reviews} 98 (1): 352--75.
-\url{https://doi.org/10.1111/brv.12910}.
+Edelaar, Pim, J. Otsuka, and V. J. Luque. 2023.
+\href{https://doi.org/10.1111/brv.12910}{{A generalised approach to the
+study and understanding of adaptive evolution}}. Biological Reviews
+98:352--375.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Fisher1930}{}}%
-Fisher, Ronald A. 1930. \emph{The Genetical Theory of Natural
-Selection}. Oxford University Press, Oxford, UK.
+Fisher, R. A. 1930. The genetical theory of natural selection. Oxford
+University Press, Oxford, UK.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Fox2006}{}}%
-Fox, Jeremy W. 2006. {``{Using the price equation to partition the
-effects of biodiversity loss on ecosystem function}.''} \emph{Ecology}
-87 (11): 2687--96.
-\url{https://doi.org/10.1890/0012-9658(2006)87\%5B2687:utpetp\%5D2.0.co;2}.
+Fox, J. W. 2006.
+\href{https://doi.org/10.1890/0012-9658(2006)87\%5B2687:utpetp\%5D2.0.co;2}{{Using
+the price equation to partition the effects of biodiversity loss on
+ecosystem function}}. Ecology 87:2687--2696.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Frank1995}{}}%
-Frank, Steven A. 1995. {``{George Price's contributions to evolutionary
-genetics}.''} \emph{Journal of Theoretical Biology} 175: 373--88.
-\url{https://doi.org/10.1006/jtbi.1995.0148}.
+Frank, S. A. 1995. \href{https://doi.org/10.1006/jtbi.1995.0148}{{George
+Price's contributions to evolutionary genetics}}. Journal of Theoretical
+Biology 175:373--388.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Frank1997}{}}%
----------. 1997. {``{The Price equation, Fisher's fundamental theorem,
-kin selection, and causal analysis}.''} \emph{Evolution} 51 (6):
-1712--29. \url{https://doi.org/10.1111/j.1558-5646.1997.tb05096.x}.
+---------. 1997.
+\href{https://doi.org/10.1111/j.1558-5646.1997.tb05096.x}{{The Price
+equation, Fisher's fundamental theorem, kin selection, and causal
+analysis}}. Evolution 51:1712--1729.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Frank2012}{}}%
----------. 2012. {``{Natural selection. IV. The Price equation}.''}
-\emph{Journal of Evolutionary Biology} 25: 1002--19.
-\url{https://doi.org/10.1111/j.1420-9101.2012.02498.x}.
+---------. 2012.
+\href{https://doi.org/10.1111/j.1420-9101.2012.02498.x}{{Natural
+selection. IV. The Price equation}}. Journal of Evolutionary Biology
+25:1002--1019.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Frank2015}{}}%
----------. 2015. {``{D'Alembert's direct and inertial forces acting on
-populations: The price equation and the fundamental theorem of natural
-selection}.''} \emph{Entropy} 17 (10): 7087--7100.
-\url{https://doi.org/10.3390/e17107087}.
+---------. 2015. \href{https://doi.org/10.3390/e17107087}{{D'Alembert's
+direct and inertial forces acting on populations: The price equation and
+the fundamental theorem of natural selection}}. Entropy 17:7087--7100.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Frank2016}{}}%
----------. 2016. {``{Common probability patterns arise from simple
-invariances}.''} \emph{Entropy} 18 (5): 1--22.
-\url{https://doi.org/10.3390/e18050192}.
+---------. 2016. \href{https://doi.org/10.3390/e18050192}{{Common
+probability patterns arise from simple invariances}}. Entropy 18:1--22.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Frank2015a}{}}%
----------. 2017. {``{Universal expressions of population change by the
-Price equation: natural selection, information, and maximum entropy
-production}.''} \emph{Ecology and Evolution}, no. February: 1--16.
-\url{https://doi.org/10.1002/ece3.2922}.
+---------. 2017. \href{https://doi.org/10.1002/ece3.2922}{{Universal
+expressions of population change by the Price equation: natural
+selection, information, and maximum entropy production}}. Ecology and
+Evolution 1--16.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Gardner2008}{}}%
-Gardner, Andy. 2008. {``{The Price equation}.''} \emph{Current Biology}
-18 (5): 198--202. \url{https://doi.org/10.1016/j.cub.2008.01.005}.
+Gardner, A. 2008. \href{https://doi.org/10.1016/j.cub.2008.01.005}{{The
+Price equation}}. Current Biology 18:198--202.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Gardner2020}{}}%
----------. 2020. {``{Price's equation made clear}.''}
-\emph{Philosophical Transactions of the Royal Society B: Biological
-Sciences} 375 (1797): 20190361.
-\url{https://doi.org/10.1098/rstb.2019.0361}.
+---------. 2020. \href{https://doi.org/10.1098/rstb.2019.0361}{{Price's
+equation made clear}}. Philosophical Transactions of the Royal Society
+B: Biological Sciences 375:20190361.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Gayon2019}{}}%
-Gayon, Jean, and Philippe Huneman. 2019. {``The Modern Synthesis:
-Theoretical or Institutional Event?''} \emph{Journal of the History of
-Biology} 52 (4): 519--35.
-\url{https://doi.org/10.1007/s10739-019-09569-2}.
+Gayon, J., and P. Huneman. 2019.
+\href{https://doi.org/10.1007/s10739-019-09569-2}{The modern synthesis:
+Theoretical or institutional event?} Journal of the History of Biology
+52:519--535.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Godsoe2021}{}}%
-Godsoe, William, Katherine E. Eisen, Daniel Stanton, and Katherine M.
-Sirianni. 2021. {``Selection and Biodiversity Change.''}
-\emph{Theoretical Ecology} 14 (3): 367--79.
-\url{https://doi.org/10.1007/s12080-020-00478-3}.
+Godsoe, W., K. E. Eisen, D. Stanton, and K. M. Sirianni. 2021.
+\href{https://doi.org/10.1007/s12080-020-00478-3}{Selection and
+biodiversity change}. Theoretical Ecology 14:367--379.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Gotelli2001}{}}%
-Gotelli, Nicholas J. 2001. {``A Primer of Ecology. Sinauer Associate.''}
-\emph{Inc. Sunderland, MA}.
+Gotelli, N. J. 2001. A primer of ecology. Sinauer associate. Inc.
+Sunderland, MA.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Govaert2019}{}}%
-Govaert, Lynn, Emanuel A. Fronhofer, Sébastien Lion, Christophe
-Eizaguirre, Dries Bonte, Martijn Egas, Andrew P. Hendry, et al. 2019.
-{``{Eco-evolutionary feedbacks---Theoretical models and
-perspectives}.''} \emph{Functional Ecology} 33 (1): 13--30.
-\url{https://doi.org/10.1111/1365-2435.13241}.
+Govaert, L., E. A. Fronhofer, S. Lion, C. Eizaguirre, D. Bonte, M. Egas,
+A. P. Hendry, et al. 2019.
+\href{https://doi.org/10.1111/1365-2435.13241}{{Eco-evolutionary
+feedbacks---Theoretical models and perspectives}}. Functional Ecology
+33:13--30.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Kerr2009}{}}%
-Kerr, Benjamin, and Peter Godfrey-Smith. 2009. {``{Generalization of the
-price equation for evolutionary change}.''} \emph{Evolution} 63 (2):
-531--36. \url{https://doi.org/10.1111/j.1558-5646.2008.00570.x}.
+Kerr, B., and P. Godfrey-Smith. 2009.
+\href{https://doi.org/10.1111/j.1558-5646.2008.00570.x}{{Generalization
+of the price equation for evolutionary change}}. Evolution 63:531--536.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Kitcher1993}{}}%
-Kitcher, Philip. 1993. \emph{{The advancement of science}}. New York:
-Oxford University Press.
+Kitcher, P. 1993. {The advancement of science}. Oxford University Press,
+New York.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lande1982}{}}%
-Lande, Russell. 1982. {``{A quantitative genetic theory of life history
-evolution}.''} \emph{Ecology} 63 (3): 607--15.
-\url{https://doi.org/10.2307/1936778}.
+Lande, R. 1982. \href{https://doi.org/10.2307/1936778}{{A quantitative
+genetic theory of life history evolution}}. Ecology 63:607--615.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2016}{}}%
-Lehtonen, Jussi. 2016. {``{Multilevel selection in kin selection
-language}.''} \emph{Trends in Ecology and Evolution} xx: 1--11.
-\url{https://doi.org/10.1016/j.tree.2016.07.006}.
+Lehtonen, J. 2016.
+\href{https://doi.org/10.1016/j.tree.2016.07.006}{{Multilevel selection
+in kin selection language}}. Trends in Ecology and Evolution xx:1--11.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2018}{}}%
----------. 2018. {``{The Price equation, gradient dynamics, and
-continuous trait game theory}.''} \emph{American Naturalist} 191 (1):
-146--53. \url{https://doi.org/10.1086/694891}.
+---------. 2018. \href{https://doi.org/10.1086/694891}{{The Price
+equation, gradient dynamics, and continuous trait game theory}}.
+American Naturalist 191:146--153.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2020}{}}%
-Lehtonen, Jussi, Samir Okasha, and Heikki Helanterä. 2020. {``{Fifty
-years of the Price equation}.''} \emph{Philosophical Transactions of the
-Royal Society B: Biological Sciences} 375 (1797): 20190350.
-\url{https://doi.org/10.1098/rstb.2019.0350}.
+Lehtonen, J., S. Okasha, and H. Helanterä. 2020.
+\href{https://doi.org/10.1098/rstb.2019.0350}{{Fifty years of the Price
+equation}}. Philosophical Transactions of the Royal Society B:
+Biological Sciences 375:20190350.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lewens2015}{}}%
-Lewens, Tim. 2015. \emph{Cultural Evolution: Conceptual Challenges}.
-Oxford University Press.
-\url{https://doi.org/10.1093/acprof:oso/9780199674183.001.0001}.
+Lewens, T. 2015.
+\href{https://doi.org/10.1093/acprof:oso/9780199674183.001.0001}{Cultural
+evolution: Conceptual challenges}. Oxford University Press.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lion2018}{}}%
-Lion, Sébastien. 2018. {``{Theoretical approaches in evolutionary
-ecology: environmental feedback as a unifying perspective}.''}
-\emph{American Naturalist} 191 (1).
-\url{https://doi.org/10.1086/694865}.
+Lion, S. 2018. \href{https://doi.org/10.1086/694865}{{Theoretical
+approaches in evolutionary ecology: environmental feedback as a unifying
+perspective}}. American Naturalist 191.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Loreau2001}{}}%
-Loreau, Michel, and Andy Hector. 2001. {``{Partitioning selection and
-complementarity in biodiversity experiments}.''} \emph{Nature} 412:
-72--76. \url{https://doi.org/10.1038/35083573}.
+Loreau, M., and A. Hector. 2001.
+\href{https://doi.org/10.1038/35083573}{{Partitioning selection and
+complementarity in biodiversity experiments}}. Nature 412:72--76.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Luque2016}{}}%
-Luque, Victor J. 2017. {``{One equation to rule them all: a
-philosophical analysis of the Price equation}.''} \emph{Biology and
-Philosophy} 32 (1): 1--29.
-\url{https://doi.org/10.1007/s10539-016-9538-y}.
+Luque, V. J. 2017. \href{https://doi.org/10.1007/s10539-016-9538-y}{{One
+equation to rule them all: a philosophical analysis of the Price
+equation}}. Biology and Philosophy 32:1--29.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Luque2021}{}}%
-Luque, Victor J., and Lorenzo Baravalle. 2021. {``{The mirror of
-physics: on how the price equation can unify evolutionary biology}.''}
-\emph{Synthese} 199: 12439--62.
-\url{https://doi.org/10.1007/s11229-021-03339-6}.
+Luque, V. J., and L. Baravalle. 2021.
+\href{https://doi.org/10.1007/s11229-021-03339-6}{{The mirror of
+physics: on how the price equation can unify evolutionary biology}}.
+Synthese 199:12439--12462.
 
 \leavevmode\vadjust pre{\hypertarget{ref-MacCallum2012}{}}%
-MacCallum, Robert M., Matthias Mauch, Austin Burt, and Armand M. Leroi.
-2012. {``{Evolution of music by public choice}.''} \emph{Proceedings of
-the National Academy of Sciences} 109 (30): 12081--86.
-\url{https://doi.org/10.5061/dryad.h0228}.
+MacCallum, R. M., M. Mauch, A. Burt, and A. M. Leroi. 2012.
+\href{https://doi.org/10.5061/dryad.h0228}{{Evolution of music by public
+choice}}. Proceedings of the National Academy of Sciences
+109:12081--12086.
 
 \leavevmode\vadjust pre{\hypertarget{ref-MacColl2011}{}}%
-MacColl, Andrew D. C. 2011. {``{The ecological causes of evolution}.''}
-\emph{Trends in Ecology and Evolution} 26 (10): 514--22.
-\url{https://doi.org/10.1016/j.tree.2011.06.009}.
+MacColl, A. D. C. 2011.
+\href{https://doi.org/10.1016/j.tree.2011.06.009}{{The ecological causes
+of evolution}}. Trends in Ecology and Evolution 26:514--522.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Mateo-Tomas2017}{}}%
-Mateo-Tomás, Patricia, Pedro P. Olea, Marcos Moleón, Nuria Selva, and
-José A. Sánchez-Zapata. 2017. {``{Both rare and common species support
-ecosystem services in scavenger communities}.''} \emph{Global Ecology
-and Biogeography} 26 (12): 1459--70.
-\url{https://doi.org/10.1111/geb.12673}.
+Mateo-Tomás, P., P. P. Olea, M. Moleón, N. Selva, and J. A.
+Sánchez-Zapata. 2017. \href{https://doi.org/10.1111/geb.12673}{{Both
+rare and common species support ecosystem services in scavenger
+communities}}. Global Ecology and Biogeography 26:1459--1470.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Mayr1960}{}}%
-Mayr, Ernst. 1959. {``Where Are We? Genetics and Twentieth Century
-Darwinism.''} In \emph{Cold Spring Harbor Symposia on Quantitative
-Biology}, 24:1--14. \url{https://doi.org/10.1101/SQB.1959.024.01.003}.
+Mayr, E. 1959. \href{https://doi.org/10.1101/SQB.1959.024.01.003}{Where
+are we? Genetics and twentieth century darwinism}. Pages 1--14
+\emph{in}Cold spring harbor symposia on quantitative biology (Vol. 24).
 
 \leavevmode\vadjust pre{\hypertarget{ref-Morrison2000}{}}%
-Morrison, Margaret. 2000. \emph{Unifying Scientific Theories: Physical
-Concepts and Mathematical Structures}. Cambridge University Press.
+Morrison, M. 2000. Unifying scientific theories: Physical concepts and
+mathematical structures. Cambridge University Press.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Nagel1961}{}}%
-Nagel, Ernest. 1961. \emph{The Structure of Science: Problems in the
-Logic of Scientific Explanation}. New York, NY, USA: Harcourt, Brace \&
-World.
+Nagel, E. 1961. The structure of science: Problems in the logic of
+scientific explanation. Harcourt, Brace \& World, New York, NY, USA.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Odling-smee2003}{}}%
-Odling-Smee, F. John, Kevin N. Laland, and Marcus W. Feldman. 2003.
-\emph{Niche Construction: The Neglected Process in Evolution}. Princeton
-University Press.
+Odling-Smee, F. J., K. N. Laland, and M. W. Feldman. 2003. Niche
+construction: The neglected process in evolution. Princeton University
+Press.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Okasha2006}{}}%
-Okasha, Samir. 2006. \emph{Evolution and the Levels of Selection}.
-Oxford University Press.
-\url{https://doi.org/10.1093/acprof:oso/9780199267972.001.0001}.
+Okasha, S. 2006.
+\href{https://doi.org/10.1093/acprof:oso/9780199267972.001.0001}{Evolution
+and the levels of selection}. Oxford University Press.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Okasha2020}{}}%
-Okasha, Samir, and Jun Otsuka. 2020. {``{The Price equation and the
-causal analysis of evolutionary change}.''} \emph{Philosophical
-Transactions of the Royal Society B: Biological Sciences} 375 (1797):
-20190365. \url{https://doi.org/10.1098/rstb.2019.0365}.
+Okasha, S., and J. Otsuka. 2020.
+\href{https://doi.org/10.1098/rstb.2019.0365}{{The Price equation and
+the causal analysis of evolutionary change}}. Philosophical Transactions
+of the Royal Society B: Biological Sciences 375:20190365.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Page2002}{}}%
-Page, Karen M., and Martin A. Nowak. 2002. {``Unifying Evolutionary
-Dynamics.''} \emph{Journal of Theoretical Biology} 219 (1): 93--98.
-\url{https://doi.org/10.1006/jtbi.2002.3112}.
+Page, K. M., and M. A. Nowak. 2002.
+\href{https://doi.org/10.1006/jtbi.2002.3112}{Unifying evolutionary
+dynamics}. Journal of theoretical biology 219:93--98.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Patel2018}{}}%
-Patel, Swati, Michael H. Cortez, and Sebastian J. Schreiber. 2018.
-{``{Partitioning the effects of eco-evolutionary feedbacks on community
-stability}.''} \emph{American Naturalist} 191 (3): 1--29.
-\url{https://doi.org/10.1101/104505}.
+Patel, S., M. H. Cortez, and S. J. Schreiber. 2018.
+\href{https://doi.org/10.1101/104505}{{Partitioning the effects of
+eco-evolutionary feedbacks on community stability}}. American Naturalist
+191:1--29.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Pigliucci2009}{}}%
-Pigliucci, Massimo. 2009. {``{An extended synthesis for evolutionary
-biology}.''} \emph{Annals of the New York Academy of Sciences} 1168:
-218--28. \url{https://doi.org/10.1111/j.1749-6632.2009.04578.x}.
+Pigliucci, M. 2009.
+\href{https://doi.org/10.1111/j.1749-6632.2009.04578.x}{{An extended
+synthesis for evolutionary biology}}. Annals of the New York Academy of
+Sciences 1168:218--228.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Price1970}{}}%
-Price, George R. 1970. {``{Selection and covariance}.''}
-\url{https://doi.org/10.1038/227520a0}.
+Price, G. R. 1970. \href{https://doi.org/10.1038/227520a0}{{Selection
+and covariance}}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Price1972}{}}%
----------. 1972. {``{Extension of covariance selection mathematics}.''}
-\emph{Annals of Human Genetics} 35 (4): 485--90.
-\url{https://doi.org/10.1111/j.1469-1809.1957.tb01874.x}.
+---------. 1972.
+\href{https://doi.org/10.1111/j.1469-1809.1957.tb01874.x}{{Extension of
+covariance selection mathematics}}. Annals of Human Genetics
+35:485--490.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Price1995}{}}%
----------. 1995. {``{The nature of selection}.''} \emph{Journal of
-Theoretical Biology} 175 (3): 389--96.
-\url{https://doi.org/10.1006/jtbi.1995.0149}.
+---------. 1995. \href{https://doi.org/10.1006/jtbi.1995.0149}{{The
+nature of selection}}. Journal of Theoretical Biology 175:389--396.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Queller2017}{}}%
-Queller, David C. 2017. {``{Fundamental theorems of evolution}.''}
-\emph{American Naturalist} 189 (4): 000--000.
-\url{https://doi.org/10.1086/690937}.
+Queller, D. C. 2017. \href{https://doi.org/10.1086/690937}{{Fundamental
+theorems of evolution}}. American Naturalist 189:000--000.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Rice2004}{}}%
-Rice, Sean H. 2004. \emph{{Evolutionary theory: mathematical and
-conceptual foundations}}. Sinauer Associates Incorporated.
+Rice, S. H. 2004. {Evolutionary theory: mathematical and conceptual
+foundations}. Sinauer Associates Incorporated.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Rice2020}{}}%
----------. 2020. {``{Universal rules for the interaction of selection
-and transmission in evolution}.''} \emph{Philosophical Transactions of
-the Royal Society B: Biological Sciences} 375 (1797).
-\url{https://doi.org/10.1098/rstb.2019.0353}.
+---------. 2020.
+\href{https://doi.org/10.1098/rstb.2019.0353}{{Universal rules for the
+interaction of selection and transmission in evolution}}. Philosophical
+Transactions of the Royal Society B: Biological Sciences 375.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Rice2009}{}}%
-Rice, Sean H., and Anthony Papadopoulos. 2009. {``{Evolution with
-stochastic fitness and stochastic migration}.''} \emph{PLoS One} 4 (10).
-\url{https://doi.org/10.1371/journal.pone.0007130}.
+Rice, S. H., and A. Papadopoulos. 2009.
+\href{https://doi.org/10.1371/journal.pone.0007130}{{Evolution with
+stochastic fitness and stochastic migration}}. PLoS One 4.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Robertson1966}{}}%
-Robertson, Alan. 1966. {``{A mathematical model of the culling process
-in dairy cattle}.''} \emph{Animal Science} 8 (1): 95--108.
-\url{https://doi.org/10.1017/S0003356100037752}.
+Robertson, A. 1966. \href{https://doi.org/10.1017/S0003356100037752}{{A
+mathematical model of the culling process in dairy cattle}}. Animal
+Science 8:95--108.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Roff2008}{}}%
-Roff, Derek A. 2008. {``{Defining fitness in evolutionary models}.''}
-\emph{Journal of Genetics} 87 (4): 339--48.
-\url{https://doi.org/10.1007/s12041-008-0056-9}.
+Roff, D. A. 2008.
+\href{https://doi.org/10.1007/s12041-008-0056-9}{{Defining fitness in
+evolutionary models}}. Journal of Genetics 87:339--348.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Shennan2011}{}}%
-Shennan, Stephen. 2011. {``{Descent with modification and the
-archaeological record}.''} \emph{Philosophical Transactions of the Royal
-Society B: Biological Sciences} 366 (1567): 1070--79.
-\url{https://doi.org/10.1098/rstb.2010.0380}.
+Shennan, S. 2011. \href{https://doi.org/10.1098/rstb.2010.0380}{{Descent
+with modification and the archaeological record}}. Philosophical
+Transactions of the Royal Society B: Biological Sciences 366:1070--1079.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Turchin2001}{}}%
-Turchin, Peter. 2001. {``Does Population Ecology Have General Laws?''}
-\emph{Oikos} 94 (1): 17--26.
-\url{https://doi.org/10.1034/j.1600-0706.2001.11310.x}.
+Turchin, P. 2001.
+\href{https://doi.org/10.1034/j.1600-0706.2001.11310.x}{Does population
+ecology have general laws?} Oikos 94:17--26.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Ulrich2024}{}}%
-Ulrich, Werner, Nicholas J. Gotelli, Giovanni Strona, and William
-Godsoe. 2024. {``Reconsidering the Price Equation: Benchmarking the
-Analytical Power of Additive Partitioning in Ecology.''}
-\emph{Ecological Modelling} 491: 110695.
-\url{https://doi.org/10.1016/j.ecolmodel.2024.110695}.
+Ulrich, W., N. J. Gotelli, G. Strona, and W. Godsoe. 2024.
+\href{https://doi.org/10.1016/j.ecolmodel.2024.110695}{Reconsidering the
+price equation: Benchmarking the analytical power of additive
+partitioning in ecology}. Ecological Modelling 491:110695.
 
 \leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2005}{}}%
-van Veelen, Matthijs. 2005. {``{On the use of the Price equation}.''}
-\emph{Journal of Theoretical Biology} 237 (4): 412--26.
-\url{https://doi.org/10.1016/j.jtbi.2005.04.026}.
+van Veelen, M. 2005.
+\href{https://doi.org/10.1016/j.jtbi.2005.04.026}{{On the use of the
+Price equation}}. Journal of Theoretical Biology 237:412--426.
 
 \leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2020}{}}%
----------. 2020. {``{The problem with the Price equation}.''}
-\emph{Philosophical Transactions of the Royal Society B: Biological
-Sciences} 375 (1797): 20190355.
-\url{https://doi.org/10.1098/rstb.2019.0355}.
+---------. 2020. \href{https://doi.org/10.1098/rstb.2019.0355}{{The
+problem with the Price equation}}. Philosophical Transactions of the
+Royal Society B: Biological Sciences 375:20190355.
 
 \leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2012}{}}%
-van Veelen, Matthijs, Julián García, Maurice W. Sabelis, and Martijn
-Egas. 2012. {``{Group selection and inclusive fitness are not
-equivalent; the Price equation vs. models and statistics}.''}
-\emph{Journal of Theoretical Biology} 299 (April): 64--80.
-\url{https://doi.org/10.1016/j.jtbi.2011.07.025}.
+van Veelen, M., J. García, M. W. Sabelis, and M. Egas. 2012.
+\href{https://doi.org/10.1016/j.jtbi.2011.07.025}{{Group selection and
+inclusive fitness are not equivalent; the Price equation vs. models and
+statistics}}. Journal of Theoretical Biology 299:64--80.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Winfree2015}{}}%
-Winfree, Rachael, Jeremy W. Fox, Neal M. Williams, James R. Reilly, and
-Daniel P. Cariveau. 2015. {``{Abundance of common species, not species
-richness, drives delivery of a real-world ecosystem service}.''}
-\emph{Ecology Letters} 18: 626--35.
-\url{https://doi.org/10.1111/ele.12424}.
+Winfree, R., J. W. Fox, N. M. Williams, J. R. Reilly, and D. P.
+Cariveau. 2015. \href{https://doi.org/10.1111/ele.12424}{{Abundance of
+common species, not species richness, drives delivery of a real-world
+ecosystem service}}. Ecology Letters 18:626--635.
 

### 2025-05-31 response_to_reviewers.md
Small clarification for reviewer 2

@@
 Initially, we reasoned that because the Price equation is sufficiently abstract to describe trait change in any population (including groups and higher moments of traits), its recursive properties, and how useful it is for making formal links across disciplines such as physics and information theory (Frank 2015; 2016), we should be able to use it directly to recover population density change.
-This ended up being a dead end for reasons we will not dwell on (briefly, conservation of total probability -- that entity frequencies need to sum to 1 -- precludes the Price equation from being used to directly to recover absolute population change).
+This ended up being a dead end for reasons we will not dwell on (briefly, conservation of total probability -- that entity frequencies need to sum to 1 -- precludes the Price equation from being used to directly to recover absolute population change -- it's not a simple matter of, e.g., setting 'z' to population density).
 We had different issues when starting from population change and trying to recover the Price equation.

### 2025-05-31 foundations_of_ecoevo.Rmd
Fix the misspelling of Fisher's name

@@
 Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness.
-The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fischer's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018].
+The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018].
 This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023].

### 2025-05-31 foundations_of_ecoevo.Rmd
change the title
Draft a rewrite of paragraphs 1 to 3

@@
 ---
-title: "Darwin's Dream: Unifying ecological and evolutionary change"
+title: "Foundations of ecological and evolutionary change"
 author: |
@@
 output:
-  word_document: default
   pdf_document:
       keep_tex: yes
+  word_document: default
   html_document:
@@
 We thereby unify the fundamental equations of population ecology and biological evolution under a general framework. 
-Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness, shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness, respectively, and links this change to ecosystem function.
+Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness and links this change to ecosystem function. 
 Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.
@@
 Such unification has been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions.
-In theoretical physics, Newton's unification of celestial and terrestrial motion, and Maxwell's unification of electricity and magnetism, are among the most important advances in scientific history. 
-Likewise, in the life sciences, the modern synthesis of natural selection and Mendelian inheritance revolutionised how mechanisms underlying evolution and genetics are understood and applied [@Gayon2019].
-The analogy to physics is imperfect because evolution encompasses non-biological systems with non-genetic inheritance [@Price1995; @Shennan2011; @MacCallum2012; @Lewens2015], but to illustrate our point, we restrict our focus to biological evolution.
-Following the modern synthesis, patterns in evolutionary biology and genetics could be interpreted through a common framework.
+In evolutionary biology, the Price equation (Box 1) provides a unifying framework for evolutionary theory by exhaustively and exactly describing evolutionary change for any closed evolving population [@Price1970; @Luque2016; @Lehtonen2020].
+The Price equation is therefore fundamental, in the sense that it binds together all of evolutionary theory by formally defining what evolutionary change is and is not [@Price1970; @Rice2004; @Gardner2008; @Frank2015a; @Luque2016; @Luque2021].
+Using this formal definition, the scope of, and relationships across, sub-disciplines within evolutionary theory can be clarified.
+For example, fundamental equations of both population and quantitative genetics can be derived from the Price equation [@Queller2017].
+This provides conceptual clarity by demonstrating the logical consistency of different theoretical frameworks within evolutionary biology.
+Our aim here is to propose an equation that extends this conceptual clarity to include population change and thereby provide a formal and exact definition of joint ecological and evolutionary change.
 
-Selection, mutation, recombination, migration, and drift all have predictable consequences for both phenotypes and genotypes in a population.
-These mechanisms are therefore foundational to understanding evolution and genetics, and their roles in explaining evolutionary and genetic phenomena are no more conceptually pluralistic than that of gravity in explaining celestial and terrestrial motion.
-Instead, biologists understand that each of these mechanisms will jointly affect genetic composition and evolutionary change in a population.
-For example, all else being equal, drift will lead to a loss of both heterozygosity and adaptive phenotypes.
-This constrains the possible relations that can exist between genetic composition and evolutionary change and therefore clarifies the scope of hypotheses that are biologically coherent [@Caswell1988].
-It would obviously be mistaken to model a population in which drift had inconsistent effects on evolution and genetics, but such mistakes would surely be common without the foundational theory required to unify the two subjects.
-The absence of this theory is almost unthinkable in modern biology.
-Evolution and genetics are arguably as intertwined in biology as electricity and magnetism are in physics.
+In biological populations, ecological change is caused by the same processes of individual birth and death that cause changes in allele frequencies and phenotypes [@Turchin2001; @Connor2004].
+As with evolution, a fundamental equation can exhaustively and exactly define population change.
+Unlike the Price equation, this fundamental equation is perhaps self-evident; population change is simply the addition of births and subtraction of deaths from current population size ($N_{t}$) to recover the new population size ($N_{t+1}$; Box 2).
+By definition, this relationship of $N_{t+1} = N_{t} + Births - Deaths$ applies to any closed population.
+@Turchin2001 argues that general principles are needed to establish a logical foundation for population ecology and argues for this simple birth and death model and the consequences that logically follow from it (e.g., exponential population growth) as fundamental laws of population ecology.
+Any unifying definition of joint ecological and evolutionary change must be able to derive both the Price equation and this birth and death model.
 
-In contrast, theory underlying ecology and evolution has not been so tightly intertwined.
-Until recently, population ecology and evolution in particular have taken almost parallel paths for developing theory.
-Although many revelatory models exist to connect ecological and evolutionary phenomena [@Lion2018; @Patel2018; @Govaert2019], the foundational assumptions and equations of ecology and evolution in these models remain mostly separated conceptually.
+Our approach is distinct from eco-evolutionary modelling, which instead focuses on feedbacks between ecological and evolutionary dynamics.
+While the interdependence of ecological and evolutionary processes has long been recognised [e.g., @Pelletier2009; @Fisher1958; @Darwin1859], the rise of models incorporating both is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Yamamichi2023].
+Eco-evolutionary models can be very general in their mathematical structure, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Govaert2019; @Lion2023].
+Simplifying assumptions are necessary to make eco-evolutionary models tractable, but model assumptions can be developed from, or linked back to, fundamental equations such as the Price equation [e.g., @Coulson2008; @Barfield2011; @Lion2018].
+
+[Okay, now state how a couple of these papers did cool things, and end by saying something about how we're trying to go a step back and establish foundations]
+
+**Eco-evolutionary** is typically more broadly defined to mean any kind of feedback between ecology and evolution [Lion]
+
 But like the mechanisms driving evolutionary and genetic change, those that drive ecological and evolutionary change are well-established with predictable joint consequences.
-In evolution, natural selection, drift, and gene flow are all instantiated through a combination of different ecological factors [@Connor2004].
-Most notably, change in population density and phenotype  are both caused by individual birth and death [@Turchin2001].
+
 This should allow biologists to place ecological and evolutionary change on a shared theoretical foundation.
@@
 We believe that such a foundation exists for population ecology and evolutionary biology, and that it can bridge the fundamental equations of both subjects and serve as a starting point for understanding change in any population.
-@Turchin2001 argues for a simple birth and death model (Box 1) as a fundamental equation of population ecology, and for exponential growth to consequently be as fundamental to ecology as Newton's first law of motion is to physics.
-@Turchin2001 argues that general principles are needed to establish a logical foundation for population ecology. 
-In evolutionary biology, the Price equation (Box 2) is almost universally regarded as fundamental [@Price1970; @Rice2004; @Gardner2008; @Frank2015a;  @Luque2016; @Luque2021].
-It exhaustively and exactly describes evolutionary change for any closed evolving population [@Price1970;  @Luque2016; @Lehtonen2020], and it serves as the starting point for all other fundamental equations in evolution [@Queller2017].
+
 We believe that this logical foundation can be made sufficiently broad to encompass both population and evolutionary change.
@@
 
-> **Box 1:** The number of individuals in any closed population ($N$) at any given time ($t + 1$) is determined by the existing number of individuals ($N_{t}$), plus the number of births ($Births$) minus the number of deaths ($Deaths$), $$N_{t+1} = N_{t} + Births - Deaths.$$ This equation is necessarily true for any closed population.
-Despite its simplicity, it is a general equation for defining population change and a starting point for understanding population ecology.
-@Turchin2001 notes that a consequence of this fundamental equation is the tendency for populations to grow exponentially (technically geometrically in the above case where time is discrete).
-This inherent underlying tendency towards exponential growth persists even as the complexities of real populations, such as structure, stochasticity, or density-dependent effects are added to population models [@Turchin2001].
-Given the assumption that all individuals in the population are identical, a per capita rate of birth, $Births = bN_{t}$ and death, $Deaths = dN_{t}$ can be defined.
-Rearranging and defining $\lambda = 1 + b - d$ gives, $N_{t+1} = N_{t}\lambda$.
-Here $\lambda$ is the finite rate of increase [@Gotelli2001], and note that because $0 \leq d \leq 1$, $\lambda \geq 0$.
-Verbally, the change in size of any closed population equals its existing size times its finite rate of increase.
-
-*******************
-
-*******************
-
-> **Box 2:** The Price equation is an abstract formula to represent evolutionary change.
+> **Box 1:** The Price equation is an abstract formula to represent evolutionary change.
 Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004].
@@
 
+*******************
+
+> **Box 2:** The number of individuals in any closed population ($N$) at any given time ($t + 1$) is determined by the existing number of individuals ($N_{t}$), plus the number of births ($Births$) minus the number of deaths ($Deaths$), $$N_{t+1} = N_{t} + Births - Deaths.$$ This equation is necessarily true for any closed population.
+Despite its simplicity, it is a general equation for defining population change and a starting point for understanding population ecology.
+@Turchin2001 notes that a consequence of this fundamental equation is the tendency for populations to grow exponentially (technically geometrically in the above case where time is discrete).
+This inherent underlying tendency towards exponential growth persists even as the complexities of real populations, such as structure, stochasticity, or density-dependent effects are added to population models [@Turchin2001].
+Given the assumption that all individuals in the population are identical, a per capita rate of birth, $Births = bN_{t}$ and death, $Deaths = dN_{t}$ can be defined.
+Rearranging and defining $\lambda = 1 + b - d$ gives, $N_{t+1} = N_{t}\lambda$.
+Here $\lambda$ is the finite rate of increase [@Gotelli2001], and note that because $0 \leq d \leq 1$, $\lambda \geq 0$.
+Verbally, the change in size of any closed population equals its existing size times its finite rate of increase.
+
+*******************
+
 A foundation for biological evolution and population ecology
@@
 
-To fully unify biological evolution and population ecology, we must reconcile the general equation for population change (Box 1) with the Price equation (Box 2).
+To fully unify biological evolution and population ecology, we must reconcile the the Price equation (Box 1) with the general equation for population change (Box 2).
 Previous theory has produced mathematical synthesis between ecological and evolutionary equations [@Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
@@
 
-From eqn 1, we can derive the most fundamental equations of population ecology (Box 1) and evolutionary biology (Box 2) through an appropriate interpretation of $z_{i}$.
+From eqn 1, we can derive the most fundamental equations of population ecology (Box 2) and evolutionary biology (Box 1) through an appropriate interpretation of $z_{i}$.
 
@@
 
-To recover the general equation for population ecology (Box 1), we define $z_{i}$ as the identity of $i$ belonging to the population.
+To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
 In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
@@
 
-If we define $\lambda = 1 + \beta - \delta$ (Box 1), then we can rewrite,
+If we define $\lambda = 1 + \beta - \delta$ (Box 2), then we can rewrite,
 
@@
 
-From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 1) and the fundamental equation of evolution (eqn 10; Box 2).
+From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).
 We might also consider $\Omega$ to be a quantity of ecosystem function, in that it is the summed effect of the traits ($z_{i}$) of individual entities ($i$) in the population.
@@
 We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
-From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 1) and principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
-By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 2).
+From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 2) and principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
+By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 1).
 We therefore deliver a general framework to unify fundamental equations of ecological and evolutionary change.
@@
 
-Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ (Box 1) [@Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 2).
+Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ (Box 2) [@Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 1).
 The population growth equation $N_{t+1} = N_{t}\lambda$ could therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
@@
 We have focused on the dynamics of a closed population, and in doing so leave ecological and evolutionary change attributable to migration for future work.
-In population ecology, immigration and emigration can be incorporated by adding a term for each to the right-hand side of the equation in Box 1 [@Gotelli2001].
+In population ecology, immigration and emigration can be incorporated by adding a term for each to the right-hand side of the equation in Box 2 [@Gotelli2001].
 In evolution, because the Price equation relies on mapping ancestor-descendant relationships, accounting for migration is more challenging.
@@
 Our fundamental equation is complete and exact.
-It therefore implicitly includes any effects of density dependence on population growth (see Box 1), or any social effects on evolutionary change (see Box 2).
+It therefore implicitly includes any effects of density dependence on population growth (see Box 2), or any social effects on evolutionary change (see Box 1).
 We can make both of these effects explicit with the same partition in the summation on the right-hand side of eqn 1.

### 2025-05-31 foundations_of_ecoevo.tex
Draft a rewrite of paragraphs 1 to 3

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{setspace}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
-  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
+  pdftitle={Foundations of ecological and evolutionary change},
   colorlinks=true,
@@
 
-\title{Darwin's Dream: Unifying ecological and evolutionary change}
+\title{Foundations of ecological and evolutionary change}
 \author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
@@
 \setstretch{2}
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
+\section{Abstract}\label{abstract}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 advancement. Such unification has been a major goal in scientific
-research throughout history (\protect\hyperlink{ref-Kitcher1993}{Kitcher
-1993}), and its value is perhaps most evident in reconciling unconnected
-models and revealing new and unexpected empirical predictions. In
-theoretical physics, Newton's unification of celestial and terrestrial
-motion, and Maxwell's unification of electricity and magnetism, are
-among the most important advances in scientific history. Likewise, in
-the life sciences, the modern synthesis of natural selection and
-Mendelian inheritance revolutionised how mechanisms underlying evolution
-and genetics are understood and applied
-(\protect\hyperlink{ref-Gayon2019}{Gayon and Huneman 2019}). The analogy
-to physics is imperfect because evolution encompasses non-biological
-systems with non-genetic inheritance
-(\protect\hyperlink{ref-Price1995}{Price 1995};
-\protect\hyperlink{ref-Shennan2011}{Shennan 2011};
-\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
-\protect\hyperlink{ref-Lewens2015}{Lewens 2015}), but to illustrate our
-point, we restrict our focus to biological evolution. Following the
-modern synthesis, patterns in evolutionary biology and genetics could be
+research throughout history (\citeproc{ref-Kitcher1993}{Kitcher 1993}),
+and its value is perhaps most evident in reconciling unconnected models
+and revealing new and unexpected empirical predictions. In theoretical
+physics, Newton's unification of celestial and terrestrial motion, and
+Maxwell's unification of electricity and magnetism, are among the most
+important advances in scientific history. Likewise, in the life
+sciences, the modern synthesis of natural selection and Mendelian
+inheritance revolutionised how mechanisms underlying evolution and
+genetics are understood and applied (\citeproc{ref-Gayon2019}{Gayon and
+Huneman 2019}). The analogy to physics is imperfect because evolution
+encompasses non-biological systems with non-genetic inheritance
+(\citeproc{ref-Price1995}{Price 1995};
+\citeproc{ref-Shennan2011}{Shennan 2011};
+\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
+\citeproc{ref-Lewens2015}{Lewens 2015}), but to illustrate our point, we
+restrict our focus to biological evolution. Following the modern
+synthesis, patterns in evolutionary biology and genetics could be
 interpreted through a common framework.
@@
 change and therefore clarifies the scope of hypotheses that are
-biologically coherent (\protect\hyperlink{ref-Caswell1988}{Caswell
-1988}). It would obviously be mistaken to model a population in which
-drift had inconsistent effects on evolution and genetics, but such
-mistakes would surely be common without the foundational theory required
-to unify the two subjects. The absence of this theory is almost
-unthinkable in modern biology. Evolution and genetics are arguably as
-intertwined in biology as electricity and magnetism are in physics.
+biologically coherent (\citeproc{ref-Caswell1988}{Caswell 1988}). It
+would obviously be mistaken to model a population in which drift had
+inconsistent effects on evolution and genetics, but such mistakes would
+surely be common without the foundational theory required to unify the
+two subjects. The absence of this theory is almost unthinkable in modern
+biology. Evolution and genetics are arguably as intertwined in biology
+as electricity and magnetism are in physics.
 
@@
 Although many revelatory models exist to connect ecological and
-evolutionary phenomena (\protect\hyperlink{ref-Lion2018}{Lion 2018};
-\protect\hyperlink{ref-Patel2018}{Patel et al. 2018};
-\protect\hyperlink{ref-Govaert2019}{Govaert et al. 2019}), the
-foundational assumptions and equations of ecology and evolution in these
-models remain mostly separated conceptually. But like the mechanisms
-driving evolutionary and genetic change, those that drive ecological and
+evolutionary phenomena (\citeproc{ref-Lion2018}{Lion 2018};
+\citeproc{ref-Patel2018}{Patel et al. 2018};
+\citeproc{ref-Govaert2019}{Govaert et al. 2019}), the foundational
+assumptions and equations of ecology and evolution in these models
+remain mostly separated conceptually. But like the mechanisms driving
+evolutionary and genetic change, those that drive ecological and
 evolutionary change are well-established with predictable joint
@@
 all instantiated through a combination of different ecological factors
-(\protect\hyperlink{ref-Connor2004}{Connor and Hartl 2004}). Most
-notably, change in population density and phenotype are both caused by
-individual birth and death (\protect\hyperlink{ref-Turchin2001}{Turchin
-2001}). This should allow biologists to place ecological and
-evolutionary change on a shared theoretical foundation.
+(\citeproc{ref-Connor2004}{Connor and Hartl 2004}). Most notably, change
+in population density and phenotype are both caused by individual birth
+and death (\citeproc{ref-Turchin2001}{Turchin 2001}). This should allow
+biologists to place ecological and evolutionary change on a shared
+theoretical foundation.
 
@@
 of both subjects and serve as a starting point for understanding change
-in any population. Turchin (\protect\hyperlink{ref-Turchin2001}{2001})
-argues for a simple birth and death model (Box 1) as a fundamental
-equation of population ecology, and for exponential growth to
-consequently be as fundamental to ecology as Newton's first law of
-motion is to physics. Turchin
-(\protect\hyperlink{ref-Turchin2001}{2001}) argues that general
+in any population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues for
+a simple birth and death model (Box 1) as a fundamental equation of
+population ecology, and for exponential growth to consequently be as
+fundamental to ecology as Newton's first law of motion is to physics.
+Turchin (\citeproc{ref-Turchin2001}{2001}) argues that general
 principles are needed to establish a logical foundation for population
 ecology. In evolutionary biology, the Price equation (Box 2) is almost
-universally regarded as fundamental
-(\protect\hyperlink{ref-Price1970}{Price 1970};
-\protect\hyperlink{ref-Rice2004}{Rice 2004};
-\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
-\protect\hyperlink{ref-Frank2015a}{Frank 2017};
-\protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021}). It
+universally regarded as fundamental (\citeproc{ref-Price1970}{Price
+1970}; \citeproc{ref-Rice2004}{Rice 2004};
+\citeproc{ref-Gardner2008}{Gardner 2008};
+\citeproc{ref-Frank2015a}{Frank 2017}; \citeproc{ref-Luque2016}{Luque
+2017}; \citeproc{ref-Luque2021}{Luque and Baravalle 2021}). It
 exhaustively and exactly describes evolutionary change for any closed
-evolving population (\protect\hyperlink{ref-Price1970}{Price 1970};
-\protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it
-serves as the starting point for all other fundamental equations in
-evolution (\protect\hyperlink{ref-Queller2017}{Queller 2017}). We
-believe that this logical foundation can be made sufficiently broad to
-encompass both population and evolutionary change.
+evolving population (\citeproc{ref-Price1970}{Price 1970};
+\citeproc{ref-Luque2016}{Luque 2017};
+\citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it serves as the
+starting point for all other fundamental equations in evolution
+(\citeproc{ref-Queller2017}{Queller 2017}). We believe that this logical
+foundation can be made sufficiently broad to encompass both population
+and evolutionary change.
 
@@
 excess) or quantitative genetics (e.g., Breeder's equation) can be
-derived from the Price equation
-(\protect\hyperlink{ref-Queller2017}{Queller 2017}). We hope that our
-equation can provide a starting point for building a shared logical
-framework between ecology and evolution.
+derived from the Price equation (\citeproc{ref-Queller2017}{Queller
+2017}). We hope that our equation can provide a starting point for
+building a shared logical framework between ecology and evolution.
 
@@
 understanding population ecology. Turchin
-(\protect\hyperlink{ref-Turchin2001}{2001}) notes that a consequence of
-this fundamental equation is the tendency for populations to grow
+(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
+fundamental equation is the tendency for populations to grow
 exponentially (technically geometrically in the above case where time is
@@
 structure, stochasticity, or density-dependent effects are added to
-population models (\protect\hyperlink{ref-Turchin2001}{Turchin 2001}).
-Given the assumption that all individuals in the population are
-identical, a per capita rate of birth, \(Births = bN_{t}\) and death,
-\(Deaths = dN_{t}\) can be defined. Rearranging and defining
-\(\lambda = 1 + b - d\) gives, \(N_{t+1} = N_{t}\lambda\). Here
-\(\lambda\) is the finite rate of increase
-(\protect\hyperlink{ref-Gotelli2001}{Gotelli 2001}), and note that
+population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
+assumption that all individuals in the population are identical, a per
+capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
+can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
+\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
+increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
 because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
@@
 evolutionary change. Formulated originally in the early 1970s by George
-Price (\protect\hyperlink{ref-Price1970}{Price 1970},
-\protect\hyperlink{ref-Price1972}{1972}), it postulates some basic
-properties that all evolutionary systems must satisfy: change over time,
-ancestor and descendant relations, and a character or phenotype
-(\protect\hyperlink{ref-Rice2004}{Rice 2004}). Using simple algebraic
-language, the Price equation represents evolutionary change with the
-predominant notation,
+Price (\citeproc{ref-Price1970}{Price 1970},
+\citeproc{ref-Price1972}{1972}), it postulates some basic properties
+that all evolutionary systems must satisfy: change over time, ancestor
+and descendant relations, and a character or phenotype
+(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
+the Price equation represents evolutionary change with the predominant
+notation,
 \[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
@@
 which reflects the extent to which offspring deviate from parents in
-\(z\) (\protect\hyperlink{ref-Rice2004}{Rice 2004};
-\protect\hyperlink{ref-Okasha2006}{Okasha 2006};
-\protect\hyperlink{ref-Frank2012}{Frank 2012}). A more specific version
-of the covariance term was already known within the quantitative and
-population genetics tradition
-(\protect\hyperlink{ref-Robertson1966}{Robertson 1966}), usually
-representing the action of natural selection. The Price equation adds an
-expectation term and abstracts away from any specific mechanisms of
-replication or reproduction, or mechanisms of inheritance. Its
-definitional nature and lack of substantive biological assumptions has
-been portrayed both as a strength (\protect\hyperlink{ref-Rice2004}{Rice
-2004}; \protect\hyperlink{ref-Frank2012}{Frank 2012};
-\protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Baravalle2022}{Baravalle and Luque 2022}), and
-its greatest weakness. The abstract nature of the Price equation places
-it at the top of the hierarchy of fundamental theorems of evolution from
-which the rest (Robertson's theorem, Fischer's fundamental theorem,
+\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
+\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
+2012}). A more specific version of the covariance term was already known
+within the quantitative and population genetics tradition
+(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
+action of natural selection. The Price equation adds an expectation term
+and abstracts away from any specific mechanisms of replication or
+reproduction, or mechanisms of inheritance. Its definitional nature and
+lack of substantive biological assumptions has been portrayed both as a
+strength (\citeproc{ref-Rice2004}{Rice 2004};
+\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
+2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
+greatest weakness. The abstract nature of the Price equation places it
+at the top of the hierarchy of fundamental theorems of evolution from
+which the rest (Robertson's theorem, Fisher's fundamental theorem,
 breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
-easily derived (\protect\hyperlink{ref-Lehtonen2016}{Lehtonen 2016},
-\protect\hyperlink{ref-Lehtonen2018}{2018};
-\protect\hyperlink{ref-Queller2017}{Queller 2017}). This abstractness is
-also key to developing a more general view of evolution
-(\protect\hyperlink{ref-Rice2020}{Rice 2020};
-\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021};
-\protect\hyperlink{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast,
-some researchers consider the Price equation just a triviality (even
+easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
+\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
+2017}). This abstractness is also key to developing a more general view
+of evolution (\citeproc{ref-Rice2020}{Rice 2020};
+\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
+\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
+researchers consider the Price equation just a triviality (even
 tautological), and useless without further modelling assumptions
-(\protect\hyperlink{ref-VanVeelen2005}{van Veelen 2005};
-\protect\hyperlink{ref-VanVeelen2012}{van Veelen et al. 2012}). The
-debate remains open (\protect\hyperlink{ref-VanVeelen2020}{van Veelen
-2020}; \protect\hyperlink{ref-Baravalle2024}{Baravalle et al. 2024}).
+(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
+\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
+remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
+\citeproc{ref-Baravalle2024}{Baravalle et al. 2024}).
 \end{quote}
@@
 
-\hypertarget{a-foundation-for-biological-evolution-and-population-ecology}{%
 \section{A foundation for biological evolution and population
-ecology}\label{a-foundation-for-biological-evolution-and-population-ecology}}
+ecology}\label{a-foundation-for-biological-evolution-and-population-ecology}
 
@@
 synthesis between ecological and evolutionary equations
-(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002};
-\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but not
+(\citeproc{ref-Page2002}{Page and Nowak 2002};
+\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but not
 established a unified conceptual framework for joint ecological and
 evolutionary change. For example, Page and Nowak
-(\protect\hyperlink{ref-Page2002}{2002}) demonstrate a mathematical
-equivalence between the Price equation and a general equation for
-community dynamics. In both equations, the relative frequencies of
-different ecological (species) or evolutionary (phenotype) variants map
-to their respective ecological (population growth) or evolutionary
-(fitness) change in a way that is mathematically equivalent. Collins and
-Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use the Price
-equation to build a general quantitative framework that partitions
-phenotypic change into ecological, evolutionary, and physiological
-components. The average change in a community-wide trait (e.g., carbon
-uptake) is exactly and generally partitioned into change attributable to
-changes in species composition, expected evolution of lineages within
-species, and expected change in lineage physiology across all species.
-Both Page and Nowak (\protect\hyperlink{ref-Page2002}{2002}) and Collins
-and Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use relative
-species abundances, meaning that species frequencies must sum to one.
-This is useful for demonstrating mathematical equivalence
-(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002}) or expected
-change in a community-wide trait
-(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but a
-different approach is needed to recover absolute change in population
-size and thereby recover a joint foundation for ecology and evolution.
+(\citeproc{ref-Page2002}{2002}) demonstrate a mathematical equivalence
+between the Price equation and a general equation for community
+dynamics. In both equations, the relative frequencies of different
+ecological (species) or evolutionary (phenotype) variants map to their
+respective ecological (population growth) or evolutionary (fitness)
+change in a way that is mathematically equivalent. Collins and Gardner
+(\citeproc{ref-Collins2009}{2009}) use the Price equation to build a
+general quantitative framework that partitions phenotypic change into
+ecological, evolutionary, and physiological components. The average
+change in a community-wide trait (e.g., carbon uptake) is exactly and
+generally partitioned into change attributable to changes in species
+composition, expected evolution of lineages within species, and expected
+change in lineage physiology across all species. Both Page and Nowak
+(\citeproc{ref-Page2002}{2002}) and Collins and Gardner
+(\citeproc{ref-Collins2009}{2009}) use relative species abundances,
+meaning that species frequencies must sum to one. This is useful for
+demonstrating mathematical equivalence (\citeproc{ref-Page2002}{Page and
+Nowak 2002}) or expected change in a community-wide trait
+(\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but a different
+approach is needed to recover absolute change in population size and
+thereby recover a joint foundation for ecology and evolution.
 
 The Price equation is critical for partitioning different components of
-biological change (\protect\hyperlink{ref-Price1970}{Price 1970};
-\protect\hyperlink{ref-Frank1997}{Frank 1997};
-\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
-\protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Queller2017}{Queller 2017};
-\protect\hyperlink{ref-Lehtonen2018}{Lehtonen 2018}). It has also been
-highly useful for integrating evolutionary theory across disciplines
-(\protect\hyperlink{ref-Fox2006}{Fox 2006};
-\protect\hyperlink{ref-Brantingham2007}{Brantingham 2007};
-\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
-\protect\hyperlink{ref-Frank2015}{Frank 2015};
-\protect\hyperlink{ref-Godsoe2021}{Godsoe et al. 2021};
-\protect\hyperlink{ref-Ulrich2024}{Ulrich et al. 2024}). These
+biological change (\citeproc{ref-Price1970}{Price 1970};
+\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Gardner2008}{Gardner
+2008}; \citeproc{ref-Luque2016}{Luque 2017};
+\citeproc{ref-Queller2017}{Queller 2017};
+\citeproc{ref-Lehtonen2018}{Lehtonen 2018}). It has also been highly
+useful for integrating evolutionary theory across disciplines
+(\citeproc{ref-Fox2006}{Fox 2006};
+\citeproc{ref-Brantingham2007}{Brantingham 2007};
+\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
+\citeproc{ref-Frank2015}{Frank 2015}; \citeproc{ref-Godsoe2021}{Godsoe
+et al. 2021}; \citeproc{ref-Ulrich2024}{Ulrich et al. 2024}). These
 properties would seem to make it an intuitive starting point for a
 logical foundation of ecology and evolution, perhaps through some kind
-of mathematical equivalence (\protect\hyperlink{ref-Page2002}{Page and
-Nowak 2002}) or addition of terms
-(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), or
-through the use of its recursive structure
-(\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
-\protect\hyperlink{ref-Frank2012}{Frank 2012}). But despite its
-flexibility, the Price equation still relies on relative frequencies,
-which must by definition sum to one
-(\protect\hyperlink{ref-Frank2015}{Frank 2015}). This is because the
-Price equation describes the average change in a population; the
-frequency of entities is scaled thereby conserving total probability
-(\protect\hyperlink{ref-Frank2015}{Frank 2015},
-\protect\hyperlink{ref-Frank2016}{2016}). But to recover the fundamental
-principle of exponential population growth
-(\protect\hyperlink{ref-Turchin2001}{Turchin 2001}), this scaling must
-be avoided in a fundamental equation of ecology and evolution.
+of mathematical equivalence (\citeproc{ref-Page2002}{Page and Nowak
+2002}) or addition of terms (\citeproc{ref-Collins2009}{Collins and
+Gardner 2009}), or through the use of its recursive structure
+(\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
+\citeproc{ref-Frank2012}{Frank 2012}). But despite its flexibility, the
+Price equation still relies on relative frequencies, which must by
+definition sum to one (\citeproc{ref-Frank2015}{Frank 2015}). This is
+because the Price equation describes the average change in a population;
+the frequency of entities is scaled thereby conserving total probability
+(\citeproc{ref-Frank2015}{Frank 2015}, \citeproc{ref-Frank2016}{2016}).
+But to recover the fundamental principle of exponential population
+growth (\citeproc{ref-Turchin2001}{Turchin 2001}), this scaling must be
+avoided in a fundamental equation of ecology and evolution.
 
 We therefore begin with the most fundamental axioms underlying the
-ecology and evolution of living systems
-(\protect\hyperlink{ref-Rice2004}{Rice 2004};
-\protect\hyperlink{ref-Rice2009}{Rice and Papadopoulos 2009}). In such
+ecology and evolution of living systems (\citeproc{ref-Rice2004}{Rice
+2004}; \citeproc{ref-Rice2009}{Rice and Papadopoulos 2009}). In such
 systems, diversity is discontinuous and can be defined in terms of
-discrete entities (\protect\hyperlink{ref-Dobzhansky1970}{Dobzhansky
-1970}). Our framework is general enough that entities can be anything
-discrete, but we will focus on each entity \(i\) as an individual
-organism. Individuals give rise to new individuals through birth such
-that \(\beta_{i}\) is the number of births attributable to \(i\).
-Individuals are removed from the population through death such that
-\(\delta_{i}\) is an indicator variable that takes a value of 1 (death
-of \(i\)) or 0 (persistence of \(i\)). All individuals are defined by
-some characteristic \(z_{i}\), and \(\Delta z_{i}\) defines any change
-in \(z_{i}\) from one time step \(t\) to the next \(t + 1\). The total
+discrete entities (\citeproc{ref-Dobzhansky1970}{Dobzhansky 1970}). Our
+framework is general enough that entities can be anything discrete, but
+we will focus on each entity \(i\) as an individual organism.
+Individuals give rise to new individuals through birth such that
+\(\beta_{i}\) is the number of births attributable to \(i\). Individuals
+are removed from the population through death such that \(\delta_{i}\)
+is an indicator variable that takes a value of 1 (death of \(i\)) or 0
+(persistence of \(i\)). All individuals are defined by some
+characteristic \(z_{i}\), and \(\Delta z_{i}\) defines any change in
+\(z_{i}\) from one time step \(t\) to the next \(t + 1\). The total
 number of individuals in the population is \(N\). From this foundation,
@@
 
-\hypertarget{population-ecology}{%
-\section{Population ecology}\label{population-ecology}}
+\section{Population ecology}\label{population-ecology}
 
@@
 and the fundamental property of exponential growth in populations
-(\protect\hyperlink{ref-Turchin2001}{Turchin 2001}) (eqn 3).
-
-\hypertarget{evolutionary-biology}{%
-\section{Evolutionary biology}\label{evolutionary-biology}}
+(\citeproc{ref-Turchin2001}{Turchin 2001}) (eqn 3).
+
+\section{Evolutionary biology}\label{evolutionary-biology}
 
@@
 Now we must consider the total conservation of probability
-(\protect\hyperlink{ref-Frank2015}{Frank 2015},
-\protect\hyperlink{ref-Frank2016}{2016}). In eqn 7, \(\Omega\) is the
-total sum trait values (\(z_{i}\)) across the entire population at
-\(t + 1\) divided by the number of individuals (\(N\)) in the population
-at \(t\). But the size of the population can change from \(t\) to
-\(t + 1\). To recover mean trait change for the Price equation (and
-therefore conserve total probability), we need to account for this
-change in population size. The left-hand side of eqn 7 describes
-contributions to the sum trait value from the new population at
-\(t + 1\). But we cannot treat \(\Omega/N\) as the mean of \(z\) at
+(\citeproc{ref-Frank2015}{Frank 2015}, \citeproc{ref-Frank2016}{2016}).
+In eqn 7, \(\Omega\) is the total sum trait values (\(z_{i}\)) across
+the entire population at \(t + 1\) divided by the number of individuals
+(\(N\)) in the population at \(t\). But the size of the population can
+change from \(t\) to \(t + 1\). To recover mean trait change for the
+Price equation (and therefore conserve total probability), we need to
+account for this change in population size. The left-hand side of eqn 7
+describes contributions to the sum trait value from the new population
+at \(t + 1\). But we cannot treat \(\Omega/N\) as the mean of \(z\) at
 \(t+1\) (\(\bar{z}'\)) because we need to weigh \(N\) by the mean
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}
 
@@
 models are logical (mathematical) consequences of more fundamental ones
-(\protect\hyperlink{ref-Nagel1961}{Nagel 1961}). Conceptual unification
-therefore has a critical role in advancing scientific theory
-(\protect\hyperlink{ref-Morrison2000}{Morrison 2000}). Showing how
-apparently disparate phenomena follow from the same shared principles
-reveals what is most fundamental about the natural world and provides a
-foundation for further scientific investigation and the construction of
-more coherent and predictive models. Our eqn 1 provides a foundation for
-the conceptual unification of the fundamental equations of population
+(\citeproc{ref-Nagel1961}{Nagel 1961}). Conceptual unification therefore
+has a critical role in advancing scientific theory
+(\citeproc{ref-Morrison2000}{Morrison 2000}). Showing how apparently
+disparate phenomena follow from the same shared principles reveals what
+is most fundamental about the natural world and provides a foundation
+for further scientific investigation and the construction of more
+coherent and predictive models. Our eqn 1 provides a foundation for the
+conceptual unification of the fundamental equations of population
 ecology and evolution.
@@
 and applying the total conservation of probability to individual
-frequencies (\protect\hyperlink{ref-Frank2015}{Frank 2015},
-\protect\hyperlink{ref-Frank2016}{2016}), we recover the most
-fundamental equation of evolution (Box 2). We therefore deliver a
-general framework to unify fundamental equations of ecological and
-evolutionary change.
+frequencies (\citeproc{ref-Frank2015}{Frank 2015},
+\citeproc{ref-Frank2016}{2016}), we recover the most fundamental
+equation of evolution (Box 2). We therefore deliver a general framework
+to unify fundamental equations of ecological and evolutionary change.
 
 The Price equation provides a complete and exact description of
-evolution in any closed evolving system
-(\protect\hyperlink{ref-Price1970}{Price 1970};
-\protect\hyperlink{ref-Frank2012}{Frank 2012}). It is derived by
+evolution in any closed evolving system (\citeproc{ref-Price1970}{Price
+1970}; \citeproc{ref-Frank2012}{Frank 2012}). It is derived by
 rearranging the mathematical notation defining changes in the
 frequencies and characteristics of any type of entity
-(\protect\hyperlink{ref-Price1970}{Price 1970};
-\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
-\protect\hyperlink{ref-Luque2016}{Luque 2017}) (e.g., individuals,
-alleles). This derivation partitions total characteristic change into
-different components, making it possible to isolate evolutionary
-mechanisms (e.g., selection) and levels of biological organisation
-(e.g., group, individual) (\protect\hyperlink{ref-Frank1995}{Frank
-1995}, \protect\hyperlink{ref-Frank2012}{2012};
-\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
-\protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Okasha2020}{Okasha and Otsuka 2020}). Because of
-its abstract nature and lack of any system-specific assumptions, the
-Price equation makes no predictions about what will happen in any
-particular system (\protect\hyperlink{ref-Gardner2020}{Gardner 2020}).
-Its role is not to predict, but to formally and completely define and
-separate components of evolutionary change. The same is true of the
-general equation for population change (eqns 2 and 3), at least as we
-have used it here where it serves to define what population growth means
-in ecology. This equation formally and completely describes population
-change in terms of births and deaths. In eqn 1, we therefore have a
-fundamental equation from which we can derive complete ecological and
-evolutionary change in any closed biological population. We anticipate
-that this will be useful for eco-evolutionary theory in the same way
-that the Price equation is useful for evolutionary theory: facilitating
-specific model development and identifying new conceptual insights,
-unresolved errors, and sources of model disagreements.
+(\citeproc{ref-Price1970}{Price 1970};
+\citeproc{ref-Gardner2008}{Gardner 2008}; \citeproc{ref-Luque2016}{Luque
+2017}) (e.g., individuals, alleles). This derivation partitions total
+characteristic change into different components, making it possible to
+isolate evolutionary mechanisms (e.g., selection) and levels of
+biological organisation (e.g., group, individual)
+(\citeproc{ref-Frank1995}{Frank 1995}, \citeproc{ref-Frank2012}{2012};
+\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
+\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Okasha2020}{Okasha
+and Otsuka 2020}). Because of its abstract nature and lack of any
+system-specific assumptions, the Price equation makes no predictions
+about what will happen in any particular system
+(\citeproc{ref-Gardner2020}{Gardner 2020}). Its role is not to predict,
+but to formally and completely define and separate components of
+evolutionary change. The same is true of the general equation for
+population change (eqns 2 and 3), at least as we have used it here where
+it serves to define what population growth means in ecology. This
+equation formally and completely describes population change in terms of
+births and deaths. In eqn 1, we therefore have a fundamental equation
+from which we can derive complete ecological and evolutionary change in
+any closed biological population. We anticipate that this will be useful
+for eco-evolutionary theory in the same way that the Price equation is
+useful for evolutionary theory: facilitating specific model development
+and identifying new conceptual insights, unresolved errors, and sources
+of model disagreements.
 
 Our unification demonstrates the equivalence between the finite rate of
-increase \(\lambda\) (Box 1)
-(\protect\hyperlink{ref-Gotelli2001}{Gotelli 2001}) and population mean
-evolutionary fitness \(\bar{w}\) (Box 2). The population growth equation
-\(N_{t+1} = N_{t}\lambda\) could therefore be rewritten as
-\(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first to notice
-this equivalence between population growth and mean population fitness.
-Indeed, the relationship between population growth rate and evolutionary
-fitness has been proposed and applied many times before
-(\protect\hyperlink{ref-Fisher1930}{Fisher 1930};
-\protect\hyperlink{ref-Charlesworth1980}{Charlesworth 1980};
-\protect\hyperlink{ref-Lande1982}{Lande 1982};
-\protect\hyperlink{ref-Roff2008}{Roff 2008};
-\protect\hyperlink{ref-Lion2018}{Lion 2018}). For example, Lande
-(\protect\hyperlink{ref-Lande1982}{1982}) explicitly concludes mean
-absolute fitness per unit time is \(\bar{w} = e^{r}\), where
-\(r = ln(\lambda)\). We show this from first principles and clarify the
-relationship between fitness and population growth. Over an arbitrary
-length of time, fitness is properly defined as
-\(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an individual's lifetime
-(which, by definition, includes death), fitness is therefore
-\(\beta_{i}\). Interestingly, the rate of change in ecology and
-evolution are reflected in the first and second statistical moments of
-fitness, respectively. Population growth rate reflects mean fitness
+increase \(\lambda\) (Box 1) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
+and population mean evolutionary fitness \(\bar{w}\) (Box 2). The
+population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
+rewritten as \(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first
+to notice this equivalence between population growth and mean population
+fitness. Indeed, the relationship between population growth rate and
+evolutionary fitness has been proposed and applied many times before
+(\citeproc{ref-Fisher1930}{Fisher 1930};
+\citeproc{ref-Charlesworth1980}{Charlesworth 1980};
+\citeproc{ref-Lande1982}{Lande 1982}; \citeproc{ref-Roff2008}{Roff
+2008}; \citeproc{ref-Lion2018}{Lion 2018}). For example, Lande
+(\citeproc{ref-Lande1982}{1982}) explicitly concludes mean absolute
+fitness per unit time is \(\bar{w} = e^{r}\), where \(r = ln(\lambda)\).
+We show this from first principles and clarify the relationship between
+fitness and population growth. Over an arbitrary length of time, fitness
+is properly defined as \(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an
+individual's lifetime (which, by definition, includes death), fitness is
+therefore \(\beta_{i}\). Interestingly, the rate of change in ecology
+and evolution are reflected in the first and second statistical moments
+of fitness, respectively. Population growth rate reflects mean fitness
 \(\bar{w}\), while the rate of evolutionary change reflects the variance
 in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem)
-(\protect\hyperlink{ref-Frank1997}{Frank 1997};
-\protect\hyperlink{ref-Rice2004}{Rice 2004};
-\protect\hyperlink{ref-Queller2017}{Queller 2017}).
+(\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
+2004}; \citeproc{ref-Queller2017}{Queller 2017}).
 
@@
 environments (the ecological interactions)
-(\protect\hyperlink{ref-MacColl2011}{MacColl 2011}). This has been a
-line of argumentation by some defenders of the so-called Extended
-Synthesis (\protect\hyperlink{ref-Pigliucci2009}{Pigliucci 2009}),
-especially in relation to niche construction
-(\protect\hyperlink{ref-Odling-smee2003}{Odling-Smee et al. 2003}).
-Famously, Mayr (\protect\hyperlink{ref-Mayr1960}{1959}) characterized
+(\citeproc{ref-MacColl2011}{MacColl 2011}). This has been a line of
+argumentation by some defenders of the so-called Extended Synthesis
+(\citeproc{ref-Pigliucci2009}{Pigliucci 2009}), especially in relation
+to niche construction (\citeproc{ref-Odling-smee2003}{Odling-Smee et al.
+2003}). Famously, Mayr (\citeproc{ref-Mayr1960}{1959}) characterized
 population genetics as a simple input and output of genes, analogous to
@@
 classical population and quantitative genetic models
-(\protect\hyperlink{ref-Queller2017}{Queller 2017}) and develop new ones
-(\protect\hyperlink{ref-Rice2004}{Rice 2004},
-\protect\hyperlink{ref-Rice2020}{2020};
-\protect\hyperlink{ref-Luque2016}{Luque 2017};
-\protect\hyperlink{ref-Lion2018}{Lion 2018}). Our eqn 1 contains ecology
-at its core, and we show how the Price equation logically follows from
-it after accounting for absolute population growth (eqn 8). We therefore
-conclude that population and quantitative genetic equations contain
-ecology (no matter how hidden), and the ecological nature of evolution
-is implicit in population and quantitative genetic models.
+(\citeproc{ref-Queller2017}{Queller 2017}) and develop new ones
+(\citeproc{ref-Rice2004}{Rice 2004}, \citeproc{ref-Rice2020}{2020};
+\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Lion2018}{Lion
+2018}). Our eqn 1 contains ecology at its core, and we show how the
+Price equation logically follows from it after accounting for absolute
+population growth (eqn 8). We therefore conclude that population and
+quantitative genetic equations contain ecology (no matter how hidden),
+and the ecological nature of evolution is implicit in population and
+quantitative genetic models.
 
@@
 incorporated by adding a term for each to the right-hand side of the
-equation in Box 1 (\protect\hyperlink{ref-Gotelli2001}{Gotelli 2001}).
-In evolution, because the Price equation relies on mapping
+equation in Box 1 (\citeproc{ref-Gotelli2001}{Gotelli 2001}). In
+evolution, because the Price equation relies on mapping
 ancestor-descendant relationships, accounting for migration is more
-challenging. Kerr and Godfrey-Smith
-(\protect\hyperlink{ref-Kerr2009}{2009}) demonstrate how the Price
-equation can be extended to allow for arbitrary links between ancestors
-and descendants, thereby extending the Price equation to allow for
-immigration and emigration. Frank
-(\protect\hyperlink{ref-Frank2012}{2012}) presents a simplified version
-of Kerr and Godfrey-Smith (\protect\hyperlink{ref-Kerr2009}{2009}) that
-allows some fraction of descendants to be unconnected to ancestors. In
-both ecology and evolution, accounting for migration is done through the
-use of additional terms on the right-hand side of the equations.
+challenging. Kerr and Godfrey-Smith (\citeproc{ref-Kerr2009}{2009})
+demonstrate how the Price equation can be extended to allow for
+arbitrary links between ancestors and descendants, thereby extending the
+Price equation to allow for immigration and emigration. Frank
+(\citeproc{ref-Frank2012}{2012}) presents a simplified version of Kerr
+and Godfrey-Smith (\citeproc{ref-Kerr2009}{2009}) that allows some
+fraction of descendants to be unconnected to ancestors. In both ecology
+and evolution, accounting for migration is done through the use of
+additional terms on the right-hand side of the equations.
 
@@
 been used to investigate ecosystem function
-(\protect\hyperlink{ref-Loreau2001}{Loreau and Hector 2001};
-\protect\hyperlink{ref-Fox2006}{Fox 2006}), but not with any attempt
-towards conceptual unification with evolutionary biology. For example,
-Fox (\protect\hyperlink{ref-Fox2006}{2006}) applied the abstract
-properties of the Price equation to partition total change in ecosystem
-function into separate components attributable to species richness,
-species composition, and context dependent effects. This approach
-provides a framework for comparing the effects of biodiversity on
-ecosystem function in empirical systems
-(\protect\hyperlink{ref-Fox2006}{Fox 2006};
-\protect\hyperlink{ref-Winfree2015}{Winfree et al. 2015};
-\protect\hyperlink{ref-Mateo-Tomas2017}{Mateo-Tomás et al. 2017}).
-Instead, our eqn 1 defines \(\Omega\) as total ecosystem function
-contributed by a focal population (e.g., biomass production,
-decomposition, photosynthetic rate). It is therefore possible to
-investigate ecological, evolutionary, and ecosystem function change from
-the same shared framework.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+(\citeproc{ref-Loreau2001}{Loreau and Hector 2001};
+\citeproc{ref-Fox2006}{Fox 2006}), but not with any attempt towards
+conceptual unification with evolutionary biology. For example, Fox
+(\citeproc{ref-Fox2006}{2006}) applied the abstract properties of the
+Price equation to partition total change in ecosystem function into
+separate components attributable to species richness, species
+composition, and context dependent effects. This approach provides a
+framework for comparing the effects of biodiversity on ecosystem
+function in empirical systems (\citeproc{ref-Fox2006}{Fox 2006};
+\citeproc{ref-Winfree2015}{Winfree et al. 2015};
+\citeproc{ref-Mateo-Tomas2017}{Mateo-Tomás et al. 2017}). Instead, our
+eqn 1 defines \(\Omega\) as total ecosystem function contributed by a
+focal population (e.g., biomass production, decomposition,
+photosynthetic rate). It is therefore possible to investigate
+ecological, evolutionary, and ecosystem function change from the same
+shared framework.
+
+\section{Acknowledgements}\label{acknowledgements}
 
@@
 
-\hypertarget{author-contributions}{%
-\section{Author Contributions}\label{author-contributions}}
+\section{Author Contributions}\label{author-contributions}
 
@@
 
-\hypertarget{competing-interests}{%
-\section{Competing Interests}\label{competing-interests}}
+\section{Competing Interests}\label{competing-interests}
 
@@
 
-\hypertarget{data-availability}{%
-\section{Data Availability}\label{data-availability}}
+\section{Data Availability}\label{data-availability}
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-Baravalle2022}{}}%
+\bibitem[\citeproctext]{ref-Baravalle2022}
 Baravalle, L., and V. J. Luque. 2022.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Baravalle2024}{}}%
+\bibitem[\citeproctext]{ref-Baravalle2024}
 Baravalle, L., A. Roffé, V. J. Luque, and S. Ginnobili. 2024.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Brantingham2007}{}}%
+\bibitem[\citeproctext]{ref-Brantingham2007}
 Brantingham, P. J. 2007. \href{https://doi.org/10.2307/40035853}{{A
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Caswell1988}{}}%
+\bibitem[\citeproctext]{ref-Caswell1988}
 Caswell, H. 1988.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Charlesworth1980}{}}%
+\bibitem[\citeproctext]{ref-Charlesworth1980}
 Charlesworth, B. 1980. Evolution in age-structured populations.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Collins2009}{}}%
+\bibitem[\citeproctext]{ref-Collins2009}
 Collins, S., and A. Gardner. 2009.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Connor2004}{}}%
+\bibitem[\citeproctext]{ref-Connor2004}
 Connor, J., and D. L. Hartl. 2004. {A premier of ecological genetics}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Dobzhansky1970}{}}%
+\bibitem[\citeproctext]{ref-Dobzhansky1970}
 Dobzhansky, T. 1970. Genetics of the evolutionary process (Vol. 139).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Edelaar2023}{}}%
+\bibitem[\citeproctext]{ref-Edelaar2023}
 Edelaar, Pim, J. Otsuka, and V. J. Luque. 2023.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Fisher1930}{}}%
+\bibitem[\citeproctext]{ref-Fisher1930}
 Fisher, R. A. 1930. The genetical theory of natural selection. Oxford
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Fox2006}{}}%
+\bibitem[\citeproctext]{ref-Fox2006}
 Fox, J. W. 2006.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Frank1995}{}}%
+\bibitem[\citeproctext]{ref-Frank1995}
 Frank, S. A. 1995. \href{https://doi.org/10.1006/jtbi.1995.0148}{{George
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Frank1997}{}}%
+\bibitem[\citeproctext]{ref-Frank1997}
 ---------. 1997.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Frank2012}{}}%
+\bibitem[\citeproctext]{ref-Frank2012}
 ---------. 2012.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Frank2015}{}}%
+\bibitem[\citeproctext]{ref-Frank2015}
 ---------. 2015. \href{https://doi.org/10.3390/e17107087}{{D'Alembert's
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Frank2016}{}}%
+\bibitem[\citeproctext]{ref-Frank2016}
 ---------. 2016. \href{https://doi.org/10.3390/e18050192}{{Common
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Frank2015a}{}}%
+\bibitem[\citeproctext]{ref-Frank2015a}
 ---------. 2017. \href{https://doi.org/10.1002/ece3.2922}{{Universal
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Gardner2008}{}}%
+\bibitem[\citeproctext]{ref-Gardner2008}
 Gardner, A. 2008. \href{https://doi.org/10.1016/j.cub.2008.01.005}{{The
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Gardner2020}{}}%
+\bibitem[\citeproctext]{ref-Gardner2020}
 ---------. 2020. \href{https://doi.org/10.1098/rstb.2019.0361}{{Price's
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Gayon2019}{}}%
+\bibitem[\citeproctext]{ref-Gayon2019}
 Gayon, J., and P. Huneman. 2019.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Godsoe2021}{}}%
+\bibitem[\citeproctext]{ref-Godsoe2021}
 Godsoe, W., K. E. Eisen, D. Stanton, and K. M. Sirianni. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Gotelli2001}{}}%
+\bibitem[\citeproctext]{ref-Gotelli2001}
 Gotelli, N. J. 2001. A primer of ecology. Sinauer associate. Inc.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Govaert2019}{}}%
+\bibitem[\citeproctext]{ref-Govaert2019}
 Govaert, L., E. A. Fronhofer, S. Lion, C. Eizaguirre, D. Bonte, M. Egas,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Kerr2009}{}}%
+\bibitem[\citeproctext]{ref-Kerr2009}
 Kerr, B., and P. Godfrey-Smith. 2009.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Kitcher1993}{}}%
+\bibitem[\citeproctext]{ref-Kitcher1993}
 Kitcher, P. 1993. {The advancement of science}. Oxford University Press,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lande1982}{}}%
+\bibitem[\citeproctext]{ref-Lande1982}
 Lande, R. 1982. \href{https://doi.org/10.2307/1936778}{{A quantitative
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2016}{}}%
+\bibitem[\citeproctext]{ref-Lehtonen2016}
 Lehtonen, J. 2016.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2018}{}}%
+\bibitem[\citeproctext]{ref-Lehtonen2018}
 ---------. 2018. \href{https://doi.org/10.1086/694891}{{The Price
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lehtonen2020}{}}%
+\bibitem[\citeproctext]{ref-Lehtonen2020}
 Lehtonen, J., S. Okasha, and H. Helanterä. 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lewens2015}{}}%
+\bibitem[\citeproctext]{ref-Lewens2015}
 Lewens, T. 2015.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lion2018}{}}%
+\bibitem[\citeproctext]{ref-Lion2018}
 Lion, S. 2018. \href{https://doi.org/10.1086/694865}{{Theoretical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Loreau2001}{}}%
+\bibitem[\citeproctext]{ref-Loreau2001}
 Loreau, M., and A. Hector. 2001.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Luque2016}{}}%
+\bibitem[\citeproctext]{ref-Luque2016}
 Luque, V. J. 2017. \href{https://doi.org/10.1007/s10539-016-9538-y}{{One
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Luque2021}{}}%
+\bibitem[\citeproctext]{ref-Luque2021}
 Luque, V. J., and L. Baravalle. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-MacCallum2012}{}}%
+\bibitem[\citeproctext]{ref-MacCallum2012}
 MacCallum, R. M., M. Mauch, A. Burt, and A. M. Leroi. 2012.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-MacColl2011}{}}%
+\bibitem[\citeproctext]{ref-MacColl2011}
 MacColl, A. D. C. 2011.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Mateo-Tomas2017}{}}%
+\bibitem[\citeproctext]{ref-Mateo-Tomas2017}
 Mateo-Tomás, P., P. P. Olea, M. Moleón, N. Selva, and J. A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Mayr1960}{}}%
+\bibitem[\citeproctext]{ref-Mayr1960}
 Mayr, E. 1959. \href{https://doi.org/10.1101/SQB.1959.024.01.003}{Where
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Morrison2000}{}}%
+\bibitem[\citeproctext]{ref-Morrison2000}
 Morrison, M. 2000. Unifying scientific theories: Physical concepts and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Nagel1961}{}}%
+\bibitem[\citeproctext]{ref-Nagel1961}
 Nagel, E. 1961. The structure of science: Problems in the logic of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Odling-smee2003}{}}%
+\bibitem[\citeproctext]{ref-Odling-smee2003}
 Odling-Smee, F. J., K. N. Laland, and M. W. Feldman. 2003. Niche
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Okasha2006}{}}%
+\bibitem[\citeproctext]{ref-Okasha2006}
 Okasha, S. 2006.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Okasha2020}{}}%
+\bibitem[\citeproctext]{ref-Okasha2020}
 Okasha, S., and J. Otsuka. 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Page2002}{}}%
+\bibitem[\citeproctext]{ref-Page2002}
 Page, K. M., and M. A. Nowak. 2002.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Patel2018}{}}%
+\bibitem[\citeproctext]{ref-Patel2018}
 Patel, S., M. H. Cortez, and S. J. Schreiber. 2018.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Pigliucci2009}{}}%
+\bibitem[\citeproctext]{ref-Pigliucci2009}
 Pigliucci, M. 2009.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Price1970}{}}%
+\bibitem[\citeproctext]{ref-Price1970}
 Price, G. R. 1970. \href{https://doi.org/10.1038/227520a0}{{Selection
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Price1972}{}}%
+\bibitem[\citeproctext]{ref-Price1972}
 ---------. 1972.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Price1995}{}}%
+\bibitem[\citeproctext]{ref-Price1995}
 ---------. 1995. \href{https://doi.org/10.1006/jtbi.1995.0149}{{The
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Queller2017}{}}%
+\bibitem[\citeproctext]{ref-Queller2017}
 Queller, D. C. 2017. \href{https://doi.org/10.1086/690937}{{Fundamental
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Rice2004}{}}%
+\bibitem[\citeproctext]{ref-Rice2004}
 Rice, S. H. 2004. {Evolutionary theory: mathematical and conceptual
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Rice2020}{}}%
+\bibitem[\citeproctext]{ref-Rice2020}
 ---------. 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Rice2009}{}}%
+\bibitem[\citeproctext]{ref-Rice2009}
 Rice, S. H., and A. Papadopoulos. 2009.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Robertson1966}{}}%
+\bibitem[\citeproctext]{ref-Robertson1966}
 Robertson, A. 1966. \href{https://doi.org/10.1017/S0003356100037752}{{A
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Roff2008}{}}%
+\bibitem[\citeproctext]{ref-Roff2008}
 Roff, D. A. 2008.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Shennan2011}{}}%
+\bibitem[\citeproctext]{ref-Shennan2011}
 Shennan, S. 2011. \href{https://doi.org/10.1098/rstb.2010.0380}{{Descent
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Turchin2001}{}}%
+\bibitem[\citeproctext]{ref-Turchin2001}
 Turchin, P. 2001.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Ulrich2024}{}}%
+\bibitem[\citeproctext]{ref-Ulrich2024}
 Ulrich, W., N. J. Gotelli, G. Strona, and W. Godsoe. 2024.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2005}{}}%
+\bibitem[\citeproctext]{ref-VanVeelen2005}
 van Veelen, M. 2005.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2020}{}}%
+\bibitem[\citeproctext]{ref-VanVeelen2020}
 ---------. 2020. \href{https://doi.org/10.1098/rstb.2019.0355}{{The
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-VanVeelen2012}{}}%
+\bibitem[\citeproctext]{ref-VanVeelen2012}
 van Veelen, M., J. García, M. W. Sabelis, and M. Egas. 2012.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Winfree2015}{}}%
+\bibitem[\citeproctext]{ref-Winfree2015}
 Winfree, R., J. W. Fox, N. M. Williams, J. R. Reilly, and D. P.

### 2025-06-01 foundations_of_ecoevo.Rmd
Rewrite the introduction along reviewer recommendations; cut some things to the first section, and prepare for a better verbal explanation of eqn 1

@@
 The Price equation is therefore fundamental, in the sense that it binds together all of evolutionary theory by formally defining what evolutionary change is and is not [@Price1970; @Rice2004; @Gardner2008; @Frank2015a; @Luque2016; @Luque2021].
-Using this formal definition, the scope of, and relationships across, sub-disciplines within evolutionary theory can be clarified.
+Using this formal definition, the scope of, and relationships among, sub-disciplines within evolutionary theory can be clarified.
 For example, fundamental equations of both population and quantitative genetics can be derived from the Price equation [@Queller2017].
@@
 
-In biological populations, ecological change is caused by the same processes of individual birth and death that cause changes in allele frequencies and phenotypes [@Turchin2001; @Connor2004].
+In biological populations, ecological change is caused by the same processes of individual birth and death that cause changes in allele frequencies and phenotypes [@Turchin2001; @Connor2004; @Barfield2011].
 As with evolution, a fundamental equation can exhaustively and exactly define population change.
@@
 By definition, this relationship of $N_{t+1} = N_{t} + Births - Deaths$ applies to any closed population.
-@Turchin2001 argues that general principles are needed to establish a logical foundation for population ecology and argues for this simple birth and death model and the consequences that logically follow from it (e.g., exponential population growth) as fundamental laws of population ecology.
+@Turchin2001 argues that general principles are needed to establish a logical foundation for population ecology, and this simple birth and death model and the consequences that logically follow from it (e.g., exponential population growth) is fundamental to population ecology.
 Any unifying definition of joint ecological and evolutionary change must be able to derive both the Price equation and this birth and death model.
 
-Our approach is distinct from eco-evolutionary modelling, which instead focuses on feedbacks between ecological and evolutionary dynamics.
-While the interdependence of ecological and evolutionary processes has long been recognised [e.g., @Pelletier2009; @Fisher1958; @Darwin1859], the rise of models incorporating both is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Yamamichi2023].
-Eco-evolutionary models can be very general in their mathematical structure, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Govaert2019; @Lion2023].
-Simplifying assumptions are necessary to make eco-evolutionary models tractable, but model assumptions can be developed from, or linked back to, fundamental equations such as the Price equation [e.g., @Coulson2008; @Barfield2011; @Lion2018].
+The interdependence of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
+Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some authors broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
+@Bassar2021 recognise two types of eco-evolutionary models that follow from these interpretations.
+The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
+The second type models population demographics as functions of quantitative traits, with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024].
+Both model types can be very general, but like all predictive models, they rely on simplifying assumptions for tractability [@Levins1966; @Luque2016].
+These simplifying assumptions can be grounded in the Price equation to demonstrate accuracy and logical consistency when modelling evolutionary change [e.g., @Coulson2008; @Barfield2011; @Rees2016; @Lion2018].
+For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory.
+The role of fundamental equations is therefore important for unifying theory [@Luque2016], and we believe that a fundamental equation of eco-evolutionary change is needed.
 
-[Okay, now state how a couple of these papers did cool things, and end by saying something about how we're trying to go a step back and establish foundations]
-
-**Eco-evolutionary** is typically more broadly defined to mean any kind of feedback between ecology and evolution [Lion]
-
-But like the mechanisms driving evolutionary and genetic change, those that drive ecological and evolutionary change are well-established with predictable joint consequences.
-
-This should allow biologists to place ecological and evolutionary change on a shared theoretical foundation.
-
-We believe that such a foundation exists for population ecology and evolutionary biology, and that it can bridge the fundamental equations of both subjects and serve as a starting point for understanding change in any population.
-
-We believe that this logical foundation can be made sufficiently broad to encompass both population and evolutionary change.
-
-We propose an equation from which the fundamental ecological and evolutionary equations can be derived.
+We present an equation from which the fundamental equations of ecology and evolutionary biology can be derived.
 Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics (e.g., average excess) or quantitative genetics (e.g., Breeder's equation) can be derived from the Price equation [@Queller2017].
-We hope that our equation can provide a starting point for building a shared logical framework between ecology and evolution.
+We propose our equation as a formal definition of eco-evolutionary change.
 
@@
 To fully unify biological evolution and population ecology, we must reconcile the the Price equation (Box 1) with the general equation for population change (Box 2).
+The Price equation is critical for partitioning different components of biological change [@Price1970; @Frank1997; @Gardner2008; @Luque2016; @Lehtonen2018; @Queller2017].
+It has also been highly useful for integrating evolutionary theory across disciplines [@Fox2006; @Brantingham2007; @MacCallum2012; @Frank2015; @Godsoe2021; @Ulrich2024].
+These properties would seem to make it an intuitive starting point for a logical foundation of ecology and evolution, perhaps through some kind of mathematical equivalence [@Page2002] or addition of terms [@Collins2009], or through the use of its recursive structure [@Kerr2009; @Frank2012].
+But despite its flexibility, the Price equation still relies on relative frequencies, which must by definition sum to one [@Frank2015].
+This is because the Price equation describes the average change in a population; the frequency of entities is scaled thereby conserving total probability [@Frank2015; @Frank2016].
+But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided in a fundamental equation of ecology and evolution.
+
+<!---
+
 Previous theory has produced mathematical synthesis between ecological and evolutionary equations [@Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
@@
 
-The Price equation is critical for partitioning different components of biological change [@Price1970; @Frank1997; @Gardner2008; @Luque2016; @Lehtonen2018; @Queller2017].
-It has also been highly useful for integrating evolutionary theory across disciplines [@Fox2006; @Brantingham2007; @MacCallum2012; @Frank2015; @Godsoe2021; @Ulrich2024].
-These properties would seem to make it an intuitive starting point for a logical foundation of ecology and evolution, perhaps through some kind of mathematical equivalence [@Page2002] or addition of terms [@Collins2009], or through the use of its recursive structure [@Kerr2009; @Frank2012].
-But despite its flexibility, the Price equation still relies on relative frequencies, which must by definition sum to one [@Frank2015].
-This is because the Price equation describes the average change in a population; the frequency of entities is scaled thereby conserving total probability [@Frank2015; @Frank2016].
-But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided in a fundamental equation of ecology and evolution.
+--->
 
@@
 $$
+
+<!---
+
+Something around here verbally presenting the logic of eqn 1, discussing how we are talking about an interaction between birth-death (individual fitness) and a individual's characteristic.
+
+--->
 

### 2025-06-01 foundations_of_ecoevo.tex
Rewrite the introduction along reviewer recommendations; cut some things to the first section, and prepare for a better verbal explanation of eqn 1

@@
 unification further demonstrates the equivalence between mean population
-growth rate and evolutionary fitness, shows how ecological and
-evolutionary change are reflected in the first and second statistical
-moments of fitness, respectively, and links this change to ecosystem
+growth rate and evolutionary fitness and links this change to ecosystem
 function. Finally, we outline how our proposed framework can be used to
@@
 and its value is perhaps most evident in reconciling unconnected models
-and revealing new and unexpected empirical predictions. In theoretical
-physics, Newton's unification of celestial and terrestrial motion, and
-Maxwell's unification of electricity and magnetism, are among the most
-important advances in scientific history. Likewise, in the life
-sciences, the modern synthesis of natural selection and Mendelian
-inheritance revolutionised how mechanisms underlying evolution and
-genetics are understood and applied (\citeproc{ref-Gayon2019}{Gayon and
-Huneman 2019}). The analogy to physics is imperfect because evolution
-encompasses non-biological systems with non-genetic inheritance
-(\citeproc{ref-Price1995}{Price 1995};
-\citeproc{ref-Shennan2011}{Shennan 2011};
-\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
-\citeproc{ref-Lewens2015}{Lewens 2015}), but to illustrate our point, we
-restrict our focus to biological evolution. Following the modern
-synthesis, patterns in evolutionary biology and genetics could be
-interpreted through a common framework.
-
-Selection, mutation, recombination, migration, and drift all have
-predictable consequences for both phenotypes and genotypes in a
-population. These mechanisms are therefore foundational to understanding
-evolution and genetics, and their roles in explaining evolutionary and
-genetic phenomena are no more conceptually pluralistic than that of
-gravity in explaining celestial and terrestrial motion. Instead,
-biologists understand that each of these mechanisms will jointly affect
-genetic composition and evolutionary change in a population. For
-example, all else being equal, drift will lead to a loss of both
-heterozygosity and adaptive phenotypes. This constrains the possible
-relations that can exist between genetic composition and evolutionary
-change and therefore clarifies the scope of hypotheses that are
-biologically coherent (\citeproc{ref-Caswell1988}{Caswell 1988}). It
-would obviously be mistaken to model a population in which drift had
-inconsistent effects on evolution and genetics, but such mistakes would
-surely be common without the foundational theory required to unify the
-two subjects. The absence of this theory is almost unthinkable in modern
-biology. Evolution and genetics are arguably as intertwined in biology
-as electricity and magnetism are in physics.
-
-In contrast, theory underlying ecology and evolution has not been so
-tightly intertwined. Until recently, population ecology and evolution in
-particular have taken almost parallel paths for developing theory.
-Although many revelatory models exist to connect ecological and
-evolutionary phenomena (\citeproc{ref-Lion2018}{Lion 2018};
-\citeproc{ref-Patel2018}{Patel et al. 2018};
-\citeproc{ref-Govaert2019}{Govaert et al. 2019}), the foundational
-assumptions and equations of ecology and evolution in these models
-remain mostly separated conceptually. But like the mechanisms driving
-evolutionary and genetic change, those that drive ecological and
-evolutionary change are well-established with predictable joint
-consequences. In evolution, natural selection, drift, and gene flow are
-all instantiated through a combination of different ecological factors
-(\citeproc{ref-Connor2004}{Connor and Hartl 2004}). Most notably, change
-in population density and phenotype are both caused by individual birth
-and death (\citeproc{ref-Turchin2001}{Turchin 2001}). This should allow
-biologists to place ecological and evolutionary change on a shared
-theoretical foundation.
-
-We believe that such a foundation exists for population ecology and
-evolutionary biology, and that it can bridge the fundamental equations
-of both subjects and serve as a starting point for understanding change
-in any population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues for
-a simple birth and death model (Box 1) as a fundamental equation of
-population ecology, and for exponential growth to consequently be as
-fundamental to ecology as Newton's first law of motion is to physics.
-Turchin (\citeproc{ref-Turchin2001}{2001}) argues that general
-principles are needed to establish a logical foundation for population
-ecology. In evolutionary biology, the Price equation (Box 2) is almost
-universally regarded as fundamental (\citeproc{ref-Price1970}{Price
-1970}; \citeproc{ref-Rice2004}{Rice 2004};
-\citeproc{ref-Gardner2008}{Gardner 2008};
-\citeproc{ref-Frank2015a}{Frank 2017}; \citeproc{ref-Luque2016}{Luque
-2017}; \citeproc{ref-Luque2021}{Luque and Baravalle 2021}). It
-exhaustively and exactly describes evolutionary change for any closed
-evolving population (\citeproc{ref-Price1970}{Price 1970};
-\citeproc{ref-Luque2016}{Luque 2017};
-\citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it serves as the
-starting point for all other fundamental equations in evolution
-(\citeproc{ref-Queller2017}{Queller 2017}). We believe that this logical
-foundation can be made sufficiently broad to encompass both population
-and evolutionary change.
-
-We propose an equation from which the fundamental ecological and
-evolutionary equations can be derived. Derivation follows by adding
+and revealing new and unexpected empirical predictions. In evolutionary
+biology, the Price equation (Box 1) provides a unifying framework for
+evolutionary theory by exhaustively and exactly describing evolutionary
+change for any closed evolving population
+(\citeproc{ref-Price1970}{Price 1970}; \citeproc{ref-Luque2016}{Luque
+2017}; \citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}). The Price
+equation is therefore fundamental, in the sense that it binds together
+all of evolutionary theory by formally defining what evolutionary change
+is and is not (\citeproc{ref-Price1970}{Price 1970};
+\citeproc{ref-Rice2004}{Rice 2004}; \citeproc{ref-Gardner2008}{Gardner
+2008}; \citeproc{ref-Frank2015a}{Frank 2017};
+\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Luque2021}{Luque and
+Baravalle 2021}). Using this formal definition, the scope of, and
+relationships among, sub-disciplines within evolutionary theory can be
+clarified. For example, fundamental equations of both population and
+quantitative genetics can be derived from the Price equation
+(\citeproc{ref-Queller2017}{Queller 2017}). This provides conceptual
+clarity by demonstrating the logical consistency of different
+theoretical frameworks within evolutionary biology. Our aim here is to
+propose an equation that extends this conceptual clarity to include
+population change and thereby provide a formal and exact definition of
+joint ecological and evolutionary change.
+
+In biological populations, ecological change is caused by the same
+processes of individual birth and death that cause changes in allele
+frequencies and phenotypes (\citeproc{ref-Turchin2001}{Turchin 2001};
+\citeproc{ref-Connor2004}{Connor and Hartl 2004};
+\citeproc{ref-Barfield2011}{Barfield et al. 2011}). As with evolution, a
+fundamental equation can exhaustively and exactly define population
+change. Unlike the Price equation, this fundamental equation is perhaps
+self-evident; population change is simply the addition of births and
+subtraction of deaths from current population size (\(N_{t}\)) to
+recover the new population size (\(N_{t+1}\); Box 2). By definition,
+this relationship of \(N_{t+1} = N_{t} + Births - Deaths\) applies to
+any closed population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues
+that general principles are needed to establish a logical foundation for
+population ecology, and this simple birth and death model and the
+consequences that logically follow from it (e.g., exponential population
+growth) is fundamental to population ecology. Any unifying definition of
+joint ecological and evolutionary change must be able to derive both the
+Price equation and this birth and death model.
+
+The interdependence of ecological and evolutionary processes has long
+been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
+\citeproc{ref-Fisher1958}{Fisher 1958};
+\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
+eco-evolutionary models, which incorporate both, is relatively recent
+following a widespread recognition that ecology and evolution can happen
+on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
+\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
+universally recognised formal definition of eco-evolutionary change is
+lacking, with some authors broadly interpreting ``eco-evolutionary
+dynamics'' to allow for a separation of ecological and evolutionary
+timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
+advocating for a more narrow interpretation in which no such separation
+is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
+al. (\citeproc{ref-Bassar2021}{2021}) recognise two types of
+eco-evolutionary models that follow from these interpretations. The
+first type uses separate equations to model population change versus
+evolutionary change, thereby allowing for any number of ecological,
+evolutionary, or environmental feedbacks (e.g.,
+\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
+al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
+models population demographics as functions of quantitative traits, with
+ecological and evolutionary change following from demographic processes
+and trait distributions operating on the same timescale (e.g.,
+\citeproc{ref-Barfield2011}{Barfield et al. 2011};
+\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
+\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
+very general, but like all predictive models, they rely on simplifying
+assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
+\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions can
+be grounded in the Price equation to demonstrate accuracy and logical
+consistency when modelling evolutionary change (e.g.,
+\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
+\citeproc{ref-Barfield2011}{Barfield et al. 2011};
+\citeproc{ref-Rees2016}{Rees and Ellner 2016};
+\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
+(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
+(\citeproc{ref-Price1970}{1970}), which they consider to be a
+``universal law of evolution'', to place their conclusions concerning
+stage structured evolution in the broader context of evolutionary
+theory. The role of fundamental equations is therefore important for
+unifying theory (\citeproc{ref-Luque2016}{Luque 2017}), and we believe
+that a fundamental equation of eco-evolutionary change is needed.
+
+We present an equation from which the fundamental equations of ecology
+and evolutionary biology can be derived. Derivation follows by adding
 assumptions that are specific to population ecology or evolution in the
@@
 derived from the Price equation (\citeproc{ref-Queller2017}{Queller
-2017}). We hope that our equation can provide a starting point for
-building a shared logical framework between ecology and evolution.
+2017}). We propose our equation as a formal definition of
+eco-evolutionary change.
 
@@
 \begin{quote}
-\textbf{Box 1:} The number of individuals in any closed population
-(\(N\)) at any given time (\(t + 1\)) is determined by the existing
-number of individuals (\(N_{t}\)), plus the number of births
-(\(Births\)) minus the number of deaths (\(Deaths\)),
-\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
-for any closed population. Despite its simplicity, it is a general
-equation for defining population change and a starting point for
-understanding population ecology. Turchin
-(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
-fundamental equation is the tendency for populations to grow
-exponentially (technically geometrically in the above case where time is
-discrete). This inherent underlying tendency towards exponential growth
-persists even as the complexities of real populations, such as
-structure, stochasticity, or density-dependent effects are added to
-population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
-assumption that all individuals in the population are identical, a per
-capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
-can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
-\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
-increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
-because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
-size of any closed population equals its existing size times its finite
-rate of increase.
-\end{quote}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\begin{quote}
-\textbf{Box 2:} The Price equation is an abstract formula to represent
+\textbf{Box 1:} The Price equation is an abstract formula to represent
 evolutionary change. Formulated originally in the early 1970s by George
@@
 
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\begin{quote}
+\textbf{Box 2:} The number of individuals in any closed population
+(\(N\)) at any given time (\(t + 1\)) is determined by the existing
+number of individuals (\(N_{t}\)), plus the number of births
+(\(Births\)) minus the number of deaths (\(Deaths\)),
+\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
+for any closed population. Despite its simplicity, it is a general
+equation for defining population change and a starting point for
+understanding population ecology. Turchin
+(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
+fundamental equation is the tendency for populations to grow
+exponentially (technically geometrically in the above case where time is
+discrete). This inherent underlying tendency towards exponential growth
+persists even as the complexities of real populations, such as
+structure, stochasticity, or density-dependent effects are added to
+population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
+assumption that all individuals in the population are identical, a per
+capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
+can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
+\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
+increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
+because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
+size of any closed population equals its existing size times its finite
+rate of increase.
+\end{quote}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
 \section{A foundation for biological evolution and population
@@
 To fully unify biological evolution and population ecology, we must
-reconcile the general equation for population change (Box 1) with the
-Price equation (Box 2). Previous theory has produced mathematical
-synthesis between ecological and evolutionary equations
-(\citeproc{ref-Page2002}{Page and Nowak 2002};
-\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but not
-established a unified conceptual framework for joint ecological and
-evolutionary change. For example, Page and Nowak
-(\citeproc{ref-Page2002}{2002}) demonstrate a mathematical equivalence
-between the Price equation and a general equation for community
-dynamics. In both equations, the relative frequencies of different
-ecological (species) or evolutionary (phenotype) variants map to their
-respective ecological (population growth) or evolutionary (fitness)
-change in a way that is mathematically equivalent. Collins and Gardner
-(\citeproc{ref-Collins2009}{2009}) use the Price equation to build a
-general quantitative framework that partitions phenotypic change into
-ecological, evolutionary, and physiological components. The average
-change in a community-wide trait (e.g., carbon uptake) is exactly and
-generally partitioned into change attributable to changes in species
-composition, expected evolution of lineages within species, and expected
-change in lineage physiology across all species. Both Page and Nowak
-(\citeproc{ref-Page2002}{2002}) and Collins and Gardner
-(\citeproc{ref-Collins2009}{2009}) use relative species abundances,
-meaning that species frequencies must sum to one. This is useful for
-demonstrating mathematical equivalence (\citeproc{ref-Page2002}{Page and
-Nowak 2002}) or expected change in a community-wide trait
-(\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but a different
-approach is needed to recover absolute change in population size and
-thereby recover a joint foundation for ecology and evolution.
-
-The Price equation is critical for partitioning different components of
-biological change (\citeproc{ref-Price1970}{Price 1970};
-\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Gardner2008}{Gardner
-2008}; \citeproc{ref-Luque2016}{Luque 2017};
-\citeproc{ref-Queller2017}{Queller 2017};
-\citeproc{ref-Lehtonen2018}{Lehtonen 2018}). It has also been highly
-useful for integrating evolutionary theory across disciplines
+reconcile the the Price equation (Box 1) with the general equation for
+population change (Box 2). The Price equation is critical for
+partitioning different components of biological change
+(\citeproc{ref-Price1970}{Price 1970}; \citeproc{ref-Frank1997}{Frank
+1997}; \citeproc{ref-Gardner2008}{Gardner 2008};
+\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Queller2017}{Queller
+2017}; \citeproc{ref-Lehtonen2018}{Lehtonen 2018}). It has also been
+highly useful for integrating evolutionary theory across disciplines
 (\citeproc{ref-Fox2006}{Fox 2006};
@@
 From eqn 1, we can derive the most fundamental equations of population
-ecology (Box 1) and evolutionary biology (Box 2) through an appropriate
+ecology (Box 2) and evolutionary biology (Box 1) through an appropriate
 interpretation of \(z_{i}\).
@@
 
-To recover the general equation for population ecology (Box 1), we
+To recover the general equation for population ecology (Box 2), we
 define \(z_{i}\) as the identity of \(i\) belonging to the population.
@@
 
-If we define \(\lambda = 1 + \beta - \delta\) (Box 1), then we can
+If we define \(\lambda = 1 + \beta - \delta\) (Box 2), then we can
 rewrite,
@@
 population, we can derive both the most fundamental model of population
-ecology (eqn 2; Box 1) and the fundamental equation of evolution (eqn
-10; Box 2). We might also consider \(\Omega\) to be a quantity of
+ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn
+10; Box 1). We might also consider \(\Omega\) to be a quantity of
 ecosystem function, in that it is the summed effect of the traits
@@
 (\(\Delta z_{i} = 0\)), we recover the most general equation of
-population ecology (Box 1) and principle of exponential growth
+population ecology (Box 2) and principle of exponential growth
 (\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
@@
 \citeproc{ref-Frank2016}{2016}), we recover the most fundamental
-equation of evolution (Box 2). We therefore deliver a general framework
+equation of evolution (Box 1). We therefore deliver a general framework
 to unify fundamental equations of ecological and evolutionary change.
@@
 Our unification demonstrates the equivalence between the finite rate of
-increase \(\lambda\) (Box 1) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
-and population mean evolutionary fitness \(\bar{w}\) (Box 2). The
+increase \(\lambda\) (Box 2) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
+and population mean evolutionary fitness \(\bar{w}\) (Box 1). The
 population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
@@
 incorporated by adding a term for each to the right-hand side of the
-equation in Box 1 (\citeproc{ref-Gotelli2001}{Gotelli 2001}). In
+equation in Box 2 (\citeproc{ref-Gotelli2001}{Gotelli 2001}). In
 evolution, because the Price equation relies on mapping
@@
 includes any effects of density dependence on population growth (see Box
-1), or any social effects on evolutionary change (see Box 2). We can
+2), or any social effects on evolutionary change (see Box 1). We can
 make both of these effects explicit with the same partition in the
@@
 
+\bibitem[\citeproctext]{ref-Barfield2011}
+Barfield, M., R. D. Holt, and R. Gomulkiewicz. 2011.
+\href{https://doi.org/10.1086/658903}{Evolution in stage-structured
+populations}. American Naturalist 177:397--409.
+
+\bibitem[\citeproctext]{ref-Bassar2021}
+Bassar, R. D., T. Coulson, J. Travis, and D. N. Reznick. 2021.
+\href{https://doi.org/10.1111/ele.13712}{Towards a more precise -- and
+accurate -- view of eco-evolution}. Ecology Letters 24:623--625.
+
 \bibitem[\citeproctext]{ref-Brantingham2007}
@@
 the Price equation}}. American Antiquity 72:395--416.
-
-\bibitem[\citeproctext]{ref-Caswell1988}
-Caswell, H. 1988.
-\href{https://doi.org/10.1016/0304-3800(88)90071-3}{Theory and models in
-ecology: A different perspective}. Ecological modelling 43:33--44.
 
@@
 
+\bibitem[\citeproctext]{ref-Coulson2008}
+Coulson, T., and S. Tuljapurkar. 2008.
+\href{https://doi.org/10.1086/591693}{The dynamics of a quantitative
+trait in an age-structured population living in a variable environment}.
+American Naturalist 172:599--612.
+
+\bibitem[\citeproctext]{ref-Darwin1859}
+Darwin, C. 1859. The origin of species. Penguin.
+
 \bibitem[\citeproctext]{ref-Dobzhansky1970}
@@
 
+\bibitem[\citeproctext]{ref-Fisher1958}
+Fisher, R. A. 1958. The genetical theory of natural selection (2nd ed.).
+Dover.
+
 \bibitem[\citeproctext]{ref-Fox2006}
@@
 B: Biological Sciences 375:20190361.
-
-\bibitem[\citeproctext]{ref-Gayon2019}
-Gayon, J., and P. Huneman. 2019.
-\href{https://doi.org/10.1007/s10739-019-09569-2}{The modern synthesis:
-Theoretical or institutional event?} Journal of the History of Biology
-52:519--535.
 
@@
 
+\bibitem[\citeproctext]{ref-Jaggi2024}
+Jaggi, H., W. Zuo, R. Kentie, J. M. Gaillard, T. Coulson, and S.
+Tuljapurkar. 2024. \href{https://doi.org/10.1111/ele.14551}{Density
+dependence shapes life-history trade-offs in a food-limited population}.
+Ecology letters 27:e14551.
+
 \bibitem[\citeproctext]{ref-Kerr2009}
@@
 
-\bibitem[\citeproctext]{ref-Lewens2015}
-Lewens, T. 2015.
-\href{https://doi.org/10.1093/acprof:oso/9780199674183.001.0001}{Cultural
-evolution: Conceptual challenges}. Oxford University Press.
+\bibitem[\citeproctext]{ref-Levins1966}
+Levins, R. 1966. \href{https://doi.org/10.2307/27836590}{The strategy of
+model building in population biology}. American Naturalist.
 
@@
 perspective}}. American Naturalist 191.
+
+\bibitem[\citeproctext]{ref-Lion2023}
+Lion, S., A. Sasaki, and M. Boots. 2023.
+\href{https://doi.org/10.1111/ele.14183}{Extending eco-evolutionary
+theory with oligomorphic dynamics}. Ecology Letters 26:S22--S46.
 
@@
 
+\bibitem[\citeproctext]{ref-Pelletier2009}
+Pelletier, F., D. Garant, and A. P. Hendry. 2009.
+\href{https://doi.org/10.1098/rstb.2009.0027}{Eco-evolutionary
+dynamics}. Philosophical Transactions of the Royal Society London B
+364:1483--1489.
+
 \bibitem[\citeproctext]{ref-Pigliucci2009}
@@
 
-\bibitem[\citeproctext]{ref-Price1995}
----------. 1995. \href{https://doi.org/10.1006/jtbi.1995.0149}{{The
-nature of selection}}. Journal of Theoretical Biology 175:389--396.
-
 \bibitem[\citeproctext]{ref-Queller2017}
@@
 theorems of evolution}}. American Naturalist 189:000--000.
+
+\bibitem[\citeproctext]{ref-Rees2016}
+Rees, M., and S. P. Ellner. 2016.
+\href{https://doi.org/10.1111/2041-210X.12487}{Evolving integral
+projection models: Evolutionary demography meets eco-evolutionary
+dynamics}. Methods in Ecology and Evolution 7:157--170.
 
@@
 
-\bibitem[\citeproctext]{ref-Shennan2011}
-Shennan, S. 2011. \href{https://doi.org/10.1098/rstb.2010.0380}{{Descent
-with modification and the archaeological record}}. Philosophical
-Transactions of the Royal Society B: Biological Sciences 366:1070--1079.
+\bibitem[\citeproctext]{ref-Simmonds2020}
+Simmonds, E. G., E. F. Cole, B. C. Sheldon, and T. Coulson. 2020.
+\href{https://doi.org/10.1111/ele.13603}{Phenological asynchrony: A
+ticking time-bomb for seemingly stable populations?} Ecology Letters.
+Blackwell Publishing Ltd.
 
@@
 
+\bibitem[\citeproctext]{ref-Yamamichi2023}
+Yamamichi, M., S. P. Ellner, and N. G. Hairston. 2023.
+\href{https://doi.org/10.1111/ele.14197}{Beyond simple adaptation:
+Incorporating other evolutionary processes and concepts into
+eco-evolutionary dynamics}. Ecology Letters 26:S16--S21.
+
 \end{CSLReferences}

### 2025-06-02 foundations_of_ecoevo.Rmd
Edits up to the Discussion

@@
 The interdependence of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
-Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some authors broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
-@Bassar2021 recognise two types of eco-evolutionary models that follow from these interpretations.
+Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some theoreticians broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
+@Bassar2021 identify two types of eco-evolutionary models that follow from these interpretations.
 The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
@@
 Both model types can be very general, but like all predictive models, they rely on simplifying assumptions for tractability [@Levins1966; @Luque2016].
-These simplifying assumptions can be grounded in the Price equation to demonstrate accuracy and logical consistency when modelling evolutionary change [e.g., @Coulson2008; @Barfield2011; @Rees2016; @Lion2018].
+These simplifying assumptions are often grounded in the Price equation to demonstrate accuracy and logical consistency when modelling evolutionary change [e.g., @Coulson2008; @Barfield2011; @Rees2016; @Lion2018].
 For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory.
@@
 
+Also need to explain units of Omega, since there is confusion here.
+
 --->
 
-From eqn 1, we can derive the most fundamental equations of population ecology (Box 2) and evolutionary biology (Box 1) through an appropriate interpretation of $z_{i}$.
+The foundation of eco-evolutionary change is an interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
+The value of $\Omega$ is a summation of individual characteristics, which takes the same units as $z$.
+From eqn 1, we can derive the most fundamental equations of population ecology (Box 2) and evolutionary biology (Box 1) through an appropriate interpretation of $z$.
+Under many interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.
 
@@
 In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
+In this case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit; @NewellEtAl2019].
 Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
@@
 $$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right).$$
-
-Since we assume that individuals are identical, we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
-Summing from 1 to $N$, we can rewrite the above,
-
-$$\Omega = N\left(1 + \beta - \delta \right).$$
 
@@
 $$
+
+If we assume that individuals are identical, then we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
+Summing from 1 to $N$, we can rewrite the above,
+
+$$\Omega = N\left(1 + \beta - \delta \right).$$
 
@@
 
-Equation 8 conserves the total probability and now clarifies $\Omega$ as a summed trait value, which equals expected population growth at $t$ times mean trait value at $t + 1$.
+Equation 8 conserves the total probability and now clarifies $\Omega$ as a summed trait value, which has the same units as $z$ and equals expected population growth at $t$ times mean trait value at $t + 1$.
 This is consistent with the population ecology derivation from the previous section where $z_{i} = 1$ by definition, and $\Omega = N_{t+1}$ (note $\lambda = \bar{w}$).
@@
 From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).
-We might also consider $\Omega$ to be a quantity of ecosystem function, in that it is the summed effect of the traits ($z_{i}$) of individual entities ($i$) in the population.
-For example, $z$ might be interpreted as biomass or photosynthetic rate.
+
+Ecosystem function
+====================================================
+
+In some cases, we might also consider $\Omega$ to be the total contribution of a population to ecosystem function.
+For example, $z$ might be interpreted as trait that underlies a function of interest such as carbon capture or photosynthetic rate.
+When that the relevant characteristic ($z_{i}$) is additive across individuals  (e.g., the total amount of carbon capture in a population is the sum of carbon captured across individuals), $\Omega$ could be an appropriate metric of ecosystem function.
+This will not always be the case.
+The sum of some measurable traits such as leaf pH [@Cornelissen2011] will not be ecologically meaningful when summed across individuals.
+And in other cases, the relationship between $\Omega$ and some ecosystem function of interest is likely to be subadditive.
+For example, we could interpret $z_{i}$ as the frequency at which a pollinator $i$ visits plants (which might change from $t$ to $t+1$ for any number of reasons, including ontogenic or environmental, as reflected in $\Delta z_{i}$).
+While $\Omega$ would then be the total frequency of pollinator visits in a population, pollinator saturation effects would likely have a logarithmic relationship between $\Omega$ and total seed set [e.g., @Mu2018].
+While we have argued that ecological and evolutionary change can both be derived from the same fundamental relationship between birth, death, and individual characteristics, the same is not true for ecosystem function.
+Nevertheless, Box 3 provides a toy example of four plants with different fitnesses and carbon captures.
+We use this toy example to demonstrate a concrete example of our proposed framework.
 

### 2025-06-02 foundations_of_ecoevo.tex
Edits up to the Discussion

@@
 universally recognised formal definition of eco-evolutionary change is
-lacking, with some authors broadly interpreting ``eco-evolutionary
+lacking, with some theoreticians broadly interpreting ``eco-evolutionary
 dynamics'' to allow for a separation of ecological and evolutionary
@@
 is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
-al. (\citeproc{ref-Bassar2021}{2021}) recognise two types of
+al. (\citeproc{ref-Bassar2021}{2021}) identify two types of
 eco-evolutionary models that follow from these interpretations. The
@@
 assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
-\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions can
-be grounded in the Price equation to demonstrate accuracy and logical
+\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions are
+often grounded in the Price equation to demonstrate accuracy and logical
 consistency when modelling evolutionary change (e.g.,
@@
 
-From eqn 1, we can derive the most fundamental equations of population
-ecology (Box 2) and evolutionary biology (Box 1) through an appropriate
-interpretation of \(z_{i}\).
+The foundation of eco-evolutionary change is an interaction between the
+demographic processes of birth and death
+(\(\beta_{i} - \delta_{i} + 1\)) and some characteristic of an
+individual (\(z_{i} + \Delta z_{i}\)). The value of \(\Omega\) is a
+summation of individual characteristics, which takes the same units as
+\(z\). From eqn 1, we can derive the most fundamental equations of
+population ecology (Box 2) and evolutionary biology (Box 1) through an
+appropriate interpretation of \(z\). Under many interpretations of
+\(z\), we can also interpret \(\Omega\) as a metric of ecosystem
+function.
 
@@
 In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
-member of the population. Further, we assume that individuals do not
-change species by setting \(\Delta z_{i} = 0\). In this case,
+member of the population. In this case, \(z\) is count, which takes the
+unit 1 {[}note that `individual' is a label, not a unit; Newell and
+Tiesinga (\citeproc{ref-NewellEtAl2019}{2019}){]}. Further, we assume
+that individuals do not change species by setting \(\Delta z_{i} = 0\).
+In this case,
 
 \[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right).\]
-
-Since we assume that individuals are identical, we can drop the
-subscript \(i\) such that \(\beta_{i} = \beta\) and
-\(\delta_{i} = \delta\). Summing from 1 to \(N\), we can rewrite the
-above,
-
-\[\Omega = N\left(1 + \beta - \delta \right).\]
 
@@
 \]
+
+If we assume that individuals are identical, then we can drop the
+subscript \(i\) such that \(\beta_{i} = \beta\) and
+\(\delta_{i} = \delta\). Summing from 1 to \(N\), we can rewrite the
+above,
+
+\[\Omega = N\left(1 + \beta - \delta \right).\]
 
@@
 Equation 8 conserves the total probability and now clarifies \(\Omega\)
-as a summed trait value, which equals expected population growth at
-\(t\) times mean trait value at \(t + 1\). This is consistent with the
-population ecology derivation from the previous section where
-\(z_{i} = 1\) by definition, and \(\Omega = N_{t+1}\) (note
-\(\lambda = \bar{w}\)). We can therefore rewrite eqn 7,
+as a summed trait value, which has the same units as \(z\) and equals
+expected population growth at \(t\) times mean trait value at \(t + 1\).
+This is consistent with the population ecology derivation from the
+previous section where \(z_{i} = 1\) by definition, and
+\(\Omega = N_{t+1}\) (note \(\lambda = \bar{w}\)). We can therefore
+rewrite eqn 7,
 
@@
 ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn
-10; Box 1). We might also consider \(\Omega\) to be a quantity of
-ecosystem function, in that it is the summed effect of the traits
-(\(z_{i}\)) of individual entities (\(i\)) in the population. For
-example, \(z\) might be interpreted as biomass or photosynthetic rate.
+10; Box 1).
+
+\section{Ecosystem function}\label{ecosystem-function}
+
+In some cases, we might also consider \(\Omega\) to be the total
+contribution of a population to ecosystem function. For example, \(z\)
+might be interpreted as trait that underlies a function of interest such
+as carbon capture or photosynthetic rate. When that the relevant
+characteristic (\(z_{i}\)) is additive across individuals (e.g., the
+total amount of carbon capture in a population is the sum of carbon
+captured across individuals), \(\Omega\) could be an appropriate metric
+of ecosystem function. This will not always be the case. The sum of some
+measurable traits such as leaf pH
+(\citeproc{ref-Cornelissen2011}{Cornelissen et al. 2011}) will not be
+ecologically meaningful when summed across individuals. And in other
+cases, the relationship between \(\Omega\) and some ecosystem function
+of interest is likely to be subadditive. For example, we could interpret
+\(z_{i}\) as the frequency at which a pollinator \(i\) visits plants
+(which might change from \(t\) to \(t+1\) for any number of reasons,
+including ontogenic or environmental, as reflected in \(\Delta z_{i}\)).
+While \(\Omega\) would then be the total frequency of pollinator visits
+in a population, pollinator saturation effects would likely have a
+logarithmic relationship between \(\Omega\) and total seed set (e.g.,
+\citeproc{ref-Mu2018}{Mu et al. 2018}). While we have argued that
+ecological and evolutionary change can both be derived from the same
+fundamental relationship between birth, death, and individual
+characteristics, the same is not true for ecosystem function.
+Nevertheless, Box 3 provides a toy example of four plants with different
+fitnesses and carbon captures. We use this toy example to demonstrate a
+concrete example of our proposed framework.
 
@@
 
+\bibitem[\citeproctext]{ref-Cornelissen2011}
+Cornelissen, J. H. C., F. Sibma, R. S. P. V. Logtestijn, R. A. Broekman,
+and K. Thompson. 2011.
+\href{https://doi.org/10.1111/j.1365-2435.2010.01765.x}{Leaf pH as a
+plant trait: Species-driven rather than soil-driven variation}.
+Functional Ecology 25:449--455.
+
 \bibitem[\citeproctext]{ref-Coulson2008}
@@
 
+\bibitem[\citeproctext]{ref-Mu2018}
+Mu, J., Q. Wu, Y. Yang, M. Huang, and C. M. Grozinger. 2018.
+\href{https://doi.org/10.1111/oik.04711}{Plant reproductive strategies
+vary under low and high pollinator densities}. Oikos 127:1081--1094.
+
 \bibitem[\citeproctext]{ref-Nagel1961}
@@
 scientific explanation. Harcourt, Brace \& World, New York, NY, USA.
+
+\bibitem[\citeproctext]{ref-NewellEtAl2019}
+Newell, D. B., and E. Tiesinga. 2019.
+\href{https://doi.org/10.6028/NIST.SP.330-2019}{The international system
+of units (SI)}.
 

### 2025-06-05 foundations_of_ecoevo.Rmd
UPdate two of the references

@@
 For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory.
-The role of fundamental equations is therefore important for unifying theory [@Luque2016], and we believe that a fundamental equation of eco-evolutionary change is needed.
+The role of fundamental equations is therefore important for unifying theory [@Luque2021], and we believe that a fundamental equation of eco-evolutionary change is needed.
 
@@
 Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004].
-Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's absolute fitness, and $\bar{w}$ average population fitness.
+Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's fitness, and $\bar{w}$ average population fitness.
 On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction.
@@
 
-We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$ and note that $N\beta$, $N\delta$, and $N$ are the total births, total deaths, and size of the population at $t$, respectively,
+We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$,
 
@@
 
-If we assume that individuals are identical, then we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
+If we assume that individuals are identical, then we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$ and note that $N\beta$, $N\delta$, and $N$ are the total births, total deaths, and size of the population at $t$, respectively,
+
 Summing from 1 to $N$, we can rewrite the above,
@@
 In some cases, we might also consider $\Omega$ to be the total contribution of a population to ecosystem function.
-For example, $z$ might be interpreted as trait that underlies a function of interest such as carbon capture or photosynthetic rate.
+For example, $z$ might be interpreted as a trait that underlies an ecosystem function of interest such as carbon capture or photosynthetic rate.
 When that the relevant characteristic ($z_{i}$) is additive across individuals  (e.g., the total amount of carbon capture in a population is the sum of carbon captured across individuals), $\Omega$ could be an appropriate metric of ecosystem function.
 This will not always be the case.
-The sum of some measurable traits such as leaf pH [@Cornelissen2011] will not be ecologically meaningful when summed across individuals.
+Some measurable traits such as leaf pH [@Cornelissen2011] will not be ecologically meaningful when summed across individuals.
 And in other cases, the relationship between $\Omega$ and some ecosystem function of interest is likely to be subadditive.
@@
 While $\Omega$ would then be the total frequency of pollinator visits in a population, pollinator saturation effects would likely have a logarithmic relationship between $\Omega$ and total seed set [e.g., @Mu2018].
-While we have argued that ecological and evolutionary change can both be derived from the same fundamental relationship between birth, death, and individual characteristics, the same is not true for ecosystem function.
+Hence, while we have argued that ecological and evolutionary change can both be derived from the same fundamental relationship between birth, death, and individual characteristics, the same is not true for ecosystem function.
 Nevertheless, Box 3 provides a toy example of four plants with different fitnesses and carbon captures.

### 2025-06-08 foundations_of_ecoevo.Rmd
Manuscript updates

@@
 output:
+  word_document: default
   pdf_document:
       keep_tex: yes
-  word_document: default
   html_document:
@@
 From eqn 1, we can derive the most fundamental equations of population ecology (Box 2) and evolutionary biology (Box 1) through an appropriate interpretation of $z$.
-Under many interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.
+Under some limited interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.
 
@@
 
-In some cases, we might also consider $\Omega$ to be the total contribution of a population to ecosystem function.
-For example, $z$ might be interpreted as a trait that underlies an ecosystem function of interest such as carbon capture or photosynthetic rate.
-When that the relevant characteristic ($z_{i}$) is additive across individuals  (e.g., the total amount of carbon capture in a population is the sum of carbon captured across individuals), $\Omega$ could be an appropriate metric of ecosystem function.
-This will not always be the case.
-Some measurable traits such as leaf pH [@Cornelissen2011] will not be ecologically meaningful when summed across individuals.
-And in other cases, the relationship between $\Omega$ and some ecosystem function of interest is likely to be subadditive.
-For example, we could interpret $z_{i}$ as the frequency at which a pollinator $i$ visits plants (which might change from $t$ to $t+1$ for any number of reasons, including ontogenic or environmental, as reflected in $\Delta z_{i}$).
-While $\Omega$ would then be the total frequency of pollinator visits in a population, pollinator saturation effects would likely have a logarithmic relationship between $\Omega$ and total seed set [e.g., @Mu2018].
-Hence, while we have argued that ecological and evolutionary change can both be derived from the same fundamental relationship between birth, death, and individual characteristics, the same is not true for ecosystem function.
-Nevertheless, Box 3 provides a toy example of four plants with different fitnesses and carbon captures.
+In some cases, $\Omega$ could also be intepreted as the total contribution of a population to ecosystem function.
+This will be restricted to cases in which $z$ is an individual characteristic defining an absolute quantity measured at the whole organism level, such as biomass, seed production, carbon capture, flower visits, or nutrient consumption.
+In these cases, the sum across individuals gives a meaningful total quantity for the population.
+When $z$ is instead defined by relative organism-level measurements such wing loading, nutrient ratio, or diet composition, or when $z$ is measured at a level of biological organisation below the organism (e.g., average cell volume or leaf surface area), $\Omega$ does not have a clear population-level interpretation.
+
+Box 3 provides a toy example of three plants with different fitnesses and carbon captures.
 We use this toy example to demonstrate a concrete example of our proposed framework.
+
+*******************
+
+> **Box 3:** As a toy example of our framework, consider a population of $N_{t} = 3$ annual plants in which individual fruit mass (kg) is measured, and change is observed over a year.
+For all plants, $\delta_{i} = 1$, and let plant fecundities be $\beta_{1} = 1$, $\beta_{2} = 1$, and $\beta_{3} = 2$.
+Applying eqn 1 to population change such that $\Omega = N_{t+1}$, $z_{i} = 1$ and $\Delta z_{i} = 0$, $N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4$ (note $\bar{w} = 4/3$, so $N_{t}\bar{w} = 4$).
+Focusing next on the characteristic of total fruit mass, let $z_{1} = 0.8$, $z_{2} = 1.0$, and $z_{3} = 1.5$.
+Also let $\Delta z_{i} = 0.1$ for all plants to reflect a change in soil environment from $t$ to $t + 1$.
+In this case, total fruit yield at $t+1$ is $\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2$ kg.
+At $t$, mean fruit yield per plant was 1.1 kg, but at $t+1$, mean fruit yield per plant is $1.3$ kg.
+Note that $\mathrm{Cov}(w, z) = 2/15$ and $\mathrm{E}(w \Delta z) = 2/15$, so applying the Price equation, $\bar{w}\Delta z = 2/15 + 2/15 = 4/15$.
+Since $\bar{w} = 4/3$, multiplying both sides of the equation by $3/4$ returns $\Delta z = 0.2$, which is the mean difference in fruit yield between $t+1$ and $t$.
+The framework expressed in eqn 1 thereby links population change, evolutionary change, and ecosystem function.
+
+*******************
 
@@
 
-A classical sign of scientific progress is the ability to connect disparate theories and models to show how empirical and theoretical models are logical (mathematical) consequences of more fundamental ones [@Nagel1961].
-Conceptual unification therefore has a critical role in advancing scientific theory [@Morrison2000].
-Showing how apparently disparate phenomena follow from the same shared principles reveals what is most fundamental about the natural world and provides a foundation for further scientific investigation and the construction of more coherent and predictive models.
-Our eqn 1 provides a foundation for the conceptual unification of the fundamental equations of population ecology and evolution.
-
+A classical sign of scientific progress is the ability to connect disparate theories and models to show how empirical and theoretical models are logical (mathematical) consequences of more fundamental ones [@Nagel1961; @Morrison2000].
 Rather than making simplifying assumptions, as is the approach for most ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
@@
 By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 1).
-We therefore deliver a general framework to unify fundamental equations of ecological and evolutionary change.
+Our eqn 1 thereby provides a foundation for defining eco-evolutionary change in any population.
 
@@
 This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation (e.g., group, individual) [@Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
-Because of its abstract nature and lack of any system-specific assumptions, the Price equation makes no predictions about what will happen in any particular system [@Gardner2020].
+Because of its abstract nature and lack of any system-specific assumptions, the Price equation is not dynamically sufficient and makes no predictions about what will happen in any particular system [@Gardner2020].
 Its role is not to predict, but to formally and completely define and separate components of evolutionary change. 
@@
 This equation formally and completely describes population change in terms of births and deaths.
-In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population.
+In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population, but one that is revealed to be necessarily abstract and not dynamically sufficient.
 We anticipate that this will be useful for eco-evolutionary theory in the same way that the Price equation is useful for evolutionary theory: facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.
 
-Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ (Box 2) [@Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 1).
-The population growth equation $N_{t+1} = N_{t}\lambda$ could therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
-We do not claim to be the first to notice this equivalence between population growth and mean population fitness.
-Indeed, the relationship between population growth rate and evolutionary fitness has been proposed and applied many times before [@Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
-For example, @Lande1982 explicitly concludes mean absolute fitness per unit time is $\bar{w} = e^{r}$, where $r = ln(\lambda)$.
+Our unification recovers the equivalence between the finite rate of increase $\lambda$ (Box 2) [@Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 1).
+The population growth equation $N_{t+1} = N_{t}\lambda$ can therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
+This specific equivalence has been proposed before [@Lande1976], as has the broader relationship between population growth rate and evolutionary fitness [@Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
 We show this from first principles and clarify the relationship between fitness and population growth.
@@
 Over an individual's lifetime (which, by definition, includes death), fitness is therefore $\beta_{i}$. 
-Interestingly, the rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
-Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)$ (i.e., Fisher's fundamental theorem) [@Frank1997; @Rice2004; @Queller2017].
+The rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
+Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)/\bar{w}$ (i.e., Fisher's fundamental theorem) [@Frank1997; @Rice2004; @Queller2017].
 
 Our unification may also help explain, at least partially, some of the success of classical population genetic models. 
-For decades, population genetics (and to some extent quantitative genetics) has been accused of being a reductionistic view of evolution, reducing everything to changes in allele frequencies and abstracting away from individuals and their environments (the ecological interactions) [@MacColl2011].
+For decades, population genetics (and to some extent quantitative genetics) has been accused of being a reductionist view of evolution, reducing everything to changes in allele frequencies and abstracting away from individuals and their environments (the ecological interactions) [@MacColl2011].
 This has been a line of argumentation by some defenders of the so-called Extended Synthesis [@Pigliucci2009], especially in relation to niche construction [@Odling-smee2003].
 Famously, @Mayr1960 characterized population genetics as a simple input and output of genes, analogous to "the adding of certain beans to a beanbag and the withdrawing of others" (also called "beanbag genetics").
-
 Historical critics of population genetics could not articulate a clear explanation for why it works so well despite all of its idealisations and simplifications.
@@
 It therefore implicitly includes any effects of density dependence on population growth (see Box 2), or any social effects on evolutionary change (see Box 1).
-We can make both of these effects explicit with the same partition in the summation on the right-hand side of eqn 1.
-To do this, we multiply the expression $\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$ by $\left(1 - \sum_{j=1}^{N} a_{ij}\left(z_{j} + \Delta z_{j}\right)\right)$, where $j$ indexes the same individuals as $i$, and $a_{ij}$ defines the effect $j$ has on the contribution of $i$ to $\Omega$.
-In evolutionary terms, $a_{ij}$ defines how $j$ modulates the fitness of $i$ ($w_{i}$).
-In ecological terms, $a_{ij}$ defines how $j$ modulates the contribution of $i$ to total population growth.
-If we define $\alpha$ as the summed effect of all individuals on a focal $i$ (i.e., $\alpha = \sum_{j=1}^{N} a_{ij}$), then we can partition ecological change into density-independent and density-dependent effects and derive a discrete time logistic growth equation.
-We save these derivations for a future investigations focused on evolutionary and community ecology.
+Both of these effects can be made explicit by specifying how other individuals in a population affect birth and death of a focal individual.
+Assuming that effects of individuals are additive, independent, and identical, we can derive well-established logistic growth equations and Hamilton's rule from eqn 1 (Supporting Information S1).
 
@@
 This approach provides a framework for comparing the effects of biodiversity on ecosystem function in empirical systems [@Fox2006; @Winfree2015; @Mateo-Tomas2017].
-Instead, our eqn 1 defines $\Omega$ as total ecosystem function contributed by a focal population (e.g., biomass production, decomposition, photosynthetic rate).
+Instead, our eqn 1 defines $\Omega$ as total ecosystem function contributed by a focal population.
 It is therefore possible to investigate ecological, evolutionary, and ecosystem function change from the same shared framework.

### 2025-06-08 foundations_of_ecoevo.tex
Manuscript updates

@@
 theory. The role of fundamental equations is therefore important for
-unifying theory (\citeproc{ref-Luque2016}{Luque 2017}), and we believe
-that a fundamental equation of eco-evolutionary change is needed.
+unifying theory (\citeproc{ref-Luque2021}{Luque and Baravalle 2021}),
+and we believe that a fundamental equation of eco-evolutionary change is
+needed.
 
@@
 character value \(z\) over a time step of arbitrary length, \(w\) is an
-individual's absolute fitness, and \(\bar{w}\) average population
-fitness. On the right-hand side of the equation, the first term is the
-covariance between a character value \(z\) and fitness \(w\), which
-reflects \(\bar{z}\) change attributable to differential survival and
+individual's fitness, and \(\bar{w}\) average population fitness. On the
+right-hand side of the equation, the first term is the covariance
+between a character value \(z\) and fitness \(w\), which reflects
+\(\bar{z}\) change attributable to differential survival and
 reproduction. The second term is the expected value of \(\Delta z\),
@@
 remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
-\citeproc{ref-Baravalle2024}{Baravalle et al. 2024}).
+\citeproc{ref-Baravalle2024}{Baravalle et al. 2025}).
 \end{quote}
@@
 population ecology (Box 2) and evolutionary biology (Box 1) through an
-appropriate interpretation of \(z\). Under many interpretations of
-\(z\), we can also interpret \(\Omega\) as a metric of ecosystem
+appropriate interpretation of \(z\). Under some limited interpretations
+of \(z\), we can also interpret \(\Omega\) as a metric of ecosystem
 function.
@@
 We can now interpret \(\Omega\) as the population size at \(t+1\),
-\(N_{t+1}\) and note that \(N\beta\), \(N\delta\), and \(N\) are the
-total births, total deaths, and size of the population at \(t\),
-respectively,
+\(N_{t+1}\),
 
@@
 subscript \(i\) such that \(\beta_{i} = \beta\) and
-\(\delta_{i} = \delta\). Summing from 1 to \(N\), we can rewrite the
-above,
+\(\delta_{i} = \delta\) and note that \(N\beta\), \(N\delta\), and \(N\)
+are the total births, total deaths, and size of the population at \(t\),
+respectively,
+
+Summing from 1 to \(N\), we can rewrite the above,
 
@@
 
-In some cases, we might also consider \(\Omega\) to be the total
-contribution of a population to ecosystem function. For example, \(z\)
-might be interpreted as trait that underlies a function of interest such
-as carbon capture or photosynthetic rate. When that the relevant
-characteristic (\(z_{i}\)) is additive across individuals (e.g., the
-total amount of carbon capture in a population is the sum of carbon
-captured across individuals), \(\Omega\) could be an appropriate metric
-of ecosystem function. This will not always be the case. The sum of some
-measurable traits such as leaf pH
-(\citeproc{ref-Cornelissen2011}{Cornelissen et al. 2011}) will not be
-ecologically meaningful when summed across individuals. And in other
-cases, the relationship between \(\Omega\) and some ecosystem function
-of interest is likely to be subadditive. For example, we could interpret
-\(z_{i}\) as the frequency at which a pollinator \(i\) visits plants
-(which might change from \(t\) to \(t+1\) for any number of reasons,
-including ontogenic or environmental, as reflected in \(\Delta z_{i}\)).
-While \(\Omega\) would then be the total frequency of pollinator visits
-in a population, pollinator saturation effects would likely have a
-logarithmic relationship between \(\Omega\) and total seed set (e.g.,
-\citeproc{ref-Mu2018}{Mu et al. 2018}). While we have argued that
-ecological and evolutionary change can both be derived from the same
-fundamental relationship between birth, death, and individual
-characteristics, the same is not true for ecosystem function.
-Nevertheless, Box 3 provides a toy example of four plants with different
-fitnesses and carbon captures. We use this toy example to demonstrate a
-concrete example of our proposed framework.
+In some cases, \(\Omega\) could also be intepreted as the total
+contribution of a population to ecosystem function. This will be
+restricted to cases in which \(z\) is an individual characteristic
+defining an absolute quantity measured at the whole organism level, such
+as biomass, seed production, carbon capture, flower visits, or nutrient
+consumption. In these cases, the sum across individuals gives a
+meaningful total quantity for the population. When \(z\) is instead
+defined by relative organism-level measurements such wing loading,
+nutrient ratio, or diet composition, or when \(z\) is measured at a
+level of biological organisation below the organism (e.g., average cell
+volume or leaf surface area), \(\Omega\) does not have a clear
+population-level interpretation.
+
+Box 3 provides a toy example of three plants with different fitnesses
+and carbon captures. We use this toy example to demonstrate a concrete
+example of our proposed framework.
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\begin{quote}
+\textbf{Box 3:} As a toy example of our framework, consider a population
+of \(N_{t} = 3\) annual plants in which individual fruit mass (kg) is
+measured, and change is observed over a year. For all plants,
+\(\delta_{i} = 1\), and let plant fecundities be \(\beta_{1} = 1\),
+\(\beta_{2} = 1\), and \(\beta_{3} = 2\). Applying eqn 1 to population
+change such that \(\Omega = N_{t+1}\), \(z_{i} = 1\) and
+\(\Delta z_{i} = 0\),
+\(N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4\)
+(note \(\bar{w} = 4/3\), so \(N_{t}\bar{w} = 4\)). Focusing next on the
+characteristic of total fruit mass, let \(z_{1} = 0.8\),
+\(z_{2} = 1.0\), and \(z_{3} = 1.5\). Also let \(\Delta z_{i} = 0.1\)
+for all plants to reflect a change in soil environment from \(t\) to
+\(t + 1\). In this case, total fruit yield at \(t+1\) is
+\(\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2\)
+kg. At \(t\), mean fruit yield per plant was 1.1 kg, but at \(t+1\),
+mean fruit yield per plant is \(1.3\) kg. Note that
+\(\mathrm{Cov}(w, z) = 2/15\) and \(\mathrm{E}(w \Delta z) = 2/15\), so
+applying the Price equation, \(\bar{w}\Delta z = 2/15 + 2/15 = 4/15\).
+Since \(\bar{w} = 4/3\), multiplying both sides of the equation by
+\(3/4\) returns \(\Delta z = 0.2\), which is the mean difference in
+fruit yield between \(t+1\) and \(t\). The framework expressed in eqn 1
+thereby links population change, evolutionary change, and ecosystem
+function.
+\end{quote}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
 
@@
 models are logical (mathematical) consequences of more fundamental ones
-(\citeproc{ref-Nagel1961}{Nagel 1961}). Conceptual unification therefore
-has a critical role in advancing scientific theory
-(\citeproc{ref-Morrison2000}{Morrison 2000}). Showing how apparently
-disparate phenomena follow from the same shared principles reveals what
-is most fundamental about the natural world and provides a foundation
-for further scientific investigation and the construction of more
-coherent and predictive models. Our eqn 1 provides a foundation for the
-conceptual unification of the fundamental equations of population
-ecology and evolution.
-
-Rather than making simplifying assumptions, as is the approach for most
-ecological and evolutionary models, we focus on fundamental axioms that
-are universal to closed biological systems: discrete individuals, birth,
-death, and change over time. We define an abstract sum (\(\Omega\)), to
-which all individuals within the population contribute. From the simple
+(\citeproc{ref-Nagel1961}{Nagel 1961};
+\citeproc{ref-Morrison2000}{Morrison 2000}). Rather than making
+simplifying assumptions, as is the approach for most ecological and
+evolutionary models, we focus on fundamental axioms that are universal
+to closed biological systems: discrete individuals, birth, death, and
+change over time. We define an abstract sum (\(\Omega\)), to which all
+individuals within the population contribute. From the simple
 assumptions of population identity (\(z_{i} = 1\)) and invariability
@@
 \citeproc{ref-Frank2016}{2016}), we recover the most fundamental
-equation of evolution (Box 1). We therefore deliver a general framework
-to unify fundamental equations of ecological and evolutionary change.
+equation of evolution (Box 1). Our eqn 1 thereby provides a foundation
+for defining eco-evolutionary change in any population.
 
@@
 and Otsuka 2020}). Because of its abstract nature and lack of any
-system-specific assumptions, the Price equation makes no predictions
-about what will happen in any particular system
-(\citeproc{ref-Gardner2020}{Gardner 2020}). Its role is not to predict,
-but to formally and completely define and separate components of
-evolutionary change. The same is true of the general equation for
-population change (eqns 2 and 3), at least as we have used it here where
-it serves to define what population growth means in ecology. This
-equation formally and completely describes population change in terms of
-births and deaths. In eqn 1, we therefore have a fundamental equation
-from which we can derive complete ecological and evolutionary change in
-any closed biological population. We anticipate that this will be useful
-for eco-evolutionary theory in the same way that the Price equation is
-useful for evolutionary theory: facilitating specific model development
-and identifying new conceptual insights, unresolved errors, and sources
-of model disagreements.
-
-Our unification demonstrates the equivalence between the finite rate of
+system-specific assumptions, the Price equation is not dynamically
+sufficient and makes no predictions about what will happen in any
+particular system (\citeproc{ref-Gardner2020}{Gardner 2020}). Its role
+is not to predict, but to formally and completely define and separate
+components of evolutionary change. The same is true of the general
+equation for population change (eqns 2 and 3), at least as we have used
+it here where it serves to define what population growth means in
+ecology. This equation formally and completely describes population
+change in terms of births and deaths. In eqn 1, we therefore have a
+fundamental equation from which we can derive complete ecological and
+evolutionary change in any closed biological population, but one that is
+revealed to be necessarily abstract and not dynamically sufficient. We
+anticipate that this will be useful for eco-evolutionary theory in the
+same way that the Price equation is useful for evolutionary theory:
+facilitating specific model development and identifying new conceptual
+insights, unresolved errors, and sources of model disagreements.
+
+Our unification recovers the equivalence between the finite rate of
 increase \(\lambda\) (Box 2) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
 and population mean evolutionary fitness \(\bar{w}\) (Box 1). The
-population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
-rewritten as \(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first
-to notice this equivalence between population growth and mean population
-fitness. Indeed, the relationship between population growth rate and
-evolutionary fitness has been proposed and applied many times before
-(\citeproc{ref-Fisher1930}{Fisher 1930};
+population growth equation \(N_{t+1} = N_{t}\lambda\) can therefore be
+rewritten as \(N_{t+1} = N_{t}\bar{w}\). This specific equivalence has
+been proposed before (\citeproc{ref-Lande1976}{Lande 1976}), as has the
+broader relationship between population growth rate and evolutionary
+fitness (\citeproc{ref-Fisher1930}{Fisher 1930};
 \citeproc{ref-Charlesworth1980}{Charlesworth 1980};
 \citeproc{ref-Lande1982}{Lande 1982}; \citeproc{ref-Roff2008}{Roff
-2008}; \citeproc{ref-Lion2018}{Lion 2018}). For example, Lande
-(\citeproc{ref-Lande1982}{1982}) explicitly concludes mean absolute
-fitness per unit time is \(\bar{w} = e^{r}\), where \(r = ln(\lambda)\).
-We show this from first principles and clarify the relationship between
-fitness and population growth. Over an arbitrary length of time, fitness
-is properly defined as \(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an
-individual's lifetime (which, by definition, includes death), fitness is
-therefore \(\beta_{i}\). Interestingly, the rate of change in ecology
-and evolution are reflected in the first and second statistical moments
-of fitness, respectively. Population growth rate reflects mean fitness
-\(\bar{w}\), while the rate of evolutionary change reflects the variance
-in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem)
+2008}; \citeproc{ref-Lion2018}{Lion 2018}). We show this from first
+principles and clarify the relationship between fitness and population
+growth. Over an arbitrary length of time, fitness is properly defined as
+\(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an individual's lifetime
+(which, by definition, includes death), fitness is therefore
+\(\beta_{i}\). The rate of change in ecology and evolution are reflected
+in the first and second statistical moments of fitness, respectively.
+Population growth rate reflects mean fitness \(\bar{w}\), while the rate
+of evolutionary change reflects the variance in fitness
+\(\mathrm{Var}(w)/\bar{w}\) (i.e., Fisher's fundamental theorem)
 (\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
@@
 genetics (and to some extent quantitative genetics) has been accused of
-being a reductionistic view of evolution, reducing everything to changes
+being a reductionist view of evolution, reducing everything to changes
 in allele frequencies and abstracting away from individuals and their
@@
 ``the adding of certain beans to a beanbag and the withdrawing of
-others'' (also called ``beanbag genetics'').
-
-Historical critics of population genetics could not articulate a clear
-explanation for why it works so well despite all of its idealisations
-and simplifications. From the Price equation, we are able to recover
-classical population and quantitative genetic models
-(\citeproc{ref-Queller2017}{Queller 2017}) and develop new ones
-(\citeproc{ref-Rice2004}{Rice 2004}, \citeproc{ref-Rice2020}{2020};
-\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Lion2018}{Lion
-2018}). Our eqn 1 contains ecology at its core, and we show how the
-Price equation logically follows from it after accounting for absolute
-population growth (eqn 8). We therefore conclude that population and
-quantitative genetic equations contain ecology (no matter how hidden),
-and the ecological nature of evolution is implicit in population and
-quantitative genetic models.
+others'' (also called ``beanbag genetics''). Historical critics of
+population genetics could not articulate a clear explanation for why it
+works so well despite all of its idealisations and simplifications. From
+the Price equation, we are able to recover classical population and
+quantitative genetic models (\citeproc{ref-Queller2017}{Queller 2017})
+and develop new ones (\citeproc{ref-Rice2004}{Rice 2004},
+\citeproc{ref-Rice2020}{2020}; \citeproc{ref-Luque2016}{Luque 2017};
+\citeproc{ref-Lion2018}{Lion 2018}). Our eqn 1 contains ecology at its
+core, and we show how the Price equation logically follows from it after
+accounting for absolute population growth (eqn 8). We therefore conclude
+that population and quantitative genetic equations contain ecology (no
+matter how hidden), and the ecological nature of evolution is implicit
+in population and quantitative genetic models.
 
@@
 includes any effects of density dependence on population growth (see Box
-2), or any social effects on evolutionary change (see Box 1). We can
-make both of these effects explicit with the same partition in the
-summation on the right-hand side of eqn 1. To do this, we multiply the
-expression
-\(\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\)
-by
-\(\left(1 - \sum_{j=1}^{N} a_{ij}\left(z_{j} + \Delta z_{j}\right)\right)\),
-where \(j\) indexes the same individuals as \(i\), and \(a_{ij}\)
-defines the effect \(j\) has on the contribution of \(i\) to \(\Omega\).
-In evolutionary terms, \(a_{ij}\) defines how \(j\) modulates the
-fitness of \(i\) (\(w_{i}\)). In ecological terms, \(a_{ij}\) defines
-how \(j\) modulates the contribution of \(i\) to total population
-growth. If we define \(\alpha\) as the summed effect of all individuals
-on a focal \(i\) (i.e., \(\alpha = \sum_{j=1}^{N} a_{ij}\)), then we can
-partition ecological change into density-independent and
-density-dependent effects and derive a discrete time logistic growth
-equation. We save these derivations for a future investigations focused
-on evolutionary and community ecology.
+2), or any social effects on evolutionary change (see Box 1). Both of
+these effects can be made explicit by specifying how other individuals
+in a population affect birth and death of a focal individual. Assuming
+that effects of individuals are additive, independent, and identical, we
+can derive well-established logistic growth equations and Hamilton's
+rule from eqn 1 (Supporting Information S1).
 
@@
 eqn 1 defines \(\Omega\) as total ecosystem function contributed by a
-focal population (e.g., biomass production, decomposition,
-photosynthetic rate). It is therefore possible to investigate
-ecological, evolutionary, and ecosystem function change from the same
-shared framework.
+focal population. It is therefore possible to investigate ecological,
+evolutionary, and ecosystem function change from the same shared
+framework.
 
@@
 \bibitem[\citeproctext]{ref-Baravalle2024}
-Baravalle, L., A. Roffé, V. J. Luque, and S. Ginnobili. 2024.
+Baravalle, L., A. Roffé, V. J. Luque, and S. Ginnobili. 2025.
 \href{https://doi.org/10.1007/s13752-024-00482-4}{The value of price}.
-Biological Theory.
+Pages 12--24 \emph{in}Biological Theory (Vol. 20).
 
@@
 Sinauer Associates Incorporated.
-
-\bibitem[\citeproctext]{ref-Cornelissen2011}
-Cornelissen, J. H. C., F. Sibma, R. S. P. V. Logtestijn, R. A. Broekman,
-and K. Thompson. 2011.
-\href{https://doi.org/10.1111/j.1365-2435.2010.01765.x}{Leaf pH as a
-plant trait: Species-driven rather than soil-driven variation}.
-Functional Ecology 25:449--455.
 
@@
 
+\bibitem[\citeproctext]{ref-Lande1976}
+Lande, R. 1976. Natural selection and random genetic drift in phenotypic
+evolution. Evolution 30:314--334.
+
 \bibitem[\citeproctext]{ref-Lande1982}
-Lande, R. 1982. \href{https://doi.org/10.2307/1936778}{{A quantitative
+---------. 1982. \href{https://doi.org/10.2307/1936778}{{A quantitative
 genetic theory of life history evolution}}. Ecology 63:607--615.
@@
 
-\bibitem[\citeproctext]{ref-Mu2018}
-Mu, J., Q. Wu, Y. Yang, M. Huang, and C. M. Grozinger. 2018.
-\href{https://doi.org/10.1111/oik.04711}{Plant reproductive strategies
-vary under low and high pollinator densities}. Oikos 127:1081--1094.
-
 \bibitem[\citeproctext]{ref-Nagel1961}

### 2025-06-08 Supporting_Information_1.Rmd
Supporting informatin

---
title: "Foundations of community ecology: Supporting Information 1"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{fancyhdr}
bibliography: refs.bib
link-citations: yes
linkcolor: blue
biblio-style: apalike
---

We start with the fundamental equation that provides a complete description of eco-evolutionary change in any system,

$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
\tag{1}
$$

In the main text, we derived both the Price equation and the birth-death model from the above.
Here we integrate interactions between individuals and demonstrate how this affects population and evolutionary dynamics simultaneously.
We can do this in a way that clarifies model assumptions by allowing an individual $j$ to modulate the birth or death of the focal individual $i$.
We will first illustrate how this leads to two separate discrete time models of density depedent population growth, then illustrate how the same starting point can be used to recover Hamilton's rule in evolutionary ecology.

Density-dependent population growth
===========================================

There are two potential ways to model the incorporation of density dependence into population growth.
First note that here we set $\Omega = N_{t+1}$, and $z_{i} = 1$ and $\Delta z_{i} = 0$ for all individuals.
We can define $r_{i} = \beta_{i} - \delta_{i}$ as an individual growth rate for $i$ [@Lion2018; @Lion2023].
In this case,

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i} + 1\right)
\tag{S1}
$$

Mathematically, the most general approach here would be to define individual growth as a function of the entire system $\mathbf{E}$, $r_{i}(\mathbf{E})$, where $\mathbf{E}$ is a vector with elements including any parameters potentially relevant to $r_{i}$.
Taking this approach would recover a version of eqn 2 in @Lion2018 and permit any relationship between the system and a focal individual's growth.
Focusing on effects of other individuals ($j$) and assuming that the effects of these individuals are additive, let $a_{ij|\cdot}$ be the effect of individual $j$ on the growth rate attributable to $i$ conditioned on all other individuals within the population such that $r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)$ defines the realised growth rate of $i$,

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij|\cdot} \right) + 1\right).
\tag{S2}
$$

Assuming that individual effects of $j$ on $i$ are also independent, we can remove the condition,

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij} \right) + 1\right).$$

Further assuming that all individuals have the same per capita effect such that $\alpha = a_{ij}$ for any $i$ and $j$ pair (as might be reasonable given resource competition in a well-mixed population),

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \alpha N_{t} \right) + 1\right).$$

If $r_{i}$ values are identical,

$$N_{t+1} = N_{t} + r N_{t}\left(1 - \alpha N_{t} \right).
\tag{S3}
$$

Equation S3 therefore recovers a classic version of a discrete time logistic growth by making assumptions from an exact model of eco-evolutionary change.
An alternative approach would be to define model the effects $e_{ij}$ an individual $j$ on the fitness of $i$ ($w_{i}$), thereby replacing eqn S1 with $N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}$ and replacing eqn S2 with,

$$N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}e_{ij|\cdot} \right).
$$

By making the same assumptions of additivity, independence, and identical effects such that $e_{ij} = \epsilon$ for all $j$ on $i$, and assuming fitness is equal ($w_{i} = w$), we can derive,

$$N_{t+1} = wN_{t}(1 - \epsilon N_{t}).
\tag{S4}
$$

Multi-level selection
=============================

We can recover multi-level selection from our eqn 1. 
Here we derive the original form of the multilevel @Price1972 equation as appears in eqn 3.1 of @Lehtonen2020a.
Individuals $i$ belong to one of $K$ total groups where $j$ indexes $K$ and groups do not overlap. 
The size of group $j$ is denoted as $N_{j}$.
Equation S5 below uses summations to partition how individuals within each group contribute to $\Omega$,

$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}\left(\beta_{j,i} - \delta_{j,i} + 1 \right)\left(z_{j, i} + \Delta z_{j, i}\right).
\tag{S5}
$$

In S5, indices $\beta_{j, i}$, $\delta_{j, i}$, and $z_{j, i}$ identify individual $i$ in group $j$.
For simplicity, we set $w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1$ and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias),

$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}w_{j,i}z_{j, i}.$$

For ease of presentation, with no loss of generality, we assume all group sizes are equal with a group size of $N_{j} = n$ for all $j$.
If group sizes differ, then weighted expectations and covariances are instead needed [@Lehtonen2020a].
Given equal group sizes, the total number of individuals ($N$) equals $K \times n$, and,

$$\frac{\Omega}{K n} = \left(\frac{1}{K}\right)\left(\frac{1}{n}\right)\sum_{j=1}^{K}\sum_{i=1}^{n}w_{j,i}z_{j, i}.$$

Rearranging,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\frac{1}{n}\sum_{i=1}^{n}w_{j,i}z_{j, i}.$$

The inner summation can be rewritten as an expectation,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i} z_{j,i}\right).$$

As in the main text, we note $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$, therefore,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(w_{j,i}\right)E\left(z_{j,i}\right).$$

We can separate the summation for each term,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i}\right)E\left(z_{j,i}\right).$$

Using the notation $\bar{w}_{j} = E(w_{j,i})$ and $\bar{z}_{j} = E(z_{j,i})$,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).$$

Similarly, defining $Cov_{j}(w_{j}, z_{j})$ as the covariance between $w_{j,i}$ and $z_{j, i}$ for group $j$,

$$\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).$$

We can rearrange the second term on the right-hand side ($\bar{\bar{w}}$ indicates grand mean over all groups),

$$\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.$$

We can set $\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}$, so,

$$\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right).$$

Because $\Delta \bar{\bar{z}} = \bar{\bar{z'}} - \bar{\bar{z}}$,

$$\bar{\bar{w}}\Delta = Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right).$$

This recovers the  multilevel @Price1972 equation [@Lehtonen2020a] from a starting point of eco-evolutionary change in different groups.

References 
==========================

### 2025-06-08 Supporting_Information_1.tex
Supporting informatin

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Foundations of community ecology: Supporting Information 1},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Foundations of community ecology: Supporting Information 1}
\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
\date{{[}1{]} Department of Biological and Environmental Sciences,
University of Stirling, Stirling, Scotland {[}2{]} Department of
Philosophy, University of Valencia, Valencia, Spain {[}*{]}
Corresponding author:
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
{[}a{]} Equal contribution}

\begin{document}
\maketitle

We start with the fundamental equation that provides a complete
description of eco-evolutionary change in any system,

\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
\tag{1}
\]

In the main text, we derived both the Price equation and the birth-death
model from the above. Here we integrate interactions between individuals
and demonstrate how this affects population and evolutionary dynamics
simultaneously. We can do this in a way that clarifies model assumptions
by allowing an individual \(j\) to modulate the birth or death of the
focal individual \(i\). We will first illustrate how this leads to two
separate discrete time models of density depedent population growth,
then illustrate how the same starting point can be used to recover
Hamilton's rule in evolutionary ecology.

\section{Density-dependent population
growth}\label{density-dependent-population-growth}

There are two potential ways to model the incorporation of density
dependence into population growth. First note that here we set
\(\Omega = N_{t+1}\), and \(z_{i} = 1\) and \(\Delta z_{i} = 0\) for all
individuals. We can define \(r_{i} = \beta_{i} - \delta_{i}\) as an
individual growth rate for \(i\) (\citeproc{ref-Lion2018}{Lion 2018};
\citeproc{ref-Lion2023}{Lion, Sasaki, and Boots 2023}). In this case,

\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i} + 1\right)
\tag{S1}
\]

Mathematically, the most general approach here would be to define
individual growth as a function of the entire system \(\mathbf{E}\),
\(r_{i}(\mathbf{E})\), where \(\mathbf{E}\) is a vector with elements
including any parameters potentially relevant to \(r_{i}\). Taking this
approach would recover a version of eqn 2 in Lion
(\citeproc{ref-Lion2018}{2018}) and permit any relationship between the
system and a focal individual's growth. Focusing on effects of other
individuals (\(j\)) and assuming that the effects of these individuals
are additive, let \(a_{ij|\cdot}\) be the effect of individual \(j\) on
the growth rate attributable to \(i\) conditioned on all other
individuals within the population such that
\(r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)\) defines the
realised growth rate of \(i\),

\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij|\cdot} \right) + 1\right).
\tag{S2}
\]

Assuming that individual effects of \(j\) on \(i\) are also independent,
we can remove the condition,

\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij} \right) + 1\right).\]

Further assuming that all individuals have the same per capita effect
such that \(\alpha = a_{ij}\) for any \(i\) and \(j\) pair (as might be
reasonable given resource competition in a well-mixed population),

\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \alpha N_{t} \right) + 1\right).\]

If \(r_{i}\) values are identical,

\[N_{t+1} = N_{t} + r N_{t}\left(1 - \alpha N_{t} \right).
\tag{S3}
\]

Equation S3 therefore recovers a classic version of a discrete time
logistic growth by making assumptions from an exact model of
eco-evolutionary change. An alternative approach would be to define
model the effects \(e_{ij}\) an individual \(j\) on the fitness of \(i\)
(\(w_{i}\)), thereby replacing eqn S1 with
\(N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\) and replacing eqn S2 with,

\[N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}e_{ij|\cdot} \right).
\]

By making the same assumptions of additivity, independence, and
identical effects such that \(e_{ij} = \epsilon\) for all \(j\) on
\(i\), and assuming fitness is equal (\(w_{i} = w\)), we can derive,

\[N_{t+1} = wN_{t}(1 - \epsilon N_{t}).
\tag{S4}
\]

\section{Multi-level selection}\label{multi-level-selection}

We can recover multi-level selection from our eqn 1. Here we derive the
original form of the multilevel Price (\citeproc{ref-Price1972}{1972})
equation as appears in eqn 3.1 of Lehtonen
(\citeproc{ref-Lehtonen2020a}{2020}). Individuals \(i\) belong to one of
\(K\) total groups where \(j\) indexes \(K\) and groups do not overlap.
The size of group \(j\) is denoted as \(N_{j}\). Equation S5 below uses
summations to partition how individuals within each group contribute to
\(\Omega\),

\[\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}\left(\beta_{j,i} - \delta_{j,i} + 1 \right)\left(z_{j, i} + \Delta z_{j, i}\right).
\tag{S5}
\]

In S5, indices \(\beta_{j, i}\), \(\delta_{j, i}\), and \(z_{j, i}\)
identify individual \(i\) in group \(j\). For simplicity, we set
\(w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1\) and let
\(\Delta z_{j, i} = 0\) (i.e., no transmission bias),

\[\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}w_{j,i}z_{j, i}.\]

For ease of presentation, with no loss of generality, we assume all
group sizes are equal with a group size of \(N_{j} = n\) for all \(j\).
If group sizes differ, then weighted expectations and covariances are
instead needed (\citeproc{ref-Lehtonen2020a}{Lehtonen 2020}). Given
equal group sizes, the total number of individuals (\(N\)) equals
\(K \times n\), and,

\[\frac{\Omega}{K n} = \left(\frac{1}{K}\right)\left(\frac{1}{n}\right)\sum_{j=1}^{K}\sum_{i=1}^{n}w_{j,i}z_{j, i}.\]

Rearranging,

\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\frac{1}{n}\sum_{i=1}^{n}w_{j,i}z_{j, i}.\]

The inner summation can be rewritten as an expectation,

\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i} z_{j,i}\right).\]

As in the main text, we note
\(\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)\),
therefore,

\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(w_{j,i}\right)E\left(z_{j,i}\right).\]

We can separate the summation for each term,

\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i}\right)E\left(z_{j,i}\right).\]

Using the notation \(\bar{w}_{j} = E(w_{j,i})\) and
\(\bar{z}_{j} = E(z_{j,i})\),

\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).\]

Similarly, defining \(Cov_{j}(w_{j}, z_{j})\) as the covariance between
\(w_{j,i}\) and \(z_{j, i}\) for group \(j\),

\[\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).\]

We can rearrange the second term on the right-hand side
(\(\bar{\bar{w}}\) indicates grand mean over all groups),

\[\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.\]

We can set \(\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}\), so,

\[\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right).\]

Because \(\Delta \bar{\bar{z}} = \bar{\bar{z'}} - \bar{\bar{z}}\),

\[\bar{\bar{w}}\Delta = Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right).\]

This recovers the multilevel Price (\citeproc{ref-Price1972}{1972})
equation (\citeproc{ref-Lehtonen2020a}{Lehtonen 2020}) from a starting
point of eco-evolutionary change in different groups.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Lehtonen2020a}
Lehtonen, Jussi. 2020. {``The Price Equation and the Unity of Social
Evolution Theory.''} \emph{Philosophical Transactions of the Royal
Society B: Biological Sciences} 375: 20190362.
\url{https://doi.org/10.1098/rstb.2019.0362}.

\bibitem[\citeproctext]{ref-Lion2018}
Lion, Sébastien. 2018. {``{Theoretical approaches in evolutionary
ecology: environmental feedback as a unifying perspective}.''}
\emph{American Naturalist} 191 (1).
\url{https://doi.org/10.1086/694865}.

\bibitem[\citeproctext]{ref-Lion2023}
Lion, Sébastien, Akira Sasaki, and Mike Boots. 2023. {``Extending
Eco-Evolutionary Theory with Oligomorphic Dynamics.''} \emph{Ecology
Letters} 26 (September): S22--46.
\url{https://doi.org/10.1111/ele.14183}.

\bibitem[\citeproctext]{ref-Price1972}
Price, George R. 1972. {``{Extension of covariance selection
mathematics}.''} \emph{Annals of Human Genetics} 35 (4): 485--90.
\url{https://doi.org/10.1111/j.1469-1809.1957.tb01874.x}.

\end{CSLReferences}

\end{document}

### 2025-06-09 Supporting_Information_1.Rmd
Some revisions that now hit three sections responding to reviewers

@@
 
-We start with the fundamental equation that provides a complete description of eco-evolutionary change in any system,
+This supporting information demonstrates how to derive well-established models in population ecology and evolutionary biology from equation 1 in the main text,
 
@@
 In the main text, we derived both the Price equation and the birth-death model from the above.
-Here we integrate interactions between individuals and demonstrate how this affects population and evolutionary dynamics simultaneously.
+Here we integrate interactions between individuals to recover density-dependent population growth, and we integrate groups within the population to recover multi-level selection.
+Finally, we integrate both to model a system in which multi-level selection and density dependent population change occur simultaneously.
 We can do this in a way that clarifies model assumptions by allowing an individual $j$ to modulate the birth or death of the focal individual $i$.
-We will first illustrate how this leads to two separate discrete time models of density depedent population growth, then illustrate how the same starting point can be used to recover Hamilton's rule in evolutionary ecology.
 
@@
 There are two potential ways to model the incorporation of density dependence into population growth.
-First note that here we set $\Omega = N_{t+1}$, and $z_{i} = 1$ and $\Delta z_{i} = 0$ for all individuals.
+We start with what is likely the most familiar model focusing on individual growth rate $r_{i}$, then use a slightly different model focusing on fitness $w_{i}$. 
+First note that here we set $\Omega = N_{t+1}$, and $z_{i} = 1$ and $\Delta z_{i} = 0$ for all individuals as in the main text.
 We can define $r_{i} = \beta_{i} - \delta_{i}$ as an individual growth rate for $i$ [@Lion2018; @Lion2023].
@@
 Taking this approach would recover a version of eqn 2 in @Lion2018 and permit any relationship between the system and a focal individual's growth.
-Focusing on effects of other individuals ($j$) and assuming that the effects of these individuals are additive, let $a_{ij|\cdot}$ be the effect of individual $j$ on the growth rate attributable to $i$ conditioned on all other individuals within the population such that $r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)$ defines the realised growth rate of $i$,
+Limiting our focus to the effects of other individuals ($j$) and assuming that the effects of these individuals are additive, let $a_{ij|\cdot}$ be the effect of individual $j$ on the growth rate attributable to $i$ conditioned on all other individuals within the population such that $r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)$ defines the realised growth rate of $i$,
 
@@
 
-Further assuming that all individuals have the same per capita effect such that $\alpha = a_{ij}$ for any $i$ and $j$ pair (as might be reasonable given resource competition in a well-mixed population),
+Further assuming that all individuals have the same per capita effect such that $a = a_{ij}$ for any $i$ and $j$ pair (as might be reasonable given resource competition in a well-mixed population),
 
-$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \alpha N_{t} \right) + 1\right).$$
+$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - a N_{t} \right) + 1\right).$$
 
@@
 
-$$N_{t+1} = N_{t} + r N_{t}\left(1 - \alpha N_{t} \right).
+$$N_{t+1} = N_{t} + r N_{t}\left(1 - a N_{t} \right).
 \tag{S3}
@@
 Equation S3 therefore recovers a classic version of a discrete time logistic growth by making assumptions from an exact model of eco-evolutionary change.
-An alternative approach would be to define model the effects $e_{ij}$ an individual $j$ on the fitness of $i$ ($w_{i}$), thereby replacing eqn S1 with $N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}$ and replacing eqn S2 with,
 
-$$N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}e_{ij|\cdot} \right).
+An alternative approach would be to define model the effects of an individual $j$ on the fitness of $i$ ($w_{i}$), thereby replacing eqn S1 with $N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}$ and replacing eqn S2 with,
+
+$$N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}\alpha_{ij|\cdot} \right).
 $$
 
-By making the same assumptions of additivity, independence, and identical effects such that $e_{ij} = \epsilon$ for all $j$ on $i$, and assuming fitness is equal ($w_{i} = w$), we can derive,
+Note that we have used $\alpha_{ij}$ to represent the effect of $j$ on the fitness of $i$ for clarity in the sections below.
+By making the same assumptions of additivity, independence, and identical effects such that $\alpha = \alpha_{ij}$ for all $j$ on $i$, and assuming fitness is equal ($w_{i} = \bar{w}$), we can derive,
 
-$$N_{t+1} = wN_{t}(1 - \epsilon N_{t}).
+$$N_{t+1} = \bar{w}N_{t}(1 - \alpha N_{t}).
 \tag{S4}
 $$
+
+This is an alternative way to express logistic growth.
 
@@
 We can recover multi-level selection from our eqn 1. 
-Here we derive the original form of the multilevel @Price1972 equation as appears in eqn 3.1 of @Lehtonen2020a.
-Individuals $i$ belong to one of $K$ total groups where $j$ indexes $K$ and groups do not overlap. 
+Here we derive the original form of the multi-level @Price1972 equation as it appears in eqn 3.1 of @Lehtonen2020a.
+Individuals belong to one of $K$ total groups where $j$ indexes $K$ groups and $i$ indexes individuals. 
+Individuals do not overlap in group membership. 
 The size of group $j$ is denoted as $N_{j}$.
@@
 
-The inner summation can be rewritten as an expectation,
+The inner summation can be rewritten as an expectation for group $j$,
 
-$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i} z_{j,i}\right).$$
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i} z_{j,i}\right).$$
 
-As in the main text, we note $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$, therefore,
+As in the main text, we note $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
+Defining $Cov_{j}(w_{j}, z_{j})$ as the covariance between $w_{j,i}$ and $z_{j, i}$ for group $j$,
 
-$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(w_{j,i}\right)E\left(z_{j,i}\right).$$
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).$$
 
@@
 
-$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i}\right)E\left(z_{j,i}\right).$$
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).$$
 
-Using the notation $\bar{w}_{j} = E(w_{j,i})$ and $\bar{z}_{j} = E(z_{j,i})$,
+Using the notation $\bar{w}_{j} = E_{j}(w_{j,i})$ and $\bar{z}_{j} = E_{j}(z_{j,i})$ to indicate the expectation in group $j$,
 
-$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).$$
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).$$
 
-Similarly, defining $Cov_{j}(w_{j}, z_{j})$ as the covariance between $w_{j,i}$ and $z_{j, i}$ for group $j$,
+We can also rewrite the first term on the right-hand side as an expectation,
 
-$$\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).$$
+$$\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).$$
 
@@
 
-$$\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.$$
+$$\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.$$
 
-We can set $\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}$, so,
+As in the main text, note that $Kn\bar{\bar{w}}$ accounts for differences in total population size from $t$ to $t+1$, with $\bar{\bar{w}}$ being mean fitness across all groups.
+We can therefore set $\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}$, so,
 
-$$\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right).$$
+$$\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right).$$
 
@@
 
-$$\bar{\bar{w}}\Delta = Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right).$$
+$$\bar{\bar{w}}\Delta = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
+\tag{S6}
+$$
 
-This recovers the  multilevel @Price1972 equation [@Lehtonen2020a] from a starting point of eco-evolutionary change in different groups.
+This recovers the multi-level @Price1972 equation [@Lehtonen2020a] from a starting point of eco-evolutionary change in different groups.
+This starting point can be found in @Lehtonen2016 B2.I, which then derives a multi-level selection version of Hamilton's rule predicting the evolution of altruism.
+
+Integration of ecology and evolution
+========================================
+
+For simplicity, we now focus on showing an integration between ecology and evolution using a population with no multi-level selection and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias).
+As above in the section on density-dependent population growth, we define $w_{i} = \beta_{i} - \delta_{i} + 1$ and use $\alpha_{i,j}$ to represent the effect of $j$ on the fitness of $i$.
+Our starting equation is therefore,
+
+$$\Omega = \sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S7}
+$$
+
+We have already demonstrated that if we assume all individuals have the same effect on a focal individual such that $\alpha = \alpha_{i,j}$ for all $i$ and $j$ pairs, we can recover equation S4 when $z_{i} = 1$ and $\Omega$ is therefore interpreted as the count of entities,
+
+$$N_{t+1} = \bar{w} N_{t} \left(1 - \alpha N_{t}\right).$$
+
+We now start from S7 to derive $\Delta \bar{z}$.
+The objective is to use our definition of eco-evolutionary change to simultaneously recover how interactions between individuals affect population change and evolutionary change.
+
+We start by dividing both sides of S7 by $N$,
+
+$$\frac{\Omega}{N} = \frac{1}{N}\sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S8}
+$$
+
+We can express the right-hand side of eqn S8 as an expectation,
+
+$$\frac{\Omega}{N} = \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}\right).
+\tag{S9}
+$$
+
+We can rewrite the right-hand side in terms of covariances,
+
+$$\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)\right)\mathrm{E}\left(z_{i}\right).
+\tag{S9}
+$$
+
+The expectations in the second term on the right-hand side can be replaced with overbars to represent the mean,
+
+$$\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i}. 
+\tag{S9}
+$$
+
+In the main text, we noted that $\Omega = N\bar{w}\bar{z}'$.
+Here mean fitness incorporates individual interactions, therefore,
+
+$$\Omega = N\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}'.
+\tag{S10}
+$$
+
+We can therefore rewrite S9,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}' -  \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right).
+\tag{S11}
+$$
+
+Noting again $\Delta \bar{z} = \bar{z}' - \bar{z}$,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
+\tag{S11}
+$$
+
+We can rewrite the left-hand side of S11,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i} - w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S12}
+$$
+
+The covariance term can be split without any additional assumptions,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S13}
+$$
+
+If we are able to further assume that $w_{i}$ and the summation over $\alpha_{i,j}$ are independent, then we could rewrite,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right)\bar{w_{i}}. 
+\tag{S14}
+$$
+
+Partitioning fitness into different components with the Price equation is commonplace.
+But this derivation highlights, e.g., the ecological and evolutionary relationship between nonsocial and social components of fitness, and population size.
+For example, the second term on the right-hand side of S13 shows the covariance between the sum of social interactions and a trait.
+When traits covary with the interaction between sociality and fitness, they will have a stronger effect on trait change.
+The magnitude of this second term will also increase with $N$, which reflects the stronger effect of the interaction between sociality and fitness when there are more individuals interacting with a focal individual.
+
+<!---
+
+We can start again with $K$ groups and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias).
+But for simplicity, now consider the case in which $\delta_{j,i} = 1$ (complete mortality), and consider $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K}\sum_{q=1}^{n} \alpha^{j, i}_{p,q})$, where $\alpha^{j, i}_{p,q}$ is the effect that individual $q$ from group $p$ has on a focal individual $i$ from group $j$.
+To keep the model tractable, we assume that all individuals within a group $p$ have identical effects on any focal individual in the group $j$, such that $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n)$
+
+$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{n}\beta_{j, i}\left(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}.
+\tag{S7}
+$$
+
+Our objective here is to start from the above equation and recover population growth and multi-level selection given this parameterisation focused on individual birth as affected by other individuals in the population.
+If we set $z_{j, i} = 1$ as we have done for population ecology, we can recover logistic growth,
+
+$$N_{t+1} = \sum_{j=1}^{K}\sum_{i=1}^{n}\beta_{j, i}\left(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n\right).$$
+
+If we ignore groups and note $N_{t} = nK$, and we assume that the per capita effect of every individual from every group on a focal individual is equal (i.e., $\alpha = \alpha^{j}_{p}$ for all groups),
+
+$$N_{t+1} = N_{t}\beta_{j, i}\left(1 - \alpha N_{t}\right).$$
+
+We therefore derive the model for population change.
+
+We now use the same starting point to carry the effects of individuals on fitness through a model of multi-level selection.
+Rearranging S7,
+
+$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{n}\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}.$$
+
+As we have done previously, we can multiply both sides of the equation by $(1/K)(1/n)$,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\frac{1}{n}\sum_{i=1}^{n}\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}.$$
+
+We can rewrite the inner summation as an expectation for group $j$,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}E\left(\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}\right).$$
+
+We can rewrite the product within the expectation as a covariance plus a product of the expectations and split the covariance (note that $Cov(X-Y, Z) = Cov(X, Z) - Cov(Y, Z)$),
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) + E\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right) E\left(z_{j, i}\right).$$
+
+All terms are within the summation, which can therefore be separated,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}E\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right) E\left(z_{j, i}\right).$$
+
+Note that because $E(X - Y) \times E(Z) = E(X)E(Z) - E(Y)E(Z)$, the third term on the right-hand side can be split,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ \frac{1}{K}\sum_{j=1}^{K}E\left(\beta_{j, i}\right)E\left(z_{j, i}\right) - E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right).
+\end{align*}
+
+We can rewrite this third term as an expectation,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ E\left(E\left(\beta_{j, i}\right)E\left(z_{j, i}\right) - E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+Because of the linearity of expectation, we can again separate the third term on the right-hand side of the equation,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ E\left(E\left(\beta_{j, i}\right)E\left(z_{j, i}\right)\right) - E\left(E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+For convenience, we substitute $E(\beta_{j,i}) = \bar{\beta}_{j,i}$ and $E(z_{j,i}) = \bar{z}_{j,i}$,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ E\left(\bar{\beta}_{j, i} \bar{z}_{j, i}\right) - E\left(E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+We can split the third term above into a covariance and product of means,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right)\\ 
+                   &+ Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) + \bar{\bar{\beta}}_{j,i}\bar{\bar{z}}_{j,i} - E\left(E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+We can then do the same for the last term on the right-hand side,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) + \bar{\bar{\beta}}_{j,i}\bar{\bar{z}}_{j,i} - Cov(\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}, \bar{z}_{j,i}) - \overline{\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}}\times\bar{\bar{z}}_{j,i}.
+\end{align*}
+
+As in the previous section of this supporting information, we note that $\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}$.
+In this case, fitness $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n)$
+
+\begin{align*}
+\bar{\bar{w}}\Delta\bar{\bar{z}} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\ 
+                                 &+ Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) - Cov(\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}, \bar{z}_{j,i}).
+\end{align*}
+
+And finally,
+
+$$\bar{\bar{w}}\Delta\bar{\bar{z}} = E\left(Cov\left(\beta_{j}, z_{j}\right)\right) - E\left(Cov\left(\beta_{j}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right)\right) + Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) - Cov(\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}, \bar{z}_{j,i}).$$
+
+This seems intuitive in that it just adds two terms to the derivation of multi-level selection.
+And given how we have defined $\alpha$, this should relate the evolutoinary change in some trait $z$ to interactions, specifically density dependent decreases in fitness in this case.
+Note that if $\alpha = 0$, the second and fourth terms drop out, as they should because the covariance with the trait is zero.
+
+--->
 

### 2025-06-09 Supporting_Information_1.tex
Some revisions that now hit three sections responding to reviewers

@@
 
-We start with the fundamental equation that provides a complete
-description of eco-evolutionary change in any system,
+This supporting information demonstrates how to derive well-established
+models in population ecology and evolutionary biology from equation 1 in
+the main text,
 
@@
 model from the above. Here we integrate interactions between individuals
-and demonstrate how this affects population and evolutionary dynamics
-simultaneously. We can do this in a way that clarifies model assumptions
-by allowing an individual \(j\) to modulate the birth or death of the
-focal individual \(i\). We will first illustrate how this leads to two
-separate discrete time models of density depedent population growth,
-then illustrate how the same starting point can be used to recover
-Hamilton's rule in evolutionary ecology.
+to recover density-dependent population growth, and we integrate groups
+within the population to recover multi-level selection. Finally, we
+integrate both to model a system in which multi-level selection and
+density dependent population change occur simultaneously. We can do this
+in a way that clarifies model assumptions by allowing an individual
+\(j\) to modulate the birth or death of the focal individual \(i\).
 
@@
 There are two potential ways to model the incorporation of density
-dependence into population growth. First note that here we set
-\(\Omega = N_{t+1}\), and \(z_{i} = 1\) and \(\Delta z_{i} = 0\) for all
-individuals. We can define \(r_{i} = \beta_{i} - \delta_{i}\) as an
-individual growth rate for \(i\) (\citeproc{ref-Lion2018}{Lion 2018};
+dependence into population growth. We start with what is likely the most
+familiar model focusing on individual growth rate \(r_{i}\), then use a
+slightly different model focusing on fitness \(w_{i}\). First note that
+here we set \(\Omega = N_{t+1}\), and \(z_{i} = 1\) and
+\(\Delta z_{i} = 0\) for all individuals as in the main text. We can
+define \(r_{i} = \beta_{i} - \delta_{i}\) as an individual growth rate
+for \(i\) (\citeproc{ref-Lion2018}{Lion 2018};
 \citeproc{ref-Lion2023}{Lion, Sasaki, and Boots 2023}). In this case,
@@
 (\citeproc{ref-Lion2018}{2018}) and permit any relationship between the
-system and a focal individual's growth. Focusing on effects of other
-individuals (\(j\)) and assuming that the effects of these individuals
-are additive, let \(a_{ij|\cdot}\) be the effect of individual \(j\) on
-the growth rate attributable to \(i\) conditioned on all other
-individuals within the population such that
+system and a focal individual's growth. Limiting our focus to the
+effects of other individuals (\(j\)) and assuming that the effects of
+these individuals are additive, let \(a_{ij|\cdot}\) be the effect of
+individual \(j\) on the growth rate attributable to \(i\) conditioned on
+all other individuals within the population such that
 \(r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)\) defines the
@@
 Further assuming that all individuals have the same per capita effect
-such that \(\alpha = a_{ij}\) for any \(i\) and \(j\) pair (as might be
+such that \(a = a_{ij}\) for any \(i\) and \(j\) pair (as might be
 reasonable given resource competition in a well-mixed population),
 
-\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \alpha N_{t} \right) + 1\right).\]
+\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - a N_{t} \right) + 1\right).\]
 
@@
 
-\[N_{t+1} = N_{t} + r N_{t}\left(1 - \alpha N_{t} \right).
+\[N_{t+1} = N_{t} + r N_{t}\left(1 - a N_{t} \right).
 \tag{S3}
@@
 logistic growth by making assumptions from an exact model of
-eco-evolutionary change. An alternative approach would be to define
-model the effects \(e_{ij}\) an individual \(j\) on the fitness of \(i\)
-(\(w_{i}\)), thereby replacing eqn S1 with
-\(N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\) and replacing eqn S2 with,
+eco-evolutionary change.
 
-\[N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}e_{ij|\cdot} \right).
+An alternative approach would be to define model the effects of an
+individual \(j\) on the fitness of \(i\) (\(w_{i}\)), thereby replacing
+eqn S1 with \(N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\) and replacing eqn S2
+with,
+
+\[N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}\alpha_{ij|\cdot} \right).
 \]
 
-By making the same assumptions of additivity, independence, and
-identical effects such that \(e_{ij} = \epsilon\) for all \(j\) on
-\(i\), and assuming fitness is equal (\(w_{i} = w\)), we can derive,
+Note that we have used \(\alpha_{ij}\) to represent the effect of \(j\)
+on the fitness of \(i\) for clarity in the sections below. By making the
+same assumptions of additivity, independence, and identical effects such
+that \(\alpha = \alpha_{ij}\) for all \(j\) on \(i\), and assuming
+fitness is equal (\(w_{i} = \bar{w}\)), we can derive,
 
-\[N_{t+1} = wN_{t}(1 - \epsilon N_{t}).
+\[N_{t+1} = \bar{w}N_{t}(1 - \alpha N_{t}).
 \tag{S4}
 \]
+
+This is an alternative way to express logistic growth.
 
@@
 We can recover multi-level selection from our eqn 1. Here we derive the
-original form of the multilevel Price (\citeproc{ref-Price1972}{1972})
-equation as appears in eqn 3.1 of Lehtonen
-(\citeproc{ref-Lehtonen2020a}{2020}). Individuals \(i\) belong to one of
-\(K\) total groups where \(j\) indexes \(K\) and groups do not overlap.
-The size of group \(j\) is denoted as \(N_{j}\). Equation S5 below uses
-summations to partition how individuals within each group contribute to
-\(\Omega\),
+original form of the multi-level Price (\citeproc{ref-Price1972}{1972})
+equation as it appears in eqn 3.1 of Lehtonen
+(\citeproc{ref-Lehtonen2020a}{2020}). Individuals belong to one of \(K\)
+total groups where \(j\) indexes \(K\) groups and \(i\) indexes
+individuals. Individuals do not overlap in group membership. The size of
+group \(j\) is denoted as \(N_{j}\). Equation S5 below uses summations
+to partition how individuals within each group contribute to \(\Omega\),
 
@@
 
-The inner summation can be rewritten as an expectation,
+The inner summation can be rewritten as an expectation for group \(j\),
 
-\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i} z_{j,i}\right).\]
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i} z_{j,i}\right).\]
 
 As in the main text, we note
-\(\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)\),
-therefore,
+\(\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)\).
+Defining \(Cov_{j}(w_{j}, z_{j})\) as the covariance between \(w_{j,i}\)
+and \(z_{j, i}\) for group \(j\),
 
-\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(w_{j,i}\right)E\left(z_{j,i}\right).\]
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).\]
 
@@
 
-\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}E\left(w_{j,i}\right)E\left(z_{j,i}\right).\]
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).\]
 
-Using the notation \(\bar{w}_{j} = E(w_{j,i})\) and
-\(\bar{z}_{j} = E(z_{j,i})\),
+Using the notation \(\bar{w}_{j} = E_{j}(w_{j,i})\) and
+\(\bar{z}_{j} = E_{j}(z_{j,i})\) to indicate the expectation in group
+\(j\),
 
-\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(w_{j,i}, z_{j,i}\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).\]
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).\]
 
-Similarly, defining \(Cov_{j}(w_{j}, z_{j})\) as the covariance between
-\(w_{j,i}\) and \(z_{j, i}\) for group \(j\),
+We can also rewrite the first term on the right-hand side as an
+expectation,
 
-\[\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) + E\left(\bar{w}_{j} \bar{z}_{j} \right).\]
+\[\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).\]
 
@@
 
-\[\frac{\Omega}{K n} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.\]
+\[\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.\]
 
-We can set \(\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}\), so,
+As in the main text, note that \(Kn\bar{\bar{w}}\) accounts for
+differences in total population size from \(t\) to \(t+1\), with
+\(\bar{\bar{w}}\) being mean fitness across all groups. We can therefore
+set \(\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}\), so,
 
-\[\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right) +  Cov\left(\bar{w}_{j}, \bar{z}_{j} \right).\]
+\[\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right).\]
 
@@
 
-\[\bar{\bar{w}}\Delta = Cov\left(\bar{w}_{j}, \bar{z}_{j} \right) + E\left(Cov_{j}\left(w_{j}, z_{j}\right)\right).\]
+\[\bar{\bar{w}}\Delta = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
+\tag{S6}
+\]
 
-This recovers the multilevel Price (\citeproc{ref-Price1972}{1972})
+This recovers the multi-level Price (\citeproc{ref-Price1972}{1972})
 equation (\citeproc{ref-Lehtonen2020a}{Lehtonen 2020}) from a starting
-point of eco-evolutionary change in different groups.
+point of eco-evolutionary change in different groups. This starting
+point can be found in Lehtonen (\citeproc{ref-Lehtonen2016}{2016}) B2.I,
+which then derives a multi-level selection version of Hamilton's rule
+predicting the evolution of altruism.
+
+\section{Integration of ecology and
+evolution}\label{integration-of-ecology-and-evolution}
+
+For simplicity, we now focus on showing an integration between ecology
+and evolution using a population with no multi-level selection and let
+\(\Delta z_{j, i} = 0\) (i.e., no transmission bias). As above in the
+section on density-dependent population growth, we define
+\(w_{i} = \beta_{i} - \delta_{i} + 1\) and use \(\alpha_{i,j}\) to
+represent the effect of \(j\) on the fitness of \(i\). Our starting
+equation is therefore,
+
+\[\Omega = \sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S7}
+\]
+
+We have already demonstrated that if we assume all individuals have the
+same effect on a focal individual such that \(\alpha = \alpha_{i,j}\)
+for all \(i\) and \(j\) pairs, we can recover equation S4 when
+\(z_{i} = 1\) and \(\Omega\) is therefore interpreted as the count of
+entities,
+
+\[N_{t+1} = \bar{w} N_{t} \left(1 - \alpha N_{t}\right).\]
+
+We now start from S7 to derive \(\Delta \bar{z}\). The objective is to
+use our definition of eco-evolutionary change to simultaneously recover
+how interactions between individuals affect population change and
+evolutionary change.
+
+We start by dividing both sides of S7 by \(N\),
+
+\[\frac{\Omega}{N} = \frac{1}{N}\sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S8}
+\]
+
+We can express the right-hand side of eqn S8 as an expectation,
+
+\[\frac{\Omega}{N} = \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}\right).
+\tag{S9}
+\]
+
+We can rewrite the right-hand side in terms of covariances,
+
+\[\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)\right)\mathrm{E}\left(z_{i}\right).
+\tag{S9}
+\]
+
+The expectations in the second term on the right-hand side can be
+replaced with overbars to represent the mean,
+
+\[\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i}. 
+\tag{S9}
+\]
+
+In the main text, we noted that \(\Omega = N\bar{w}\bar{z}'\). Here mean
+fitness incorporates individual interactions, therefore,
+
+\[\Omega = N\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}'.
+\tag{S10}
+\]
+
+We can therefore rewrite S9,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}' -  \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right).
+\tag{S11}
+\]
+
+Noting again \(\Delta \bar{z} = \bar{z}' - \bar{z}\),
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
+\tag{S11}
+\]
+
+We can rewrite the left-hand side of S11,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i} - w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S12}
+\]
+
+The covariance term can be split without any additional assumptions,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S13}
+\]
+
+If we are able to further assume that \(w_{i}\) and the summation over
+\(\alpha_{i,j}\) are independent, then we could rewrite,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right)\bar{w_{i}}. 
+\tag{S14}
+\]
+
+Partitioning fitness into different components with the Price equation
+is commonplace. But this derivation highlights, e.g., the ecological and
+evolutionary relationship between nonsocial and social components of
+fitness, and population size. For example, the second term on the
+right-hand side of S13 shows the covariance between the sum of social
+interactions and a trait. When traits covary with the interaction
+between sociality and fitness, they will have a stronger effect on trait
+change. The magnitude of this second term will also increase with \(N\),
+which reflects the stronger effect of the interaction between sociality
+and fitness when there are more individuals interacting with a focal
+individual.
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-Lehtonen2016}
+Lehtonen, Jussi. 2016. {``{Multilevel selection in kin selection
+language}.''} \emph{Trends in Ecology and Evolution} xx: 1--11.
+\url{https://doi.org/10.1016/j.tree.2016.07.006}.
+
 \bibitem[\citeproctext]{ref-Lehtonen2020a}
-Lehtonen, Jussi. 2020. {``The Price Equation and the Unity of Social
-Evolution Theory.''} \emph{Philosophical Transactions of the Royal
-Society B: Biological Sciences} 375: 20190362.
+---------. 2020. {``The Price Equation and the Unity of Social Evolution
+Theory.''} \emph{Philosophical Transactions of the Royal Society B:
+Biological Sciences} 375: 20190362.
 \url{https://doi.org/10.1098/rstb.2019.0362}.

### 2025-06-09 foundations_of_ecoevo.Rmd
Some more minor updates right before making final edits for Victor

@@
 For example, if mean fitness at $t$ is 2 (i.e., $\bar{w} = 2$), then half as many individuals will have contributed to $\Omega$ in $t + 1$ than would have if $\bar{w} = 1$ (i.e., there are $N$ individuals at $t$ and $2N$ individuals at $t + 1$).
-We therefore need to multiply the mean trait value $\bar{z}'$ (at $t + 1$) by the mean fitness $\bar{w}$ (at $t$) to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$.
+We therefore need to multiply the mean trait value $\bar{z}'$ (at $t + 1$) by the mean fitness $\bar{w}$ (at $t$) to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$ [this is noted in the context of population genetics by @Ewens2014].
 Consequently,
@@
 Both of these effects can be made explicit by specifying how other individuals in a population affect birth and death of a focal individual.
-Assuming that effects of individuals are additive, independent, and identical, we can derive well-established logistic growth equations and Hamilton's rule from eqn 1 (Supporting Information S1).
+Assuming that effects of individuals are additive, independent, and identical, we can derive well-established logistic growth equations and multi-level selection from eqn 1 (Supporting Information S1).
 

### 2025-06-09 foundations_of_ecoevo.tex
Some more minor updates right before making final edits for Victor

@@
 mean fitness \(\bar{w}\) (at \(t\)) to recover the mean contribution of
-the \(N\) individuals at \(t\) to the total \(\Omega\). Consequently,
+the \(N\) individuals at \(t\) to the total \(\Omega\) (this is noted in
+the context of population genetics by \citeproc{ref-Ewens2014}{Ewens
+2014}). Consequently,
 
@@
 that effects of individuals are additive, independent, and identical, we
-can derive well-established logistic growth equations and Hamilton's
-rule from eqn 1 (Supporting Information S1).
+can derive well-established logistic growth equations and multi-level
+selection from eqn 1 (Supporting Information S1).
 
@@
 
+\bibitem[\citeproctext]{ref-Ewens2014}
+Ewens, W. J. 2014.
+\href{https://doi.org/10.1007/s10539-013-9412-0}{Grafen, the price
+equations, fitness maximization, optimisation and the fundamental
+theorem of natural selection}. Biology and Philosophy 29:197--205.
+
 \bibitem[\citeproctext]{ref-Fisher1930}

### 2025-06-14 foundations_of_ecoevo.Rmd
Cite Baravalle 2025

--- 2025-06-08 foundations_of_ecoevo.Rmd
@@
 Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density. 
-The deep interdependence between ecology and evolution is well-established, but much theory in each discipline has been developed in isolation. 
-While recent work has accomplished eco-evolutionary synthesis, a gap remains between the logical foundations of ecology and evolution. 
-We bridge this gap with a new equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived.
-We thereby unify the fundamental equations of population ecology and biological evolution under a general framework. 
-Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness and links this change to ecosystem function. 
-Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.
+The deep interdependence between ecology and evolution is well-established, and recent models have focused on integrating eco-evolutionary dynamics. 
+Nevertheless, a gap remains between the logical foundations of ecology and evolution. 
+We bridge this gap with an equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived.
+We thereby identify an overlooked bridge between the fundamental equations of population ecology and biological evolution. 
+Our unification formally recovers the equivalence between mean population growth rate and evolutionary fitness and links this change to ecosystem function. 
+We outline how our proposed framework can be used to further develop eco-evolutionary theory.
 
@@
 
-Theoretical unification of phenomena is a powerful tool for scientific advancement.
+Theoretical unification is a powerful tool for scientific advancement.
 Such unification has been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions.
@@
 This provides conceptual clarity by demonstrating the logical consistency of different theoretical frameworks within evolutionary biology.
-Our aim here is to propose an equation that extends this conceptual clarity to include population change and thereby provide a formal and exact definition of joint ecological and evolutionary change.
+Our aim here is to propose an equation that extends this conceptual clarity to include population change and thereby provide a formal and exact definition for joint ecological and evolutionary change.
 
@@
 As with evolution, a fundamental equation can exhaustively and exactly define population change.
-Unlike the Price equation, this fundamental equation is perhaps self-evident; population change is simply the addition of births and subtraction of deaths from current population size ($N_{t}$) to recover the new population size ($N_{t+1}$; Box 2).
+Unlike the Price equation, this fundamental equation is perhaps self-evident.
+Population change is simply the addition of individuals minus the removal of individuals from current population size ($N_{t}$), which recovers the new population size ($N_{t+1}$; Box 2).
 By definition, this relationship of $N_{t+1} = N_{t} + Births - Deaths$ applies to any closed population.
-@Turchin2001 argues that general principles are needed to establish a logical foundation for population ecology, and this simple birth and death model and the consequences that logically follow from it (e.g., exponential population growth) is fundamental to population ecology.
+@Turchin2001 argues that general principles are needed to establish a logical foundation for population ecology, and this simple birth and death model and the consequences that logically follow from it (e.g., exponential population growth) are fundamental to population ecology.
 Any unifying definition of joint ecological and evolutionary change must be able to derive both the Price equation and this birth and death model.
 
-The interdependence of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
+The union of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
 Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some theoreticians broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
@@
 The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
-The second type models population demographics as functions of quantitative traits, with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024].
+The second type models population demographics as functions of quantitative traits with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024].
 Both model types can be very general, but like all predictive models, they rely on simplifying assumptions for tractability [@Levins1966; @Luque2016].
@@
 For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory.
-The role of fundamental equations is therefore important for unifying theory [@Luque2021], and we believe that a fundamental equation of eco-evolutionary change is needed.
+The role of fundamental equations is therefore important for unifying theory [@Luque2021], and we believe that a fundamental equation of eco-evolutionary change has been curiously overlooked.
 
 We present an equation from which the fundamental equations of ecology and evolutionary biology can be derived.
-Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics (e.g., average excess) or quantitative genetics (e.g., Breeder's equation) can be derived from the Price equation [@Queller2017].
+Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics (e.g., the average excess) or quantitative genetics (e.g., the Breeder's equation) can be derived from the Price equation [@Queller2017].
 We propose our equation as a formal definition of eco-evolutionary change.
@@
 Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness.
-The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018].
+The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be derived [@Queller2017; @Lehtonen2016; @Lehtonen2018; @Lehtonen2020a].
 This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023].
@@
 
-> **Box 2:** The number of individuals in any closed population ($N$) at any given time ($t + 1$) is determined by the existing number of individuals ($N_{t}$), plus the number of births ($Births$) minus the number of deaths ($Deaths$), $$N_{t+1} = N_{t} + Births - Deaths.$$ This equation is necessarily true for any closed population.
+> **Box 2:** The number of individuals in any closed population ($N$) at any given time ($t + 1$) is determined by the number of individuals at $t$ ($N_{t}$), plus the number of births ($Births$) minus the number of deaths ($Deaths$), $$N_{t+1} = N_{t} + Births - Deaths.$$ This equation is necessarily true for any closed population.
 Despite its simplicity, it is a general equation for defining population change and a starting point for understanding population ecology.
@@
 This inherent underlying tendency towards exponential growth persists even as the complexities of real populations, such as structure, stochasticity, or density-dependent effects are added to population models [@Turchin2001].
-Given the assumption that all individuals in the population are identical, a per capita rate of birth, $Births = bN_{t}$ and death, $Deaths = dN_{t}$ can be defined.
+Given the assumption that all individuals in the population are identical, a per capita rate of birth, $Births = bN_{t}$, and death, $Deaths = dN_{t}$, can be defined.
 Rearranging and defining $\lambda = 1 + b - d$ gives, $N_{t+1} = N_{t}\lambda$.
@@
 
-To fully unify biological evolution and population ecology, we must reconcile the the Price equation (Box 1) with the general equation for population change (Box 2).
-The Price equation is critical for partitioning different components of biological change [@Price1970; @Frank1997; @Gardner2008; @Luque2016; @Lehtonen2018; @Queller2017].
+To unify biological evolution and population ecology, we must reconcile the Price equation (Box 1) with the general equation for population change (Box 2).
+The Price equation is critical for partitioning different components of biological change [@Price1970; @Frank1997; @Gardner2008; @Luque2016; @Lehtonen2018; @Lehtonen2020a; @Queller2017].
 It has also been highly useful for integrating evolutionary theory across disciplines [@Fox2006; @Brantingham2007; @MacCallum2012; @Frank2015; @Godsoe2021; @Ulrich2024].
@@
 This is because the Price equation describes the average change in a population; the frequency of entities is scaled thereby conserving total probability [@Frank2015; @Frank2016].
-But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided in a fundamental equation of ecology and evolution.
+But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided.
 
@@
 
-The foundation of eco-evolutionary change is an interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
-The value of $\Omega$ is a summation of individual characteristics, which takes the same units as $z$.
-From eqn 1, we can derive the most fundamental equations of population ecology (Box 2) and evolutionary biology (Box 1) through an appropriate interpretation of $z$.
-Under some limited interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.
+This foundation of eco-evolutionary change is an interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
+The value of $\Omega$ is a summation of individual characteristics and takes the same units as $z$.
+From eqn 1, we can derive the most fundamental equations of population ecology and evolutionary biology through an appropriate interpretation of $z$.
+Under more limited interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.
 
@@
 In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
-In this case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit; @NewellEtAl2019].
+In this case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
 Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
@@
 
-We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$,
+We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$.
+Summing up $\beta_{i}$, $\delta_{i}$, and current entities tallies up the total number of entities in the next time step,
 
@@
 
-We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [@Turchin2001] (eqn 3).
+We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [eqn 3, @Turchin2001].
 
@@
 For example, if mean fitness at $t$ is 2 (i.e., $\bar{w} = 2$), then half as many individuals will have contributed to $\Omega$ in $t + 1$ than would have if $\bar{w} = 1$ (i.e., there are $N$ individuals at $t$ and $2N$ individuals at $t + 1$).
-We therefore need to multiply the mean trait value $\bar{z}'$ (at $t + 1$) by the mean fitness $\bar{w}$ (at $t$) to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$.
+We therefore need to multiply the mean trait value $\bar{z}'$ (at $t + 1$) by the mean fitness $\bar{w}$ (at $t$) to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$ [this has been applied in the context of population genetics by @Ewens2014].
 Consequently,
@@
 
-We can rearrange eqn 9 to derive the Price equation by expressing covariance as, $\mathrm{Cov}(X,Y) = \mathrm{E}(XY) - \mathrm{E}(X)\mathrm{E}(Y)$, and therefore $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
+We can rearrange eqn 9 to derive the Price equation by expressing covariance as $\mathrm{Cov}(X,Y) = \mathrm{E}(XY) - \mathrm{E}(X)\mathrm{E}(Y)$, and therefore $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
 Substituting into eqn 9,
@@
 In some cases, $\Omega$ could also be intepreted as the total contribution of a population to ecosystem function.
-This will be restricted to cases in which $z$ is an individual characteristic defining an absolute quantity measured at the whole organism level, such as biomass, seed production, carbon capture, flower visits, or nutrient consumption.
-In these cases, the sum across individuals gives a meaningful total quantity for the population.
+This is restricted to cases in which $z$ is a characteristic defining an absolute quantity measured at the whole organism level such as biomass, seed production, carbon capture, flower visits, or nutrient consumption.
+In such cases, the sum across individuals gives a meaningful total quantity for the population.
 When $z$ is instead defined by relative organism-level measurements such wing loading, nutrient ratio, or diet composition, or when $z$ is measured at a level of biological organisation below the organism (e.g., average cell volume or leaf surface area), $\Omega$ does not have a clear population-level interpretation.
@@
 
-> **Box 3:** As a toy example of our framework, consider a population of $N_{t} = 3$ annual plants in which individual fruit mass (kg) is measured, and change is observed over a year.
+> **Box 3:** As a toy example of our framework, consider a population of $N_{t} = 3$ annual plants in which individual total fruit mass (kg) is measured, and change is observed over a year.
 For all plants, $\delta_{i} = 1$, and let plant fecundities be $\beta_{1} = 1$, $\beta_{2} = 1$, and $\beta_{3} = 2$.
 Applying eqn 1 to population change such that $\Omega = N_{t+1}$, $z_{i} = 1$ and $\Delta z_{i} = 0$, $N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4$ (note $\bar{w} = 4/3$, so $N_{t}\bar{w} = 4$).
-Focusing next on the characteristic of total fruit mass, let $z_{1} = 0.8$, $z_{2} = 1.0$, and $z_{3} = 1.5$.
+Focusing next on the characteristic of total fruit mass, let $z_{1} = 0.8$ kg, $z_{2} = 1.0$ kg, and $z_{3} = 1.5$ kg.
 Also let $\Delta z_{i} = 0.1$ for all plants to reflect a change in soil environment from $t$ to $t + 1$.
-In this case, total fruit yield at $t+1$ is $\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2$ kg.
+In this case, population fruit yield at $t+1$ is $\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2$ kg.
 At $t$, mean fruit yield per plant was 1.1 kg, but at $t+1$, mean fruit yield per plant is $1.3$ kg.
@@
 
-A classical sign of scientific progress is the ability to connect disparate theories and models to show how empirical and theoretical models are logical (mathematical) consequences of more fundamental ones [@Nagel1961; @Morrison2000].
-Rather than making simplifying assumptions, as is the approach for most ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
+An important aspect of scientific progress is the ability to connect disparate theories and models to show how specific empirical and theoretical models are logical (mathematical) consequences of more fundamental ones [@Nagel1961; @Morrison2000].
+Rather than making simplifying assumptions, as is the approach for specific ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
 We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
@@
 The Price equation provides a complete and exact description of evolution in any closed evolving system [@Price1970; @Frank2012].
-It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [@Price1970; @Gardner2008; @Luque2016] (e.g., individuals, alleles).
+It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [e.g., individuals, alleles, @Price1970; @Gardner2008; @Luque2016].
 This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation (e.g., group, individual) [@Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
 Because of its abstract nature and lack of any system-specific assumptions, the Price equation is not dynamically sufficient and makes no predictions about what will happen in any particular system [@Gardner2020].
-Its role is not to predict, but to formally and completely define and separate components of evolutionary change. 
-The same is true of the general equation for population change (eqns 2 and 3), at least as we have used it here where it serves to define what population growth means in ecology.
+Its role is not to predict, but to formally and completely define and separate components of evolutionary change [@Baravalle2024]. 
+The same is true of the general equation for population change (eqn 2), at least as we have used it here where it serves to define what population growth means in ecology.
 This equation formally and completely describes population change in terms of births and deaths.
 In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population, but one that is revealed to be necessarily abstract and not dynamically sufficient.
-We anticipate that this will be useful for eco-evolutionary theory in the same way that the Price equation is useful for evolutionary theory: facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.
+We hope that this will be useful for eco-evolutionary theory in a similar way that the Price equation is useful for evolutionary theory: potentially facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.
 
-Our unification recovers the equivalence between the finite rate of increase $\lambda$ (Box 2) [@Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 1).
-The population growth equation $N_{t+1} = N_{t}\lambda$ can therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
-This specific equivalence has been proposed before [@Lande1976], as has the broader relationship between population growth rate and evolutionary fitness [@Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
+Our unification recovers the equivalence between the finite rate of increase $\lambda$ (Box 2) and population mean evolutionary fitness $\bar{w}$ (Box 1).
+The population growth equation $N_{t+1} = N_{t}\lambda$ can always be rewritten as $N_{t+1} = N_{t}\bar{w}$.
+This specific equivalence has been proposed before [e.g., @Lande1976], as has the broader relationship between population growth rate and evolutionary fitness [@Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
 We show this from first principles and clarify the relationship between fitness and population growth.
@@
 Over an individual's lifetime (which, by definition, includes death), fitness is therefore $\beta_{i}$. 
-The rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
-Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)/\bar{w}$ (i.e., Fisher's fundamental theorem) [@Frank1997; @Rice2004; @Queller2017].
+Rates of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
+Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)/\bar{w}$ [i.e., Fisher's fundamental theorem, @Frank1997; @Rice2004; @Queller2017].
 
 Our unification may also help explain, at least partially, some of the success of classical population genetic models. 
-For decades, population genetics (and to some extent quantitative genetics) has been accused of being a reductionist view of evolution, reducing everything to changes in allele frequencies and abstracting away from individuals and their environments (the ecological interactions) [@MacColl2011].
+For decades, population genetics (and to some extent quantitative genetics) has been accused of being a reductionist view of evolution, reducing everything to changes in allele frequencies and abstracting away from individuals and their environments [the ecological interactions, @MacColl2011].
 This has been a line of argumentation by some defenders of the so-called Extended Synthesis [@Pigliucci2009], especially in relation to niche construction [@Odling-smee2003].
-Famously, @Mayr1960 characterized population genetics as a simple input and output of genes, analogous to "the adding of certain beans to a beanbag and the withdrawing of others" (also called "beanbag genetics").
+Famously, @Mayr1960 characterised population genetics as a simple input and output of genes, analogous to "the adding of certain beans to a beanbag and the withdrawing of others" (also called "beanbag genetics").
 Historical critics of population genetics could not articulate a clear explanation for why it works so well despite all of its idealisations and simplifications.
@@
 
-Our fundamental equation is complete and exact.
+We believe our fundamental equation to be complete and exact for any closed population.
 It therefore implicitly includes any effects of density dependence on population growth (see Box 2), or any social effects on evolutionary change (see Box 1).
 Both of these effects can be made explicit by specifying how other individuals in a population affect birth and death of a focal individual.
-Assuming that effects of individuals are additive, independent, and identical, we can derive well-established logistic growth equations and Hamilton's rule from eqn 1 (Supporting Information S1).
+We demonstrate this by deriving more specific models of density dependent population growth and multi-level selection in Supporting Information S1.
+In Supporting Information S1, we also suggest how new theoretical insights might be gained by integrating density-dependent growth and social interactions into our framework.
 

### 2025-06-14 foundations_of_ecoevo.Rmd
Make some edits based on Matt's first read and comments

@@
 This provides conceptual clarity by demonstrating the logical consistency of different theoretical frameworks within evolutionary biology.
-Our aim here is to propose an equation that extends this conceptual clarity to include population change and thereby provide a formal and exact definition for joint ecological and evolutionary change.
+Our aim here is to propose an equation that extends this conceptual clarity to include population size change and thereby provide a formal and exact definition for joint ecological and evolutionary change.
 
-In biological populations, ecological change is caused by the same processes of individual birth and death that cause changes in allele frequencies and phenotypes [@Turchin2001; @Connor2004; @Barfield2011].
+In biological populations, ecological change is caused by the same processes of individual birth and death that cause evolutionary changes in allele frequencies and phenotypes [@Turchin2001; @Connor2004; @Barfield2011].
 As with evolution, a fundamental equation can exhaustively and exactly define population change.
@@
 @Bassar2021 identify two types of eco-evolutionary models that follow from these interpretations.
-The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
+The first type uses separate equations to model population size change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
 The second type models population demographics as functions of quantitative traits with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024].
@@
 We therefore begin with the most fundamental axioms underlying the ecology and evolution of living systems [@Rice2004; @Rice2009].
-In such systems, diversity is discontinuous and can be defined in terms of discrete entities [@Dobzhansky1970].
+In such systems, diversity is discontinuous, in the sense that living systems are composed of discrete entities including individual organisms and groups of organisms [@Dobzhansky1970].
 Our framework is general enough that entities can be anything discrete, but we will focus on each entity $i$ as an individual organism.
@@
 In this definition, the longevity of the individual matters.
-An individual that survives from $t$ to $t + 1$ has a higher fitness than one that dies, even if both have the same reproductive output.
+By definition, an individual that survives from $t$ to $t + 1$ has a higher fitness than one that dies, even if both have the same reproductive output.
 With this definition of fitness (eqn 4), we substitute,
@@
 We are grateful for many conversations with S$\'{e}$bastien Lion, Kelsey Lyberger, Swati Patel, and especially Lynn Govaert, whose questions helped us clarify the relationship between population growth and fitness.
-Brad Duthie would also like to thank Brent Danielson and Stan Harpole.
+Brad Duthie would also like to thank Matt Tinsley, Brent Danielson, and Stan Harpole.
 Victor J. Luque would also like to thank Lorenzo Baravalle, Pau Carazo, Santiago Ginnobili, Manuel Serra, and Ariel Roff$\'{e}$.

### 2025-06-18 response_to_reviewers.md
General response and add checkboxes
Fix the misspelling of Fisher's name
Edits for resubmission of revision to ecology and evolution

@@
+General response to reviewer comments
+================================================
+
 We thank both reviewers for their valuable comments and suggestions.
@@
 
+Line numbers below refer to the new clean draft of our manuscript.
+
 Reviewer 1 comment 1 
@@
 
-We have revised our introduction extensively to address this comment and better communicate the goals of our work.
+We have completely rewritten our introduction (retaining only the first sentence) to address this comment and better communicate the goals of our work.
 
@@
 Reviewer 1 is not correct in stating that these models retrieve the Price equation; the Price equation is not a predictive model but a mathematical identity that formally defines evolutionary change.
-Hence, it must always be possible, at least in theory, to retrieve the Price equation from any evolutionary model.
 These models don't retrieve the Price equation so much as the Price equation is necessarily embedded within them.
@@
 The whole point of introducing the paper the way that we did was to illustrate that we are trying to ask a different question than the ones in those papers.
-We obviously need to try a different approach, and we have done this using extensive revisions in our introduction that address the points made above.
+We obviously need to try a different approach, and we have done this by completely rewriting the introduction to address the points made above.
 
@@
 
-We now state this more clearly in the text (Line XXX).
+We now state this more clearly in the text (L245-246).
 
@@
 Models need to make (simplifying, specific system) assumptions to be predictive (Levins 1966, Frank 2012), and a fundamental equation, by its nature, needs to be free of such assumptions and focus only on what is essential.
-The worthiness of our equation rests entirely on its power to unify fundamental equations of ecology and evolution and provide conceptual clarity, and its limitations are the same as any other fundamental equation.
+The relevance of our equation rests entirely on its power to unify fundamental equations of ecology and evolution and provide conceptual clarity, and its limitations are the same as any other fundamental equation.
 
@@
 
-We have changed the title and reframed the introduction.
+We have changed the title and rewritten the introduction.
 Our goal was to explain the importance of conceptual unification in the sciences (which, in our experience, is often underappreciated in the life sciences) without overselling our present results.
-In our revised introduction, we have attempted to tone down our own contribution while also not losing our central point that this kind of theory is important in science (i.e., *useful*, but because it demonstrates logical consistency and provides conceptual clarity -- not just because you can use it in the same way as a predictive model).
+In our revised introduction and discussion, we have toned down our own contribution while also not losing our central point that this kind of theory is important in science (i.e., *useful*, but because it demonstrates logical consistency and provides conceptual clarity -- not just because you can use it in the same way as a predictive model).
 
@@
 It therefore defines what eco-evolutionary change is (and is not), mathematically, at its most fundamental.
-Verbally, it is the interaction between absolute fitness and some characteristic of an individual.
+Verbally, it is the statistical interaction between absolute fitness and some characteristic of an individual.
+We now state this more explicitly (L141).
 
@@
 
-In our revision, we try to clarify the need for our equation more clearly.
+In our revision, we try to clarify the need for our equation more clearly. Our introduction has been completely rewritten in this effort.
 
@@
 We thank Reviewer 2 for raising this query; it gave us pause to reflect and be even more deliberate about what our equation is describing.
-We now mention in our manuscript why this is not an issue by pointing out that individuals are unitless (LINE).
+We now mention in our manuscript why this is not an issue by pointing out that individuals are unitless (L148-L150).
+We also introduce a new Box 3 (L203-218) walking the reader through a concrete toy example with actual numbers, calculating ecological and evolutionary change, and ecosystem function.
 
@@
 
-We now emphasise more clearly, in words, the interpretation of equation 1 and how it is meaningful even prior to the derivation of population change or the Price equation (LINE XXX).
+We now emphasise more clearly, in words, the interpretation of equation 1 and how it is meaningful even prior to the derivation of population change or the Price equation (L141-142, L14-16).
 
@@
 
-We now remove this from the abstract. In the manuscript, we now clarify that our derivation recovers this relationship, which has been pointed out previously.
+We now remove this from the abstract. In the manuscript, we now clarify that our derivation recovers this relationship, which has been pointed out previously (L255-258).
 
@@
 We are grateful to Reviewer 2 for pointing this out, and we have revised to make our point about ecosystem function clearer with a more concise explanation of our point in our response to Reviewer 2 comment 2.
-We now elaborate on how Omega can be interpreted as a measure of ecosystem function for a population (LINE XXX).
+We now elaborate on how Omega can be interpreted as a measure of ecosystem function for a population in a new section (L190-201), and we include an example in a new Box 3 (L203-218).
 
@@
 
-This is a good suggestion; we have cited this paper now (Line XXX).
+This is a good suggestion. We have rewritten the entire introduction, so this sentence no longer exists, but we cite Smocovitis (1992) now in the first sentence of the manuscript (L25).
 
-Reviewer 2 minor comment 3
+Reviewer 2 minor comment 3 [ ]
 =========================
@@
 =========================
+
+We have rewritten the introduction, and this sentence no longer exists.
+In our revision, we acknowledge the long-recognised union between the two disciplines while also pointing out that the rise of eco-evolutionary models is relatively recent following a recognition that ecology and evolution can occur on similar timescales, with citations (L49-52).
 
@@
 Finally, we also agree with Reviewer 2 that classic evolutionary works (like those developed by Haldane) had ecological ingredients.
-Actually, as we said in our manuscript, our claim is that, contrary to a spread view among evolutionary biologists (see page XXX of our manuscript for references), population genetics has ecology at its core, though not in an obvious way.
-
-We now clarify this historical view (Line XXX)
+Actually, as we said in our manuscript, our claim is that, contrary to a view among evolutionary biologists (see L259-272), population genetics has ecology at its core, though not in an obvious way.
 
@@
 
-Fixed (Line XXX).
+Fixed (Line 95).
 
-Reviewer 2 minor comment 5 
+Reviewer 2 minor comment 5 [ ]
 =========================
@@
 We are reluctant to include too much work still in preparation, or distract from what we believe is the central message of the current manuscript.
-But we have now included some additional derivations in supporting information.
+But we have now included additional derivations in supporting information.
+
+In this new supporting information, we use our eqn 1 to derive (1) density-dependent population growth, (2) multi-level selection (and therefore a version of Hamilton's rule for when altruism is predicted to evolve in a population), and (3) a novel integration of ecology and evolution that demonstrates the relationship between how sociality and fitness (i.e., the extent to which conspecific interactions can muldulate fitness) affect trait change, and how this effect depends on population size -- while simultaneously deriving a classic logistic growth equation.
 

### 2025-06-18 Supporting_Information_1.Rmd
Edits for resubmission of revision to ecology and evolution

@@
 In S5, indices $\beta_{j, i}$, $\delta_{j, i}$, and $z_{j, i}$ identify individual $i$ in group $j$.
-For simplicity, we set $w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1$ and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias),
+We set $w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1$, and for simplicity let $\Delta z_{j, i} = 0$ (i.e., no transmission bias),
 
@@
 
-$$\bar{\bar{w}}\Delta = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
+$$\bar{\bar{w}}\Delta \bar{\bar{z}} = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
 \tag{S6}
@@
 This recovers the multi-level @Price1972 equation [@Lehtonen2020a] from a starting point of eco-evolutionary change in different groups.
-This starting point can be found in @Lehtonen2016 B2.I, which then derives a multi-level selection version of Hamilton's rule predicting the evolution of altruism.
+Equation S6 can be found in @Lehtonen2016 B2.I, who then derives a multi-level selection version of Hamilton's rule predicting the evolution of altruism.
 
@@
 $$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
-\tag{S11}
+\tag{S12}
 $$
 
-We can rewrite the left-hand side of S11,
+We can rewrite the left-hand side of S12,
 
 $$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i} - w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
-\tag{S12}
+\tag{S13}
 $$
@@
 $$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
-\tag{S13}
+\tag{S14}
 $$
@@
 $$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right)\bar{w_{i}}. 
-\tag{S14}
+\tag{S15}
 $$

### 2025-06-18 Supporting_Information_1.tex
Edits for resubmission of revision to ecology and evolution

@@
 In S5, indices \(\beta_{j, i}\), \(\delta_{j, i}\), and \(z_{j, i}\)
-identify individual \(i\) in group \(j\). For simplicity, we set
-\(w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1\) and let
+identify individual \(i\) in group \(j\). We set
+\(w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1\), and for simplicity let
 \(\Delta z_{j, i} = 0\) (i.e., no transmission bias),
@@
 
-\[\bar{\bar{w}}\Delta = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
+\[\bar{\bar{w}}\Delta \bar{\bar{z}} = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
 \tag{S6}
@@
 equation (\citeproc{ref-Lehtonen2020a}{Lehtonen 2020}) from a starting
-point of eco-evolutionary change in different groups. This starting
-point can be found in Lehtonen (\citeproc{ref-Lehtonen2016}{2016}) B2.I,
-which then derives a multi-level selection version of Hamilton's rule
-predicting the evolution of altruism.
+point of eco-evolutionary change in different groups. Equation S6 can be
+found in Lehtonen (\citeproc{ref-Lehtonen2016}{2016}) B2.I, who then
+derives a multi-level selection version of Hamilton's rule predicting
+the evolution of altruism.
 
@@
 \[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
-\tag{S11}
+\tag{S12}
 \]
 
-We can rewrite the left-hand side of S11,
+We can rewrite the left-hand side of S12,
 
 \[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i} - w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
-\tag{S12}
+\tag{S13}
 \]
@@
 \[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
-\tag{S13}
+\tag{S14}
 \]
@@
 \[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right)\bar{w_{i}}. 
-\tag{S14}
+\tag{S15}
 \]

### 2025-06-18 foundations_of_ecoevo_arxiv.Rmd
Edits for resubmission of revision to ecology and evolution

--- 2025-06-14 foundations_of_ecoevo.Rmd
@@
 - \usepackage{lineno}
-- \linenumbers
-- \modulolinenumbers[2]
-linestretch: 2
+linestretch: 1
 bibliography: refs.bib
@@
 Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density. 
-The deep interdependence between ecology and evolution is well-established, and recent models have focused on integrating eco-evolutionary dynamics. 
+The deep interdependence between ecology and evolution is well-established, and recent models focus on integrating eco-evolutionary dynamics to demonstrate how ecological and evolutionary processes interact and feed back upon each other.
 Nevertheless, a gap remains between the logical foundations of ecology and evolution. 
-We bridge this gap with an equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived.
-We thereby identify an overlooked bridge between the fundamental equations of population ecology and biological evolution. 
+Both population ecology and evolution have fundamental equations that define how the size of a population (ecology) and the average characteristic within a population (evolution) change, respectively, over time.
+These fundamental equations are a complete and exact description of change for any closed population, but how they are formally linked remains unclear.
+We link the fundamental equations of population ecology and evolution with an equation that sums how individual characteristics interact with individual fitness in a population.
+From this equation, we derive the fundamental equations of population ecology and evolutionary biology (the Price equation). 
+We thereby identify an overlooked bridge between ecology and biological evolution. 
 Our unification formally recovers the equivalence between mean population growth rate and evolutionary fitness and links this change to ecosystem function. 
-We outline how our proposed framework can be used to further develop eco-evolutionary theory.
+We outline how our framework can be used to further develop eco-evolutionary theory.
 
@@
 Theoretical unification is a powerful tool for scientific advancement.
-Such unification has been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions.
+Such unification has been a major goal in scientific research throughout history [@Smocovitis1992; @Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions.
 In evolutionary biology, the Price equation (Box 1) provides a unifying framework for evolutionary theory by exhaustively and exactly describing evolutionary change for any closed evolving population [@Price1970; @Luque2016; @Lehtonen2020].
@@
 We present an equation from which the fundamental equations of ecology and evolutionary biology can be derived.
-Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics (e.g., the average excess) or quantitative genetics (e.g., the Breeder's equation) can be derived from the Price equation [@Queller2017].
+Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics or quantitative genetics can be derived from the Price equation by restricting the domain of interest [e.g., to allele frequencies in the case of population genetics, or to continuous phenotypes in the case of quantitative genetics, @Queller2017].
 We propose our equation as a formal definition of eco-evolutionary change.
@@
 
-This foundation of eco-evolutionary change is an interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
+This foundation of eco-evolutionary change is a statistical interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
 The value of $\Omega$ is a summation of individual characteristics and takes the same units as $z$.
@@
 To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
-In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
-In this case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
+In other words, we set $z_{i} = 1$ to simply indicate that $i$ is one member of the population.
+In this restricted case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
 Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
@@
 We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$.
-Summing up $\beta_{i}$, $\delta_{i}$, and current entities tallies up the total number of entities in the next time step,
+Summing up $\beta_{i}$, $\delta_{i}$, and current individuals tallies up the total number of individuals in the next time step,
 
@@
 The left-hand side of eqn 7 describes contributions to the sum trait value from the new population at $t + 1$.
-But we cannot treat $\Omega/N$ as the mean of $z$ at $t+1$ ($\bar{z}'$) because we need to weigh $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.
+But we cannot treat $\Omega/N$ as the mean of $z$ at $t+1$ ($\bar{z}'$) because we need to weight $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.
 
@@
 
-From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).
+Equation 10 is the Price equation.
+From eqn 1, which describes fundamental birth and death processes in a population, we can therefore derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).
 
@@
 
-In some cases, $\Omega$ could also be intepreted as the total contribution of a population to ecosystem function.
+In some cases, $\Omega$ could also be interpreted as the total contribution of a population to ecosystem function.
 This is restricted to cases in which $z$ is a characteristic defining an absolute quantity measured at the whole organism level such as biomass, seed production, carbon capture, flower visits, or nutrient consumption.
-In such cases, the sum across individuals gives a meaningful total quantity for the population.
+In such cases, the sum across individuals gives a meaningful total quantity for the population (e.g., the total biomass or seeds produced in the population).
 When $z$ is instead defined by relative organism-level measurements such wing loading, nutrient ratio, or diet composition, or when $z$ is measured at a level of biological organisation below the organism (e.g., average cell volume or leaf surface area), $\Omega$ does not have a clear population-level interpretation.
+In such cases, eqn 1 still defines eco-evolutionary change; the interpretation of $\Omega$ by itself is just not as interesting, biologically.
 
-Box 3 provides a toy example of three plants with different fitnesses and carbon captures.
-We use this toy example to demonstrate a concrete example of our proposed framework.
+Box 3 provides an instructive example of three plants with different fitnesses and carbon captures.
+We use this example to demonstrate a concrete example of our proposed framework.
 
@@
 
-> **Box 3:** As a toy example of our framework, consider a population of $N_{t} = 3$ annual plants in which individual total fruit mass (kg) is measured, and change is observed over a year.
-For all plants, $\delta_{i} = 1$, and let plant fecundities be $\beta_{1} = 1$, $\beta_{2} = 1$, and $\beta_{3} = 2$.
-Applying eqn 1 to population change such that $\Omega = N_{t+1}$, $z_{i} = 1$ and $\Delta z_{i} = 0$, $N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4$ (note $\bar{w} = 4/3$, so $N_{t}\bar{w} = 4$).
+> **Box 3:** As an instructive example of our framework, consider a population of $N_{t} = 3$ annual plants in which individual total fruit mass (kg) is measured, and change is observed over a year.
+For all plants, $\delta_{i} = 1$ (because annual plants die between $t$ and $t+1$), and let plant fecundities be $\beta_{1} = 1$, $\beta_{2} = 1$, and $\beta_{3} = 2$.
+Applying eqn 1 to population change such that $\Omega = N_{t+1}$, $z_{i} = 1$ and $\Delta z_{i} = 0$, $N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4$ (note $\bar{w} = 4/3$, so $N_{t}\bar{w} = 3(4/3) = 4$).
 Focusing next on the characteristic of total fruit mass, let $z_{1} = 0.8$ kg, $z_{2} = 1.0$ kg, and $z_{3} = 1.5$ kg.
@@
 In this case, population fruit yield at $t+1$ is $\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2$ kg.
-At $t$, mean fruit yield per plant was 1.1 kg, but at $t+1$, mean fruit yield per plant is $1.3$ kg.
-Note that $\mathrm{Cov}(w, z) = 2/15$ and $\mathrm{E}(w \Delta z) = 2/15$, so applying the Price equation, $\bar{w}\Delta z = 2/15 + 2/15 = 4/15$.
+At $t$, mean fruit yield per plant was $(0.8 + 1.0 + 1.5)/3 = 1.1$ kg, but at $t+1$, mean fruit yield per plant is $[(0.8 + 0.1) + (1.0 + 0.1) + 2(1.5 + 0.1)]/4 = 1.3$ kg.
+Note that $\mathrm{Cov}(w, z) = (1/N)\sum_{i}^{N}(z_{i} - \bar{z})(w_{i} - \bar{w}) = (1/3)\left[(0.8 - 1.1)(1-4/3) + (1 - 1.1)(1-4/3) + (1.5 - 1.1)(2-4/3) \right]  = 2/15$ and $\mathrm{E}(w \Delta z) = 4/3 \times 0.1 = 2/15$, so applying the Price equation, $\bar{w}\Delta \bar{z} = 2/15 + 2/15 = 4/15$.
 Since $\bar{w} = 4/3$, multiplying both sides of the equation by $3/4$ returns $\Delta z = 0.2$, which is the mean difference in fruit yield between $t+1$ and $t$.
@@
 We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
-From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 2) and principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
-By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 1).
+From the basic axiom that each individual is one member of a specific population ($z_{i} = 1$ and $\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 2) and the principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
+By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution, the Price equation (Box 1).
 Our eqn 1 thereby provides a foundation for defining eco-evolutionary change in any population.
@@
 This equation formally and completely describes population change in terms of births and deaths.
-In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population, but one that is revealed to be necessarily abstract and not dynamically sufficient.
-We hope that this will be useful for eco-evolutionary theory in a similar way that the Price equation is useful for evolutionary theory: potentially facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.
+In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population.
+Like all fundamental equations, our equation is necessarily abstract and not dynamically sufficient.
+We believe that it will be useful for eco-evolutionary theory in a similar way that the Price equation is useful for evolutionary theory: potentially facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements (see below and Supporting Information S1).
 
@@
 ====================================================
+
+<div id="refs"></div>
+
+\clearpage
+
+Supporting Information S1
+================================================
+
+This supporting information demonstrates how to derive well-established models in population ecology and evolutionary biology from equation 1 in the main text,
+
+$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
+\tag{1}
+$$
+
+In the main text, we derived both the Price equation and the birth-death model from the above.
+Here we integrate interactions between individuals to recover density-dependent population growth, and we integrate groups within the population to recover multi-level selection.
+Finally, we integrate both to model a system in which multi-level selection and density dependent population change occur simultaneously.
+We can do this in a way that clarifies model assumptions by allowing an individual $j$ to modulate the birth or death of the focal individual $i$.
+
+Density-dependent population growth
+===========================================
+
+There are two potential ways to model the incorporation of density dependence into population growth.
+We start with what is likely the most familiar model focusing on individual growth rate $r_{i}$, then use a slightly different model focusing on fitness $w_{i}$. 
+First note that here we set $\Omega = N_{t+1}$, and $z_{i} = 1$ and $\Delta z_{i} = 0$ for all individuals as in the main text.
+We can define $r_{i} = \beta_{i} - \delta_{i}$ as an individual growth rate for $i$ [@Lion2018; @Lion2023].
+In this case,
+
+$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i} + 1\right)
+\tag{S1}
+$$
+
+Mathematically, the most general approach here would be to define individual growth as a function of the entire system $\mathbf{E}$, $r_{i}(\mathbf{E})$, where $\mathbf{E}$ is a vector with elements including any parameters potentially relevant to $r_{i}$.
+Taking this approach would recover a version of eqn 2 in @Lion2018 and permit any relationship between the system and a focal individual's growth.
+Limiting our focus to the effects of other individuals ($j$) and assuming that the effects of these individuals are additive, let $a_{ij|\cdot}$ be the effect of individual $j$ on the growth rate attributable to $i$ conditioned on all other individuals within the population such that $r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)$ defines the realised growth rate of $i$,
+
+$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij|\cdot} \right) + 1\right).
+\tag{S2}
+$$
+
+Assuming that individual effects of $j$ on $i$ are also independent, we can remove the condition,
+
+$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij} \right) + 1\right).$$
+
+Further assuming that all individuals have the same per capita effect such that $a = a_{ij}$ for any $i$ and $j$ pair (as might be reasonable given resource competition in a well-mixed population),
+
+$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - a N_{t} \right) + 1\right).$$
+
+If $r_{i}$ values are identical,
+
+$$N_{t+1} = N_{t} + r N_{t}\left(1 - a N_{t} \right).
+\tag{S3}
+$$
+
+Equation S3 therefore recovers a classic version of a discrete time logistic growth by making assumptions from an exact model of eco-evolutionary change.
+
+An alternative approach would be to define model the effects of an individual $j$ on the fitness of $i$ ($w_{i}$), thereby replacing eqn S1 with $N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}$ and replacing eqn S2 with,
+
+$$N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}\alpha_{ij|\cdot} \right).
+$$
+
+Note that we have used $\alpha_{ij}$ to represent the effect of $j$ on the fitness of $i$ for clarity in the sections below.
+By making the same assumptions of additivity, independence, and identical effects such that $\alpha = \alpha_{ij}$ for all $j$ on $i$, and assuming fitness is equal ($w_{i} = \bar{w}$), we can derive,
+
+$$N_{t+1} = \bar{w}N_{t}(1 - \alpha N_{t}).
+\tag{S4}
+$$
+
+This is an alternative way to express logistic growth.
+
+Multi-level selection
+=============================
+
+We can recover multi-level selection from our eqn 1. 
+Here we derive the original form of the multi-level @Price1972 equation as it appears in eqn 3.1 of @Lehtonen2020a.
+Individuals belong to one of $K$ total groups where $j$ indexes $K$ groups and $i$ indexes individuals. 
+Individuals do not overlap in group membership. 
+The size of group $j$ is denoted as $N_{j}$.
+Equation S5 below uses summations to partition how individuals within each group contribute to $\Omega$,
+
+$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}\left(\beta_{j,i} - \delta_{j,i} + 1 \right)\left(z_{j, i} + \Delta z_{j, i}\right).
+\tag{S5}
+$$
+
+In S5, indices $\beta_{j, i}$, $\delta_{j, i}$, and $z_{j, i}$ identify individual $i$ in group $j$.
+We set $w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1$, and for simplicity let $\Delta z_{j, i} = 0$ (i.e., no transmission bias),
+
+$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}w_{j,i}z_{j, i}.$$
+
+For ease of presentation, with no loss of generality, we assume all group sizes are equal with a group size of $N_{j} = n$ for all $j$.
+If group sizes differ, then weighted expectations and covariances are instead needed [@Lehtonen2020a].
+Given equal group sizes, the total number of individuals ($N$) equals $K \times n$, and,
+
+$$\frac{\Omega}{K n} = \left(\frac{1}{K}\right)\left(\frac{1}{n}\right)\sum_{j=1}^{K}\sum_{i=1}^{n}w_{j,i}z_{j, i}.$$
+
+Rearranging,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\frac{1}{n}\sum_{i=1}^{n}w_{j,i}z_{j, i}.$$
+
+The inner summation can be rewritten as an expectation for group $j$,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i} z_{j,i}\right).$$
+
+As in the main text, we note $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
+Defining $Cov_{j}(w_{j}, z_{j})$ as the covariance between $w_{j,i}$ and $z_{j, i}$ for group $j$,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).$$
+
+We can separate the summation for each term,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).$$
+
+Using the notation $\bar{w}_{j} = E_{j}(w_{j,i})$ and $\bar{z}_{j} = E_{j}(z_{j,i})$ to indicate the expectation in group $j$,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).$$
+
+We can also rewrite the first term on the right-hand side as an expectation,
+
+$$\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).$$
+
+We can rearrange the second term on the right-hand side ($\bar{\bar{w}}$ indicates grand mean over all groups),
+
+$$\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.$$
+
+As in the main text, note that $Kn\bar{\bar{w}}$ accounts for differences in total population size from $t$ to $t+1$, with $\bar{\bar{w}}$ being mean fitness across all groups.
+We can therefore set $\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}$, so,
+
+$$\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right).$$
+
+Because $\Delta \bar{\bar{z}} = \bar{\bar{z'}} - \bar{\bar{z}}$,
+
+$$\bar{\bar{w}}\Delta \bar{\bar{z}} = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
+\tag{S6}
+$$
+
+This recovers the multi-level @Price1972 equation [@Lehtonen2020a] from a starting point of eco-evolutionary change in different groups.
+Equation S6 can be found in @Lehtonen2016 B2.I, who then derives a multi-level selection version of Hamilton's rule predicting the evolution of altruism.
+
+Integration of ecology and evolution
+========================================
+
+For simplicity, we now focus on showing an integration between ecology and evolution using a population with no multi-level selection and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias).
+As above in the section on density-dependent population growth, we define $w_{i} = \beta_{i} - \delta_{i} + 1$ and use $\alpha_{i,j}$ to represent the effect of $j$ on the fitness of $i$.
+Our starting equation is therefore,
+
+$$\Omega = \sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S7}
+$$
+
+We have already demonstrated that if we assume all individuals have the same effect on a focal individual such that $\alpha = \alpha_{i,j}$ for all $i$ and $j$ pairs, we can recover equation S4 when $z_{i} = 1$ and $\Omega$ is therefore interpreted as the count of entities,
+
+$$N_{t+1} = \bar{w} N_{t} \left(1 - \alpha N_{t}\right).$$
+
+We now start from S7 to derive $\Delta \bar{z}$.
+The objective is to use our definition of eco-evolutionary change to simultaneously recover how interactions between individuals affect population change and evolutionary change.
+
+We start by dividing both sides of S7 by $N$,
+
+$$\frac{\Omega}{N} = \frac{1}{N}\sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S8}
+$$
+
+We can express the right-hand side of eqn S8 as an expectation,
+
+$$\frac{\Omega}{N} = \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}\right).
+\tag{S9}
+$$
+
+We can rewrite the right-hand side in terms of covariances,
+
+$$\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)\right)\mathrm{E}\left(z_{i}\right).
+\tag{S9}
+$$
+
+The expectations in the second term on the right-hand side can be replaced with overbars to represent the mean,
+
+$$\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i}. 
+\tag{S9}
+$$
+
+In the main text, we noted that $\Omega = N\bar{w}\bar{z}'$.
+Here mean fitness incorporates individual interactions, therefore,
+
+$$\Omega = N\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}'.
+\tag{S10}
+$$
+
+We can therefore rewrite S9,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}' -  \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right).
+\tag{S11}
+$$
+
+Noting again $\Delta \bar{z} = \bar{z}' - \bar{z}$,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
+\tag{S12}
+$$
+
+We can rewrite the left-hand side of S12,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i} - w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S13}
+$$
+
+The covariance term can be split without any additional assumptions,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S14}
+$$
+
+If we are able to further assume that $w_{i}$ and the summation over $\alpha_{i,j}$ are independent, then we could rewrite,
+
+$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right)\bar{w_{i}}. 
+\tag{S15}
+$$
+
+Partitioning fitness into different components with the Price equation is commonplace.
+But this derivation highlights, e.g., the ecological and evolutionary relationship between nonsocial and social components of fitness, and population size.
+For example, the second term on the right-hand side of S13 shows the covariance between the sum of social interactions and a trait.
+When traits covary with the interaction between sociality and fitness, they will have a stronger effect on trait change.
+The magnitude of this second term will also increase with $N$, which reflects the stronger effect of the interaction between sociality and fitness when there are more individuals interacting with a focal individual.
+
+<!---
+
+We can start again with $K$ groups and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias).
+But for simplicity, now consider the case in which $\delta_{j,i} = 1$ (complete mortality), and consider $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K}\sum_{q=1}^{n} \alpha^{j, i}_{p,q})$, where $\alpha^{j, i}_{p,q}$ is the effect that individual $q$ from group $p$ has on a focal individual $i$ from group $j$.
+To keep the model tractable, we assume that all individuals within a group $p$ have identical effects on any focal individual in the group $j$, such that $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n)$
+
+$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{n}\beta_{j, i}\left(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}.
+\tag{S7}
+$$
+
+Our objective here is to start from the above equation and recover population growth and multi-level selection given this parameterisation focused on individual birth as affected by other individuals in the population.
+If we set $z_{j, i} = 1$ as we have done for population ecology, we can recover logistic growth,
+
+$$N_{t+1} = \sum_{j=1}^{K}\sum_{i=1}^{n}\beta_{j, i}\left(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n\right).$$
+
+If we ignore groups and note $N_{t} = nK$, and we assume that the per capita effect of every individual from every group on a focal individual is equal (i.e., $\alpha = \alpha^{j}_{p}$ for all groups),
+
+$$N_{t+1} = N_{t}\beta_{j, i}\left(1 - \alpha N_{t}\right).$$
+
+We therefore derive the model for population change.
+
+We now use the same starting point to carry the effects of individuals on fitness through a model of multi-level selection.
+Rearranging S7,
+
+$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{n}\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}.$$
+
+As we have done previously, we can multiply both sides of the equation by $(1/K)(1/n)$,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\frac{1}{n}\sum_{i=1}^{n}\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}.$$
+
+We can rewrite the inner summation as an expectation for group $j$,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}E\left(\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)z_{j, i}\right).$$
+
+We can rewrite the product within the expectation as a covariance plus a product of the expectations and split the covariance (note that $Cov(X-Y, Z) = Cov(X, Z) - Cov(Y, Z)$),
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) + E\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right) E\left(z_{j, i}\right).$$
+
+All terms are within the summation, which can therefore be separated,
+
+$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}E\left(\beta_{j, i} - \beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right) E\left(z_{j, i}\right).$$
+
+Note that because $E(X - Y) \times E(Z) = E(X)E(Z) - E(Y)E(Z)$, the third term on the right-hand side can be split,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ \frac{1}{K}\sum_{j=1}^{K}E\left(\beta_{j, i}\right)E\left(z_{j, i}\right) - E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right).
+\end{align*}
+
+We can rewrite this third term as an expectation,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ E\left(E\left(\beta_{j, i}\right)E\left(z_{j, i}\right) - E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+Because of the linearity of expectation, we can again separate the third term on the right-hand side of the equation,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ E\left(E\left(\beta_{j, i}\right)E\left(z_{j, i}\right)\right) - E\left(E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+For convenience, we substitute $E(\beta_{j,i}) = \bar{\beta}_{j,i}$ and $E(z_{j,i}) = \bar{z}_{j,i}$,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ E\left(\bar{\beta}_{j, i} \bar{z}_{j, i}\right) - E\left(E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+We can split the third term above into a covariance and product of means,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right)\\ 
+                   &+ Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) + \bar{\bar{\beta}}_{j,i}\bar{\bar{z}}_{j,i} - E\left(E\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n\right)E\left(z_{j, i}\right)\right).
+\end{align*}
+
+We can then do the same for the last term on the right-hand side,
+
+\begin{align*}
+\frac{\Omega}{K n} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\
+                   &+ Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) + \bar{\bar{\beta}}_{j,i}\bar{\bar{z}}_{j,i} - Cov(\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}, \bar{z}_{j,i}) - \overline{\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}}\times\bar{\bar{z}}_{j,i}.
+\end{align*}
+
+As in the previous section of this supporting information, we note that $\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}$.
+In this case, fitness $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n)$
+
+\begin{align*}
+\bar{\bar{w}}\Delta\bar{\bar{z}} &= \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}, z_{j,i}\right) - \frac{1}{K}\sum_{j=1}^{K}Cov\left(\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right) \\ 
+                                 &+ Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) - Cov(\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}, \bar{z}_{j,i}).
+\end{align*}
+
+And finally,
+
+$$\bar{\bar{w}}\Delta\bar{\bar{z}} = E\left(Cov\left(\beta_{j}, z_{j}\right)\right) - E\left(Cov\left(\beta_{j}\sum_{p=1}^{K} \alpha^{j}_{p} n, z_{j,i}\right)\right) + Cov\left(\bar{\beta}_{j, i}, \bar{z}_{j, i}\right) - Cov(\overline{\beta_{j, i}\sum_{p=1}^{K} \alpha^{j}_{p} n}, \bar{z}_{j,i}).$$
+
+This seems intuitive in that it just adds two terms to the derivation of multi-level selection.
+And given how we have defined $\alpha$, this should relate the evolutoinary change in some trait $z$ to interactions, specifically density dependent decreases in fitness in this case.
+Note that if $\alpha = 0$, the second and fourth terms drop out, as they should because the covariance with the trait is zero.
+
+--->
+
+Supporting Information References 
+==========================
+
+\addcontentsline{toc}{section}{References}
+
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-Lehtonen2016}
+Lehtonen, Jussi. 2016. {``{Multilevel selection in kin selection
+language}.''} \emph{Trends in Ecology and Evolution} xx: 1--11.
+\url{https://doi.org/10.1016/j.tree.2016.07.006}.
+
+\bibitem[\citeproctext]{ref-Lehtonen2020a}
+---------. 2020. {``The Price Equation and the Unity of Social Evolution
+Theory.''} \emph{Philosophical Transactions of the Royal Society B:
+Biological Sciences} 375: 20190362.
+\url{https://doi.org/10.1098/rstb.2019.0362}.
+
+\bibitem[\citeproctext]{ref-Lion2018}
+Lion, Sébastien. 2018. {``{Theoretical approaches in evolutionary
+ecology: environmental feedback as a unifying perspective}.''}
+\emph{American Naturalist} 191 (1).
+\url{https://doi.org/10.1086/694865}.
+
+\bibitem[\citeproctext]{ref-Lion2023}
+Lion, Sébastien, Akira Sasaki, and Mike Boots. 2023. {``Extending
+Eco-Evolutionary Theory with Oligomorphic Dynamics.''} \emph{Ecology
+Letters} 26 (September): S22--46.
+\url{https://doi.org/10.1111/ele.14183}.
+
+\bibitem[\citeproctext]{ref-Price1972}
+Price, George R. 1972. {``{Extension of covariance selection
+mathematics}.''} \emph{Annals of Human Genetics} 35 (4): 485--90.
+\url{https://doi.org/10.1111/j.1469-1809.1957.tb01874.x}.
+
+\end{CSLReferences}

### 2025-06-18 foundations_of_ecoevo_arxiv.tex
Edits for resubmission of revision to ecology and evolution

--- 2025-06-09 foundations_of_ecoevo.tex
@@
 \usepackage{lineno}
-\linenumbers
-\modulolinenumbers[2]
 \ifLuaTeX
@@
 
-\setstretch{2}
+\setstretch{1}
 \section{Abstract}\label{abstract}
@@
 and death that underlie change in population density. The deep
-interdependence between ecology and evolution is well-established, but
-much theory in each discipline has been developed in isolation. While
-recent work has accomplished eco-evolutionary synthesis, a gap remains
-between the logical foundations of ecology and evolution. We bridge this
-gap with a new equation that defines a summed value for a characteristic
-across individuals in a population, from which the fundamental equations
-of population ecology and evolutionary biology (the Price equation) are
-derived. We thereby unify the fundamental equations of population
-ecology and biological evolution under a general framework. Our
-unification further demonstrates the equivalence between mean population
-growth rate and evolutionary fitness and links this change to ecosystem
-function. Finally, we outline how our proposed framework can be used to
-unify social evolution and density-dependent population growth.
+interdependence between ecology and evolution is well-established, and
+recent models focus on integrating eco-evolutionary dynamics to
+demonstrate how ecological and evolutionary processes interact and feed
+back upon each other. Nevertheless, a gap remains between the logical
+foundations of ecology and evolution. Both population ecology and
+evolution have fundamental equations that define how the size of a
+population (ecology) and the average characteristic within a population
+(evolution) change, respectively, over time. These fundamental equations
+are a complete and exact description of change for any closed
+population, but how they are formally linked remains unclear. We link
+the fundamental equations of population ecology and evolution with an
+equation that sums how individual characteristics interact with
+individual fitness in a population. From this equation, we derive the
+fundamental equations of population ecology and evolutionary biology
+(the Price equation). We thereby identify an overlooked bridge between
+ecology and biological evolution. Our unification formally recovers the
+equivalence between mean population growth rate and evolutionary fitness
+and links this change to ecosystem function. We outline how our
+framework can be used to further develop eco-evolutionary theory.
 
@@
 
-Theoretical unification of phenomena is a powerful tool for scientific
-advancement. Such unification has been a major goal in scientific
-research throughout history (\citeproc{ref-Kitcher1993}{Kitcher 1993}),
-and its value is perhaps most evident in reconciling unconnected models
-and revealing new and unexpected empirical predictions. In evolutionary
-biology, the Price equation (Box 1) provides a unifying framework for
-evolutionary theory by exhaustively and exactly describing evolutionary
-change for any closed evolving population
-(\citeproc{ref-Price1970}{Price 1970}; \citeproc{ref-Luque2016}{Luque
-2017}; \citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}). The Price
-equation is therefore fundamental, in the sense that it binds together
-all of evolutionary theory by formally defining what evolutionary change
-is and is not (\citeproc{ref-Price1970}{Price 1970};
+Theoretical unification is a powerful tool for scientific advancement.
+Such unification has been a major goal in scientific research throughout
+history (\citeproc{ref-Smocovitis1992}{Smocovitis 1992};
+\citeproc{ref-Kitcher1993}{Kitcher 1993}), and its value is perhaps most
+evident in reconciling unconnected models and revealing new and
+unexpected empirical predictions. In evolutionary biology, the Price
+equation (Box 1) provides a unifying framework for evolutionary theory
+by exhaustively and exactly describing evolutionary change for any
+closed evolving population (\citeproc{ref-Price1970}{Price 1970};
+\citeproc{ref-Luque2016}{Luque 2017};
+\citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}). The Price equation
+is therefore fundamental, in the sense that it binds together all of
+evolutionary theory by formally defining what evolutionary change is and
+is not (\citeproc{ref-Price1970}{Price 1970};
 \citeproc{ref-Rice2004}{Rice 2004}; \citeproc{ref-Gardner2008}{Gardner
@@
 propose an equation that extends this conceptual clarity to include
-population change and thereby provide a formal and exact definition of
-joint ecological and evolutionary change.
+population size change and thereby provide a formal and exact definition
+for joint ecological and evolutionary change.
 
 In biological populations, ecological change is caused by the same
-processes of individual birth and death that cause changes in allele
-frequencies and phenotypes (\citeproc{ref-Turchin2001}{Turchin 2001};
-\citeproc{ref-Connor2004}{Connor and Hartl 2004};
+processes of individual birth and death that cause evolutionary changes
+in allele frequencies and phenotypes (\citeproc{ref-Turchin2001}{Turchin
+2001}; \citeproc{ref-Connor2004}{Connor and Hartl 2004};
 \citeproc{ref-Barfield2011}{Barfield et al. 2011}). As with evolution, a
@@
 change. Unlike the Price equation, this fundamental equation is perhaps
-self-evident; population change is simply the addition of births and
-subtraction of deaths from current population size (\(N_{t}\)) to
-recover the new population size (\(N_{t+1}\); Box 2). By definition,
-this relationship of \(N_{t+1} = N_{t} + Births - Deaths\) applies to
-any closed population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues
-that general principles are needed to establish a logical foundation for
-population ecology, and this simple birth and death model and the
-consequences that logically follow from it (e.g., exponential population
-growth) is fundamental to population ecology. Any unifying definition of
-joint ecological and evolutionary change must be able to derive both the
-Price equation and this birth and death model.
-
-The interdependence of ecological and evolutionary processes has long
-been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
+self-evident. Population change is simply the addition of individuals
+minus the removal of individuals from current population size
+(\(N_{t}\)), which recovers the new population size (\(N_{t+1}\); Box
+2). By definition, this relationship of
+\(N_{t+1} = N_{t} + Births - Deaths\) applies to any closed population.
+Turchin (\citeproc{ref-Turchin2001}{2001}) argues that general
+principles are needed to establish a logical foundation for population
+ecology, and this simple birth and death model and the consequences that
+logically follow from it (e.g., exponential population growth) are
+fundamental to population ecology. Any unifying definition of joint
+ecological and evolutionary change must be able to derive both the Price
+equation and this birth and death model.
+
+The union of ecological and evolutionary processes has long been
+recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
 \citeproc{ref-Fisher1958}{Fisher 1958};
@@
 eco-evolutionary models that follow from these interpretations. The
-first type uses separate equations to model population change versus
-evolutionary change, thereby allowing for any number of ecological,
-evolutionary, or environmental feedbacks (e.g.,
+first type uses separate equations to model population size change
+versus evolutionary change, thereby allowing for any number of
+ecological, evolutionary, or environmental feedbacks (e.g.,
 \citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
 al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
-models population demographics as functions of quantitative traits, with
+models population demographics as functions of quantitative traits with
 ecological and evolutionary change following from demographic processes
@@
 unifying theory (\citeproc{ref-Luque2021}{Luque and Baravalle 2021}),
-and we believe that a fundamental equation of eco-evolutionary change is
-needed.
+and we believe that a fundamental equation of eco-evolutionary change
+has been curiously overlooked.
 
@@
 assumptions that are specific to population ecology or evolution in the
-same way that key equations of population genetics (e.g., average
-excess) or quantitative genetics (e.g., Breeder's equation) can be
-derived from the Price equation (\citeproc{ref-Queller2017}{Queller
-2017}). We propose our equation as a formal definition of
-eco-evolutionary change.
+same way that key equations of population genetics or quantitative
+genetics can be derived from the Price equation by restricting the
+domain of interest (e.g., to allele frequencies in the case of
+population genetics, or to continuous phenotypes in the case of
+quantitative genetics, \citeproc{ref-Queller2017}{Queller 2017}). We
+propose our equation as a formal definition of eco-evolutionary change.
 
@@
 breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
-easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
-\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
-2017}). This abstractness is also key to developing a more general view
-of evolution (\citeproc{ref-Rice2020}{Rice 2020};
-\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
-\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
-researchers consider the Price equation just a triviality (even
-tautological), and useless without further modelling assumptions
+derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
+\citeproc{ref-Lehtonen2018}{2018}, \citeproc{ref-Lehtonen2020a}{2020};
+\citeproc{ref-Queller2017}{Queller 2017}). This abstractness is also key
+to developing a more general view of evolution
+(\citeproc{ref-Rice2020}{Rice 2020}; \citeproc{ref-Luque2021}{Luque and
+Baravalle 2021}; \citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In
+contrast, some researchers consider the Price equation just a triviality
+(even tautological), and useless without further modelling assumptions
 (\citeproc{ref-VanVeelen2005}{van Veelen 2005};
@@
 \textbf{Box 2:} The number of individuals in any closed population
-(\(N\)) at any given time (\(t + 1\)) is determined by the existing
-number of individuals (\(N_{t}\)), plus the number of births
-(\(Births\)) minus the number of deaths (\(Deaths\)),
+(\(N\)) at any given time (\(t + 1\)) is determined by the number of
+individuals at \(t\) (\(N_{t}\)), plus the number of births (\(Births\))
+minus the number of deaths (\(Deaths\)),
 \[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
@@
 assumption that all individuals in the population are identical, a per
-capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
-can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
-\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
-increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
-because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
-size of any closed population equals its existing size times its finite
-rate of increase.
+capita rate of birth, \(Births = bN_{t}\), and death,
+\(Deaths = dN_{t}\), can be defined. Rearranging and defining
+\(\lambda = 1 + b - d\) gives, \(N_{t+1} = N_{t}\lambda\). Here
+\(\lambda\) is the finite rate of increase
+(\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that because
+\(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in size of
+any closed population equals its existing size times its finite rate of
+increase.
 \end{quote}
@@
 
-To fully unify biological evolution and population ecology, we must
-reconcile the the Price equation (Box 1) with the general equation for
-population change (Box 2). The Price equation is critical for
-partitioning different components of biological change
+To unify biological evolution and population ecology, we must reconcile
+the Price equation (Box 1) with the general equation for population
+change (Box 2). The Price equation is critical for partitioning
+different components of biological change
 (\citeproc{ref-Price1970}{Price 1970}; \citeproc{ref-Frank1997}{Frank
@@
 \citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Queller2017}{Queller
-2017}; \citeproc{ref-Lehtonen2018}{Lehtonen 2018}). It has also been
-highly useful for integrating evolutionary theory across disciplines
+2017}; \citeproc{ref-Lehtonen2018}{Lehtonen 2018},
+\citeproc{ref-Lehtonen2020a}{2020}). It has also been highly useful for
+integrating evolutionary theory across disciplines
 (\citeproc{ref-Fox2006}{Fox 2006};
@@
 growth (\citeproc{ref-Turchin2001}{Turchin 2001}), this scaling must be
-avoided in a fundamental equation of ecology and evolution.
+avoided.
 
@@
 2004}; \citeproc{ref-Rice2009}{Rice and Papadopoulos 2009}). In such
-systems, diversity is discontinuous and can be defined in terms of
-discrete entities (\citeproc{ref-Dobzhansky1970}{Dobzhansky 1970}). Our
-framework is general enough that entities can be anything discrete, but
-we will focus on each entity \(i\) as an individual organism.
+systems, diversity is discontinuous, in the sense that living systems
+are composed of discrete entities including individual organisms and
+groups of organisms (\citeproc{ref-Dobzhansky1970}{Dobzhansky 1970}).
+Our framework is general enough that entities can be anything discrete,
+but we will focus on each entity \(i\) as an individual organism.
 Individuals give rise to new individuals through birth such that
@@
 
-The foundation of eco-evolutionary change is an interaction between the
-demographic processes of birth and death
+This foundation of eco-evolutionary change is a statistical interaction
+between the demographic processes of birth and death
 (\(\beta_{i} - \delta_{i} + 1\)) and some characteristic of an
 individual (\(z_{i} + \Delta z_{i}\)). The value of \(\Omega\) is a
-summation of individual characteristics, which takes the same units as
+summation of individual characteristics and takes the same units as
 \(z\). From eqn 1, we can derive the most fundamental equations of
-population ecology (Box 2) and evolutionary biology (Box 1) through an
-appropriate interpretation of \(z\). Under some limited interpretations
-of \(z\), we can also interpret \(\Omega\) as a metric of ecosystem
-function.
+population ecology and evolutionary biology through an appropriate
+interpretation of \(z\). Under more limited interpretations of \(z\), we
+can also interpret \(\Omega\) as a metric of ecosystem function.
 
@@
 define \(z_{i}\) as the identity of \(i\) belonging to the population.
-In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
-member of the population. In this case, \(z\) is count, which takes the
-unit 1 {[}note that `individual' is a label, not a unit; Newell and
-Tiesinga (\citeproc{ref-NewellEtAl2019}{2019}){]}. Further, we assume
-that individuals do not change species by setting \(\Delta z_{i} = 0\).
-In this case,
+In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is
+one member of the population. In this restricted case, \(z\) is count,
+which takes the unit 1 (note that `individual' is a label, not a unit,
+see \citeproc{ref-NewellEtAl2019}{Newell and Tiesinga 2019}). Further,
+we assume that individuals do not change species by setting
+\(\Delta z_{i} = 0\). In this case,
 
@@
 We can now interpret \(\Omega\) as the population size at \(t+1\),
-\(N_{t+1}\),
+\(N_{t+1}\). Summing up \(\beta_{i}\), \(\delta_{i}\), and current
+individuals tallies up the total number of individuals in the next time
+step,
 
@@
 We therefore recover the general equation for population ecology (eqn 2)
-and the fundamental property of exponential growth in populations
-(\citeproc{ref-Turchin2001}{Turchin 2001}) (eqn 3).
+and the fundamental property of exponential growth in populations (eqn
+3, \citeproc{ref-Turchin2001}{Turchin 2001}).
 
@@
 
-In this definition, the longevity of the individual matters. An
-individual that survives from \(t\) to \(t + 1\) has a higher fitness
-than one that dies, even if both have the same reproductive output. With
-this definition of fitness (eqn 4), we substitute,
+In this definition, the longevity of the individual matters. By
+definition, an individual that survives from \(t\) to \(t + 1\) has a
+higher fitness than one that dies, even if both have the same
+reproductive output. With this definition of fitness (eqn 4), we
+substitute,
 
@@
 at \(t + 1\). But we cannot treat \(\Omega/N\) as the mean of \(z\) at
-\(t+1\) (\(\bar{z}'\)) because we need to weigh \(N\) by the mean
+\(t+1\) (\(\bar{z}'\)) because we need to weight \(N\) by the mean
 fitness of the population at \(t\) to account for any change in
@@
 mean fitness \(\bar{w}\) (at \(t\)) to recover the mean contribution of
-the \(N\) individuals at \(t\) to the total \(\Omega\) (this is noted in
-the context of population genetics by \citeproc{ref-Ewens2014}{Ewens
-2014}). Consequently,
+the \(N\) individuals at \(t\) to the total \(\Omega\) (this has been
+applied in the context of population genetics by
+\citeproc{ref-Ewens2014}{Ewens 2014}). Consequently,
 
@@
 We can rearrange eqn 9 to derive the Price equation by expressing
-covariance as,
+covariance as
 \(\mathrm{Cov}(X,Y) = \mathrm{E}(XY) - \mathrm{E}(X)\mathrm{E}(Y)\), and
@@
 
-From eqn 1, which describes fundamental birth and death processes in a
-population, we can derive both the most fundamental model of population
-ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn
-10; Box 1).
+Equation 10 is the Price equation. From eqn 1, which describes
+fundamental birth and death processes in a population, we can therefore
+derive both the most fundamental model of population ecology (eqn 2; Box
+2) and the fundamental equation of evolution (eqn 10; Box 1).
 
@@
 
-In some cases, \(\Omega\) could also be intepreted as the total
-contribution of a population to ecosystem function. This will be
-restricted to cases in which \(z\) is an individual characteristic
-defining an absolute quantity measured at the whole organism level, such
-as biomass, seed production, carbon capture, flower visits, or nutrient
-consumption. In these cases, the sum across individuals gives a
-meaningful total quantity for the population. When \(z\) is instead
-defined by relative organism-level measurements such wing loading,
-nutrient ratio, or diet composition, or when \(z\) is measured at a
-level of biological organisation below the organism (e.g., average cell
-volume or leaf surface area), \(\Omega\) does not have a clear
-population-level interpretation.
-
-Box 3 provides a toy example of three plants with different fitnesses
-and carbon captures. We use this toy example to demonstrate a concrete
-example of our proposed framework.
+In some cases, \(\Omega\) could also be interpreted as the total
+contribution of a population to ecosystem function. This is restricted
+to cases in which \(z\) is a characteristic defining an absolute
+quantity measured at the whole organism level such as biomass, seed
+production, carbon capture, flower visits, or nutrient consumption. In
+such cases, the sum across individuals gives a meaningful total quantity
+for the population (e.g., the total biomass or seeds produced in the
+population). When \(z\) is instead defined by relative organism-level
+measurements such wing loading, nutrient ratio, or diet composition, or
+when \(z\) is measured at a level of biological organisation below the
+organism (e.g., average cell volume or leaf surface area), \(\Omega\)
+does not have a clear population-level interpretation. In such cases,
+eqn 1 still defines eco-evolutionary change; the interpretation of
+\(\Omega\) by itself is just not as interesting, biologically.
+
+Box 3 provides an instructive example of three plants with different
+fitnesses and carbon captures. We use this example to demonstrate a
+concrete example of our proposed framework.
 
@@
 \begin{quote}
-\textbf{Box 3:} As a toy example of our framework, consider a population
-of \(N_{t} = 3\) annual plants in which individual fruit mass (kg) is
-measured, and change is observed over a year. For all plants,
-\(\delta_{i} = 1\), and let plant fecundities be \(\beta_{1} = 1\),
+\textbf{Box 3:} As an instructive example of our framework, consider a
+population of \(N_{t} = 3\) annual plants in which individual total
+fruit mass (kg) is measured, and change is observed over a year. For all
+plants, \(\delta_{i} = 1\) (because annual plants die between \(t\) and
+\(t+1\)), and let plant fecundities be \(\beta_{1} = 1\),
 \(\beta_{2} = 1\), and \(\beta_{3} = 2\). Applying eqn 1 to population
@@
 \(N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4\)
-(note \(\bar{w} = 4/3\), so \(N_{t}\bar{w} = 4\)). Focusing next on the
-characteristic of total fruit mass, let \(z_{1} = 0.8\),
-\(z_{2} = 1.0\), and \(z_{3} = 1.5\). Also let \(\Delta z_{i} = 0.1\)
-for all plants to reflect a change in soil environment from \(t\) to
-\(t + 1\). In this case, total fruit yield at \(t+1\) is
+(note \(\bar{w} = 4/3\), so \(N_{t}\bar{w} = 3(4/3) = 4\)). Focusing
+next on the characteristic of total fruit mass, let \(z_{1} = 0.8\) kg,
+\(z_{2} = 1.0\) kg, and \(z_{3} = 1.5\) kg. Also let
+\(\Delta z_{i} = 0.1\) for all plants to reflect a change in soil
+environment from \(t\) to \(t + 1\). In this case, population fruit
+yield at \(t+1\) is
 \(\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2\)
-kg. At \(t\), mean fruit yield per plant was 1.1 kg, but at \(t+1\),
-mean fruit yield per plant is \(1.3\) kg. Note that
-\(\mathrm{Cov}(w, z) = 2/15\) and \(\mathrm{E}(w \Delta z) = 2/15\), so
-applying the Price equation, \(\bar{w}\Delta z = 2/15 + 2/15 = 4/15\).
-Since \(\bar{w} = 4/3\), multiplying both sides of the equation by
-\(3/4\) returns \(\Delta z = 0.2\), which is the mean difference in
-fruit yield between \(t+1\) and \(t\). The framework expressed in eqn 1
-thereby links population change, evolutionary change, and ecosystem
-function.
+kg. At \(t\), mean fruit yield per plant was
+\((0.8 + 1.0 + 1.5)/3 = 1.1\) kg, but at \(t+1\), mean fruit yield per
+plant is \([(0.8 + 0.1) + (1.0 + 0.1) + 2(1.5 + 0.1)]/4 = 1.3\) kg. Note
+that
+\(\mathrm{Cov}(w, z) = (1/N)\sum_{i}^{N}(z_{i} - \bar{z})(w_{i} - \bar{w}) = (1/3)\left[(0.8 - 1.1)(1-4/3) + (1 - 1.1)(1-4/3) + (1.5 - 1.1)(2-4/3) \right]  = 2/15\)
+and \(\mathrm{E}(w \Delta z) = 4/3 \times 0.1 = 2/15\), so applying the
+Price equation, \(\bar{w}\Delta \bar{z} = 2/15 + 2/15 = 4/15\). Since
+\(\bar{w} = 4/3\), multiplying both sides of the equation by \(3/4\)
+returns \(\Delta z = 0.2\), which is the mean difference in fruit yield
+between \(t+1\) and \(t\). The framework expressed in eqn 1 thereby
+links population change, evolutionary change, and ecosystem function.
 \end{quote}
@@
 
-A classical sign of scientific progress is the ability to connect
-disparate theories and models to show how empirical and theoretical
-models are logical (mathematical) consequences of more fundamental ones
-(\citeproc{ref-Nagel1961}{Nagel 1961};
+An important aspect of scientific progress is the ability to connect
+disparate theories and models to show how specific empirical and
+theoretical models are logical (mathematical) consequences of more
+fundamental ones (\citeproc{ref-Nagel1961}{Nagel 1961};
 \citeproc{ref-Morrison2000}{Morrison 2000}). Rather than making
-simplifying assumptions, as is the approach for most ecological and
+simplifying assumptions, as is the approach for specific ecological and
 evolutionary models, we focus on fundamental axioms that are universal
@@
 change over time. We define an abstract sum (\(\Omega\)), to which all
-individuals within the population contribute. From the simple
-assumptions of population identity (\(z_{i} = 1\)) and invariability
-(\(\Delta z_{i} = 0\)), we recover the most general equation of
-population ecology (Box 2) and principle of exponential growth
+individuals within the population contribute. From the basic axiom that
+each individual is one member of a specific population (\(z_{i} = 1\)
+and \(\Delta z_{i} = 0\)), we recover the most general equation of
+population ecology (Box 2) and the principle of exponential growth
 (\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
@@
 \citeproc{ref-Frank2016}{2016}), we recover the most fundamental
-equation of evolution (Box 1). Our eqn 1 thereby provides a foundation
-for defining eco-evolutionary change in any population.
+equation of evolution, the Price equation (Box 1). Our eqn 1 thereby
+provides a foundation for defining eco-evolutionary change in any
+population.
 
@@
 rearranging the mathematical notation defining changes in the
-frequencies and characteristics of any type of entity
-(\citeproc{ref-Price1970}{Price 1970};
+frequencies and characteristics of any type of entity (e.g.,
+individuals, alleles, \citeproc{ref-Price1970}{Price 1970};
 \citeproc{ref-Gardner2008}{Gardner 2008}; \citeproc{ref-Luque2016}{Luque
-2017}) (e.g., individuals, alleles). This derivation partitions total
-characteristic change into different components, making it possible to
-isolate evolutionary mechanisms (e.g., selection) and levels of
-biological organisation (e.g., group, individual)
-(\citeproc{ref-Frank1995}{Frank 1995}, \citeproc{ref-Frank2012}{2012};
-\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
-\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Okasha2020}{Okasha
-and Otsuka 2020}). Because of its abstract nature and lack of any
-system-specific assumptions, the Price equation is not dynamically
-sufficient and makes no predictions about what will happen in any
-particular system (\citeproc{ref-Gardner2020}{Gardner 2020}). Its role
-is not to predict, but to formally and completely define and separate
-components of evolutionary change. The same is true of the general
-equation for population change (eqns 2 and 3), at least as we have used
-it here where it serves to define what population growth means in
-ecology. This equation formally and completely describes population
-change in terms of births and deaths. In eqn 1, we therefore have a
-fundamental equation from which we can derive complete ecological and
-evolutionary change in any closed biological population, but one that is
-revealed to be necessarily abstract and not dynamically sufficient. We
-anticipate that this will be useful for eco-evolutionary theory in the
-same way that the Price equation is useful for evolutionary theory:
-facilitating specific model development and identifying new conceptual
-insights, unresolved errors, and sources of model disagreements.
+2017}). This derivation partitions total characteristic change into
+different components, making it possible to isolate evolutionary
+mechanisms (e.g., selection) and levels of biological organisation
+(e.g., group, individual) (\citeproc{ref-Frank1995}{Frank 1995},
+\citeproc{ref-Frank2012}{2012}; \citeproc{ref-Kerr2009}{Kerr and
+Godfrey-Smith 2009}; \citeproc{ref-Luque2016}{Luque 2017};
+\citeproc{ref-Okasha2020}{Okasha and Otsuka 2020}). Because of its
+abstract nature and lack of any system-specific assumptions, the Price
+equation is not dynamically sufficient and makes no predictions about
+what will happen in any particular system
+(\citeproc{ref-Gardner2020}{Gardner 2020}). Its role is not to predict,
+but to formally and completely define and separate components of
+evolutionary change (\citeproc{ref-Baravalle2024}{Baravalle et al.
+2025}). The same is true of the general equation for population change
+(eqn 2), at least as we have used it here where it serves to define what
+population growth means in ecology. This equation formally and
+completely describes population change in terms of births and deaths. In
+eqn 1, we therefore have a fundamental equation from which we can derive
+complete ecological and evolutionary change in any closed biological
+population. Like all fundamental equations, our equation is necessarily
+abstract and not dynamically sufficient. We believe that it will be
+useful for eco-evolutionary theory in a similar way that the Price
+equation is useful for evolutionary theory: potentially facilitating
+specific model development and identifying new conceptual insights,
+unresolved errors, and sources of model disagreements (see below and
+Supporting Information S1).
 
 Our unification recovers the equivalence between the finite rate of
-increase \(\lambda\) (Box 2) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
-and population mean evolutionary fitness \(\bar{w}\) (Box 1). The
-population growth equation \(N_{t+1} = N_{t}\lambda\) can therefore be
-rewritten as \(N_{t+1} = N_{t}\bar{w}\). This specific equivalence has
-been proposed before (\citeproc{ref-Lande1976}{Lande 1976}), as has the
-broader relationship between population growth rate and evolutionary
-fitness (\citeproc{ref-Fisher1930}{Fisher 1930};
+increase \(\lambda\) (Box 2) and population mean evolutionary fitness
+\(\bar{w}\) (Box 1). The population growth equation
+\(N_{t+1} = N_{t}\lambda\) can always be rewritten as
+\(N_{t+1} = N_{t}\bar{w}\). This specific equivalence has been proposed
+before (e.g., \citeproc{ref-Lande1976}{Lande 1976}), as has the broader
+relationship between population growth rate and evolutionary fitness
+(\citeproc{ref-Fisher1930}{Fisher 1930};
 \citeproc{ref-Charlesworth1980}{Charlesworth 1980};
@@
 (which, by definition, includes death), fitness is therefore
-\(\beta_{i}\). The rate of change in ecology and evolution are reflected
-in the first and second statistical moments of fitness, respectively.
+\(\beta_{i}\). Rates of change in ecology and evolution are reflected in
+the first and second statistical moments of fitness, respectively.
 Population growth rate reflects mean fitness \(\bar{w}\), while the rate
 of evolutionary change reflects the variance in fitness
-\(\mathrm{Var}(w)/\bar{w}\) (i.e., Fisher's fundamental theorem)
-(\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
+\(\mathrm{Var}(w)/\bar{w}\) (i.e., Fisher's fundamental theorem,
+\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
 2004}; \citeproc{ref-Queller2017}{Queller 2017}).
@@
 in allele frequencies and abstracting away from individuals and their
-environments (the ecological interactions)
-(\citeproc{ref-MacColl2011}{MacColl 2011}). This has been a line of
+environments (the ecological interactions,
+\citeproc{ref-MacColl2011}{MacColl 2011}). This has been a line of
 argumentation by some defenders of the so-called Extended Synthesis
@@
 to niche construction (\citeproc{ref-Odling-smee2003}{Odling-Smee et al.
-2003}). Famously, Mayr (\citeproc{ref-Mayr1960}{1959}) characterized
+2003}). Famously, Mayr (\citeproc{ref-Mayr1960}{1959}) characterised
 population genetics as a simple input and output of genes, analogous to
@@
 
-Our fundamental equation is complete and exact. It therefore implicitly
-includes any effects of density dependence on population growth (see Box
-2), or any social effects on evolutionary change (see Box 1). Both of
-these effects can be made explicit by specifying how other individuals
-in a population affect birth and death of a focal individual. Assuming
-that effects of individuals are additive, independent, and identical, we
-can derive well-established logistic growth equations and multi-level
-selection from eqn 1 (Supporting Information S1).
+We believe our fundamental equation to be complete and exact for any
+closed population. It therefore implicitly includes any effects of
+density dependence on population growth (see Box 2), or any social
+effects on evolutionary change (see Box 1). Both of these effects can be
+made explicit by specifying how other individuals in a population affect
+birth and death of a focal individual. We demonstrate this by deriving
+more specific models of density dependent population growth and
+multi-level selection in Supporting Information S1. In Supporting
+Information S1, we also suggest how new theoretical insights might be
+gained by integrating density-dependent growth and social interactions
+into our framework.
 
@@
 helped us clarify the relationship between population growth and
-fitness. Brad Duthie would also like to thank Brent Danielson and Stan
-Harpole. Victor J. Luque would also like to thank Lorenzo Baravalle, Pau
-Carazo, Santiago Ginnobili, Manuel Serra, and Ariel Roff\(\'{e}\).
+fitness. Brad Duthie would also like to thank Matt Tinsley, Brent
+Danielson, and Stan Harpole. Victor J. Luque would also like to thank
+Lorenzo Baravalle, Pau Carazo, Santiago Ginnobili, Manuel Serra, and
+Ariel Roff\(\'{e}\).
 
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
+\section{References}\label{references}
 
 \phantomsection\label{refs}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
 \bibitem[\citeproctext]{ref-Baravalle2022}
@@
 
+\bibitem[\citeproctext]{ref-Lehtonen2020a}
+---------. 2020. \href{https://doi.org/10.1098/rstb.2019.0362}{The price
+equation and the unity of social evolution theory}. Philosophical
+Transactions of the Royal Society B: Biological Sciences 375:20190362.
+
 \bibitem[\citeproctext]{ref-Lehtonen2020}
@@
 
+\bibitem[\citeproctext]{ref-Smocovitis1992}
+Smocovitis, V. B. 1992. \emph{Unifying biology: The evolutionary
+synthesis and evolutionary biology}. Source: Journal of the History of
+Biology (Vol. 25).
+
 \bibitem[\citeproctext]{ref-Turchin2001}
@@
 
+\clearpage
+
+\section{Supporting Information S1}\label{supporting-information-s1}
+
+This supporting information demonstrates how to derive well-established
+models in population ecology and evolutionary biology from equation 1 in
+the main text,
+
+\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
+\tag{1}
+\]
+
+In the main text, we derived both the Price equation and the birth-death
+model from the above. Here we integrate interactions between individuals
+to recover density-dependent population growth, and we integrate groups
+within the population to recover multi-level selection. Finally, we
+integrate both to model a system in which multi-level selection and
+density dependent population change occur simultaneously. We can do this
+in a way that clarifies model assumptions by allowing an individual
+\(j\) to modulate the birth or death of the focal individual \(i\).
+
+\section{Density-dependent population
+growth}\label{density-dependent-population-growth}
+
+There are two potential ways to model the incorporation of density
+dependence into population growth. We start with what is likely the most
+familiar model focusing on individual growth rate \(r_{i}\), then use a
+slightly different model focusing on fitness \(w_{i}\). First note that
+here we set \(\Omega = N_{t+1}\), and \(z_{i} = 1\) and
+\(\Delta z_{i} = 0\) for all individuals as in the main text. We can
+define \(r_{i} = \beta_{i} - \delta_{i}\) as an individual growth rate
+for \(i\) (\citeproc{ref-Lion2018}{Lion 2018};
+\citeproc{ref-Lion2023}{Lion et al. 2023}). In this case,
+
+\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i} + 1\right)
+\tag{S1}
+\]
+
+Mathematically, the most general approach here would be to define
+individual growth as a function of the entire system \(\mathbf{E}\),
+\(r_{i}(\mathbf{E})\), where \(\mathbf{E}\) is a vector with elements
+including any parameters potentially relevant to \(r_{i}\). Taking this
+approach would recover a version of eqn 2 in Lion
+(\citeproc{ref-Lion2018}{2018}) and permit any relationship between the
+system and a focal individual's growth. Limiting our focus to the
+effects of other individuals (\(j\)) and assuming that the effects of
+these individuals are additive, let \(a_{ij|\cdot}\) be the effect of
+individual \(j\) on the growth rate attributable to \(i\) conditioned on
+all other individuals within the population such that
+\(r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)\) defines the
+realised growth rate of \(i\),
+
+\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij|\cdot} \right) + 1\right).
+\tag{S2}
+\]
+
+Assuming that individual effects of \(j\) on \(i\) are also independent,
+we can remove the condition,
+
+\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij} \right) + 1\right).\]
+
+Further assuming that all individuals have the same per capita effect
+such that \(a = a_{ij}\) for any \(i\) and \(j\) pair (as might be
+reasonable given resource competition in a well-mixed population),
+
+\[N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - a N_{t} \right) + 1\right).\]
+
+If \(r_{i}\) values are identical,
+
+\[N_{t+1} = N_{t} + r N_{t}\left(1 - a N_{t} \right).
+\tag{S3}
+\]
+
+Equation S3 therefore recovers a classic version of a discrete time
+logistic growth by making assumptions from an exact model of
+eco-evolutionary change.
+
+An alternative approach would be to define model the effects of an
+individual \(j\) on the fitness of \(i\) (\(w_{i}\)), thereby replacing
+eqn S1 with \(N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\) and replacing eqn S2
+with,
+
+\[N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}\alpha_{ij|\cdot} \right).
+\]
+
+Note that we have used \(\alpha_{ij}\) to represent the effect of \(j\)
+on the fitness of \(i\) for clarity in the sections below. By making the
+same assumptions of additivity, independence, and identical effects such
+that \(\alpha = \alpha_{ij}\) for all \(j\) on \(i\), and assuming
+fitness is equal (\(w_{i} = \bar{w}\)), we can derive,
+
+\[N_{t+1} = \bar{w}N_{t}(1 - \alpha N_{t}).
+\tag{S4}
+\]
+
+This is an alternative way to express logistic growth.
+
+\section{Multi-level selection}\label{multi-level-selection}
+
+We can recover multi-level selection from our eqn 1. Here we derive the
+original form of the multi-level Price (\citeproc{ref-Price1972}{1972})
+equation as it appears in eqn 3.1 of Lehtonen
+(\citeproc{ref-Lehtonen2020a}{2020}). Individuals belong to one of \(K\)
+total groups where \(j\) indexes \(K\) groups and \(i\) indexes
+individuals. Individuals do not overlap in group membership. The size of
+group \(j\) is denoted as \(N_{j}\). Equation S5 below uses summations
+to partition how individuals within each group contribute to \(\Omega\),
+
+\[\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}\left(\beta_{j,i} - \delta_{j,i} + 1 \right)\left(z_{j, i} + \Delta z_{j, i}\right).
+\tag{S5}
+\]
+
+In S5, indices \(\beta_{j, i}\), \(\delta_{j, i}\), and \(z_{j, i}\)
+identify individual \(i\) in group \(j\). We set
+\(w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1\), and for simplicity let
+\(\Delta z_{j, i} = 0\) (i.e., no transmission bias),
+
+\[\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}w_{j,i}z_{j, i}.\]
+
+For ease of presentation, with no loss of generality, we assume all
+group sizes are equal with a group size of \(N_{j} = n\) for all \(j\).
+If group sizes differ, then weighted expectations and covariances are
+instead needed (\citeproc{ref-Lehtonen2020a}{Lehtonen 2020}). Given
+equal group sizes, the total number of individuals (\(N\)) equals
+\(K \times n\), and,
+
+\[\frac{\Omega}{K n} = \left(\frac{1}{K}\right)\left(\frac{1}{n}\right)\sum_{j=1}^{K}\sum_{i=1}^{n}w_{j,i}z_{j, i}.\]
+
+Rearranging,
+
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\frac{1}{n}\sum_{i=1}^{n}w_{j,i}z_{j, i}.\]
+
+The inner summation can be rewritten as an expectation for group \(j\),
+
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i} z_{j,i}\right).\]
+
+As in the main text, we note
+\(\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)\).
+Defining \(Cov_{j}(w_{j}, z_{j})\) as the covariance between \(w_{j,i}\)
+and \(z_{j, i}\) for group \(j\),
+
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).\]
+
+We can separate the summation for each term,
+
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).\]
+
+Using the notation \(\bar{w}_{j} = E_{j}(w_{j,i})\) and
+\(\bar{z}_{j} = E_{j}(z_{j,i})\) to indicate the expectation in group
+\(j\),
+
+\[\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).\]
+
+We can also rewrite the first term on the right-hand side as an
+expectation,
+
+\[\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).\]
+
+We can rearrange the second term on the right-hand side
+(\(\bar{\bar{w}}\) indicates grand mean over all groups),
+
+\[\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.\]
+
+As in the main text, note that \(Kn\bar{\bar{w}}\) accounts for
+differences in total population size from \(t\) to \(t+1\), with
+\(\bar{\bar{w}}\) being mean fitness across all groups. We can therefore
+set \(\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}\), so,
+
+\[\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right).\]
+
+Because \(\Delta \bar{\bar{z}} = \bar{\bar{z'}} - \bar{\bar{z}}\),
+
+\[\bar{\bar{w}}\Delta \bar{\bar{z}} = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
+\tag{S6}
+\]
+
+This recovers the multi-level Price (\citeproc{ref-Price1972}{1972})
+equation (\citeproc{ref-Lehtonen2020a}{Lehtonen 2020}) from a starting
+point of eco-evolutionary change in different groups. Equation S6 can be
+found in Lehtonen (\citeproc{ref-Lehtonen2016}{2016}) B2.I, who then
+derives a multi-level selection version of Hamilton's rule predicting
+the evolution of altruism.
+
+\section{Integration of ecology and
+evolution}\label{integration-of-ecology-and-evolution}
+
+For simplicity, we now focus on showing an integration between ecology
+and evolution using a population with no multi-level selection and let
+\(\Delta z_{j, i} = 0\) (i.e., no transmission bias). As above in the
+section on density-dependent population growth, we define
+\(w_{i} = \beta_{i} - \delta_{i} + 1\) and use \(\alpha_{i,j}\) to
+represent the effect of \(j\) on the fitness of \(i\). Our starting
+equation is therefore,
+
+\[\Omega = \sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S7}
+\]
+
+We have already demonstrated that if we assume all individuals have the
+same effect on a focal individual such that \(\alpha = \alpha_{i,j}\)
+for all \(i\) and \(j\) pairs, we can recover equation S4 when
+\(z_{i} = 1\) and \(\Omega\) is therefore interpreted as the count of
+entities,
+
+\[N_{t+1} = \bar{w} N_{t} \left(1 - \alpha N_{t}\right).\]
+
+We now start from S7 to derive \(\Delta \bar{z}\). The objective is to
+use our definition of eco-evolutionary change to simultaneously recover
+how interactions between individuals affect population change and
+evolutionary change.
+
+We start by dividing both sides of S7 by \(N\),
+
+\[\frac{\Omega}{N} = \frac{1}{N}\sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
+\tag{S8}
+\]
+
+We can express the right-hand side of eqn S8 as an expectation,
+
+\[\frac{\Omega}{N} = \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}\right).
+\tag{S9}
+\]
+
+We can rewrite the right-hand side in terms of covariances,
+
+\[\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)\right)\mathrm{E}\left(z_{i}\right).
+\tag{S9}
+\]
+
+The expectations in the second term on the right-hand side can be
+replaced with overbars to represent the mean,
+
+\[\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i}. 
+\tag{S9}
+\]
+
+In the main text, we noted that \(\Omega = N\bar{w}\bar{z}'\). Here mean
+fitness incorporates individual interactions, therefore,
+
+\[\Omega = N\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}'.
+\tag{S10}
+\]
+
+We can therefore rewrite S9,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}' -  \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right).
+\tag{S11}
+\]
+
+Noting again \(\Delta \bar{z} = \bar{z}' - \bar{z}\),
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
+\tag{S12}
+\]
+
+We can rewrite the left-hand side of S12,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i} - w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S13}
+\]
+
+The covariance term can be split without any additional assumptions,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
+\tag{S14}
+\]
+
+If we are able to further assume that \(w_{i}\) and the summation over
+\(\alpha_{i,j}\) are independent, then we could rewrite,
+
+\[\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right)\bar{w_{i}}. 
+\tag{S15}
+\]
+
+Partitioning fitness into different components with the Price equation
+is commonplace. But this derivation highlights, e.g., the ecological and
+evolutionary relationship between nonsocial and social components of
+fitness, and population size. For example, the second term on the
+right-hand side of S13 shows the covariance between the sum of social
+interactions and a trait. When traits covary with the interaction
+between sociality and fitness, they will have a stronger effect on trait
+change. The magnitude of this second term will also increase with \(N\),
+which reflects the stronger effect of the interaction between sociality
+and fitness when there are more individuals interacting with a focal
+individual.
+
+\section{Supporting Information
+References}\label{supporting-information-references}
+
+\addcontentsline{toc}{section}{References}
+
+\phantomsection\label{refs}
+
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-Lehtonen2016}
+Lehtonen, Jussi. 2016. {``{Multilevel selection in kin selection
+language}.''} \emph{Trends in Ecology and Evolution} xx: 1--11.
+\url{https://doi.org/10.1016/j.tree.2016.07.006}.
+
+\bibitem[\citeproctext]{ref-Lehtonen2020a}
+---------. 2020. {``The Price Equation and the Unity of Social Evolution
+Theory.''} \emph{Philosophical Transactions of the Royal Society B:
+Biological Sciences} 375: 20190362.
+\url{https://doi.org/10.1098/rstb.2019.0362}.
+
+\bibitem[\citeproctext]{ref-Lion2018}
+Lion, Sébastien. 2018. {``{Theoretical approaches in evolutionary
+ecology: environmental feedback as a unifying perspective}.''}
+\emph{American Naturalist} 191 (1).
+\url{https://doi.org/10.1086/694865}.
+
+\bibitem[\citeproctext]{ref-Lion2023}
+Lion, Sébastien, Akira Sasaki, and Mike Boots. 2023. {``Extending
+Eco-Evolutionary Theory with Oligomorphic Dynamics.''} \emph{Ecology
+Letters} 26 (September): S22--46.
+\url{https://doi.org/10.1111/ele.14183}.
+
+\bibitem[\citeproctext]{ref-Price1972}
+Price, George R. 1972. {``{Extension of covariance selection
+mathematics}.''} \emph{Annals of Human Genetics} 35 (4): 485--90.
+\url{https://doi.org/10.1111/j.1469-1809.1957.tb01874.x}.
+
+\end{CSLReferences}
+
 \end{document}

### 2025-06-18 foundations_of_community_ecology.Rmd
Initialise commentary point

@@
 
+**IDEAS**
+
+Do we have an issue because *only* a trait of j's should affect both i's contribution to the sum population size and the sum trait quantity omega, right?
+The existence of j shouldn't matter, unless we're multiplying 1 by j's trait.
+But we shouldn't double count j's existence and its trait value to predict how it affects i (or anything about i).
+
+Or maybe the original formula isn't quite right. We need alpha to actually be imaginary. So maybe if we have 0 + i\alpha_{ij}, something like $(0 + i1)\alpha * (i\tau_{ij})$
+
+Well, could try to have an individual with the property $(1 + 0.5i)$ and define ($\alpha = 1 + 0.1i$, but something seems off here.
+Then again, if we have fitness as 2, then set an individual and trait to $(1 + 0.5i)$, affected by competition of a j individual, $a_{ij} = (0 + 0.5i)$.
+I'm not sure.
+I can't get the density to change while also getting the trait to change.
+Unless it's something like $z = 1 + 0.5i$, then $a_{ij} = (0 + 0.1i)$, so $z + za$.
+
+One last try before bed, if w = 2, let's say, $2 (1 + 0.5i) * (1 - (0 - 0i) * (1 + 0.4i))$, where $\alpha = (0 - 0i)$ and j is $(1 + 0.4i)$.
+But alpha is the effect of a *trait* on fitness, so something isn't quite right here.
+
+Okay, try making a_{ij} into an imaginary number. This should work.
+
+Deriving Price like
+=====================
+
+$$\frac{\Omega}{N} = \underbrace{\mathbb{E}[w z]}_{\text{Term 1}} + \underbrace{\mathbb{E}[w \Delta z]}_{\text{Term 2}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N a_{ij} z_i^2 \right]}_{\text{Term 3}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N \left( 2 a_{ij} z_i \Delta z_i + a_{ij} (\Delta z_i)^2 \right) \right]}_{\text{Term 4}}.$$
+
+Then,
+
+$$\bar{w} \bar{z}' = \underbrace{\text{Cov}(w, z) + \bar{w} \bar{z}}_{\text{Term 1}} + \underbrace{\mathbb{E}[w \Delta z]}_{\text{Term 2}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N a_{ij} z_i^2 \right]}_{\text{Term 3}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N \left( 2 a_{ij} z_i \Delta z_i + a_{ij} (\Delta z_i)^2 \right) \right]}_{\text{Term 4}}.$$
+
+Then,
+
+$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - \mathbb{E}\left[w \sum_{j=1}^N a_{ij} z_i^2 \right] - \mathbb{E}\left[w \sum_{j=1}^N \left( 2 a_{ij} z_i \Delta z_i + a_{ij} (\Delta z_i)^2 \right) \right].$$
+If we are willing to assume that all individuals $j$ have the same effect on a focal $i$, then we can rewrite as,
+
+$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}[w \alpha_i z_i^2] - N \mathbb{E}\left[w \alpha_i \left( 2 z_i \Delta z_i + (\Delta z_i)^2 \right) \right].$
+
+This simplifies again,
+
+$$\begin{align*}
+\bar{w} \Delta \bar{z} &= \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}\left[ w \alpha_i \left( z_i^2 + 2 z_i \Delta z_i + (\Delta z_i)^2 \right) \right] \\
+&= \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}\left[ w \alpha_i (z_i + \Delta z_i)^2 \right].
+\end{align*}$$
+
+$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}\left[ w \alpha_i (z_i + \Delta z_i)^2 \right].$$
+Let's assume that $\Delta z_{i} = 0$, where the middle term will drop out and we can focus on population level effects on trait change (selection and drift) and the effect of interaction coefficients,
+
+$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) - N \mathbb{E}[w \alpha_i z_i^2].$$
+
+Let's just drop the subscripts in the second term as in the first,
+
+$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) - N \mathbb{E}[w \alpha z^2].$$
+
+It's reassuring that the sign of z doesn't affect the second term. Verbally, 
+I *think* the first term represents the change in a trait attributable to selection (and drift) in the absence of interspecific interactions, while the second term represents trait change attributable to selection attributable to interactions?
+
+Note that the second term is the expected interaction between fitness and interaction strength $w \alpha$ times $z^2$, which embeds the variance of $z^{2}$ because,
+
+$$E[z_{i}^{2}] = Var(z) + \bar{z}^{2}$$
+
 References 

### 2025-06-18 foundations_of_community_ecology.Rmd
Basic structure to later be written out

--- 2024-12-20 foundations_of_community_ecology.Rmd
@@
 ---
-title: "Foundations of community ecology"
+title: "Foundations of community ecology: Derivation of community Price equation"
 author: |
-  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
-date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
+  | UNICOP Group
+date: '18 JUN 2025'
 output:
@@
 
-We have discovered the fundamental equation that provides a complete description of eco-evolutionary change in any system,
-
-$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
-\tag{1}
-$$
-
-In our previous efforts, we derived both the Price equation and the birth-death model from the above.
-Here we can derive a fundamental model of community ecology, a discrete time generalised Lotka-Volterra equations leading to modern coexistence theory.
-We can do this by partitioning the sum total change $\Omega$ into change caused by density-independent and density-dependent effects,
-
-$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\left(1 - \sum_{j = 1}^{N}\left(a_{ij}\left(z_{i} + \Delta z_{i} \right) \right) \right).
-\tag{2}
-$$
-
-In eqn 2, $a_{ij}$ is the effect that $j$ has on the contribution of $i$ to $\Omega$.
-Note that we are, again, not relaxing an assumption in eqn 1, nor are we adding something that did not already exist to the model.
-Equation 1 is already a complete description of eco-evolutionary change, so it included both density-independent and density-dependent effects. 
-The new expression in parentheses is only splitting the total change into density-independent versus density-dependent effects.
-As with the previous case of ecology, we can set $z_{i} = 1$ and $\Delta z_{i} = 0$, which allows us to simplify,
-
-$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
-\tag{3}
-$$
-
-Now, assume that every individual $j$ in the population has an identical effect on $i$, such that $a_{ij} = \alpha$.
-With the conceptual understanding that $\alpha$ is then the per capita effect of $j$ on $i$, we can rewrite eqn 3,
-
-$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}\alpha \right).
-$$
-
-Because we then sum up $\alpha$ a total of $N$ times, it then follows,
-
-$$\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \alpha N \right).
-$$
-
-Now, we can again assume that individual births ($b = \beta_{i}$) and deaths ($d = \delta_{i}$) are identical for all $i$,
-
-$$\Omega = \sum_{i=1}^{N} \left(b -d + 1 \right)\left(1 - \alpha N \right).
-$$
-
-Because we define $\lambda = b - d + 1$,
-
-$$\Omega = \sum_{i=1}^{N} \lambda\left(1 - \alpha N \right).
-$$
-
-We now have removed all subscripts $i$ from the right-hand side, and we can rewrite,
-
-$$\Omega =  N \lambda\left(1 - \alpha N \right).
-$$
-
-Now is a good time to define $\Omega$ again and add subscripts to $N$,
-
-$$N_{t+1} =  N_{t} \lambda\left(1 - \alpha N_{t} \right).
-\tag{4}
-$$
-
-Equation 4 gives us a discrete time logistic growth equation. 
-This is not typically the way a discrete time equation is expressed, which made me a bit nervous at first, but there is nothing wrong with it either.
-
-We can do some simple simulations to confirm that it behaves as it should.
-I am going to make a lot of common simplifying assumptions in what follows (rather, I already have in eqn 4).
-I assume that $\alpha$ and $\lambda$ are constant and do not change with density.
-Suppose then that we have a population of just two individuals, $N_{t} = 2$. 
-With density independent effects, assume that the population is doubling from $t$ to $t + 1$, so $\lambda = 2$.
-Now assume that the effect of each individual in the population is to decrease each other individual's growth by $\alpha = 0.01$.
-We then have $N_{t+1} =  2 N_{t} \left(1 - 0.01 N_{t} \right)$.
-We can simulate this over 10 time steps.
-
-```{r}
-NN <- 2;
-LL <- 2;
-aa <- 0.01;
-tt <- 10;
-i  <- 1;
-while(i < tt){
-    Nt        <- NN[i]*LL*(1 - aa * NN[i]);
-    NN[i + 1] <- Nt;
-    i         <- i + 1;
-}
-print(NN);
-plot(x = 1:10, y = NN, type = "b", pch = 20, lwd = 2,
-     xlab = "Time step",
-     ylab = "Population density (N)");
-points(x = 1:10, y = NN, type = "l", pch = 20, lwd = 2);
-```
-
-The population increases exponentially, then levels off at $N = 50$, at which point the population growth $N\lambda = 2(50) = 100$ is balanced by the total effect of competition $\alpha N = 50^{2}(2)(0.01) = 50$, such that $100 - 50$ keeps the population at $50$.
-
-I was a bit bothered by the unusual formulation at first, but now I am feeling better about it.
-I am convinced it is correct, albeit not the typically way that researchers tend to model population change.
-But usually modellers will switch to continuous time models for logistic growth and generalised Lotka-Volterra. 
-And for discrete growth, they will usually do something like $N_{t+1} = N_{t} + N_{t}\lambda(1 - N_{t}/K)$, where $K$ is the carrying capacity.
-But this is effort is not derived from first principles, I do not think, in the same way as the above eqn 4.
-
-Recovering proper logistic growth
-=======================================
-
-Note that we can also recover a different logistic growth equation if we start at eqn 3, but then assume that $r = b - d$, and density dependent effects only apply to birth and death,
-
-$$\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
-\tag{5}
-$$
-
-This seems more reasonable.
-The effect of an individual $j$ on a focal individual $i$ has to be either on birth or death.
-The 1 is a constant, and I do not think it should be affected.
-Mathematically, we can expand eqn 5.
-
-$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + \left(1 - \sum_{j = 1}^{N}a_{ij} \right).$$
-
-Biologically, there is a clear way to resolve this, though I think some caution is needed here with the mathematics.
-It is biologically reasonable that $a_{ij} = 0$ in the second term for all $j$.
-If $a_{ij}$ only applies to $r_{i}$ such that $a_{ij} = 0$ for all $i$ and $j$ in the second term, then,
-
-$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + 1.$$
-
-If we can get to this point, then we can make the same assumption that we did around eqn 3, such that every individual $j$ has an identical effect on $i$, so $a_{ij} = \alpha$.
-It then follows,
-
-$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}\alpha \right) + 1.$$
-
-From here, we then sum up the $\alpha$ values,
-
-$$\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \alpha N \right) + 1.$$
-
-If we assume that $r_{i} = r$ for all $i$, then we are also summing $r(1 - \alpha N) + 1$ a total of $N$ times,
-
-$$\Omega = N \left( r_{i}\left(1 - \alpha N \right) + 1\right).$$
-
-By setting $\Omega = N_{t+1}$ and $N = N_{t}$, we can then recover,
-
-$$N_{t+1} = N_{t} + rN_{t}\left(1 - \alpha N_{t}\right)$$
-
-Again, in this case, $\alpha$ reflects interactions among individuals only affecting birth and death.
-This is a standard discrete time generalised Lotka-Volterra equation, as appears in @Chesson2000b.
-To recover community ecology, it should be a standard matter of defining two sets $J$ and $K$ representing different species,
-
-$$\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j \in J}a_{ij} - \sum_{k \in K}a_{ik} \right).$$
-
-Summations can be made over different scales too, but I think that this gets us solidly to @Chesson2000b.
-
 References 

### 2025-06-18 foundations_of_community_ecology.tex
Basic structure to later be written out

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Foundations of community ecology},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx}
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 \usepackage{fancyhdr}
@@
 \fi
+\usepackage{bookmark}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Foundations of community ecology: Derivation of community Price equation},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
 
-\title{Foundations of community ecology}
-\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
-\date{{[}1{]} Department of Biological and Environmental Sciences,
-University of Stirling, Stirling, Scotland {[}2{]} Department of
-Philosophy, University of Valencia, Valencia, Spain {[}*{]}
-Corresponding author:
-\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
-{[}a{]} Equal contribution}
+\title{Foundations of community ecology: Derivation of community Price
+equation}
+\author{UNICOP Group}
+\date{18 JUN 2025}
 
@@
 
-We have discovered the fundamental equation that provides a complete
-description of eco-evolutionary change in any system,
-
-\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right).
-\tag{1}
-\]
-
-In our previous efforts, we derived both the Price equation and the
-birth-death model from the above. Here we can derive a fundamental model
-of community ecology, a discrete time generalised Lotka-Volterra
-equations leading to modern coexistence theory. We can do this by
-partitioning the sum total change \(\Omega\) into change caused by
-density-independent and density-dependent effects,
-
-\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\left(1 - \sum_{j = 1}^{N}\left(a_{ij}\left(z_{i} + \Delta z_{i} \right) \right) \right).
-\tag{2}
-\]
-
-In eqn 2, \(a_{ij}\) is the effect that \(j\) has on the contribution of
-\(i\) to \(\Omega\). Note that we are, again, not relaxing an assumption
-in eqn 1, nor are we adding something that did not already exist to the
-model. Equation 1 is already a complete description of eco-evolutionary
-change, so it included both density-independent and density-dependent
-effects. The new expression in parentheses is only splitting the total
-change into density-independent versus density-dependent effects. As
-with the previous case of ecology, we can set \(z_{i} = 1\) and
-\(\Delta z_{i} = 0\), which allows us to simplify,
-
-\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
-\tag{3}
-\]
-
-Now, assume that every individual \(j\) in the population has an
-identical effect on \(i\), such that \(a_{ij} = \alpha\). With the
-conceptual understanding that \(\alpha\) is then the per capita effect
-of \(j\) on \(i\), we can rewrite eqn 3,
-
-\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}\alpha \right).
-\]
-
-Because we then sum up \(\alpha\) a total of \(N\) times, it then
-follows,
-
-\[\Omega = \sum_{i=1}^{N} \left(\beta_{i} - \delta_{i} + 1 \right)\left(1 - \alpha N \right).
-\]
-
-Now, we can again assume that individual births (\(b = \beta_{i}\)) and
-deaths (\(d = \delta_{i}\)) are identical for all \(i\),
-
-\[\Omega = \sum_{i=1}^{N} \left(b -d + 1 \right)\left(1 - \alpha N \right).
-\]
-
-Because we define \(\lambda = b - d + 1\),
-
-\[\Omega = \sum_{i=1}^{N} \lambda\left(1 - \alpha N \right).
-\]
-
-We now have removed all subscripts \(i\) from the right-hand side, and
-we can rewrite,
-
-\[\Omega =  N \lambda\left(1 - \alpha N \right).
-\]
-
-Now is a good time to define \(\Omega\) again and add subscripts to
-\(N\),
-
-\[N_{t+1} =  N_{t} \lambda\left(1 - \alpha N_{t} \right).
-\tag{4}
-\]
-
-Equation 4 gives us a discrete time logistic growth equation. This is
-not typically the way a discrete time equation is expressed, which made
-me a bit nervous at first, but there is nothing wrong with it either.
-
-We can do some simple simulations to confirm that it behaves as it
-should. I am going to make a lot of common simplifying assumptions in
-what follows (rather, I already have in eqn 4). I assume that \(\alpha\)
-and \(\lambda\) are constant and do not change with density. Suppose
-then that we have a population of just two individuals, \(N_{t} = 2\).
-With density independent effects, assume that the population is doubling
-from \(t\) to \(t + 1\), so \(\lambda = 2\). Now assume that the effect
-of each individual in the population is to decrease each other
-individual's growth by \(\alpha = 0.01\). We then have
-\(N_{t+1} = 2 N_{t} \left(1 - 0.01 N_{t} \right)\). We can simulate this
-over 10 time steps.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{NN }\OtherTok{\textless{}{-}} \DecValTok{2}\NormalTok{;}
-\NormalTok{LL }\OtherTok{\textless{}{-}} \DecValTok{2}\NormalTok{;}
-\NormalTok{aa }\OtherTok{\textless{}{-}} \FloatTok{0.01}\NormalTok{;}
-\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{10}\NormalTok{;}
-\NormalTok{i  }\OtherTok{\textless{}{-}} \DecValTok{1}\NormalTok{;}
-\ControlFlowTok{while}\NormalTok{(i }\SpecialCharTok{\textless{}}\NormalTok{ tt)\{}
-\NormalTok{    Nt        }\OtherTok{\textless{}{-}}\NormalTok{ NN[i]}\SpecialCharTok{*}\NormalTok{LL}\SpecialCharTok{*}\NormalTok{(}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ aa }\SpecialCharTok{*}\NormalTok{ NN[i]);}
-\NormalTok{    NN[i }\SpecialCharTok{+} \DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ Nt;}
-\NormalTok{    i         }\OtherTok{\textless{}{-}}\NormalTok{ i }\SpecialCharTok{+} \DecValTok{1}\NormalTok{;}
-\NormalTok{\}}
-\FunctionTok{print}\NormalTok{(NN);}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1]  2.000000  3.920000  7.532672 13.930521 23.979854 36.459040 46.332848
-##  [8] 49.731040 49.998553 50.000000
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{plot}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10}\NormalTok{, }\AttributeTok{y =}\NormalTok{ NN, }\AttributeTok{type =} \StringTok{"b"}\NormalTok{, }\AttributeTok{pch =} \DecValTok{20}\NormalTok{, }\AttributeTok{lwd =} \DecValTok{2}\NormalTok{,}
-     \AttributeTok{xlab =} \StringTok{"Time step"}\NormalTok{,}
-     \AttributeTok{ylab =} \StringTok{"Population density (N)"}\NormalTok{);}
-\FunctionTok{points}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10}\NormalTok{, }\AttributeTok{y =}\NormalTok{ NN, }\AttributeTok{type =} \StringTok{"l"}\NormalTok{, }\AttributeTok{pch =} \DecValTok{20}\NormalTok{, }\AttributeTok{lwd =} \DecValTok{2}\NormalTok{);}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{foundations_of_community_ecology_files/figure-latex/unnamed-chunk-1-1.pdf}
-
-The population increases exponentially, then levels off at \(N = 50\),
-at which point the population growth \(N\lambda = 2(50) = 100\) is
-balanced by the total effect of competition
-\(\alpha N = 50^{2}(2)(0.01) = 50\), such that \(100 - 50\) keeps the
-population at \(50\).
-
-I was a bit bothered by the unusual formulation at first, but now I am
-feeling better about it. I am convinced it is correct, albeit not the
-typically way that researchers tend to model population change. But
-usually modellers will switch to continuous time models for logistic
-growth and generalised Lotka-Volterra. And for discrete growth, they
-will usually do something like
-\(N_{t+1} = N_{t} + N_{t}\lambda(1 - N_{t}/K)\), where \(K\) is the
-carrying capacity. But this is effort is not derived from first
-principles, I do not think, in the same way as the above eqn 4.
-
-\hypertarget{recovering-proper-logistic-growth}{%
-\section{Recovering proper logistic
-growth}\label{recovering-proper-logistic-growth}}
-
-Note that we can also recover a different logistic growth equation if we
-start at eqn 3, but then assume that \(r = b - d\), and density
-dependent effects only apply to birth and death,
-
-\[\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j = 1}^{N}a_{ij} \right).
-\tag{5}
-\]
-
-This seems more reasonable. The effect of an individual \(j\) on a focal
-individual \(i\) has to be either on birth or death. The 1 is a
-constant, and I do not think it should be affected. Mathematically, we
-can expand eqn 5.
-
-\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + \left(1 - \sum_{j = 1}^{N}a_{ij} \right).\]
-
-Biologically, there is a clear way to resolve this, though I think some
-caution is needed here with the mathematics. It is biologically
-reasonable that \(a_{ij} = 0\) in the second term for all \(j\). If
-\(a_{ij}\) only applies to \(r_{i}\) such that \(a_{ij} = 0\) for all
-\(i\) and \(j\) in the second term, then,
-
-\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}a_{ij} \right) + 1.\]
-
-If we can get to this point, then we can make the same assumption that
-we did around eqn 3, such that every individual \(j\) has an identical
-effect on \(i\), so \(a_{ij} = \alpha\). It then follows,
-
-\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \sum_{j = 1}^{N}\alpha \right) + 1.\]
-
-From here, we then sum up the \(\alpha\) values,
-
-\[\Omega = \sum_{i=1}^{N} r_{i}\left(1 - \alpha N \right) + 1.\]
-
-If we assume that \(r_{i} = r\) for all \(i\), then we are also summing
-\(r(1 - \alpha N) + 1\) a total of \(N\) times,
-
-\[\Omega = N \left( r_{i}\left(1 - \alpha N \right) + 1\right).\]
-
-By setting \(\Omega = N_{t+1}\) and \(N = N_{t}\), we can then recover,
-
-\[N_{t+1} = N_{t} + rN_{t}\left(1 - \alpha N_{t}\right)\]
-
-Again, in this case, \(\alpha\) reflects interactions among individuals
-only affecting birth and death. This is a standard discrete time
-generalised Lotka-Volterra equation, as appears in Chesson
-(\protect\hyperlink{ref-Chesson2000b}{2000}). To recover community
-ecology, it should be a standard matter of defining two sets \(J\) and
-\(K\) representing different species,
-
-\[\Omega = \sum_{i=1}^{N} \left(r_{i} + 1 \right)\left(1 - \sum_{j \in J}a_{ij} - \sum_{k \in K}a_{ik} \right).\]
-
-Summations can be made over different scales too, but I think that this
-gets us solidly to Chesson (\protect\hyperlink{ref-Chesson2000b}{2000}).
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-Chesson2000b}{}}%
-Chesson, Peter L. 2000. {``{Mechanisms of maintenance of species
-diversity}.''} \emph{Annual Review of Ecology and Systematics} 31:
-343--66.
-
-\end{CSLReferences}
+\section{References}\label{references}
 

### 2025-08-04 foundations_of_ecoevo.Rmd
Some notes before a new update

@@
 
+<!---
+
+NOTE: ADD CASE AND TAPER (2000) AS A REFERENCE FOR Delta N = N bar(w)
+
+--->
+
 Abstract
@@
 
-Something around here verbally presenting the logic of eqn 1, discussing how we are talking about an interaction between birth-death (individual fitness) and a individual's characteristic.
+There's an odd technicality if the time between t and t + 1 is large.
+I defined beta as 'births attributable to i', which is correct, but maybe it would be more correct to define it as 'descendents of i'.
+Then, delta needs to be more technically 'deaths of i and its descendents').
+Nothing changes in effect, but now if a focal individual i has a bunch of descendents that die between t and t + 1, the correct fitness is still preserved.
 
-Also need to explain units of Omega, since there is confusion here.
+Put this after the sentence about Omega?
 
@@
 To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
-In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
-In this case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
+In other words, we set $z_{i} = 1$ to simply indicate that $i$ is one member of the population. <!--- and therefore contributes one unit toward the total population size. --->
+In this restricted case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
 Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
+<!--- Further, we assume that individual membership and contribution to population size does not change by...
+--->
 In this case,
@@
 
-Summing from 1 to $N$, we can rewrite the above,
+Summing from 1 to $N_{t}$, we can rewrite the above,
 
-$$\Omega = N\left(1 + \beta - \delta \right).$$
+$$N_{t+1} = N_{t}\left(1 + \beta - \delta \right).$$
 
@@
 
-We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [eqn 3, @Turchin2001].
+We therefore recover the general equation for population ecology (eqn 2) and the fundamental tendency for populations to grow (or decline) exponentially [eqn 3, @Turchin2001].
 
@@
 
-$$\Omega = N\bar{w}\bar{z}'
+$$\Omega = N\bar{w}\bar{z}'.
 \tag{8}
@@
 
-In some cases, $\Omega$ could also be intepreted as the total contribution of a population to ecosystem function.
+In some cases, $\Omega$ could also be interpreted as the total contribution of a population to ecosystem function.
 This is restricted to cases in which $z$ is a characteristic defining an absolute quantity measured at the whole organism level such as biomass, seed production, carbon capture, flower visits, or nutrient consumption.
-In such cases, the sum across individuals gives a meaningful total quantity for the population.
+In such cases, the sum across individuals gives a meaningful total quantity for the population (e.g., the total biomass or seeds produced in the population).
 When $z$ is instead defined by relative organism-level measurements such wing loading, nutrient ratio, or diet composition, or when $z$ is measured at a level of biological organisation below the organism (e.g., average cell volume or leaf surface area), $\Omega$ does not have a clear population-level interpretation.
+In such cases, eqn 1 still defines eco-evolutionary change; the interpretation of $\Omega$ by itself is just not as interesting, biologically.
 
-Box 3 provides a toy example of three plants with different fitnesses and carbon captures.
-We use this toy example to demonstrate a concrete example of our proposed framework.
+Box 3 provides an instructive example of three plants with different fitnesses and fruit masses.
+We use this to demonstrate a concrete example of our proposed framework.
 
@@
 Rather than making simplifying assumptions, as is the approach for specific ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
-We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
-From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 2) and principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
-By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 1).
+We define an abstract sum ($\Omega$), to which all individuals in the population contribute.
+From the basic axiom that each individual is one member of a specific population ($z_{i} = 1$ and $\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 2) and the principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
+By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution, the Price equation (Box 1).
 Our eqn 1 thereby provides a foundation for defining eco-evolutionary change in any population.
@@
 It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [e.g., individuals, alleles, @Price1970; @Gardner2008; @Luque2016].
-This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation (e.g., group, individual) [@Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
+This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation [e.g., group, individual, @Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
 Because of its abstract nature and lack of any system-specific assumptions, the Price equation is not dynamically sufficient and makes no predictions about what will happen in any particular system [@Gardner2020].
@@
 The population growth equation $N_{t+1} = N_{t}\lambda$ can always be rewritten as $N_{t+1} = N_{t}\bar{w}$.
-This specific equivalence has been proposed before [e.g., @Lande1976], as has the broader relationship between population growth rate and evolutionary fitness [@Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
+This specific equivalence has been proposed before [e.g., @Lande1976], as has the broader relationship between population growth rate and evolutionary fitness [e.g., @Fisher1930; @Charlesworth1980; @Lande1982; @Case2000; @Roff2008; @Lion2018].
 We show this from first principles and clarify the relationship between fitness and population growth.
 Over an arbitrary length of time, fitness is properly defined as $w_{i} = \beta_{i} - \delta_{i} + 1$.
+<!--- For simplicity, we defined beta and delta as births and deaths, but we could more generally define beta as the count of descendents of i, and delta as the count of deaths in i and all of its descendents --->
 Over an individual's lifetime (which, by definition, includes death), fitness is therefore $\beta_{i}$. 
@@
 
+<!---
+
+Over 120 years ago, @Needham1904 described natural history as "the study of the phenomena of fitness".
+Fundamentally, we show why eco-evolutionary change is a statistical interaction between fitness and an individual characteristic.
+From this definition, we can recover both population size change and evolutionary change.
+Our $\Omega$ can be therefore be interpreted as characteristic-scaled population change, with the special case of $z = 1$ recovering population size change by recognising that each individual contributes one to the total.
+
+--->
+
 Acknowledgements
@@
 Victor J. Luque would also like to thank Lorenzo Baravalle, Pau Carazo, Santiago Ginnobili, Manuel Serra, and Ariel Roff$\'{e}$.
+Lastly, we thank two anonymous reviewers.
 
@@
 ====================================================
+
+<!---
+
+Note, for blog and appendix: Eqn 1 constrains ecology and evolution.
+What if we want a model in which there is a strong density-dependent effect of conspecifics but not evolutionary effect on traits?
+This means that the covariance between z and w must be low.
+But social effects cannot decrease another conspecific's fitness in a population without also decreasing population growth.
+Covariance between fitness and a trait implies that there is variance in fitness, which should slow population growth?
+
+--->

### 2025-08-04 foundations_of_ecoevo.Rmd
Line 99 reviewer suggested edit. The w was just missing here, so this was a typo

@@
 On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction.
-The second term is the expected value of $\Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012].
+The second term is the expected value of $w \Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012].
 A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection.
@@
 But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided.
-
-<!---
-
-Previous theory has produced mathematical synthesis between ecological and evolutionary equations [@Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
-For example, @Page2002 demonstrate a mathematical equivalence between the Price equation and a general equation for community dynamics. 
-In both equations, the relative frequencies of different ecological (species) or evolutionary (phenotype) variants map to their respective ecological (population growth) or evolutionary (fitness) change in a way that is mathematically equivalent.
-@Collins2009 use the Price equation to build a general quantitative framework that partitions phenotypic change into ecological, evolutionary, and physiological components.
-The average change in a community-wide trait (e.g., carbon uptake) is exactly and generally partitioned into change attributable to changes in species composition, expected evolution of lineages within species, and expected change in lineage physiology across all species.
-Both @Page2002 and @Collins2009 use relative species abundances, meaning that species frequencies must sum to one.
-This is useful for demonstrating mathematical equivalence [@Page2002] or expected change in a community-wide trait [@Collins2009], but a different approach is needed to recover absolute change in population size and thereby recover a joint foundation for ecology and evolution.
-
---->
 

### 2025-08-04 foundations_of_ecoevo.Rmd
Line 165-167 address the comment about the description of Omega, which reviewer points out can more simply be described as the sum of character values

@@
 The total number of individuals in the population is $N$.
-From this foundation, we can define $\Omega$ to be a summed characteristic across $N$ entities,
+From this foundation, we can define $\Omega$ to be the sum of character values in $t+1$,
 
@@
 This foundation of eco-evolutionary change is a statistical interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
-The value of $\Omega$ is a summation of individual characteristics and takes the same units as $z$.
+The value of $\Omega$ takes the same units as $z$.
 From eqn 1, we can derive the most fundamental equations of population ecology and evolutionary biology through an appropriate interpretation of $z$.

### 2025-08-04 foundations_of_ecoevo.Rmd
Lines 165-167 changing things to more directly define the average characteristic of an individuals offspring, but also more formally defining beta and delta for arbitrary lengths of time steps

@@
 Our framework is general enough that entities can be anything discrete, but we will focus on each entity $i$ as an individual organism.
-Individuals give rise to new individuals through birth such that $\beta_{i}$ is the number of births attributable to $i$.
-Individuals are removed from the population through death such that $\delta_{i}$ is an indicator variable that takes a value of 1 (death of $i$) or 0 (persistence of $i$).
+Change occurs between the current time step $t$ and a future time step $t + 1$, and time steps can be arbitrarily short or long.
+Let $\beta_{i}$ be the count of direct descendants of $i$ at $t + 1$ (e.g., offspring if a time step is a single generation).
+Similarly, let $\delta_{i}$ be the count of deaths summed across $i$ and any of its descendants from $t$ to $t+1$.
+For example, if $i$ and all of its descendants persist at $t + 1$, then $\delta_{i} = 0$, or if a time step is a single non-overlapping generation, then $\delta_{i} = 1$.
 All individuals are defined by some characteristic $z_{i}$, and $\Delta z_{i}$ defines any change in $z_{i}$ from one time step $t$ to the next $t + 1$.
+The value $z_{i} + \Delta z_{i}$ is the average characteristic across extant individuals and their descendants at $t+1$.
 The total number of individuals in the population is $N$.
-From this foundation, we can define $\Omega$ to be the sum of character values in $t+1$,
+From this foundation, we can define $\Omega$ to be the sum of characteristic values in $t+1$,
 
@@
 
-<!---
+The value $\Omega$ will have the same units as $z$.
 
-There's an odd technicality if the time between t and t + 1 is large.
-I defined beta as 'births attributable to i', which is correct, but maybe it would be more correct to define it as 'descendents of i'.
-Then, delta needs to be more technically 'deaths of i and its descendents').
-Nothing changes in effect, but now if a focal individual i has a bunch of descendents that die between t and t + 1, the correct fitness is still preserved.
-
-Put this after the sentence about Omega?
-
---->
-
-This foundation of eco-evolutionary change is a statistical interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
-The value of $\Omega$ takes the same units as $z$.
+The foundation of eco-evolutionary change defined by eqn 1 is therefore a statistical interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and the average characteristic of an individual's offspring ($z_{i} + \Delta z_{i}$).
 From eqn 1, we can derive the most fundamental equations of population ecology and evolutionary biology through an appropriate interpretation of $z$.
 Under more limited interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.
-
-<!---
-
-19 MAY: Point out here that by itself, Omega has the same terms as z.
-
-- Verbally, our equation shows interaction between fitness and an individual's characteristic, summed
-- Because it is summing across individuals, some characteristics can be interpreted as measure of ecosystem function
-- Ecosystem function will make the assumption of additivity of z_i across all N
-- We will see that mean fitness is population ecology
-- Variance is evolutionary change
-- Summation is ecosystem function
-
-Eqn 1 captures an invariant relationship between individual fitness and an individual's characteristic
-
-Also note that directly tying fitness to demography
-
---->
 

### 2025-08-04 foundations_of_ecoevo.Rmd
Line 182 issue when moving to the assumptions of population ecology, as the reviewer pointed out that there is an inconsistency with the definition of delta as an indicator function (which we're no longer defining it as, but this was unnecessary anyway). Fixing equation numbers and removing some comments

@@
 To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
-In other words, we set $z_{i} = 1$ to simply indicate that $i$ is one member of the population. <!--- and therefore contributes one unit toward the total population size. --->
+In other words, we set $z_{i} = 1$ to simply indicate that $i$ is one member of the population and therefore contributes one unit to the total population size.
 In this restricted case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
-Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
-<!--- Further, we assume that individual membership and contribution to population size does not change by...
---->
+Further, we assume that individual membership and the unit contribution to population size does not change by setting $\Delta z_{i} = 0$
 In this case,
@@
 
-If we assume that individuals are identical, then we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$ and note that $N\beta$, $N\delta$, and $N$ are the total births, total deaths, and size of the population at $t$, respectively,
-
-Summing from 1 to $N_{t}$, we can rewrite the above,
-
-$$N_{t+1} = N_{t}\left(1 + \beta - \delta \right).$$
-
-If we define $\lambda = 1 + \beta - \delta$ (Box 2), then we can rewrite,
-
-$$N_{t+1} = N_{t}\lambda.
-\tag{3}
-$$
-
-We therefore recover the general equation for population ecology (eqn 2) and the fundamental tendency for populations to grow (or decline) exponentially [eqn 3, @Turchin2001].
+Using the classical assumptions of population ecology [Box 2, @Gotelli2001], we can recover the fundamental tendency for populations to grow (or decline) exponentially [eqn 3, @Turchin2001].
 
@@
 $$w_{i} = \beta_{i} - \delta_{i} + 1.
-\tag{4}
+\tag{3}
 $$
@@
 By definition, an individual that survives from $t$ to $t + 1$ has a higher fitness than one that dies, even if both have the same reproductive output.
-With this definition of fitness (eqn 4), we substitute,
+With this definition of fitness (eqn 3), we substitute,
 
 $$\Omega = \sum_{i=1}^{N} \left(w_{i}z_{i} + w_{i}\Delta z_{i} \right).
+\tag{4}
+$$
+
+We can break eqn 4 down further and multiply each side by $1/N$,
+
+$$\frac{1}{N}\Omega = \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}z_{i} \right) + \frac{1}{N}\sum_{i=1}^{N}\left( w_{i}\Delta z_{i} \right).
 \tag{5}
@@
 
-We can break eqn 5 down further and multiply each side by $1/N$,
+We can rewrite the terms on the right-hand side of eqn 5 as expected values and remove the subscripts,
 
-$$\frac{1}{N}\Omega = \frac{1}{N}\sum_{i=1}^{N} \left(w_{i}z_{i} \right) + \frac{1}{N}\sum_{i=1}^{N}\left( w_{i}\Delta z_{i} \right).
+$$\frac{1}{N}\Omega = \mathrm{E}\left(w z \right) + \mathrm{E}\left( w \Delta z  \right).
 \tag{6}
@@
 
-We can rewrite the terms on the right-hand side of eqn 6 as expected values and remove the subscripts,
-
-$$\frac{1}{N}\Omega = \mathrm{E}\left(w z \right) + \mathrm{E}\left( w \Delta z  \right).
-\tag{7}
-$$
-
 Now we must consider the total conservation of probability [@Frank2015; @Frank2016].
-In eqn 7, $\Omega$ is the total sum trait values ($z_{i}$) across the entire population at $t + 1$ divided by the number of individuals ($N$) in the population at $t$.
+In eqn 6, $\Omega$ is the total sum trait values ($z_{i}$) across the entire population at $t + 1$ divided by the number of individuals ($N$) in the population at $t$.
 But the size of the population can change from $t$ to $t + 1$.
 To recover mean trait change for the Price equation (and therefore conserve total probability), we need to account for this change in population size.
-The left-hand side of eqn 7 describes contributions to the sum trait value from the new population at $t + 1$.
+The left-hand side of eqn 6 describes contributions to the sum trait value from the new population at $t + 1$.
 But we cannot treat $\Omega/N$ as the mean of $z$ at $t+1$ ($\bar{z}'$) because we need to weight $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.
@@
 $$\Omega = N\bar{w}\bar{z}'.
-\tag{8}
+\tag{7}
 $$
@@
 This is consistent with the population ecology derivation from the previous section where $z_{i} = 1$ by definition, and $\Omega = N_{t+1}$ (note $\lambda = \bar{w}$).
-We can therefore rewrite eqn 7,
+We can therefore rewrite eqn 6,
 
 $$\bar{w}\bar{z}' = \mathrm{E}\left(w z \right) + \mathrm{E}\left( w \Delta z  \right).
-\tag{9}
+\tag{8}
 $$
 
-We can rearrange eqn 9 to derive the Price equation by expressing covariance as $\mathrm{Cov}(X,Y) = \mathrm{E}(XY) - \mathrm{E}(X)\mathrm{E}(Y)$, and therefore $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
-Substituting into eqn 9,
+We can rearrange eqn 8 to derive the Price equation by expressing covariance as $\mathrm{Cov}(X,Y) = \mathrm{E}(XY) - \mathrm{E}(X)\mathrm{E}(Y)$, and therefore $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
+Substituting into eqn 8,
 
@@
 $$\bar{w}\Delta \bar{z} = \mathrm{Cov}\left(w ,z \right) + \mathrm{E}\left( w \Delta z  \right).
-\tag{10}
+\tag{9}
 $$
@@
 Equation 10 is the Price equation.
-From eqn 1, which describes fundamental birth and death processes in a population, we can therefore derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).
+From eqn 1, which describes fundamental birth and death processes in a population, we can therefore derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 9; Box 1).
 
@@
 From the Price equation, we are able to recover classical population and quantitative genetic models [@Queller2017] and develop new ones [@Rice2004; @Rice2020; @Luque2016; @Lion2018].
-Our eqn 1 contains ecology at its core, and we show how the Price equation logically follows from it after accounting for absolute population growth (eqn 8).
+Our eqn 1 contains ecology at its core, and we show how the Price equation logically follows from it after accounting for absolute population growth (eqn 7).
 We therefore conclude that population and quantitative genetic equations contain ecology (no matter how hidden), and the ecological nature of evolution is implicit in population and quantitative genetic models.

### 2025-08-04 foundations_of_ecoevo.Rmd
Line 193 Make it clear that this statement means if all else are being equal

@@
 In this definition, the longevity of the individual matters.
-By definition, an individual that survives from $t$ to $t + 1$ has a higher fitness than one that dies, even if both have the same reproductive output.
+All else being equal, an individual that survives from $t$ to $t + 1$ has a higher fitness than one that dies, even if both have the same reproductive output.
 With this definition of fitness (eqn 3), we substitute,

### 2025-08-04 foundations_of_ecoevo.Rmd
Lines 210-211 removing that sentence that was causing confusion, since it's really not necessary

@@
 But we cannot treat $\Omega/N$ as the mean of $z$ at $t+1$ ($\bar{z}'$) because we need to weight $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.
-
-For example, if mean fitness at $t$ is 2 (i.e., $\bar{w} = 2$), then half as many individuals will have contributed to $\Omega$ in $t + 1$ than would have if $\bar{w} = 1$ (i.e., there are $N$ individuals at $t$ and $2N$ individuals at $t + 1$).
-We therefore need to multiply the mean trait value $\bar{z}'$ (at $t + 1$) by the mean fitness $\bar{w}$ (at $t$) to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$ [this has been applied in the context of population genetics by @Ewens2014].
+We need to multiply the mean trait value $\bar{z}'$ (at $t + 1$) by the mean fitness $\bar{w}$ (at $t$) to recover the mean contribution of the $N$ individuals at $t$ to the total $\Omega$ [@Case2000; @Ewens2014].
 Consequently,
@@
 
-Equation 8 conserves the total probability and now clarifies $\Omega$ as a summed trait value, which has the same units as $z$ and equals expected population growth at $t$ times mean trait value at $t + 1$.
+Equation 7 conserves the total probability and now clarifies $\Omega$ as a summed trait value, which has the same units as $z$ and equals expected population growth at $t$ times mean trait value at $t + 1$.
 This is consistent with the population ecology derivation from the previous section where $z_{i} = 1$ by definition, and $\Omega = N_{t+1}$ (note $\lambda = \bar{w}$).
@@
 
-Equation 10 is the Price equation.
+Equation 9 is the Price equation.
 From eqn 1, which describes fundamental birth and death processes in a population, we can therefore derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 9; Box 1).

### 2025-08-04 foundations_of_ecoevo.Rmd
Thanking Andy for comments

@@
 
-<!---
-
 Over 120 years ago, @Needham1904 described natural history as "the study of the phenomena of fitness".
-Fundamentally, we show why eco-evolutionary change is a statistical interaction between fitness and an individual characteristic.
+Fundamentally, we show why eco-evolutionary change is a statistical interaction between fitness and individual characteristics.
 From this definition, we can recover both population size change and evolutionary change.
-Our $\Omega$ can be therefore be interpreted as characteristic-scaled population change, with the special case of $z = 1$ recovering population size change by recognising that each individual contributes one to the total.
-
---->
 
@@
 Victor J. Luque would also like to thank Lorenzo Baravalle, Pau Carazo, Santiago Ginnobili, Manuel Serra, and Ariel Roff$\'{e}$.
-Lastly, we thank two anonymous reviewers.
+Lastly, we thank two anonymous reviewers and Andy Gardner for helpful comments.
 

### 2025-08-04 foundations_of_ecoevo.Rmd
Some minor edits

@@
 ---
-
-<!---
-
-NOTE: ADD CASE AND TAPER (2000) AS A REFERENCE FOR Delta N = N bar(w)
-
---->
 
@@
 Over an arbitrary length of time, fitness is properly defined as $w_{i} = \beta_{i} - \delta_{i} + 1$.
-<!--- For simplicity, we defined beta and delta as births and deaths, but we could more generally define beta as the count of descendents of i, and delta as the count of deaths in i and all of its descendents --->
-Over an individual's lifetime (which, by definition, includes death), fitness is therefore $\beta_{i}$. 
 Rates of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
